### 2012-03-12 dissertation-proposal/research-proposal.tex
Add draft research proposal

\documentclass[12pt, a4paper]{article}
\setlength{\oddsidemargin}{0.5cm}
\setlength{\evensidemargin}{0.5cm}
\setlength{\topmargin}{-1.6cm}
\setlength{\leftmargin}{0.5cm}
\setlength{\rightmargin}{0.5cm}
\setlength{\textheight}{24.00cm} 
\setlength{\textwidth}{15.00cm}
\parindent 0pt
\parskip 5pt
\pagestyle{plain}

\title{Research Proposal}
\author{}
\date{}

\newcommand{\namelistlabel}[1]{\mbox{#1}\hfil}
\newenvironment{namelist}[1]{%1
\begin{list}{}
    {
        \let\makelabel\namelistlabel
        \settowidth{\labelwidth}{#1}
        \setlength{\leftmargin}{1.1\labelwidth}
    }
  }{%1
\end{list}}

\begin{document}
\maketitle

\begin{namelist}{xxxxxxxxxxxx}
\item[{\bf Title:}]
	Optional Static Type Checking for Dynamic Languages
\item[{\bf Author:}]
	Ambrose Bonnaire-Sergeant
\item[{\bf Supervisor:}]
	Rowan Davies
\end{namelist}

\section*{Background}

% In this section you should give some background to your
% research area. What is the problem you are tackling, and why is it
% worthwhile solving? Who has already done some work in this area,
% and what have they achieved?

Programmers have long been divided between statically typed and 
dynamically typed languages.
Traditionally, languages have been defined as using one of these styles,
but recently languages such as Racket/Typed Racket, Dart and Qi/Shen
have borrowed from both.

The tradeoffs of both styles has been subject to much debate, but some
points have general agreement. Dynamically typed languages
are generally more convenient for writing smaller 
programs, whereas statically typed languages are usually preferred 
when writing large programs.

There has been little practical experience writing programs in a single
language that combine both styles.  One example is
Typed Racket, which provides a statically typed variation of Racket 
with the intention of safely porting untyped code to typed code. 
Typed code can interact with untyped code while preserving some properties
of typed code by, for example, delaying static type checks until runtime.
Porting untyped Racket code to Typed Racket generally involves 
adding type annotations to top level declarations with minimal (usually zero)
modification to existing code structure.

Type checking is often an operation complected with compilation. This is
because languages that have type checkers generally enforce the rule that
if a program does not satisfy the type checker, then the program should not be 
allowed to compile. When considering an optional type checker, it makes sense
to provide type checking as a simple operation, callable a la carte. 
This relieves a common restriction associated with statically typed languages;
programs can now compile regardless of the results of the 
type checker.

\section*{Aim}

% Now state explicitly the hypothesis you aim to
% test. Make references to the items listed in the Reference section
% that back up your arguments for why this is a reasonable
% hypothesis to test, for example the work of Knuth~\cite{knuth}.
% Explain what you expect will be accomplished by undertaking this
% particular project.  Moreover, is it likely to have any other
% applications?

I hypothesise that adding an optional static type system to a dynamically typed language
will increase its utility, while avoiding some disadvantages of statically typed
languages.

I expect to develop a library for Clojure which provides an optional static
type system, intended for practical use.

The rich type annotations that will be accumulated could also be
useful elsewhere, such as a function search engine like
(name of haskell search engine? TODO)

\section*{Method}

% In this section you should outline how you intend to go
% about accomplishing the aims you have set in the previous
% section. Try to break your grand aims down into small,
% achievable tasks. Try to estimate how long you will
% spend on each task, and draw up a timetable for each
% sub-task.

\begin{description}
\item[0.1]
	\begin{itemize}
	\item Concrete types (no type variables)
	\item Union types
	\item Fixed arity function types
	\item Typed deftype
	\item Uniform variable-arity function types
	\item Type checking without refinements or occurance typing
	\end{itemize}
\item[0.2]
	\begin{itemize}
	\item Occurance typing for type inference
	\item Type variables (polymorphic types)
	\end{itemize}
\item[0.3]
	\begin{itemize}
	\item Typed defrecord
	\item All Types annotated for clojure.core library
	\end{itemize}
\item[0.4]
	\begin{itemize}
	\item Manage interactions between typed namespaces
	\end{itemize}
\item[0.5]
	\begin{itemize}
	\end{itemize}
\item[0.6]
	\begin{itemize}
	\item Mutable reference types
	\item Non-uniform variable arity polymorphism
	\end{itemize}

\end{description}
\section*{Software and Hardware Requirements}
% Outline what your specific requirements will be with regard
% to software and hardware, but note that any special requests
% might need to be approved by your supervisor and the Head of
% Department.

Linux environment with Java, git, and maven installed.

% Overall, you should aim to produce roughly a two page document
% (and certainly no more than four pages)
% outlining your plan for the year.

% \begin{thebibliography}{9}
% \bibitem{knuth} D. E. Knuth. {\em The \TeX~book.}\/ Addison-Wesley,
% Reading, Massachusetts, 1984.
% \bibitem{lamport} L. Lamport. {\em \LaTeX~: A Document Preparation
% System}.\/ Addison-Wesley, Reading, Massachusetts, 1986.
% \bibitem{ken} Ken Wessen, Preparing a thesis using \LaTeX~, private
% communication, 1994.
% \bibitem{lamport2} L. Lamport. Document Production: Visual
% or Logical, {\em Notices of the Amer. Maths. Soc.},\/ Vol. 34,
% 1987, pp. 621-624.
% \end{thebibliography}

\end{document}

### 2012-03-13 dissertation-proposal/research-proposal.tex
Start on second draft

@@
 \documentclass[12pt, a4paper]{article}
+
+\usepackage{biblatex}
+\bibliography{bibliography}
+
 \setlength{\oddsidemargin}{0.5cm}
@@
 
-Programmers have long been divided between statically typed and 
-dynamically typed languages.
-Traditionally, languages have been defined as using one of these styles,
-but recently languages such as Racket/Typed Racket, Dart and Qi/Shen
-have borrowed from both.
+Dynamic languages are designed to be convenient for the programmer to
+write programs quickly and aspire to get out of the programmer's
+way as much as possible. When programs grow large and stabilize,
+the features of static languages are missed, specifically static
+type checking.
+
+Having a dynamic language that supports static properties 
+has many potential practical benefits. 
+Module-by-module porting of existing untyped code to a typed language 
+is the largest benefit cited by Tobin-Hochstadt \cite{TypedScheme:2010}.
+He developed Typed Scheme
+\cite{TypedScheme:2010} 
+to safely and incrementally port existing Scheme code, a dynamic language, to
+Typed Scheme, a static language.
+
+A recent example is Dart 
+\cite{Dart:2012:Online}, a dynamically typed language that allows optional
+static checks.
+
+Good type systems deeply integrate with the language they are written for.
+Statically typed languages usually include a type system in early releases,
+which is able to evolve with the language while it is malleable.
+
+The type system understands common idioms
+and becomes the programmer's ally.
 
@@
 
-There has been little practical experience writing programs in a single
-language that combine both styles.  One example is
-Typed Racket, which provides a statically typed variation of Racket 
+Typed Racket \cite{TypedScheme:2010}
+, which provides a statically typed variation of Racket 
 with the intention of safely porting untyped code to typed code. 
@@
 
-I hypothesise that adding an optional static type system to a dynamically typed language
-will increase its utility, while avoiding some disadvantages of statically typed
-languages.
+My goal is to develop an optional static type system for Clojure, 
+intended for practical use, to be provided as a library.
 
-I expect to develop a library for Clojure which provides an optional static
-type system, intended for practical use.
+It will be based on the lessons learnt throught the development
+of Typed Scheme, and as a response to Tobin-Hochstadt's \cite{TypedScheme:2010}
+suggestion to add type systems to dynamic languages.
 
-The rich type annotations that will be accumulated could also be
-useful elsewhere, such as a function search engine like
-(name of haskell search engine? TODO)
+I expect this library to be an important asset to Clojure's feature set.
+I expect it will become a factor in language decision and justification for programming projects.
 
@@
 
+\printbibliography
+
 \end{document}

### 2012-03-13 dissertation-proposal/research-proposal.tex
More changes for second draft. TODO method

@@
 \item[{\bf Title:}]
-	Optional Static Type Checking for Dynamic Languages
+	Optional Static Type Checking for Clojure
 \item[{\bf Author:}]
@@
 
-Dynamic languages are designed to be convenient for the programmer to
-write programs quickly and aspire to get out of the programmer's
+Dynamic languages are designed to be convenient for
+writing programs quickly, and aspire to get out of the programmer's
 way as much as possible. When programs grow large and stabilize,
-the features of static languages are missed, specifically static
+some features of static languages are missed, specifically static
 type checking.
@@
 
-A recent example is Dart 
-\cite{Dart:2012:Online}, a dynamically typed language that allows optional
-static checks.
-
-Good type systems deeply integrate with the language they are written for.
-Statically typed languages usually include a type system in early releases,
-which is able to evolve with the language while it is malleable.
-
-The type system understands common idioms
-and becomes the programmer's ally.
-
-The tradeoffs of both styles has been subject to much debate, but some
-points have general agreement. Dynamically typed languages
-are generally more convenient for writing smaller 
-programs, whereas statically typed languages are usually preferred 
-when writing large programs.
-
-Typed Racket \cite{TypedScheme:2010}
-, which provides a statically typed variation of Racket 
-with the intention of safely porting untyped code to typed code. 
-Typed code can interact with untyped code while preserving some properties
-of typed code by, for example, delaying static type checks until runtime.
-Porting untyped Racket code to Typed Racket generally involves 
-adding type annotations to top level declarations with minimal (usually zero)
-modification to existing code structure.
-
-Type checking is often an operation complected with compilation. This is
-because languages that have type checkers generally enforce the rule that
-if a program does not satisfy the type checker, then the program should not be 
-allowed to compile. When considering an optional type checker, it makes sense
-to provide type checking as a simple operation, callable a la carte. 
-This relieves a common restriction associated with statically typed languages;
-programs can now compile regardless of the results of the 
-type checker.
+Clojure is a dynamic, functional language with implementations for the Java Virtual
+Machine and Common Language Runtime, and compiling to Javascript. Clojure
+is also a Lisp, which makes it a good candidate to test the ideas developed in
+Typed Scheme \cite{TypedScheme:2010}.
 
@@
 
-My goal is to develop an optional static type system for Clojure, 
-intended for practical use, to be provided as a library.
+My goal is to develop a prototype optional static type system for Clojure, 
+eventually intended for practical use, to be provided as a library.
 
@@
 of Typed Scheme, and as a response to Tobin-Hochstadt's \cite{TypedScheme:2010}
-suggestion to add type systems to dynamic languages.
+suggestion to add type systems to existing dynamic languages.
 
-I expect this library to be an important asset to Clojure's feature set.
-I expect it will become a factor in language decision and justification for programming projects.
+There are several challenges to creating a satisfactory type system for Clojure.
+
+Multimethods play a significant role in Clojure, and a satisfactory
+type system for Clojure should understand them to some degree.
+There is some experience statically typing multimethods
+\cite{Millstein02modularstatically}
+, but this is a challenge and will probably not be completed in the
+timeframe proposed.
+
+Clojure's core library includes variable-arity functions which
+are not easily typed with current static type systems. 
+Providing satisfactory types for functions 
+like Clojure's 'map', 'filter', and 'reduce' require
+support for non-uniform variable arity polymorphism. 
+Strickland, Tobin-Hochstadt and Felleisen \cite{Strickland:2009:PVP:1532974.1532978}.
+describe their approach for non-uniform variable arity polymorphism, as used
+in Typed Racket. It is not clear if including a similar system is 
+achievable in the given timeframe, but would be a very useful feature.
+
+Occurence typing is type checking technique developed for Typed Scheme 
+\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
+and improved for Typed Racket
+\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}.
+It helps the type checker understand common programming idioms 
+with minimal type annotations.
+The general approach of occurence typing appears to fit Clojure well,
+but the implementation challenges are not clear. Occurence typing,
+or something similar, would be essential to a practical system.
+
+Ensuring vigorous type safety is an important aspect of Typed Scheme 
+\cite{TypedScheme:2010} ,
+especially when interacting between untyped and typed modules.
+I do not expect to concentrate on every combination of cross-module
+interaction. In particular, safely using typed code from untyped code
+is not important for my needs.
 

### 2012-03-14 dissertation-proposal/research-proposal.tex
2nd draft

@@
 \cite{Millstein02modularstatically}
-, but this is a challenge and will probably not be completed in the
+, but this is a challenge and may not be addressed in the
 timeframe proposed.
@@
 describe their approach for non-uniform variable arity polymorphism, as used
-in Typed Racket. It is not clear if including a similar system is 
-achievable in the given timeframe, but would be a very useful feature.
+in Typed Racket. 
+This would be a very useful feature, but may be a challenge to implement.
 
@@
 interaction. In particular, safely using typed code from untyped code
-is not important for my needs.
+is not a priority.
 
@@
 
+The milestones for this project are broken into release milestones
+for the prototype library.
+
+My goal is to finish all features listed in releases below
+and to begin work on the challenges, listed last.
+
 \begin{description}
@@
 	\item Fixed arity function types
-	\item Typed deftype
+	\item Typed deftype (class definitions)
 	\item Uniform variable-arity function types
 	\item Type checking without refinements or occurance typing
+	\item Enough functions annotated from clojure.core for proof-of-concept
 	\end{itemize}
@@
 	\begin{itemize}
-	\item Occurance typing for type inference
+	\item Typed defrecord (record structures)
 	\item Type variables (polymorphic types)
+	\item Most functions annotated from clojure.core library
 	\end{itemize}
-\item[0.3]
+\item[Challenges]
 	\begin{itemize}
-	\item Typed defrecord
-	\item All Types annotated for clojure.core library
-	\end{itemize}
-\item[0.4]
-	\begin{itemize}
+	\item Devise strategy for type inference (eg. occurence typing)
 	\item Manage interactions between typed namespaces
-	\end{itemize}
-\item[0.5]
-	\begin{itemize}
-	\end{itemize}
-\item[0.6]
-	\begin{itemize}
 	\item Mutable reference types
 	\item Non-uniform variable arity polymorphism
+	\item Typing Multimethods
 	\end{itemize}

### 2012-03-14 dissertation-proposal/research-proposal.tex
3rd Draft, after review from Rowan

@@
 \item[{\bf Title:}]
-	Optional Static Type Checking for Clojure
+	An Optional Static Type System for Clojure
 \item[{\bf Author:}]
@@
 
-Dynamic languages are designed to be convenient for
+Dynamically typed languages (also known as monotyped) are designed to be convenient for
 writing programs quickly, and aspire to get out of the programmer's
@@
 
-Having a dynamic language that supports static properties 
-has many potential practical benefits. 
-Module-by-module porting of existing untyped code to a typed language 
-is the largest benefit cited by Tobin-Hochstadt \cite{TypedScheme:2010}.
+Tobin-Hochstadt notes that ``untyped scripts are difficult to 
+maintain over the long run'' \cite{TypedScheme:2010} because types
+contain valuable design information.
 He developed Typed Scheme
@@
 
-Clojure is a dynamic, functional language with implementations for the Java Virtual
+Clojure is a dynamically typed, functional language with implementations for the Java Virtual
 Machine and Common Language Runtime, and compiling to Javascript. Clojure
-is also a Lisp, which makes it a good candidate to test the ideas developed in
-Typed Scheme \cite{TypedScheme:2010}.
+is also a Lisp, which makes it a good candidate to test the generality of
+ideas developed for Typed Scheme \cite{TypedScheme:2010}.
 
@@
 My goal is to develop a prototype optional static type system for Clojure, 
-eventually intended for practical use, to be provided as a library.
+eventually intended for practical use.
 
@@
 of Typed Scheme, and as a response to Tobin-Hochstadt's \cite{TypedScheme:2010}
-suggestion to add type systems to existing dynamic languages.
+suggestion to add type systems to other existing dynamic languages.
 
@@
 , but this is a challenge and may not be addressed in the
-timeframe proposed.
+timeframe proposed, beyond initial consideration of how it would
+fit with the rest of the system.
 
@@
 in Typed Racket. 
-This would be a very useful feature, but may be a challenge to implement.
+At a minimum, the proposed project will include broadly identifiying 
+what new issues arise when adapting this approach to Clojure, 
+with a prototype design if these issues are not too major.
 
@@
 with minimal type annotations.
-The general approach of occurence typing appears to fit Clojure well,
-but the implementation challenges are not clear. Occurence typing,
-or something similar, would be essential to a practical system.
+The proposed project will include a comparison of these approaches in
+the context of Clojure, and a prototype design if a satisfactory approach 
+is designed and no major issues are identifed.
 
@@
 interaction. In particular, safely using typed code from untyped code
-is not a priority.
+will not be designed. Safe interaction between typed modules
+however is needed for a practical type checker. I
+will attempt to design a viable strategy or identify issues that
+require further consideration.
 
@@
 
+For each of these it's not yet fully clear what novel issues may 
+arise in adapting existing techniques to Clojure.  For each item 
+below the project will involve either implemention or identification 
+of the novel issues that make implementation difficult.  
+The classification below reflects the level of uncertainity 
+regarding such issues.
+
 \begin{description}
@@
 	\item Uniform variable-arity function types
-	\item Type checking without refinements or occurance typing
+	\item Type checking without occurance typing
 	\item Enough functions annotated from clojure.core for proof-of-concept
@@
 Linux environment with Java, git, and maven installed.
+No issues are anticipated with regards to access to these.
 
@@
 
-% \begin{thebibliography}{9}
-% \bibitem{knuth} D. E. Knuth. {\em The \TeX~book.}\/ Addison-Wesley,
-% Reading, Massachusetts, 1984.
-% \bibitem{lamport} L. Lamport. {\em \LaTeX~: A Document Preparation
-% System}.\/ Addison-Wesley, Reading, Massachusetts, 1986.
-% \bibitem{ken} Ken Wessen, Preparing a thesis using \LaTeX~, private
-% communication, 1994.
-% \bibitem{lamport2} L. Lamport. Document Production: Visual
-% or Logical, {\em Notices of the Amer. Maths. Soc.},\/ Vol. 34,
-% 1987, pp. 621-624.
-% \end{thebibliography}
-
 \printbibliography

### 2012-03-14 dissertation-summary/research-summary.tex
Add research summary

--- 2012-03-14 dissertation-proposal/research-proposal.tex
@@
 
-\title{Research Proposal}
+\title{Research Summary}
 \author{}
@@
 
-\section*{Background}
-
-% In this section you should give some background to your
-% research area. What is the problem you are tackling, and why is it
-% worthwhile solving? Who has already done some work in this area,
-% and what have they achieved?
+\section*{Summary}
 
@@
 
-\section*{Aim}
-
-% Now state explicitly the hypothesis you aim to
-% test. Make references to the items listed in the Reference section
-% that back up your arguments for why this is a reasonable
-% hypothesis to test, for example the work of Knuth~\cite{knuth}.
-% Explain what you expect will be accomplished by undertaking this
-% particular project.  Moreover, is it likely to have any other
-% applications?
-
-My goal is to develop a prototype optional static type system for Clojure, 
-eventually intended for practical use.
-
+This project will explore designing and adding an optional type system to Clojure.
 It will be based on the lessons learnt throught the development
@@
 
-There are several challenges to creating a satisfactory type system for Clojure.
-
-Multimethods play a significant role in Clojure, and a satisfactory
-type system for Clojure should understand them to some degree.
-There is some experience statically typing multimethods
-\cite{Millstein02modularstatically}
-, but this is a challenge and may not be addressed in the
-timeframe proposed, beyond initial consideration of how it would
-fit with the rest of the system.
-
-Clojure's core library includes variable-arity functions which
-are not easily typed with current static type systems. 
-Providing satisfactory types for functions 
-like Clojure's 'map', 'filter', and 'reduce' require
-support for non-uniform variable arity polymorphism. 
-Strickland, Tobin-Hochstadt and Felleisen \cite{Strickland:2009:PVP:1532974.1532978}.
-describe their approach for non-uniform variable arity polymorphism, as used
-in Typed Racket. 
-At a minimum, the proposed project will include broadly identifiying 
-what new issues arise when adapting this approach to Clojure, 
-with a prototype design if these issues are not too major.
-
-Occurence typing is type checking technique developed for Typed Scheme 
-\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
-and improved for Typed Racket
-\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}.
-It helps the type checker understand common programming idioms 
-with minimal type annotations.
-The proposed project will include a comparison of these approaches in
-the context of Clojure, and a prototype design if a satisfactory approach 
-is designed and no major issues are identifed.
-
-Ensuring vigorous type safety is an important aspect of Typed Scheme 
-\cite{TypedScheme:2010} ,
-especially when interacting between untyped and typed modules.
-I do not expect to concentrate on every combination of cross-module
-interaction. In particular, safely using typed code from untyped code
-will not be designed. Safe interaction between typed modules
-however is needed for a practical type checker. I
-will attempt to design a viable strategy or identify issues that
-require further consideration.
-
-\section*{Method}
-
-% In this section you should outline how you intend to go
-% about accomplishing the aims you have set in the previous
-% section. Try to break your grand aims down into small,
-% achievable tasks. Try to estimate how long you will
-% spend on each task, and draw up a timetable for each
-% sub-task.
-
-The milestones for this project are broken into release milestones
-for the prototype library.
-
-My goal is to finish all features listed in releases below
-and to begin work on the challenges, listed last.
-
-For each of these it's not yet fully clear what novel issues may 
-arise in adapting existing techniques to Clojure.  For each item 
-below the project will involve either implemention or identification 
-of the novel issues that make implementation difficult.  
-The classification below reflects the level of uncertainity 
-regarding such issues.
-
-\begin{description}
-\item[0.1]
-	\begin{itemize}
-	\item Concrete types (no type variables)
-	\item Union types
-	\item Fixed arity function types
-	\item Typed deftype (class definitions)
-	\item Uniform variable-arity function types
-	\item Type checking without occurance typing
-	\item Enough functions annotated from clojure.core for proof-of-concept
-	\end{itemize}
-\item[0.2]
-	\begin{itemize}
-	\item Typed defrecord (record structures)
-	\item Type variables (polymorphic types)
-	\item Most functions annotated from clojure.core library
-	\end{itemize}
-\item[Challenges]
-	\begin{itemize}
-	\item Devise strategy for type inference (eg. occurence typing)
-	\item Manage interactions between typed namespaces
-	\item Mutable reference types
-	\item Non-uniform variable arity polymorphism
-	\item Typing Multimethods
-	\end{itemize}
-
-\end{description}
-\section*{Software and Hardware Requirements}
-% Outline what your specific requirements will be with regard
-% to software and hardware, but note that any special requests
-% might need to be approved by your supervisor and the Head of
-% Department.
-
-Linux environment with Java, git, and maven installed.
-No issues are anticipated with regards to access to these.
-
-% Overall, you should aim to produce roughly a two page document
-% (and certainly no more than four pages)
-% outlining your plan for the year.
-
 \printbibliography

### 2012-04-20 lit-review/lit-review-draft.tex
Add lit review draft

\documentclass[12pt, a4paper]{article}
\usepackage{biblatex}
\usepackage{fancyvrb}
\bibliography{bibliography}
\title{A Practical Optional Type System for Clojure, Literary Review}
\author{Ambrose Bonnaire-Sergeant (20350292)}

\begin{document}
\maketitle

\begin{abstract}
Clojure is a practical, dynamically typed language for the Java Virtual Machine.
I will design and implement a prototype optional static type 
system for Clojure, intended for practical use. Devising a static type system 
requires a good balance of components; they should be conceptually compatible with each other
while not significantly complicating its implementation.
I look at several novel inventions created as a result 
of the Typed Racket project.
Non-uniform variable-arity polymorphism and occurrence typing are two such
components that could be used in harmony in this project.
Typed Racket and Scala use different forms of local type inference,
which are compared and judged for applicability to this project.
Other relevant features of Scala are considered, such as its approach to bounded polymorphism,
the Scala class hierarchy, and the similarities of its methods of composition to Clojure's.
\end{abstract}

\tableofcontents

\section{Introduction}

Designing a static type system for any dynamically typed language requires
careful consideration. Dynamically typed languages usually have a particular
style, designed to be powerful yet non-intrusive. If a static type system
for such languages is to be effective, it should preserve existing idioms and
syntactic abstractions, at least until time tested idioms are devised for
the typed language \cite{SAMTH:dissertation} .

Typed Racket \cite{SAMTH:dissertation} integrates a selection of ideas to
create a typed sister language to Racket. 
Many ideas produced as a result of the Typed Racket project are relevant here.
Clojure and Racket are similar; they are both dynamically typed Lisps
and therefore have similar idioms. The inventions of
occurrence typing \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
and a systematic approach to variable-arity polymorphism
\cite{Strickland:2009:PVP:1532974.1532978}
are both applicable to this project.

Clojure supports multimethods, a mechanism to dispatch on the
results of an arbitrary function. Statically typing multimethods that support
multiple dispatch (which are more specific than Clojure's multimethods) has been tackled
in the context of statically typed languages
\cite{Millstein02modularstatically}
. Typing multimethods that support dispatch on an arbitrary function
has not been approached in the context of a dynamic language utilizing 
occurrence typing.

Clojure and Scala \cite{Odersky06anoverview},
are also similar langages.  
They could both be considered object-oriented, being built on abstractions
and classes. While there are some significant differences between them,
Scala being statically typed for example, there are still ideas that are applicable to this project.

Scala's inclusion of bounded type variables are useful in the context
of a language that makes heavy use of subtyping. There are some parallels
between Clojure protocols, which is Clojure's main abstraction, and Scala
traits, both can conceptually be useful in a language with bounded polymorphism.

Scala's class hierarchy \cite{Odersky06anoverview}
is also interesting and relevant. Scala and Clojure are
both languages with a focus on host interoperability, and both run
on the Java Virtual Machine.  Scala's and Clojure's
differing treatment of Java's \emph{null} reference is
also worth reviewing when devising the class hierarchy for this project.

\section{Variable-Arity Polymorphism}

Functions with variable-arity are abundant in dynamically typed languages;
it is common to have a large standard library of functions that utilize
variable-arity, and Clojure is no different. 
Statically typed languages tend to have limited support
for typing variable-arity functions. Strickland claims 
\cite{Strickland:2009:PVP:1532974.1532978}
``no typed language supports them in a systematic and principled manner''.

Strickland et al.  \cite{Strickland:2009:PVP:1532974.1532978} distinguish between two kinds
of variable-arity functions: uniform and non-uniform.

Uniform variable-arity functions have a homogeneous variable parameter list.
For example, the Clojure \emph{+} function takes any number of numeric values
and returns their sum.
Statically typed languages such as Java and C\# provide support to type such functions,
since access to variable arguments occurs via a homogeneous array \cite{Strickland:2009:PVP:1532974.1532978}
.

Non-uniform variable-arity functions have a heterogeneous variable parameter list.
For example, Clojure's \emph{map} function takes a function and one or more sequences,
and applies the function to each element of the sequences simultaneously.
Languages like Java and C\# lack support for typing these kinds of functions,
neither do languages with more powerful type systems such as Haskell and ML.
Strickland et al. conjecture that ``if their type systems provided variable-arity
polymorphism, Haskell and ML programmers would routinely define such functions too''
\cite{Strickland:2009:PVP:1532974.1532978}
.

\emph{map} is also interesting because it treats polymorphic functions as first
class values. This is the key innovation of Strickland et al. . Dzeng and Haynes
\cite{Dzeng94typereconstruction} work on adding variable-arity functions to
an ML style language differs mostly in this respect. Dzeng and Hayes support uniform and
non-uniform variable-arity functions, but not first class polymorphic values.  \footnotemark

\footnotetext{
First class polymorphic functions are significant in the context of local type inference (next section).
Hosoya and Pierce \cite{Hosoya99howgood} identify that synthesizing the type
of a first class polymorphic function passed to a polymorphic function is
\emph{hard to synthesize}. This is relevant because, even though
Strickland et al. provide the technology to type non-uniform variable-arity
functions with first class polymorphic values, extra type annotations could still be
required to satisfy local type inference when passing a polymorphic function
as an argument. Odersky's \cite{Odersky:2001:CLT:373243.360207} local type inference algorithm
makes manually providing these annotations less frequent.
}

A common usage of variable parameters in Clojure (and Racket) is keyword
parameters. The paper \cite {Strickland:2009:PVP:1532974.1532978}
mentions keyword parameters briefly, but it is unclear how it fits
with the given calculus. Starred pre-types and dotted pre-types are used
as placeholders for uniform and non-uniform variable parameters respectively, but
keyword parameters tend to be grouped in two. They could be regarded as ``semi-uniform''
variable parameter list. The paper does not mention such a middle-ground 
between the two styles.

\section{Local Type Inference}

Local type inference is used in both Typed Racket 
\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
and Scala \cite{Odersky06anoverview}, although both use slightly different
styles of checking. Typed Racket implements Local Type Inference
\cite{Pierce:2000:LTI:345099.345100},
while Scala implements Colored Local Type Inference \cite{Odersky:2001:CLT:373243.360207}.

Both local type inference papers describe similar algorithms. At the core
of any local type inference scheme is a \emph{bidirectional checking}
algorithm. This algorithm enables the local propagation of types both
up and down the syntax tree. Essentially, the algorithm is split into two
halves: checking mode and synthesis mode. In checking mode we are verifying
that an expression is a subtype of a given type, and in synthesis mode, the
types of expressions are propagated up when possible.

Colored Local Type Inference subsumes Local Type Inference. Colored Local Type Inference
allows the propagation of partial type information down the syntax tree,

Hosoya and Pierce \cite{Hosoya99howgood} identified two known issues with
local type inference, both related to synthesizing the type of arguments.
Unfortunately, they are unable to devise satisfactory extensions to existing
algorithms to mitigate these issues, and instead suggest avoiding
\emph{hard to synthesize} arguments. On the positive side, they identify some
common patterns for users to look out for. Odersky's \cite{Odersky:2001:CLT:373243.360207} 
algorithm is better behaved
than Piece and Turner's \cite{Pierce:2000:LTI:345099.345100} in some cases,
and is able to synthesize more \emph{hard to synthesize} expressions.

\section{Bounded and Unbounded Polymorphism}

The Local Type Inference paper \cite{Pierce:2000:LTI:345099.345100}
describes two implementations of type variables, for bounded
and unbounded type variables. The bounded implementation is presented
as an optional extension  to the unbounded implementation, which preserves all
properties described in the Local Type Inference algorithm.

An unbounded type variable does not have subtype constraints.
Bounded type variables can have subtype constraints, and 
subsume unbounded type variables \cite{Pierce:2000:LTI:345099.345100}
, as a unbounded variable can be represented as a variable bounded
by the Top type.

Still, unbounded type variables have an advantage: their implementations are
are simpler in the presence of a \emph{Bottom} type. 
The constraint resolution algorithm for bounded variables
is more subtle, due to ``some surprising interactions between bounded quantifiers
and the \emph{Bot} type'' \cite{Pierce:2000:LTI:345099.345100}, described fully
by Pierce \cite{Pierce97boundedquantification}.

Typed Racket \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
supports unbounded polymorphism, while Scala \cite{Odersky06anoverview}
supports an extended form of bounded polymorphism called
F-bounded polymorphism \cite{Canning:1989:FPO:99370.99392}, which allows the
bound variable to occur in its own bound.
F-bounded polymorphism is useful in the context of object-oriented abstractions,
as demonstrated by Odersky \cite{Odersky06anoverview} .
This is one possible explanation why Typed Racket, which is not built on abstractions like Scala,
does not support bounded quantification. Unfortunately, no Typed Racket paper mentions 
bounded quantification, so the rationale is not clear.

Clojure, like Scala, is built on object-oriented abstractions. Clojure protocols
and Java interfaces (interfaces are supported by Clojure) are good candidates
for bounds in bounded or F-bounded polymorphism.

\section{Typed Racket}

Typed Racket is a statically typed sister language of Racket. It
attempts to preserve existing Racket idioms and aims type check
existing Racket code by simply adding top level type annotations. \cite{SAMTH:dissertation}

Typing macros is an open research problem \cite{Herman10:Theory},
Typed Racket expands all macro calls before type checking, \cite{SAMTH:dissertation}
avoiding the complex semantics of pre-expanded macro calls described by Herman 
\cite{Herman10:Theory} .
My design of the type system for Clojure will follow a similar strategy. Macros
will be expanded (by the compiler), and the expanded form will be type checked in
the normal fashion. Type checking macro definitions are outside the scope of this project.

Certain macros encode invariants that are not easily extracted. For example,
\emph{define-struct} in Racket defines a structure. Typed Racket includes the
\emph{define-struct:} macro \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
, which annotates the structure's slots with types,
supports recursive types, and annotates the structure's type predicate with propositions
for compatibility with occurrence typing \cite{Strickland:2009:PVP:1532974.1532978}.
In this project, Clojure macros such as \emph{deftype} and \emph{defrecord}, used for defining a type
and a record respectively, would need similar typed versions.

Along with a full static type system, Typed Racket also comes with extensive support for
contracts to protect program invariants at runtime, \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
building on Racket's existing infrastructure for contracts.
Adding contracts is outside the scope of this project, but would be considered desirable
future work.

Two other Typed Racket features worth mentioning are recursive types and refinement types  
\cite{SAMTH:dissertation}
.
Recursive types allow a type definition to refer to itself, enabling structurally
recursive types like binary trees. Refinement types let the programmer define
new types that are subsets of an existing type, such as the type for all even integer,
which is a subset of all integers.
Both these features would fit well in a future implementation of this project.

\section{Occurrence Typing}
\label{sec:OccurrenceTyping}

Dynamically typed languages use a combination of type predicates and
selectors to reason about the type of variables at various points in a program.

For example, the following snippet shows how the type predicate
\emph{number?} and the selector \emph{first} are used to direct control flow and ensure the 
type of local variables.

\newpage

\begin{Verbatim}[frame=single]
(let [x (list (number-or-string))]
  (cond 
    (number? (first x)) (+ 1 (first x))
    :else               (str (first x))))
\end{Verbatim}

If a static type system for a dynamically typed language 
should be made to support existing programming idioms, it must
be able to use the results of these predicates and selectors to
infer the types at particular points in the program.

Occurrence typing \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
was developed for the Typed Racket project capture common Racket idioms.
Occurrence typing extends the type system with a \emph{proposition environment}
that represents the type refinements the type system has inferred
up to a point in the program.

The 2010 paper \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
reformulated occurrence typing, improving the original 2008 paper
\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
after revealing that ``three years of practical experience has revealed
serious shortcoming of our type system.'' \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
The new implementation utilizes a simple proof system to solve
propositional logic statements, in terms of type predicates and selectors.

Type predicates and selectors require two extra annotations 
for the occurrence typing system to understand it: a ``then'' proposition
when the result is a true value the, and an ``else'' proposition for a false value.
For example, \emph{number?} would have a ``then'' proposition saying its argument
is a number, and an ``else'' proposition saying its argument is \emph{not} a number.

One extension not mentioned in the papers is tracking values of
key-value pairs in hash-maps. Tobin-Hochstadt \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
uses the Racket selectors \emph{car} and \emph{cdr} to refine the type of
a list, but Clojure, which uses maps extensively, could benefit from refining the types 
of selectors like \emph{get} and \emph{find}, which extract values from maps.
This would require more fine grained types for hash-maps.

System D, a novel type system developed by Chugh et al. \cite{Chugh:2012:NRL:2103621.2103686}
, provides such fine grained hash-map types. By using \emph{nested refinements},
``System D can express complex invariants between base values and richer values''
\cite{Chugh:2012:NRL:2103621.2103686}
.
System D is much more expressive than occurrence typing, subsuming it, but it comes at a cost:
``the huge leap in expressiveness \ldots is accompanied by some unique technical challenges''
\cite{Chugh:2012:NRL:2103621.2103686}
.
An occurrence typing implementation is much more straightforward than System D, which involves
specially optimizing SMT queries to be fed into an SMT solver. By comparison, simple 
algorithms dealing with propositional logic are needed for occurrence typing.

Occurrence typing is a relatively simple, time worn technique used successfully 
in Typed Racket. Clojure is similar enough to Racket for occurrence typing to work
without issues, but further research should be carried out on more fine grained types
for hash maps to be satisfying pairing.

\section{Statically Typed Multimethods}

% challenges: methods occurring in different modules

A Clojure multimethod is slightly different to most other
abstractions of the same name. Here, a multimethod 
dispatches on the result of an \emph{arbitrary function},
rather that more traditional multimethods that dispatch on
the types or structure of its arguments.

Millstein and Chambers \cite{Millstein02modularstatically}
describe Dubious, a simple statically typed core language including multimethods that
dispatch on the type of its arguments. They tackle a key challenge for statically typing
multimethods: ``it is possible for two modules containing arbitrary multimethods to typecheck
successfully in isolation but generate type errors when linked together.'' \cite{Millstein02modularstatically}

This challenge holds when statically typing Clojure multimethods.
One additional challenge would be multimethods being imported from 
untyped modules. This would have to be handled carefully.

If we consider a multimethod as a spread out conditional, a
novel use of occurrence typing becomes apparent. We can use
occurrence typing to refine the argument types in the method bodies
by utilizing the results of the dispatch function. This is simple
when dispatching by type, but a common idiom in Clojure is to dispatch
on the value associated with a key in a hash-map.

To satisfy this idiom, we require occurrence typing with more
finely grained hash-map types (see section~\ref{sec:OccurrenceTyping} for discussion).

\section{Conclusion}

Many related components must come together in the design of a
static type system. Typed Racket achieves a satisfying balance of 
occurrence typing, local type inference and variable-arity polymorphism.
Similarly, Scala features F-bounded polymorphism, a class hierarchy
that is compatible with Java, and colored local type inference.

I intend to design a static type system for Clojure
that features a satisfying combination of features, individually taken from other
languages, when combined produce novel and interesting problems.

I have outlined an interesting possible extension to occurrence typing,
adding finer grain hash-map types.

I have proposed novel method of statically typing multimethods
using occurrence typing to consider the results of the dispatch function.

I suggested combining these two ideas to produce an even more expressive
method for typing multimethods that dispatch on the value associated with keys in hash-maps.

\printbibliography

\end{document}

### 2012-05-18 type-inference-review/assignment3.tex
add type inference

\documentclass[12pt,a4paper]{article}
%\usepackage{graphicx}
%\usepackage{cite}
%\usepackage{latexsym}
\usepackage{biblatex}

\bibliography{bibliography}

\usepackage{color}
\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  %showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=b,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

\title{Type Inference for a Clojure Type System}
\author{Ambrose Bonnaire-Sergeant}

\begin{document}

\maketitle

\begin{abstract}

  Type inference empowers programmers to work faster and safer in a statically 
  typed world. The popularity of language idioms is strongly linked
  with type inference: programmers won't use hard-to-express idioms.
  Designing a type system for a dynamic language should preserve existing idioms, 
  and this crucially relies on the design of type inference.
  I explore and evaluate the type inference strategies
  used by Typed Racket and Scala, showing their relevance to a type system for Clojure
  (an existing dynamically typed language).

\end{abstract}

{\bf Keywords:} occurrence typing, local type inference, Typed Racket, Scala, Clojure

\section{Introduction}

Statically typed languages know the types of their programs
at compile time: but how? Explicitly typed languages
put most responsibility on the programmer
to provide type annotations; implicitly typed languages attempt to
recover type information using various type inference strategies. 
A good type inference algorithm can minimise and sometimes completely
eliminate manual type annotations.

Clojure is a dynamically typed language on the Java Virtual Machine.
In designing a type system for Clojure, I explore two existing statically typed
languages: Scala and Typed Racket.
I conclude that several type inference techniques used by Typed Racket
are applicable to a Clojure type system.

At the core of Typed Racket's type inference strategy is Local Type Inference
\cite{Pierce:2000:LTI:345099.345100}.
Local Type Inference requires all top level bindings to be fully annotated,
but promises to almost never require type annotations to local bindings.
Hosoya and Pierce detail the limitations of Local Type Inference,
and note little success in improving them\cite{Hosoya99howgood}.

The Scala type inference strategy utilises Colored Local Type Inference
\cite{Odersky:2001:CLT:373243.360207}, an extension of Pierce and 
Turner's Local Type Inference
\cite{Pierce:2000:LTI:345099.345100}.
Colored Local Type Inference improves Pierce and Turner's algorithm
by allowing partial type information to propagate downwards, where
the original algorithm only allowed the propagation of complete type information.

Typed Racket's type inference is helped by occurrence typing \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
, which reasons about predicates directing program flow. Its
goal is to understand dynamic language idioms, such that they can
be preserved and type checked. Occurrence typing is crucial to the
usability of Typed Racket, and it could fit well into a Clojure
type system.

To test the applicability of Typed Racket's type inference strategies,
I port a function provided by Clojure's standard library
to Typed Racket. The function uses multiple type predicates
combined in either disjunction and conjunction, and expects two very general
parameters. This tests Typed Racket's occurrence typing implementation.
The results are quite promising, with the basic structure of the code remaining unchanged.

I also verify some known limitations of Local Type Inference 
\cite{Pierce:2000:LTI:345099.345100}\cite{Hosoya99howgood}
by type checking code known to fail type synthesis. I show the specified cases fail as expected
and then I provide correct type annotations to type check the cases.

From these experiments, I conclude that occurrence typing is
a good candidate for a Clojure type system. It allows idiomatic code to be
typed without structural modifications. I also conclude that Local Type Inference 
\cite{Pierce:2000:LTI:345099.345100}
is more appropriate than Colored Local Type Inference \cite{Odersky:2001:CLT:373243.360207}, since the benefits of the
latter are limited outside Scala, but I note that Colored Local Type Inference
would be a good starting point to improve Local Type Inference.

\section{Previous and related work}

One goal of Typed Racket \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486} is 
adding a satisfying type system to an existing, dynamic, 
language. This is exactly the use-case I explore with a Clojure type system.
The particular mix of type checking features used by Typed Racket harmonise to satisfy the 
goal of type checking idiomatic Racket code.

\subsection{Local Type Inference}

Typed Racket uses Local Type Inference \cite{Pierce:2000:LTI:345099.345100}
as an inference tool and checking tool. Pierce and Turner
\cite{Pierce:2000:LTI:345099.345100} divide Local Type Inference into
two complementary algorithms. \emph{Local type argument synthesis}
synthesises type arguments to polymorphic applications, and \emph{bidirectional
propagation} propagates type information to nodes in both directions of the source tree.

Pierce and Turner split \emph{local type argument synthesis} into two further
algorithms: bounded, and unbounded quantification \cite{Pierce:2000:LTI:345099.345100}. 
Typed Racket 
supports unbounded polymorphism \cite{SAMTH:dissertation}, implementing the latter algorithm by Piece et al. .
Scala supports bounded quantification with F-bounded polymorphism \cite{Canning:1989:FPO:99370.99392},
basing its type argument synthesis on the bounded quantification algorithm.

Pierce and Turner explicitly forbid \cite{Pierce:2000:LTI:345099.345100}
type variables with interdependent bounds, including
F-bounds, having failed to devise an algorithm to infer these cases.
Scala's type argument synthesis implementation deviates from Pierce and Turner to support these features.
I am not aware of papers specifically describing Scala's modifications, but they are at least inspired by
Scala's spiritual ancestors Generic Java \cite{Bracha98gj:extending} and Pizza \cite{Odersky97pizzainto}.

Hosoya and Pierce \cite{Hosoya99howgood} reiterate two common problems with Local Type Inference:
``hard-to-synthesise arguments'' and ``no best type argument''. The first problem occurs because
both local type argument synthesis and bidirectional propagation cannot perform synthesis
simultaneously. Cases where both algorithms can recover new type information are usually ``hard-to-synthesise''.
``No best type argument'' describes the situation where the results of local
type argument synthesis yield more than one type, and no type is better than the other. Sometimes we cannot recover and synthesis
fails.

\subsection{Colored Local Type Inference}

Scala's type checking uses Colored Local Type Inference \cite{Odersky:2001:CLT:373243.360207},
a variant of Local Type Inference \cite{Pierce:2000:LTI:345099.345100} specifically designed to
improve inference with certain kinds of pattern matching expressions in Scala. It allows
\emph{partial} type information to propagate down the syntax tree, instead of only full type information
as required by Local Type Inference.

\emph{Colored} types contain extra information than normal types, including the propagation direction
and missing parts of the type (when representing a \emph{partial type}). They are generally useful
for describing ``information flow in polymorphic type systems with propagation-based type inference''
\cite{Odersky:2001:CLT:373243.360207}. If the scope of my research permits it, I plan to investigate
using colored types to solve the polymorphic higher-order-function limitation of Local Type Inference.

\subsection{Occurrence Typing}

Occurrence typing was devised \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486} and improved
\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561} for the Typed Racket project. It uses propositional 
logic to combine type information gathered by a combination of type predicates and path selectors.
It enables recovery of type information in an ad-hoc fashion, and gives the type checker some insight
in the invariants at given points in the program.

Crucially, occurrence typing allows us to continue writing code in a dynamic style, without restructuring
our functions for the sake of type checking.

\subsection{A Clojure Type System}

Clojure is similar to both Scala and Typed Racket in a few ways. Scala and Clojure
both run on the JVM, and have extensive object-oriented standard libraries to utilise.
As is common in statically typed object-oriented languages, bounded quantification is
included, with the extension of F-bounded polymorphism.

The parallels with Typed Racket are more fruitful in a type checking context.
Occurrence typing is a good, lightweight solution to the problem faced by Typed Racket:
type check idiomatic and existing Racket code. Clojure and Racket are both Lisps, so even
visually there is a close relationship between them. It seems safe to assume occurrence
typing will bring as many benefits to a Clojure type system as it did to Typed Racket.

\section{Methodology}

Evaluating a type inference implementation should involve several
strategies. I wish to maximise the intersection of 
the set of idiomatic Clojure forms and the set of forms whos type
can be successfully inferred by the type inference algorithm.

The Typed Racket type inference implementation is a combination
of two strategies: Local Type Inference and occurrence typing. 
Local Type Inference can be further described as an interaction between
two further algorithms,
\emph{local type argument synthesis} and \emph{bidirectional checking}\cite{Hosoya99howgood}.
I identify cases that test some components individually.

Some known limitations of Local Type Inference are explored by
Hosoya and Turner \cite{Hosoya99howgood}. They describe the strengths
of Local Type Inference, in particular they claim the limitations of
Local Type Inference are well defined and easy to understand. This claim
is also mirrored by Pierce and Turner \cite{Pierce:2000:LTI:345099.345100}.

One limitation described by both sets of authors is summarised by
Hosoya and Turner as ``hard to synthesise arguments'' \cite{Hosoya99howgood}.
It refers to situations where both the \emph{local type argument synthesis} and
\emph{bidirectional checking} algorithms could help synthesise the type. Because
both algorithms cannot run together, type synthesis fails.

Any example given by Hosoya et at. \cite{Hosoya99howgood} attempts to synthesises the type of an application of a polymorphic
function that takes an unannotated function as an argument.
For example \lstinline|(map (fn [x] x) [1 2 3])|,
where the type of the polymorphic function \emph{map} is 
\lstinline|(All [a b] [[a -> b] (List a) -> (List b)])|. Here, neither 
local type argument synthesis or bidirectional checking have sufficient type information
to infer types, so type inference fails. We should be able to help type inference
by providing parameter types to the anonymous function argument \lstinline|(map (fn [x : Integer] x) [1 2 3])|
or by instantiating \emph{map} with type arguments \lstinline|((inst map Integer Integer) (fn [x] x) [1 2 3])|.

\section{Implementation}

To test the type inference in Typed Racket,
I use the version of Typed Racket packaged with DrRacket 5.2, running on Ubuntu Linux 11.10.

I chose a function from the Clojure standard library to port to Typed Racket [Listing \ref{lst:isaCLJ}].
\lstinline|clojure.core/isa?| was chosen to test the occurrence
typing implementation of Typed Racket; there are several type predicates nested
in conjunctions that guard function applications. For example, the conjunction
of \lstinline|(class? parent)| and \lstinline|(class? child)| on line 5 ensure \lstinline|parent|
and \lstinline|child| are instances of \lstinline|Class| before calling the \lstinline|Class| instance
method \lstinline|isAssignableForm|
on line 6,
which expects a \lstinline|Class| argument. These ad-hoc type predicate applications must be recognised
and utilised by occurrence typing for type inference to succeed.

\begin{lstlisting}[caption=isa? Clojure Version, label=lst:isaCLJ]
(defn isa?
  ([child parent] (isa? global-hierarchy child parent))
  ([h child parent]
   (or (= child parent)
       (and (class? parent) (class? child)
            (. ^Class parent isAssignableFrom child))
       (contains? ((:ancestors h) child) parent)
       (and (class? child) 
            (some #(contains? ((:ancestors h) %) parent)
                  (supers child)))
       (and (vector? parent) (vector? child)
            (= (count parent) (count child))
            (loop [ret true i 0]
              (if (or (not ret) (= i (count parent)))
                ret
                (recur (isa? h (child i) (parent i)) (inc i))))))))
\end{lstlisting}

I also test the Local Type Inference implementation by reproducing
a ``hard to synthesise case'' \cite{Hosoya99howgood}.
I test the often cited common weakness, where type inference
fails because of applying a polymorphic function to an unannotated anonymous function
argument.

%Describe how you went about testing the hypothesis that you have stated
%in the introduction. What equipment did you use? Were there any parameter
%values you used? Give enough detail so that others can replicate your work.

\section{Results}

\subsection{Testing Typed Racket's Occurrence Typing}

I ported \lstinline|clojure.core/isa?| [Listing \ref{lst:isaCLJ}] to Typed Racket [Listing \ref{lst:isaTR}] by
first converting Clojure functions to Racket. If there was no
equivalent, such as the \lstinline|vector-of?| predicate, I define my own
and use it.

Note that \lstinline|isa?| takes two argument of type \lstinline|Any|
and returns a \lstinline|Boolean|. At line 17 is the ported version of
the example described in the previous section. Notice is it now statically
type checked, which means type inference succeeded. We know Typed Racket
recognised the predicate \lstinline|class?| because the call 
\lstinline|class-assignable-from-class| takes two \lstinline|class|
parameters.

I ran some tests with dummy \lstinline|class|s, which reinforced
the correctness of my code.

\begin{lstlisting}[caption=isa? Typed Racket Port, label=lst:isaTR]
(struct: class ...)
(struct: hierarchy ([ancestors : (HashTable class (Setof class))]))

(: class-assignable-from-class? (class class -> Boolean))
(: some-super (hierarchy Any class -> Boolean))

(define-predicate vector-of? (Vectorof Any))

(: isa? (case-lambda
          (Any Any -> Boolean)
          (hierarchy Any Any -> Boolean)))
(define isa?
  (case-lambda
    ((child parent) (isa? global-hierarchy child parent))
    ((h child parent)
     (or (equal? child parent)
         (and (class? parent) (class? child)
              (class-assignable-from-class? child parent))
         (and (class? child)
              (let ([hs (hierarchy-ancestors h)])
                (and (hash-has-key? hs child)
                     (let ([as (hash-ref hs child)])
                       (set-member? as parent)))))
         (and (class? child) (some-super h parent child))
         (and (vector-of? child) (vector-of? parent)
              (= (vector-length child) (vector-length parent))
              (let: loop : Boolean
                ([ret : Boolean #t] 
                 [i : Integer 0])
                (if (or (not ret) (= i (vector-length parent)))
                    ret
                    (loop (isa? h (vector-ref child i)
                                  (vector-ref parent i)) 
                          (+ 1 i)))))))))
\end{lstlisting}

\subsection{Testing Typed Racket's Local Type Inference}

I performed some verification tests on Typed Racket's Local Type Inference 
implementation. I chose a polymorphic function, \lstinline|map| [Listing \ref{lst:maptype}], and applied
an anonymous function.

\begin{lstlisting}[caption=Type of Typed Racket's map, label=lst:maptype]
(:print-type map)
;(All (c a b ...) 
;  (case-> 
;    ((a -> c) (Pairof a (Listof a)) 
;     -> (Pairof c (Listof c)))
;    ((a b ... b -> c) (Listof a) (Listof b) ... b 
;     -> (Listof c))))
\end{lstlisting}

Here I omit type parameters and parameter annotations [Listing \ref{lst:hard1}]. As expected,
type inference fails, so parameter \lstinline|a| is inferred to be
of type \lstinline|Any|, which is too general to be passed to
\lstinline|+|.

\begin{lstlisting}[caption=Hard to Synthesise Polymorphic Application without Type Arguments and Annotations - Typed Racket, label=lst:hard1]
(map (lambda (a) (+ 1 a 3)) (list 1 2 3))
; Type Checker: Expected Number, but got Any in: a
\end{lstlisting}

Here I give sufficient type arguments to instantiate \lstinline|map| [Listing \ref{lst:hard2}], but
this is, perhaps surprisingly, not enough information for Local Type Inference to
infer the correct type for \lstinline|a|.

\begin{lstlisting}[caption=Hard to Synthesise Polymorphic Application with Type Arguments without Annotations - Typed Racket, label=lst:hard2]
((inst map Integer Integer) (lambda (a) (+ 1 a 3)) (list 1 2 3))
; Type Checker: Expected Number, but got Any in: a
\end{lstlisting}

Finally, I show a version compatible with the type inferencer by
providing  the parameter types of the anonymous function argument [Listing \ref{lst:welltyped}].

\begin{lstlisting}[caption=Well Typed - Typed Racket, label=lst:welltyped]
(map (lambda: ([a : Integer]) (+ 1 a 3)) (list 1 2 3))
\end{lstlisting}

%Carefully outline each experiment: what was its purpose, and what
%precisely did you find? Sometimes it is easiest to display your
%results in tabular form, nut note that tables are also floating
%environments in \LaTeX\/, so they don't always end up where you want
%them to be!
%
%\begin{table}[htbp]
%  \caption{A very interesting table of tens.}
%  \begin{center}
%    \begin{tabular}{|l|c|l|} \hline
%      item1 & 10 & integer \\
%      item2 & 10.0 & float \\
%      item3 & `ten' & char \\
%      item4 & 1010 & binary \\
%      item5 & $10 + i0$ & complex \\
%      \hline
%    \end{tabular}
%  \end{center}
%\end{table}

\section{Conclusion}

I have evaluated several type inference techniques for applicability
to a Clojure type system. A type inference technique was considered
applicable if it helped type-check existing Clojure idioms without
adding redundant type annotations.

The type inference strategy is an essential part of any type system
design. When adding type inference to a language with
existing idioms, the search for well fitting type inference techniques
should be as exhaustive as possible, because the type inferencing must
fit with the idioms. If the wrong type inferencing strategy is employed,
it may force the idioms of the language to change. It is risky 
changing time honoured idioms, and everything should be done to avoid
that possibility.

The evaluation of existing type inference techniques consisted of two activities: simulating Clojure idioms
in other languages, and verify limitations to algorithms. I ported a
standard Clojure function to Typed Racket and ensured it was well typed.
The overall structure of the code did not change in the porting process.
I also tested the limitations of Local Type Inference by testing
small, contrived, forms in Typed Racket which I expected to fail. The first
failed without surprise, and the second failed after some uncertainty.

These results indicate that Typed Racket's type inference strategy \cite{SAMTH:dissertation} seems a good
fit for a Clojure type system, and is well worth a prototype. The stress test,
which required the porting effort, was testing occurrence typing and
how it might behave when applied to Clojure code. The results
were pleasing which implies occurrence typing could fit Clojure code as well as
Racket code. The successful small tests of known edge cases for Local Type Inference tells us
that a satisfying type system can be built in spite of the limitations of Local Type Inference\cite{Hosoya99howgood}.
This reinforces my decision to utilise Local Type Inference \cite{Pierce:2000:LTI:345099.345100}
over the more complicated Colored Type Inference \cite{Odersky:2001:CLT:373243.360207}.

This evaluation of type inference techniques is part of a larger project to add a static
type system to Clojure. Type inference is an incredibly important aspect of a type system in terms
of actual day-to-day use. The programmer will have type inference at the tip of their fingers
at every keystroke. How many keystrokes of frustrating redundancy would you tolerate 
before throwing in the towel? The type inference strategy must fit the language like a glove; then
you are in with a fighting chance.

%\section*{Acknowledgements}
%
%Acknowledge any help you might have had, either technical of financial.

\printbibliography

\end{document}

### 2012-05-24 dissertation-proposal/research-proposal.tex
Final lit review and proposal

@@
 There is some experience statically typing multimethods
-\cite{Millstein02modularstatically}
-, but this is a challenge and may not be addressed in the
+\cite{Millstein02modularstatically}, 
+but this is a challenge and may not be addressed in the
 timeframe proposed, beyond initial consideration of how it would
@@
 	\begin{itemize}
-	\item Concrete types (no type variables)
 	\item Union types
+  \item Basic Local Type Inference algorithm
 	\item Fixed arity function types
@@
 	\item Uniform variable-arity function types
-	\item Type checking without occurance typing
 	\item Enough functions annotated from clojure.core for proof-of-concept
-	\end{itemize}
-\item[0.2]
-	\begin{itemize}
-	\item Typed defrecord (record structures)
-	\item Type variables (polymorphic types)
-	\item Most functions annotated from clojure.core library
+	\item Bounded Type variables
 	\end{itemize}
@@
 	\item Typing Multimethods
+	\item Fine Grained Hash-Map Types
 	\end{itemize}

### 2012-05-24 lit-review/lit-review.tex
Final lit review and proposal

\documentclass[12pt, a4paper]{article}
\usepackage{biblatex}
\usepackage{fancyvrb}

\usepackage{color}
\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  %showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=b,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

\bibliography{bibliography}
\title{A Practical Optional Type System for Clojure, Literary Review}
\author{Ambrose Bonnaire-Sergeant (20350292)}

\begin{document}
\maketitle

\begin{abstract}
Clojure is a practical, dynamically typed language running on the Java Virtual Machine.
I intend to design and implement a prototype optional static type 
system for Clojure, with practical use being a priority. Devising a static type system 
requires a combination of components that each adds some utility
without significantly complicating the overall implementation.
I look at several novel inventions created as a result 
of the Typed Racket project including
non-uniform variable-arity polymorphism and occurrence typing.
Typed Racket and Scala both use flavours of local type inference
which I compare and judge for applicability to this project.
Other relevant features of Scala are considered, such as its approach to bounded polymorphism,
the Scala class hierarchy, and the similarities of its methods of composition to Clojure's.
\end{abstract}

\tableofcontents

\section{Introduction}

Designing a static type system for any dynamically typed language requires
careful consideration. Idiomatic style in dynamically typed languages is usually
designed to be powerful yet non-intrusive. An effective type system
for such languages should preserve existing idioms and syntactic abstractions
until time tested idioms are devised for the new typed language \cite{SAMTH:dissertation}.

Typed Racket \cite{SAMTH:dissertation} integrates a selection of ideas to
create a typed sister language to Racket.
Both Clojure and Racket are dynamically typed Lisps and unsurprisingly have similar idioms. 
Consequently, a number of novel Typed Racket features are relevant to designing and implementing 
a type system for Clojure.
The inventions of occurrence typing \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
and a systematic approach to variable-arity polymorphism
\cite{Strickland:2009:PVP:1532974.1532978}
are both applicable to this project.

Clojure provides two dispatching mechanisms: multimethods, a mechanism to dispatch on the
results of an arbitrary function; and protocols, a mechanism for single dispatch.
Millstien and Chambers explore statically typing multiple dispatched multimethods in Dubious \cite{Millstein02modularstatically}.
Dubious' multimethods are less powerful than Clojure's multimethods but more powerful than protocols,
and Clojure's dispatch functions can be extended at runtime, unlike Dubious' multimethods.

Clojure and the statically typed language Scala \cite{Odersky06anoverview} 
have some interesting similarities. The recently redesigned 
Scala collection hierarchy
\cite{Odersky06anoverview} \cite{Scala:Collections}
bears similarity to Clojure's encompassing ``seq''
model, revealing an opportunity to test potential typed extensions to the Clojure 
class hierarchy.

Scala and Clojure are both languages with a focus on host interoperability, specifically
to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
relevant, particularly considering the potential of using occurrence typing \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
to prevent erroneous usage of \lstinline|null|.

\section{Variable-Arity Polymorphism}

Functions with variable-arity are abundant in dynamically typed languages;
it is common to have a large standard library of functions that utilize
variable-arity, and Clojure is no different. 
Support for complex variable-arity functions is sparse in typed languages,
Strickland claims \cite{Strickland:2009:PVP:1532974.1532978}
``no typed language supports them in a systematic and principled manner''.

Strickland et al. invented \cite{Strickland:2009:PVP:1532974.1532978} 
a calculus capable of typing non-uniform variable-arity functions, a version of which
is implemented in Typed Racket. A non-uniform variable-arity function is a function that
takes a heterogeneous variable parameter list, which usually has a complex relationship 
with other components of the function.
For example, Clojure's \lstinline|map| function takes a function and one or more sequences,
and returns the result of applying the function argument to each element of the sequences pair-wise.

\begin{lstlisting}[caption=An application of the non-uniform variable-arity function \lstinline|map|, label=lst:map]
(map + [1 2] [2.1 3.2]) 
;=> (3.1 5.2)
\end{lstlisting}

To statically check calls to \lstinline|map|, we must enforce the provided function argument can accept as many
arguments as there are sequence arguments to \lstinline|map|, and the parameter types of the provided function can accept
the pair-wise application of the elements in each sequence. This is a complex relationship between the variable parameters and
the rest of the function.
Listing \ref{lst:map} requires the first argument to \lstinline|map| to be a function of 2 parameters because
there are two sequence parameters. \lstinline|+| takes any number of \lstinline|Number| parameters, 
and applying pair-wise arguments of \lstinline|(Vector Long)| and \lstinline|(Vector Double)| 
results in types \lstinline|Long| and \lstinline|Double| being applied to \lstinline|+|. These are subtypes
of \lstinline|Number|, so the expression is well typed.

Clojure presents some opportunities to extend Strickland's calculus. It is idiomatic
to utilize hash-maps as ad-hoc objects in Clojure, and it would be useful to track the keys associated
with a hash-map.  The fundamental hash-map manipulation functions \lstinline|assoc| and \lstinline|dissoc|,
which associate and dissociate entries to (immutable) hash-maps, both take a variable number of arguments.

\begin{lstlisting}[caption=Using \lstinline|dissoc| with variable arguments, label=lst:assoc1]
(dissoc {:a 1, :b 2} ; => {:a 1, :b 2}
        :a           ; => {:b 2}
        :b}          ; => {}
;=> {}
\end{lstlisting}

One approach to typing \lstinline|dissoc| could be to include in every hash-map type how to dissociate exactly one
key from itself, and then treat subsequent entries passed to \lstinline|dissoc| individually in a similar manner.
This is a complex polymorphic relationship between the parameter and result types useful for typing
several core Clojure functions.

\section{Type Inference}

\subsection{Local Type Inference}

Typed Racket uses Local Type Inference \cite{Pierce:2000:LTI:345099.345100}
as an inference and checking tool. Pierce and Turner
\cite{Pierce:2000:LTI:345099.345100} divide Local Type Inference into
two complementary algorithms. \emph{Local type argument synthesis}
synthesises type arguments to polymorphic applications, and \emph{bidirectional
propagation} propagates type information both down and up the source tree,
known as \emph{checking} and \emph{synthesis} mode respectively.

\begin{lstlisting}[caption=Bidirectional checking algorithm with Typed Clojure pseudocode, label=lst:bidir]
(map (fn [a :- Long, b :- Float]
       (+ a b))
     [1 2]
     [2.1 3.2])
;=> (3.1 5.2)
\end{lstlisting}

The pseudocode in listing \ref{lst:bidir} show both algorithms in action. Local type argument synthesis is able
to infer the type arguments to \lstinline|map| by observing the argument types of the first
argument to \lstinline|map| and the types of subsequent sequence arguments. Bidirectional checking
then \emph{synthesises} the resulting type of the expression by \emph{checking} each element
of \lstinline|[1 2]| is a subtype of \lstinline|a|, and each element of \lstinline|[2.1 3.2]| is a subtype of
\lstinline|b|. The result of the anonymous function argument is \emph{synthesised} from the type of
\lstinline|(+ a b)| as \lstinline|Float|. We now have sufficient information to 
synthesise the type of listing \ref{lst:bidir} to be \lstinline|(List Float)|.

Pierce and Turner split \emph{local type argument synthesis} into two further
algorithms: bounded, and unbounded quantification \cite{Pierce:2000:LTI:345099.345100}. 
Typed Racket 
supports unbounded polymorphism \cite{SAMTH:dissertation}, implementing the latter algorithm by Piece et al.
Scala supports bounded quantification with F-bounded polymorphism \cite{Canning:1989:FPO:99370.99392},
basing its type argument synthesis on the bounded quantification algorithm.

Pierce and Turner explicitly forbid \cite{Pierce:2000:LTI:345099.345100}
attempting to synthesise type variables with interdependent bounds, including
F-bounds, having failed to devise an algorithm to infer these cases.
Scala's type argument synthesis implementation deviates from Pierce and Turner and supports these features.
I am not aware of papers specifically describing Scala's modifications, but they are at least inspired by
Scala's spiritual ancestors Generic Java \cite{Bracha98gj:extending} and Pizza \cite{Odersky97pizzainto}.

Hosoya and Pierce \cite{Hosoya99howgood} reiterate two common problems with Local Type Inference:
``hard-to-synthesise arguments'' and ``no best type argument''. The first problem occurs because
both local type argument synthesis and bidirectional propagation cannot perform synthesis
simultaneously.

\begin{lstlisting}[caption=Hard-to-synthesise expression, label=lst:hts]
(map (fn [a b] 
       (+ a b)) 
     [1 2] 
     [2.1 3.2])
\end{lstlisting}

Listing \ref{lst:hts} shows an example of this limitation,
here caused by both not providing type arguments to \lstinline|map| and not providing the parameter types of \lstinline|(fn [a b] (+ a b))|.
 Cases where both algorithms can simultaneously recover new type information are usually ``hard-to-synthesise''.
``No best type argument'' describes the situation where the results of local
type argument synthesis yield more than one type, and no type is better than the other. Sometimes we cannot recover and synthesis
fails.

\subsection{Colored Local Type Inference}

Scala's type checking uses Colored Local Type Inference \cite{Odersky:2001:CLT:373243.360207},
a variant of Local Type Inference \cite{Pierce:2000:LTI:345099.345100} specifically designed to
improve inference with certain kinds of Scala pattern matching expressions. It allows
\emph{partial} type information to propagate down the syntax tree, instead of only full type information
as required by Local Type Inference.

\emph{Colored} types contain extra contextual information, including the propagation direction
and missing parts of the type. They are generally useful
for describing ``information flow in polymorphic type systems with propagation-based type inference''
\cite{Odersky:2001:CLT:373243.360207}. If the scope of my research permits it, I plan to investigate
using colored types to solve the polymorphic higher-order-function limitation of Local Type Inference
(see listing \ref{lst:hts}).

\section{Bounded and Unbounded Polymorphism}

The Local Type Inference paper \cite{Pierce:2000:LTI:345099.345100}
describes two implementations of type variables, for bounded
and unbounded type variables. The bounded implementation is presented
as an optional extension  to the unbounded implementation, which preserves all
properties described in the Local Type Inference algorithm.

An unbounded type variable does not have subtype constraints.
Bounded type variables can have subtype constraints, and 
subsume unbounded type variables \cite{Pierce:2000:LTI:345099.345100}, 
as a unbounded variable can be represented as a variable bounded
by the Top type.

Still, unbounded type variables have an advantage: their implementations are
are simpler in the presence of a \emph{Bottom} type. 
The constraint resolution algorithm for bounded variables
is more subtle, due to ``some surprising interactions between bounded quantifiers
and the \emph{Bot} type'' \cite{Pierce:2000:LTI:345099.345100}, described fully
by Pierce \cite{Pierce97boundedquantification}.

Typed Racket \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
supports unbounded polymorphism, while Scala \cite{Odersky06anoverview}
supports an extended form of bounded polymorphism called
F-bounded polymorphism \cite{Canning:1989:FPO:99370.99392}, which allows the
bound variable to occur in its own bound.
F-bounded polymorphism is useful in the context of object-oriented abstractions,
as demonstrated by Odersky \cite{Odersky06anoverview}.
This is one possible explanation why Typed Racket, which is not built on abstractions like Scala,
does not support bounded quantification. Unfortunately, no Typed Racket paper mentions 
bounded quantification, so the rationale is not clear.

Clojure, like Scala, is built on object-oriented abstractions. Clojure protocols
and Java interfaces (interfaces are supported by Clojure) are good candidates
for bounds in bounded or F-bounded polymorphism.

\section{Typed Racket}

Typed Racket is a statically typed sister language of Racket. It
attempts to preserve existing Racket idioms and aims type check
existing Racket code by simply adding top level type annotations \cite{SAMTH:dissertation}.

Typed Racket fully expands all macro calls before type checking \cite{SAMTH:dissertation} to
avoid the complex semantics of type checking macro definitions, an ongoing research area summarised
 by Herman \cite{Herman10:Theory}.
The design of my Clojure type system will follow a similar strategy; only the fully macro-expanded form
will be type checked. Type checking macro definitions are outside the scope of this project, and would
be exceptionally hard.

Along with a full static type system, Typed Racket 
also uses runtime contracts to enforce type invariants at runtime \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}.
Utilising runtime contracts to aid type checking is outside the scope of this project, but would be 
considered desirable and accessible future work.

Two other Typed Racket features that will be explored are recursive types and refinement types  
\cite{SAMTH:dissertation}. Recursive types allow a type definition to refer to itself, enabling structurally
recursive types like binary trees. Refinement types let the programmer define
new types that are subsets of existing types, such as the type for even integers, a subset of all integers.
Both these features would fit well in a future implementation of this project.

\section{Occurrence Typing}
\label{sec:OccurrenceTyping}

Dynamically typed languages use an ad-hoc combination of type predicates,
selectors, and conditionals to steer execution flow and reason about runtime types of variables.
Typed Racket uses occurrence typing to capture these ad-hoc type refinements.
For example, listing \ref{lst:occ1} shows occurrence typing following the implications 
of the type predicate \emph{number?} and the selector \emph{first}, and utilises those implications to refine
the type of \lstinline|x|. If the test at line 3 succeeds, occurrence typing refines the
type of \lstinline|(first x)| to be \lstinline|Number|, which allows \lstinline|(+ 1 (first x))|
to be well typed. Similarly at line 4, we can be sure that \lstinline|(first x)| is
a \lstinline|String|, since we have ruled out the case of being a \lstinline|Number|.

\begin{lstlisting}[caption=A well typed form utilising occurrence typing with Clojure syntax, label=lst:occ1]
(let [x (list (number-or-string))]
  (cond 
    (number? (first x)) (+ 1 (first x))
    :else               (str (first x))))
\end{lstlisting}

Occurrence typing \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561} extends the type 
system with a \emph{proposition environment} that represents 
the type refinements inferred down a particular path.
The 2010 paper \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
reformulated occurrence typing, improving the original 2008 paper
\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
after revealing that ``three years of practical experience has revealed
serious shortcoming of our type system.''\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
The new implementation utilizes a simple proof system to solve
propositional logic statements, in terms of type predicates and selectors.

For occurrence typing to infer propositions from type predicate usages, it requires 
two extra annotations: a ``then'' proposition
when the result is a true value, and an ``else'' proposition for a false value.
For example, \lstinline|number?| has a ``then'' proposition that says its argument
is of type \lstinline|Number|, and an ``else'' proposition that says its argument is not of type \lstinline|Number|.

An exciting application of occurrence typing as yet unexplored is facilitating null-safe interop with Java.
By declaring \lstinline|nil| (Clojure's value of Java's \lstinline|null|) to \emph{not} be a subtype of reference types,
we can begin to statically disallow potential inconsistent usages of \lstinline|nil| as part of the type system.

\begin{lstlisting}[caption=Observing nil-checks using occurrence typing, label=lst:nil]
(let [a (ObjectFactory/getObject)]
  (when a
    (expects-non-nil a)))
\end{lstlisting}

Listing \ref{lst:nil} infers from the Java signature \lstinline|Object getObject()| that
that \lstinline|a| is of type \lstinline|(Union nil Object)|. This is equivalent to Java's
\lstinline|Object| class, as \lstinline|null| is a subtype of all reference types. By surrounding
the call \lstinline|(expects-non-nil a)| with \lstinline|(when a ...)|, we guarantee that
\lstinline|a| is non-nil when passed to \lstinline|expects-non-nil|. Occurrence typing infers
this by observing \lstinline|nil| is a false value in Clojure, therefore \lstinline|a| cannot
be \lstinline|nil| the body of the \lstinline|when|, refining \lstinline|a|'s type to \lstinline|Object|
from \lstinline|(Union nil Object)|.

Occurrence typing is a relatively simple, time worn technique used successfully 
in Typed Racket. Clojure is similar enough to Racket for occurrence typing to work
without issues, and has good potential to enable null-safe Java interop.

\section{Statically Typed Multimethods}

A Clojure multimethod is slightly different to most other
abstractions of the same name. Here, a multimethod 
dispatches on the result of an \emph{arbitrary function},
rather that more traditional multimethods that dispatch on
the types or structure of its arguments.

Millstein and Chambers \cite{Millstein02modularstatically}
describe Dubious, a simple statically typed core language including multimethods that
dispatch on the type of its arguments. They tackle a key challenge for statically typing
multimethods: ``it is possible for two modules containing arbitrary multimethods to typecheck
successfully in isolation but generate type errors when linked together.'' \cite{Millstein02modularstatically}

This challenge holds when statically typing Clojure multimethods.
One additional challenge would be multimethods being imported from 
untyped modules. This would have to be handled carefully.

If we consider a multimethod as a spread out conditional, a
novel use of occurrence typing becomes apparent. We can use
occurrence typing to refine the argument types in the method bodies
by utilizing the results of the dispatch function. This is simple
when dispatching by type, but a common idiom in Clojure is to dispatch
on the value associated with a key in a hash-map.

To satisfy this idiom, we require occurrence typing with more
finely grained hash-map types (see section~\ref{sec:OccurrenceTyping} for discussion).

\section{Conclusion}

Many related components must come together in the design of a
static type system. Typed Racket achieves a satisfying balance of 
occurrence typing, local type inference and variable-arity polymorphism.
Similarly, Scala features F-bounded polymorphism, a class hierarchy
that is compatible with Java, and colored local type inference.

I intend to design a static type system for Clojure
that features a satisfying combination of features, individually taken from other
languages, when combined produce novel and interesting problems.
I have looked at the limitations of Local Type Inference and identified Colored Local
Type Inference as potentially improving them.
I have proposed a novel use of occurrence typing to prevent erroneous use of Java's
null pointer, and identified a novel problem typing certain Clojure functions with variable
arguments.
There are some interesting problems and novel solutions to be discovered while undertaking this
project, many of which will reveal themselves during development.

\printbibliography

\end{document}

### 2012-08-26 dissertation/paper.tex
Chapter 2 draft. UWA cshonours formatting.

\documentclass{cshonours}
\bibliographystyle{acm}

%\usepackage{biblatex}

\title{Typed Clojure}
\author{Ambrose Bonnaire-Sergeant}

\keywords{Type systems}

\bibliography{bibliography}

\usepackage{color}
\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  %numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=t,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

\begin{document}
\maketitle

\chapter{Introduction}

This dissertation describes Typed Clojure, an optional static type system for Clojure.

% Based on Typed Racket

\chapter{Typed Clojure through Examples}

% Hello world
This section introduces Typed Clojure with example code. An attempt is
made to introduce some Clojure syntax and sematics to those unfamiliar or needing a refresher.
A basic knowledge of Lisp syntax is assumed.

We begin with the obligatory \emph{Hello world} example.

\begin{lstlisting}[caption=Hello world]
(ns typed.test.hello-world
  (:require [typed.core :refer [check-ns]]]))

(println "Hello world")
\end{lstlisting}

The \lstinline|clojure.core/ns|
\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
invocation declares the current namespace to be \lstinline|typed.test.hello-world|
and requires the namespace \lstinline|typed.core| to be loaded as a dependency, referring the var
\lstinline|typed.core/check-ns| to be referenced locally. Other than this
dependency, this is identical to the untyped \emph{Hello world}.
% footnote: check-ns used to initiate type checking usually at REPL

More complex code require extra annotations to type check:

\begin{lstlisting}
(ns typed.test.collatz
  (:require [typed.core :refer [check-ns ann]]))

(ann collatz [Number -> Number])
(defn collatz [n]
  (cond
    (= 1 n) 
      1
    (and (integer? n) 
         (even? n)) 
      (collatz (/ n 2))
    :else 
      (collatz (inc (* 3 n)))))
\end{lstlisting}
\footnote{Example adapted from Sam's dissertation TODO}

In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
to be a function from \lstinline{java.lang.Number} to 
\lstinline{java.lang.Number}
\footnote{All Classes in the \lstinline|java.lang| package
are automatically imported in every Clojure namespace (for those familiar with Java, the equivalent of \lstinline|import java.lang.*;|).}.

\section{Datatypes and Protocols}

We can annotate datatype and protocol definitions similarly.

\begin{lstlisting}
(ns typed.test.deftype
  (:require [typed.core 
             :refer [check-ns ann-datatype
                     tc-ignore ann-protocol AnyInteger]]))

(ann-protocol Age 
  :methods
  {age [Age -> AnyInteger]})
(tc-ignore
  (defprotocol Age
    (age [this])))

(ann-datatype Person 
  [[name :- String]
   [age :- AnyInteger]])
(deftype Person [name age]
  Age
  (age [this] age))

(age (Person. "Lucy" 34))
\end{lstlisting}

\lstinline|clojure.core/defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols}
with a set of methods. \lstinline|typed.core/ann-protocol| annotates a protocol with the types of its methods.
In this example, we define a protocol \lstinline|typed.test.person/Age| with an \lstinline|age| method.
The call to \lstinline|clojure.core/defprotocol| also defines a new var \lstinline|typed.test.person/age|, a first-class function
wrapping the \lstinline|age| method, but taking the target Object as the first parameter. The
type signature provided with \lstinline|typed.core/ann-protocol|, here \lstinline|[typed.test.person/Age -> typed.core/AnyInteger]|, 
is for this function.

Invocations of \lstinline|clojure.core/defprotocol| are currently not able to be type checked
and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.

\lstinline|clojure.core/deftype|
defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes}
in the current namespace with a number of fields and methods. 
\lstinline|typed.core/ann-datatype| annotates a datatype with its field types.
In this example, we create a datatype \lstinline|typed.test.person.Person|
with fields \lstinline|name| and \lstinline|age| and implement the \lstinline|age|
method from protocol \lstinline|typed.test.person/Age|.

Java constructors are invoked in Clojure by suffixing the Class we want to instantiate with a dot.
Datatypes are implemented as Java Classes with immutable fields (by default) and a single constructor, taking as arguments its fields 
in the order they are passed to \lstinline|deftype|
\footnote{When unambiguous, I omit the qualifying namespace/package for the remainder of the chapter.}.
\lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
instance, setting the fields to their corresponding positional arguments.
Typed Clojure checks the datatype constructor to be the equivalent of 
\lstinline|[String AnyInteger -> Person]|.

Finally, Typed Clojure checks invocations of Protocol methods. It infers \lstinline|Person|
is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.

\section{Polymorphism}

Typed Clojure supports F-bounded polymorphism. All type variables may have
upper and lower bounds.

Typed Clojure parameterises some of Clojure's data structures. For example,
the interface behind Clojure's \emph{seq} abstraction \lstinline|clojure.lang.Seqable| has one 
covariant parameter\footnote{\lstinline|(Seqable Integer)| being a subtype of \lstinline|(Seqable Number)|
because Integer is a subtype of Number.}.

\begin{lstlisting}
...
(ann to-set 
     (All [x]
       [(U nil (Seqable x)) -> (clojure.lang.PersistentHashSet x)]))
(defn to-set [a]
  (set a))
...
\end{lstlisting}

In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}
, we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
\lstinline|All| introduces a set of type variables to the body of a type,
here \lstinline|x| is used to define a relationship between the input type and return type.

\lstinline|(U nil (Seqable x))| is a common type in Clojure, read as the union
of the singleton type \lstinline|nil| and the type \lstinline|(Seqable x)|.
The vast majority of types for collection processing functions in the Clojure core library feature
this as an input type, where passing \lstinline|nil| either has some special behaviour 
or is synonymous with passing an empty \lstinline|Seqable|.

\section{Heterogeneous Maps}

Where particularly object-oriented languages reach for objects, Clojure
utilises maps. Clojure provides a map literal using curly braces. For example,
\lstinline|{:a 1, :b 2}| is a map with two key-value entries: from Keyword key \lstinline|:a|
to value \lstinline|1|, and Keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
whitespace in Clojure and are included for readability.

Typed Clojure provides a heterogeneous map type, restricted to 
maps with singleton Keyword keys. This restriction is reflected
in the syntax for defining heterogeneous map types.

\begin{lstlisting}
...
(ann config
     (HMap {:file String,
            :ns Symbol}))
(def config
  {:file "clojure/core.clj",
   :ns 'clojure.core})
...
\end{lstlisting}

This example checks \lstinline|config| to be a heterogeneous map
with \lstinline|:file| and \lstinline|:ns| keys, with values of
type \lstinline|String| and \lstinline|Symbol| respectively.

Heterogeneous vector and seq types are also provided.

\section{Multi-Arity Functions}

All Clojure functions support multiple arities, allowing different
behaviour for different number of arguments.

\begin{lstlisting}
(ns typed.test.poly
  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
  (:import [clojure.lang Seqable]))

(ann repeatedly'
     (All [x]
       (Fn [[-> x] -> (Seqable x)]
           [AnyInteger [-> x] -> (Seqable x)])))
(defn repeatedly'
  "Takes a function of no args, presumably with side effects, and
  returns an infinite (or length n if supplied) lazy sequence of calls
  to it"
  ([f] (lazy-seq (cons (f) (repeatedly' f))))
  ([n f] (take n (repeatedly' f))))
\end{lstlisting}

This example defines \lstinline|repeatedly'|\footnote{Identical to \lstinline|clojure.core/repeatedly|},
a multi-arity function taking either a zero-argument function, or an integer and a zero-argument function as arguments.
The type variable \lstinline|x| links the return type of the generator function to
the resulting \lstinline|Seqable| instance.

When checking the definition of multi-arity functions, Typed Clojure
finds the matching function types for a particular arity by number of arguments
and requires each relevant type to check successfully.

\section{Variable-Arity Functions}

Clojure functions can take an optional \emph{rest} argument.

\subsection{Uniform Variable-Arity}

\subsection{Non-uniform Variable-Arity}

\section{Occurrence Typing}

Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
Occurrence typing is used to refine the known types for immutable local bindings 
down conditional branches.

\begin{lstlisting}
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [a b]
  [(if a a 0) 
   (if b b 0)])
\end{lstlisting}

To check this example, occurrence typing infers type information based on the result of each test.
In Clojure, \lstinline|nil| and \lstinline|false| are false values and all other values are true.
The test \lstinline|(if a a 0)| follows the \emph{then} branch is \lstinline|a| is not \lstinline|nil|
or \lstinline|false|. When checking this branch, we can safely refine the type of \lstinline|a| to \lstinline|Number| from
\lstinline|(U nil Number)|. Similarly, following the \emph{else} branch refines the type of \lstinline|a|
to \lstinline|nil| from \lstinline|(U nil Number)|.

\section{Java Interoperability}

Typed Clojure currently provides safe interop\footnote{Interop is short for interoperability} with (non-Generic) Java
\footnote{Compatibility with Generic Java is planned future work}.

The key to safe Java interop is the treatment of Java's \emph{null}.
\emph{null} is a subtype to all reference types
represented in Clojure by \lstinline|nil|. Unlike Java's type system
Clojure explicitly separates \emph{null} and reference types, allowing
Typed Clojure to express
\emph{nullable}
\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
positions.

\begin{lstlisting}
(ns typed.test.interop
  (:import (java.io File))
  (:require [typed.core :refer [ann non-nil-return check-ns]]))

(ann f File)
(def f (File. "a"))

(ann prt (U nil String))
(def prt (.getParent ^File f))

(non-nil-return java.io.File/getName :all)
(ann nme String)
(def nme (.getName ^File f))

\end{lstlisting}

This example shows how Typed Clojure handles \emph{null} while creating and
using an instance of \emph{java.io.File}\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.

Typed Clojure checks calls to Java constructors by requiring the provided
arguments be acceptable input to at least one constructor for that Class.
In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
argument, so \lstinline|(File. "a")| is type safe. By design, constructors never
return \emph{null}, so Clojure assigns the return type to be \lstinline|File|.
This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.

Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
The \emph{java.io.File} instance method \emph{getParent}
is equivalent to \lstinline|[-> (U nil String]| in Typed Clojure. This happens to be
a valid approximation of the method as \emph{getParent} returns \emph{null} 
``if the pathname does not name a parent directory''\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
On the other hand, the instance method \emph{getName} always returns an
instance of \emph{java.lang.String}, so we set the return position of
\emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
\footnote{The second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
to override all arities of that method.}.

\chapter{Design Goals}

\section{Safer Java Interop}
\section{Optional Type Checking}
\section{Explicit Typing}
\section{No new idioms}

\section{Local Type Inference}

Similar to Typed Racket, type inference is based on Local Type Inference
by Pierce and Turner.

\section{F-bounded Polymorphism}

Typed Clojure includes support for F-bounds on type variables, as an extension
to Local Type Inference.

An extra environment from type variables to bounds is kept until after inference,
after which each unground bound is substituted with the inferred types and the
subtyping relationships are checked.

\section{Heterogeneous Maps}
% keyword keys for now, generally encouraged, fast
% Some heterogeneous maps have string args, eg. from web frameworks

\section{Java Interoperability}
% Args non-nil, return nilable
% Constructors and other special methods handled appropriately (eg. Constructor are non-nil return)

\subsection{Primitive Arrays}
% Primitive arrays are covariant, which is well-known to be statically unsound.
% We cannot trust the type signature of any Java method or field.
% Separate read and write types into two parameters.

\subsection{Interaction with null}
% null is explicitly expressible in my type system
% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
%

\subsection{Generic Java}
% Existential types are needed to fully express Generic Java types in Typed Clojure
% F-bounded polymorphism supported

\section{Occurrence Typing}
% interesting paths
% - count path element
% - class path element
% - keyword path element

\section{Variable-arity Polymorphism}
% Problematic areas
% - flattened paired arguments. assoc, hash-maps, array-map
% - complex argument dependencies. comp, partial

\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}

Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
representing a sequence of types, allowing non-uniform variable parameters.
This enables \lstinline|map| and other function with complex arguments to be typed.
Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
Typed Racket, and is a key feature of Typed Racket's implementation.

Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
associates key-value pairs to a target map. It takes one or more key-value pairs and
applies them left-to-right to the target map. The key problem from a typing perspective
is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
higher-order function.
\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
dotted parameters represent an unrestricted number of types, where we require an even number.

The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
Firstly, subtyping between different kinds of dotted and rest
\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
difficult to express.
Thirdly, higher-order functions taking functions with dotted arguments commonly 
The first issue requires careful design and implementation.
The second and third points require new constructs, in the form of \emph{projections}
and abstractions over \emph{dotted} parameters, respectively.

\section{Implementation}

\subsection{Composite Propositions}

\subsection{Non-hygienic Macro Expansion and Filters}
% need to remove bindings as they fall out of scope

\section{Future Work}

\subsection{Multimethods}
% Problem - typing isa?
% - probably impossible to type higher order usages of isa?.
% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
% Statically proving at least one relevant dispatch method is present.
% Warn potential method conflicts.
% Suppress warnings, similar to prefer-method syntax

\subsection{Records}
% A composition of existing features: Datatypes and Heterogeneous Maps

\subsection{Other Clojure Dialects}
% unify ASTs for easier porting

\subsection{Linter}

%
%\begin{mathpar}
%\inferrule*[right=B-Var]
%  {\rho(x) = v}
%  {\rho \vdash x \downarrow v}
%
%\inferrule*[right=B-Assoc]
%  {\rho \vdash e_m \downarrow mi \\
%   \rho \vdash e_k \downarrow k \\
%   \rho \vdash e_v \downarrow v \\
%   \rho \vdash \overrightarrow{e_j}^n \downarrow \overrightarrow{k_j}^n \\
%   \rho \vdash \overrightarrow{e_m}^n \downarrow \overrightarrow{k_m}^n \\
%   \delta(add\_entry, mi, k, v, \overrightarrow{k_j, k_m}^n) = mo}
%  {\rho \vdash (assoc\ e_m\ e_k\ e_v\ \overrightarrow{e_j\ e_m}^n) \downarrow mo}

%\infrule[B-Var]
%  {\rho(x) = v}
%  {\rho \vdash x \downarrow v}
%
%\infrule[B-Assoc]

%\end{mathpar}

\end{document}

### 2012-09-11 dissertation/design-choices.tex
Rearrange into a file per chapter

--- 2012-08-26 dissertation/paper.tex
@@
-\documentclass{cshonours}
-\bibliographystyle{acm}
+\chapter{Design Choices}
 
-%\usepackage{biblatex}
+Typed Clojure is designed to be of practical use to Clojure programmers.
+Many of the design goals are similar to Tobin-Hochstadt's \cite{SAMTH:dissertation}
+for Typed Racket.
 
-\title{Typed Clojure}
-\author{Ambrose Bonnaire-Sergeant}
+\section{Typed Racket}
 
-\keywords{Type systems}
+The majority of the design and implementation of Typed Clojure is based on Typed Racket.
 
-\bibliography{bibliography}
+\subsection{Occurrence Typing}
+% interesting paths
+% - count path element
+% - class path element
+% - keyword path element
 
-\usepackage{color}
-\usepackage{listings}
-\lstset{ %
-  language=Lisp,                % choose the language of the code
-  columns=fixed,basewidth=.5em,
-  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-  %numbers=left,                   % where to put the line-numbers
-  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
-  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-  %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  %showspaces=false,               % show spaces adding particular underscores
-  showstringspaces=false,         % underline spaces within strings
-  %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
-  %tabsize=2,          % sets default tabsize to 2 spaces
-  captionpos=t,           % sets the caption-position to bottom
-  breaklines=true,        % sets automatic line breaking
-  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
-  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
-}
+\subsection{Variable-arity Polymorphism}
+% Problematic areas
+% - flattened paired arguments. assoc, hash-maps, array-map
+% - complex argument dependencies. comp, partial
 
-\begin{document}
-\maketitle
+\section{Safer Host Interoperability}
+% Args non-nil, return nilable
+% Constructors and other special methods handled appropriately (eg. Constructor are non-nil return)
+% TODO refine language Clojure vs. Clojure JVM
 
-\chapter{Introduction}
+% what is this paragraph saying?
+Dialects of Clojure provide fast and unrestricted access to their host platform. 
+This is a intentional source of incompatibility between dialects.
+For example, Clojure is hosted on the Java Virtual Machine (JVM), ClojureCLR on the Common Language Runtime (CLR),
+and ClojureScript on Javascript VMs. These are very different platforms, and it is unlikely
+a Clojure dot call will be portable. There is no attempt at reconciling
+host interoperability differences, and it is up to the programmer to decide
+how to best abstract over different hosts.
 
-This dissertation describes Typed Clojure, an optional static type system for Clojure.
+Typed Clojure targets the JVM hosted Clojure.
+Clojure embraces the JVM as a host by sharing its runtime type system and encouraging direct
+interaction with libraries written in other JVM languages. Most commonly, Clojure programmers
+reach to Java libraries. Java is statically typed language. It follows that any interaction with
+Java will already have an annotated Java type. Typed Clojure can use these annotations
+to statically type check these interactions.
 
-% Based on Typed Racket
+Typed Clojure attempts to improve some shortcomings of Java's type system.
 
-\chapter{Typed Clojure through Examples}
+Java's static type system does not provide a type-safe construct for eliminating
+the \lstinline|null| pointer. Instead, Java programmers often rely on either testing
+for \lstinline|null| or prior domain knowledge.
 
-% Hello world
-This section introduces Typed Clojure with example code. An attempt is
-made to introduce some Clojure syntax and sematics to those unfamiliar or needing a refresher.
-A basic knowledge of Lisp syntax is assumed.
-
-We begin with the obligatory \emph{Hello world} example.
-
-\begin{lstlisting}[caption=Hello world]
-(ns typed.test.hello-world
-  (:require [typed.core :refer [check-ns]]]))
-
-(println "Hello world")
-\end{lstlisting}
-
-The \lstinline|clojure.core/ns|
-\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
-invocation declares the current namespace to be \lstinline|typed.test.hello-world|
-and requires the namespace \lstinline|typed.core| to be loaded as a dependency, referring the var
-\lstinline|typed.core/check-ns| to be referenced locally. Other than this
-dependency, this is identical to the untyped \emph{Hello world}.
-% footnote: check-ns used to initiate type checking usually at REPL
-
-More complex code require extra annotations to type check:
-
-\begin{lstlisting}
-(ns typed.test.collatz
-  (:require [typed.core :refer [check-ns ann]]))
-
-(ann collatz [Number -> Number])
-(defn collatz [n]
-  (cond
-    (= 1 n) 
-      1
-    (and (integer? n) 
-         (even? n)) 
-      (collatz (/ n 2))
-    :else 
-      (collatz (inc (* 3 n)))))
-\end{lstlisting}
-\footnote{Example adapted from Sam's dissertation TODO}
-
-In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
-used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
-to be a function from \lstinline{java.lang.Number} to 
-\lstinline{java.lang.Number}
-\footnote{All Classes in the \lstinline|java.lang| package
-are automatically imported in every Clojure namespace (for those familiar with Java, the equivalent of \lstinline|import java.lang.*;|).}.
-
-\section{Datatypes and Protocols}
-
-We can annotate datatype and protocol definitions similarly.
-
-\begin{lstlisting}
-(ns typed.test.deftype
-  (:require [typed.core 
-             :refer [check-ns ann-datatype
-                     tc-ignore ann-protocol AnyInteger]]))
-
-(ann-protocol Age 
-  :methods
-  {age [Age -> AnyInteger]})
-(tc-ignore
-  (defprotocol Age
-    (age [this])))
-
-(ann-datatype Person 
-  [[name :- String]
-   [age :- AnyInteger]])
-(deftype Person [name age]
-  Age
-  (age [this] age))
-
-(age (Person. "Lucy" 34))
-\end{lstlisting}
-
-\lstinline|clojure.core/defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols}
-with a set of methods. \lstinline|typed.core/ann-protocol| annotates a protocol with the types of its methods.
-In this example, we define a protocol \lstinline|typed.test.person/Age| with an \lstinline|age| method.
-The call to \lstinline|clojure.core/defprotocol| also defines a new var \lstinline|typed.test.person/age|, a first-class function
-wrapping the \lstinline|age| method, but taking the target Object as the first parameter. The
-type signature provided with \lstinline|typed.core/ann-protocol|, here \lstinline|[typed.test.person/Age -> typed.core/AnyInteger]|, 
-is for this function.
-
-Invocations of \lstinline|clojure.core/defprotocol| are currently not able to be type checked
-and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.
-
-\lstinline|clojure.core/deftype|
-defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes}
-in the current namespace with a number of fields and methods. 
-\lstinline|typed.core/ann-datatype| annotates a datatype with its field types.
-In this example, we create a datatype \lstinline|typed.test.person.Person|
-with fields \lstinline|name| and \lstinline|age| and implement the \lstinline|age|
-method from protocol \lstinline|typed.test.person/Age|.
-
-Java constructors are invoked in Clojure by suffixing the Class we want to instantiate with a dot.
-Datatypes are implemented as Java Classes with immutable fields (by default) and a single constructor, taking as arguments its fields 
-in the order they are passed to \lstinline|deftype|
-\footnote{When unambiguous, I omit the qualifying namespace/package for the remainder of the chapter.}.
-\lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
-instance, setting the fields to their corresponding positional arguments.
-Typed Clojure checks the datatype constructor to be the equivalent of 
-\lstinline|[String AnyInteger -> Person]|.
-
-Finally, Typed Clojure checks invocations of Protocol methods. It infers \lstinline|Person|
-is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.
-
-\section{Polymorphism}
-
-Typed Clojure supports F-bounded polymorphism. All type variables may have
-upper and lower bounds.
-
-Typed Clojure parameterises some of Clojure's data structures. For example,
-the interface behind Clojure's \emph{seq} abstraction \lstinline|clojure.lang.Seqable| has one 
-covariant parameter\footnote{\lstinline|(Seqable Integer)| being a subtype of \lstinline|(Seqable Number)|
-because Integer is a subtype of Number.}.
-
-\begin{lstlisting}
+\begin{lstlisting}[caption=null elimination in Java]
 ...
-(ann to-set 
-     (All [x]
-       [(U nil (Seqable x)) -> (clojure.lang.PersistentHashSet x)]))
-(defn to-set [a]
-  (set a))
+Object a = nextObject();
+if (a != null)
+  parseNextObject(a);
+else
+  throw new Exception("null Pointer Found!);
 ...
@@
 
-In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}
-, we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
-\lstinline|All| introduces a set of type variables to the body of a type,
-here \lstinline|x| is used to define a relationship between the input type and return type.
+In Java, \lstinline|null| is a subtype to all reference
+types, but \lstinline|null| is not expressible as a static type. In other words,
+when a Java type claims a reference type, you should also assume it can also be of 
+type \lstinline|null|.
 
-\lstinline|(U nil (Seqable x))| is a common type in Clojure, read as the union
-of the singleton type \lstinline|nil| and the type \lstinline|(Seqable x)|.
-The vast majority of types for collection processing functions in the Clojure core library feature
-this as an input type, where passing \lstinline|nil| either has some special behaviour 
-or is synonymous with passing an empty \lstinline|Seqable|.
+Scala provides the Option type for safe \lstinline|null| elimination, forcing pattern matching
+where \lstinline|null| might occur.
 
-\section{Heterogeneous Maps}
-
-Where particularly object-oriented languages reach for objects, Clojure
-utilises maps. Clojure provides a map literal using curly braces. For example,
-\lstinline|{:a 1, :b 2}| is a map with two key-value entries: from Keyword key \lstinline|:a|
-to value \lstinline|1|, and Keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
-whitespace in Clojure and are included for readability.
-
-Typed Clojure provides a heterogeneous map type, restricted to 
-maps with singleton Keyword keys. This restriction is reflected
-in the syntax for defining heterogeneous map types.
-
-\begin{lstlisting}
-...
-(ann config
-     (HMap {:file String,
-            :ns Symbol}))
-(def config
-  {:file "clojure/core.clj",
-   :ns 'clojure.core})
-...
+\begin{lstlisting}[caption=null elimination in Scala]
+TODO
 \end{lstlisting}
 
-This example checks \lstinline|config| to be a heterogeneous map
-with \lstinline|:file| and \lstinline|:ns| keys, with values of
-type \lstinline|String| and \lstinline|Symbol| respectively.
+Clojure
 
-Heterogeneous vector and seq types are also provided.
+\subsection{Primitive Arrays}
+% Primitive arrays are covariant, which is well-known to be statically unsound.
+% We cannot trust the type signature of any Java method or field.
+% Separate read and write types into two parameters.
 
-\section{Multi-Arity Functions}
+\subsection{Interaction with null}
+% null is explicitly expressible in my type system
+% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
+%
 
-All Clojure functions support multiple arities, allowing different
-behaviour for different number of arguments.
+\subsection{Generic Java}
+% Existential types are needed to fully express Generic Java types in Typed Clojure
+% F-bounded polymorphism supported
 
-\begin{lstlisting}
-(ns typed.test.poly
-  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
-  (:import [clojure.lang Seqable]))
-
-(ann repeatedly'
-     (All [x]
-       (Fn [[-> x] -> (Seqable x)]
-           [AnyInteger [-> x] -> (Seqable x)])))
-(defn repeatedly'
-  "Takes a function of no args, presumably with side effects, and
-  returns an infinite (or length n if supplied) lazy sequence of calls
-  to it"
-  ([f] (lazy-seq (cons (f) (repeatedly' f))))
-  ([n f] (take n (repeatedly' f))))
-\end{lstlisting}
-
-This example defines \lstinline|repeatedly'|\footnote{Identical to \lstinline|clojure.core/repeatedly|},
-a multi-arity function taking either a zero-argument function, or an integer and a zero-argument function as arguments.
-The type variable \lstinline|x| links the return type of the generator function to
-the resulting \lstinline|Seqable| instance.
-
-When checking the definition of multi-arity functions, Typed Clojure
-finds the matching function types for a particular arity by number of arguments
-and requires each relevant type to check successfully.
-
-\section{Variable-Arity Functions}
-
-Clojure functions can take an optional \emph{rest} argument.
-
-\subsection{Uniform Variable-Arity}
-
-\subsection{Non-uniform Variable-Arity}
-
-\section{Occurrence Typing}
-
-Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
-Occurrence typing is used to refine the known types for immutable local bindings 
-down conditional branches.
-
-\begin{lstlisting}
-(ann num-vec2 
-     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
-(defn num-vec2 [a b]
-  [(if a a 0) 
-   (if b b 0)])
-\end{lstlisting}
-
-To check this example, occurrence typing infers type information based on the result of each test.
-In Clojure, \lstinline|nil| and \lstinline|false| are false values and all other values are true.
-The test \lstinline|(if a a 0)| follows the \emph{then} branch is \lstinline|a| is not \lstinline|nil|
-or \lstinline|false|. When checking this branch, we can safely refine the type of \lstinline|a| to \lstinline|Number| from
-\lstinline|(U nil Number)|. Similarly, following the \emph{else} branch refines the type of \lstinline|a|
-to \lstinline|nil| from \lstinline|(U nil Number)|.
-
-\section{Java Interoperability}
-
-Typed Clojure currently provides safe interop\footnote{Interop is short for interoperability} with (non-Generic) Java
-\footnote{Compatibility with Generic Java is planned future work}.
-
-The key to safe Java interop is the treatment of Java's \emph{null}.
-\emph{null} is a subtype to all reference types
-represented in Clojure by \lstinline|nil|. Unlike Java's type system
-Clojure explicitly separates \emph{null} and reference types, allowing
-Typed Clojure to express
-\emph{nullable}
-\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
-which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
-positions.
-
-\begin{lstlisting}
-(ns typed.test.interop
-  (:import (java.io File))
-  (:require [typed.core :refer [ann non-nil-return check-ns]]))
-
-(ann f File)
-(def f (File. "a"))
-
-(ann prt (U nil String))
-(def prt (.getParent ^File f))
-
-(non-nil-return java.io.File/getName :all)
-(ann nme String)
-(def nme (.getName ^File f))
-
-\end{lstlisting}
-
-This example shows how Typed Clojure handles \emph{null} while creating and
-using an instance of \emph{java.io.File}\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
-
-Typed Clojure checks calls to Java constructors by requiring the provided
-arguments be acceptable input to at least one constructor for that Class.
-In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
-argument, so \lstinline|(File. "a")| is type safe. By design, constructors never
-return \emph{null}, so Clojure assigns the return type to be \lstinline|File|.
-This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.
-
-Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
-The \emph{java.io.File} instance method \emph{getParent}
-is equivalent to \lstinline|[-> (U nil String]| in Typed Clojure. This happens to be
-a valid approximation of the method as \emph{getParent} returns \emph{null} 
-``if the pathname does not name a parent directory''\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
-On the other hand, the instance method \emph{getName} always returns an
-instance of \emph{java.lang.String}, so we set the return position of
-\emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
-\footnote{The second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
-returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
-to override all arities of that method.}.
-
-\chapter{Design Goals}
-
-\section{Safer Java Interop}
 \section{Optional Type Checking}
-\section{Explicit Typing}
-\section{No new idioms}
-
 \section{Local Type Inference}
@@
 
-\section{Java Interoperability}
-% Args non-nil, return nilable
-% Constructors and other special methods handled appropriately (eg. Constructor are non-nil return)
+% Operational semantics
 
-\subsection{Primitive Arrays}
-% Primitive arrays are covariant, which is well-known to be statically unsound.
-% We cannot trust the type signature of any Java method or field.
-% Separate read and write types into two parameters.
+\subsection{Operational Semantics}
 
-\subsection{Interaction with null}
-% null is explicitly expressible in my type system
-% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
-%
+$$
+\begin{tdisplay}{Test1}
+\begin{altgrammar}
+  \s{},\t{} &::= & \int \alt \bool \alt {\proctype {\s{}} {\t{}}} 
+  &\mbox{Types}\\
+  \M{}, \N{}, \P{}, \dots &::=& \x{} \alt \abs{\x{}}{\M{}} \alt
+  \comb{\M1}{\M2}  &\mbox{Raw Terms} \\
+\end{altgrammar}
+\end{tdisplay} 
+$$
 
-\subsection{Generic Java}
-% Existential types are needed to fully express Generic Java types in Typed Clojure
-% F-bounded polymorphism supported
+$$
+\begin{tdisplay}{Syntax of Terms}
+\begin{altgrammar}
+  \exp{} &::=&  \x{} \alt \{ \overrightarrow{\exp{}\ \exp{}} \} 
+             \alt [ \overrightarrow{\exp{}} ] 
+             &\mbox{Expressions} \\ 
+  \v{} &::=& \alt \{ \overrightarrow{\v{}\ \v{}} \} 
+             \alt [ \overrightarrow{\v{}} ] 
+  &\mbox{Values} \\ 
+\end{altgrammar}
+\end{tdisplay}
+$$
 
-\section{Occurrence Typing}
-% interesting paths
-% - count path element
-% - class path element
-% - keyword path element
-
-\section{Variable-arity Polymorphism}
-% Problematic areas
-% - flattened paired arguments. assoc, hash-maps, array-map
-% - complex argument dependencies. comp, partial
+$$
+\begin{tdisplay}{Operational Semantics}
+\begin{array}{c}
+{\inferrule[E-Assoc]
+  { {\Gamma \vdash e_1 \hookrightarrow v_1} \\
+    {\Gamma \vdash e_2 \hookrightarrow v_2} \\
+    {\Gamma \vdash e_3 \hookrightarrow v_3} \\
+    {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
+  { {\Gamma \vdash (assoc\ e_1\ e_2\ e_3) \hookrightarrow v_4} }} \\
+\\
+{\inferrule[E-Dissoc]
+  { {\Gamma \vdash e_1 \hookrightarrow v_1} \\
+    {\Gamma \vdash e_2 \hookrightarrow v_2} \\
+    {v_3 = v_1\ without\ entry\ with\ key\ v_2}}
+ {\Gamma \vdash (dissoc\ e_1\ e_2) \hookrightarrow v_3} } \\
+\\
+{\inferrule[E-Var]
+  {\Gamma(x) = v}
+  {\Gamma \vdash x \hookrightarrow v}}
+\end{array} 
+\end{tdisplay} 
+$$
 
@@
 and abstractions over \emph{dotted} parameters, respectively.
-
-\section{Implementation}
-
-\subsection{Composite Propositions}
-
-\subsection{Non-hygienic Macro Expansion and Filters}
-% need to remove bindings as they fall out of scope
-
-\section{Future Work}
-
-\subsection{Multimethods}
-% Problem - typing isa?
-% - probably impossible to type higher order usages of isa?.
-% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
-% Statically proving at least one relevant dispatch method is present.
-% Warn potential method conflicts.
-% Suppress warnings, similar to prefer-method syntax
-
-\subsection{Records}
-% A composition of existing features: Datatypes and Heterogeneous Maps
-
-\subsection{Other Clojure Dialects}
-% unify ASTs for easier porting
-
-\subsection{Linter}
-
-%
-%\begin{mathpar}
-%\inferrule*[right=B-Var]
-%  {\rho(x) = v}
-%  {\rho \vdash x \downarrow v}
-%
-%\inferrule*[right=B-Assoc]
-%  {\rho \vdash e_m \downarrow mi \\
-%   \rho \vdash e_k \downarrow k \\
-%   \rho \vdash e_v \downarrow v \\
-%   \rho \vdash \overrightarrow{e_j}^n \downarrow \overrightarrow{k_j}^n \\
-%   \rho \vdash \overrightarrow{e_m}^n \downarrow \overrightarrow{k_m}^n \\
-%   \delta(add\_entry, mi, k, v, \overrightarrow{k_j, k_m}^n) = mo}
-%  {\rho \vdash (assoc\ e_m\ e_k\ e_v\ \overrightarrow{e_j\ e_m}^n) \downarrow mo}
-
-%\infrule[B-Var]
-%  {\rho(x) = v}
-%  {\rho \vdash x \downarrow v}
-%
-%\infrule[B-Assoc]
-
-%\end{mathpar}
-
-\end{document}

### 2012-09-11 dissertation/future-work.tex
Rearrange into a file per chapter

--- 2012-08-26 dissertation/paper.tex
@@
-\documentclass{cshonours}
-\bibliographystyle{acm}
+\chapter{Future Work}
 
-%\usepackage{biblatex}
-
-\title{Typed Clojure}
-\author{Ambrose Bonnaire-Sergeant}
-
-\keywords{Type systems}
-
-\bibliography{bibliography}
-
-\usepackage{color}
-\usepackage{listings}
-\lstset{ %
-  language=Lisp,                % choose the language of the code
-  columns=fixed,basewidth=.5em,
-  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-  %numbers=left,                   % where to put the line-numbers
-  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
-  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-  %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  %showspaces=false,               % show spaces adding particular underscores
-  showstringspaces=false,         % underline spaces within strings
-  %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
-  %tabsize=2,          % sets default tabsize to 2 spaces
-  captionpos=t,           % sets the caption-position to bottom
-  breaklines=true,        % sets automatic line breaking
-  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
-  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
-}
-
-\begin{document}
-\maketitle
-
-\chapter{Introduction}
-
-This dissertation describes Typed Clojure, an optional static type system for Clojure.
-
-% Based on Typed Racket
-
-\chapter{Typed Clojure through Examples}
-
-% Hello world
-This section introduces Typed Clojure with example code. An attempt is
-made to introduce some Clojure syntax and sematics to those unfamiliar or needing a refresher.
-A basic knowledge of Lisp syntax is assumed.
-
-We begin with the obligatory \emph{Hello world} example.
-
-\begin{lstlisting}[caption=Hello world]
-(ns typed.test.hello-world
-  (:require [typed.core :refer [check-ns]]]))
-
-(println "Hello world")
-\end{lstlisting}
-
-The \lstinline|clojure.core/ns|
-\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
-invocation declares the current namespace to be \lstinline|typed.test.hello-world|
-and requires the namespace \lstinline|typed.core| to be loaded as a dependency, referring the var
-\lstinline|typed.core/check-ns| to be referenced locally. Other than this
-dependency, this is identical to the untyped \emph{Hello world}.
-% footnote: check-ns used to initiate type checking usually at REPL
-
-More complex code require extra annotations to type check:
-
-\begin{lstlisting}
-(ns typed.test.collatz
-  (:require [typed.core :refer [check-ns ann]]))
-
-(ann collatz [Number -> Number])
-(defn collatz [n]
-  (cond
-    (= 1 n) 
-      1
-    (and (integer? n) 
-         (even? n)) 
-      (collatz (/ n 2))
-    :else 
-      (collatz (inc (* 3 n)))))
-\end{lstlisting}
-\footnote{Example adapted from Sam's dissertation TODO}
-
-In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
-used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
-to be a function from \lstinline{java.lang.Number} to 
-\lstinline{java.lang.Number}
-\footnote{All Classes in the \lstinline|java.lang| package
-are automatically imported in every Clojure namespace (for those familiar with Java, the equivalent of \lstinline|import java.lang.*;|).}.
-
-\section{Datatypes and Protocols}
-
-We can annotate datatype and protocol definitions similarly.
-
-\begin{lstlisting}
-(ns typed.test.deftype
-  (:require [typed.core 
-             :refer [check-ns ann-datatype
-                     tc-ignore ann-protocol AnyInteger]]))
-
-(ann-protocol Age 
-  :methods
-  {age [Age -> AnyInteger]})
-(tc-ignore
-  (defprotocol Age
-    (age [this])))
-
-(ann-datatype Person 
-  [[name :- String]
-   [age :- AnyInteger]])
-(deftype Person [name age]
-  Age
-  (age [this] age))
-
-(age (Person. "Lucy" 34))
-\end{lstlisting}
-
-\lstinline|clojure.core/defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols}
-with a set of methods. \lstinline|typed.core/ann-protocol| annotates a protocol with the types of its methods.
-In this example, we define a protocol \lstinline|typed.test.person/Age| with an \lstinline|age| method.
-The call to \lstinline|clojure.core/defprotocol| also defines a new var \lstinline|typed.test.person/age|, a first-class function
-wrapping the \lstinline|age| method, but taking the target Object as the first parameter. The
-type signature provided with \lstinline|typed.core/ann-protocol|, here \lstinline|[typed.test.person/Age -> typed.core/AnyInteger]|, 
-is for this function.
-
-Invocations of \lstinline|clojure.core/defprotocol| are currently not able to be type checked
-and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.
-
-\lstinline|clojure.core/deftype|
-defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes}
-in the current namespace with a number of fields and methods. 
-\lstinline|typed.core/ann-datatype| annotates a datatype with its field types.
-In this example, we create a datatype \lstinline|typed.test.person.Person|
-with fields \lstinline|name| and \lstinline|age| and implement the \lstinline|age|
-method from protocol \lstinline|typed.test.person/Age|.
-
-Java constructors are invoked in Clojure by suffixing the Class we want to instantiate with a dot.
-Datatypes are implemented as Java Classes with immutable fields (by default) and a single constructor, taking as arguments its fields 
-in the order they are passed to \lstinline|deftype|
-\footnote{When unambiguous, I omit the qualifying namespace/package for the remainder of the chapter.}.
-\lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
-instance, setting the fields to their corresponding positional arguments.
-Typed Clojure checks the datatype constructor to be the equivalent of 
-\lstinline|[String AnyInteger -> Person]|.
-
-Finally, Typed Clojure checks invocations of Protocol methods. It infers \lstinline|Person|
-is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.
-
-\section{Polymorphism}
-
-Typed Clojure supports F-bounded polymorphism. All type variables may have
-upper and lower bounds.
-
-Typed Clojure parameterises some of Clojure's data structures. For example,
-the interface behind Clojure's \emph{seq} abstraction \lstinline|clojure.lang.Seqable| has one 
-covariant parameter\footnote{\lstinline|(Seqable Integer)| being a subtype of \lstinline|(Seqable Number)|
-because Integer is a subtype of Number.}.
-
-\begin{lstlisting}
-...
-(ann to-set 
-     (All [x]
-       [(U nil (Seqable x)) -> (clojure.lang.PersistentHashSet x)]))
-(defn to-set [a]
-  (set a))
-...
-\end{lstlisting}
-
-In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}
-, we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
-\lstinline|All| introduces a set of type variables to the body of a type,
-here \lstinline|x| is used to define a relationship between the input type and return type.
-
-\lstinline|(U nil (Seqable x))| is a common type in Clojure, read as the union
-of the singleton type \lstinline|nil| and the type \lstinline|(Seqable x)|.
-The vast majority of types for collection processing functions in the Clojure core library feature
-this as an input type, where passing \lstinline|nil| either has some special behaviour 
-or is synonymous with passing an empty \lstinline|Seqable|.
-
-\section{Heterogeneous Maps}
-
-Where particularly object-oriented languages reach for objects, Clojure
-utilises maps. Clojure provides a map literal using curly braces. For example,
-\lstinline|{:a 1, :b 2}| is a map with two key-value entries: from Keyword key \lstinline|:a|
-to value \lstinline|1|, and Keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
-whitespace in Clojure and are included for readability.
-
-Typed Clojure provides a heterogeneous map type, restricted to 
-maps with singleton Keyword keys. This restriction is reflected
-in the syntax for defining heterogeneous map types.
-
-\begin{lstlisting}
-...
-(ann config
-     (HMap {:file String,
-            :ns Symbol}))
-(def config
-  {:file "clojure/core.clj",
-   :ns 'clojure.core})
-...
-\end{lstlisting}
-
-This example checks \lstinline|config| to be a heterogeneous map
-with \lstinline|:file| and \lstinline|:ns| keys, with values of
-type \lstinline|String| and \lstinline|Symbol| respectively.
-
-Heterogeneous vector and seq types are also provided.
-
-\section{Multi-Arity Functions}
-
-All Clojure functions support multiple arities, allowing different
-behaviour for different number of arguments.
-
-\begin{lstlisting}
-(ns typed.test.poly
-  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
-  (:import [clojure.lang Seqable]))
-
-(ann repeatedly'
-     (All [x]
-       (Fn [[-> x] -> (Seqable x)]
-           [AnyInteger [-> x] -> (Seqable x)])))
-(defn repeatedly'
-  "Takes a function of no args, presumably with side effects, and
-  returns an infinite (or length n if supplied) lazy sequence of calls
-  to it"
-  ([f] (lazy-seq (cons (f) (repeatedly' f))))
-  ([n f] (take n (repeatedly' f))))
-\end{lstlisting}
-
-This example defines \lstinline|repeatedly'|\footnote{Identical to \lstinline|clojure.core/repeatedly|},
-a multi-arity function taking either a zero-argument function, or an integer and a zero-argument function as arguments.
-The type variable \lstinline|x| links the return type of the generator function to
-the resulting \lstinline|Seqable| instance.
-
-When checking the definition of multi-arity functions, Typed Clojure
-finds the matching function types for a particular arity by number of arguments
-and requires each relevant type to check successfully.
-
-\section{Variable-Arity Functions}
-
-Clojure functions can take an optional \emph{rest} argument.
-
-\subsection{Uniform Variable-Arity}
-
-\subsection{Non-uniform Variable-Arity}
-
-\section{Occurrence Typing}
-
-Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
-Occurrence typing is used to refine the known types for immutable local bindings 
-down conditional branches.
-
-\begin{lstlisting}
-(ann num-vec2 
-     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
-(defn num-vec2 [a b]
-  [(if a a 0) 
-   (if b b 0)])
-\end{lstlisting}
-
-To check this example, occurrence typing infers type information based on the result of each test.
-In Clojure, \lstinline|nil| and \lstinline|false| are false values and all other values are true.
-The test \lstinline|(if a a 0)| follows the \emph{then} branch is \lstinline|a| is not \lstinline|nil|
-or \lstinline|false|. When checking this branch, we can safely refine the type of \lstinline|a| to \lstinline|Number| from
-\lstinline|(U nil Number)|. Similarly, following the \emph{else} branch refines the type of \lstinline|a|
-to \lstinline|nil| from \lstinline|(U nil Number)|.
-
-\section{Java Interoperability}
-
-Typed Clojure currently provides safe interop\footnote{Interop is short for interoperability} with (non-Generic) Java
-\footnote{Compatibility with Generic Java is planned future work}.
-
-The key to safe Java interop is the treatment of Java's \emph{null}.
-\emph{null} is a subtype to all reference types
-represented in Clojure by \lstinline|nil|. Unlike Java's type system
-Clojure explicitly separates \emph{null} and reference types, allowing
-Typed Clojure to express
-\emph{nullable}
-\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
-which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
-positions.
-
-\begin{lstlisting}
-(ns typed.test.interop
-  (:import (java.io File))
-  (:require [typed.core :refer [ann non-nil-return check-ns]]))
-
-(ann f File)
-(def f (File. "a"))
-
-(ann prt (U nil String))
-(def prt (.getParent ^File f))
-
-(non-nil-return java.io.File/getName :all)
-(ann nme String)
-(def nme (.getName ^File f))
-
-\end{lstlisting}
-
-This example shows how Typed Clojure handles \emph{null} while creating and
-using an instance of \emph{java.io.File}\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
-
-Typed Clojure checks calls to Java constructors by requiring the provided
-arguments be acceptable input to at least one constructor for that Class.
-In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
-argument, so \lstinline|(File. "a")| is type safe. By design, constructors never
-return \emph{null}, so Clojure assigns the return type to be \lstinline|File|.
-This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.
-
-Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
-The \emph{java.io.File} instance method \emph{getParent}
-is equivalent to \lstinline|[-> (U nil String]| in Typed Clojure. This happens to be
-a valid approximation of the method as \emph{getParent} returns \emph{null} 
-``if the pathname does not name a parent directory''\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
-On the other hand, the instance method \emph{getName} always returns an
-instance of \emph{java.lang.String}, so we set the return position of
-\emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
-\footnote{The second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
-returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
-to override all arities of that method.}.
-
-\chapter{Design Goals}
-
-\section{Safer Java Interop}
-\section{Optional Type Checking}
-\section{Explicit Typing}
-\section{No new idioms}
-
-\section{Local Type Inference}
-
-Similar to Typed Racket, type inference is based on Local Type Inference
-by Pierce and Turner.
-
-\section{F-bounded Polymorphism}
-
-Typed Clojure includes support for F-bounds on type variables, as an extension
-to Local Type Inference.
-
-An extra environment from type variables to bounds is kept until after inference,
-after which each unground bound is substituted with the inferred types and the
-subtyping relationships are checked.
-
-\section{Heterogeneous Maps}
-% keyword keys for now, generally encouraged, fast
-% Some heterogeneous maps have string args, eg. from web frameworks
-
-\section{Java Interoperability}
-% Args non-nil, return nilable
-% Constructors and other special methods handled appropriately (eg. Constructor are non-nil return)
-
-\subsection{Primitive Arrays}
-% Primitive arrays are covariant, which is well-known to be statically unsound.
-% We cannot trust the type signature of any Java method or field.
-% Separate read and write types into two parameters.
-
-\subsection{Interaction with null}
-% null is explicitly expressible in my type system
-% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
-%
-
-\subsection{Generic Java}
-% Existential types are needed to fully express Generic Java types in Typed Clojure
-% F-bounded polymorphism supported
-
-\section{Occurrence Typing}
-% interesting paths
-% - count path element
-% - class path element
-% - keyword path element
-
-\section{Variable-arity Polymorphism}
-% Problematic areas
-% - flattened paired arguments. assoc, hash-maps, array-map
-% - complex argument dependencies. comp, partial
-
-\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}
-
-Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
-representing a sequence of types, allowing non-uniform variable parameters.
-This enables \lstinline|map| and other function with complex arguments to be typed.
-Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
-Typed Racket, and is a key feature of Typed Racket's implementation.
-
-Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
-to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
-variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
-associates key-value pairs to a target map. It takes one or more key-value pairs and
-applies them left-to-right to the target map. The key problem from a typing perspective
-is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
-accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
-higher-order function.
-\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
-dotted parameters represent an unrestricted number of types, where we require an even number.
-
-The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
-new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
-Firstly, subtyping between different kinds of dotted and rest
-\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
-parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
-Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
-difficult to express.
-Thirdly, higher-order functions taking functions with dotted arguments commonly 
-The first issue requires careful design and implementation.
-The second and third points require new constructs, in the form of \emph{projections}
-and abstractions over \emph{dotted} parameters, respectively.
-
-\section{Implementation}
-
-\subsection{Composite Propositions}
-
-\subsection{Non-hygienic Macro Expansion and Filters}
-% need to remove bindings as they fall out of scope
-
-\section{Future Work}
-
-\subsection{Multimethods}
+\section{Multimethods}
 % Problem - typing isa?
@@
 
-\subsection{Records}
+\section{Records}
 % A composition of existing features: Datatypes and Heterogeneous Maps
@@
 \subsection{Linter}
-
-%
-%\begin{mathpar}
-%\inferrule*[right=B-Var]
-%  {\rho(x) = v}
-%  {\rho \vdash x \downarrow v}
-%
-%\inferrule*[right=B-Assoc]
-%  {\rho \vdash e_m \downarrow mi \\
-%   \rho \vdash e_k \downarrow k \\
-%   \rho \vdash e_v \downarrow v \\
-%   \rho \vdash \overrightarrow{e_j}^n \downarrow \overrightarrow{k_j}^n \\
-%   \rho \vdash \overrightarrow{e_m}^n \downarrow \overrightarrow{k_m}^n \\
-%   \delta(add\_entry, mi, k, v, \overrightarrow{k_j, k_m}^n) = mo}
-%  {\rho \vdash (assoc\ e_m\ e_k\ e_v\ \overrightarrow{e_j\ e_m}^n) \downarrow mo}
-
-%\infrule[B-Var]
-%  {\rho(x) = v}
-%  {\rho \vdash x \downarrow v}
-%
-%\infrule[B-Assoc]
-
-%\end{mathpar}
-
-\end{document}

### 2012-09-11 dissertation/implementation.tex
Rearrange into a file per chapter

\chapter{Implementation}

\section{Composite Propositions}

\section{Non-hygienic Macro Expansion and Filters}
% need to remove bindings as they fall out of scope
% Need hygiene to ensure correct propagation of filters
%(fn [a]
%  (if (= a 1)
%    (let [a 'foo] ; here this shadows the argument, impossible to recover filters
%      a)          ; in fact any new filters about a will be incorrectly assumed to be the argument
%      false))

### 2012-09-11 dissertation/intro.tex
Rearrange into a file per chapter

--- 2012-08-26 dissertation/paper.tex
@@
-\documentclass{cshonours}
-\bibliographystyle{acm}
-
-%\usepackage{biblatex}
-
-\title{Typed Clojure}
-\author{Ambrose Bonnaire-Sergeant}
-
-\keywords{Type systems}
-
-\bibliography{bibliography}
-
-\usepackage{color}
-\usepackage{listings}
-\lstset{ %
-  language=Lisp,                % choose the language of the code
-  columns=fixed,basewidth=.5em,
-  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-  %numbers=left,                   % where to put the line-numbers
-  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
-  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-  %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  %showspaces=false,               % show spaces adding particular underscores
-  showstringspaces=false,         % underline spaces within strings
-  %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
-  %tabsize=2,          % sets default tabsize to 2 spaces
-  captionpos=t,           % sets the caption-position to bottom
-  breaklines=true,        % sets automatic line breaking
-  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
-  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
-}
-
-\begin{document}
-\maketitle
-
 \chapter{Introduction}
 
-This dissertation describes Typed Clojure, an optional static type system for Clojure.
+\section{Thesis}
 
-% Based on Typed Racket
+\emph{It is practical and useful to design and implement an optional typing system 
+for the Clojure programming language that allows Clojure programmers to continue 
+using idioms and style found in current Clojure code.}
 
-\chapter{Typed Clojure through Examples}
+\section{Motivation}
+
+% Type Systems for lisp?
+% Why type systems?
+% Why Clojure?
+
+In the last decade it has become increasingly common to enhance
+dynamically typed languages with static type systems. This idea not new,
+but recent attempts are noteworthy for their successful use of bidirectional checking.
+Instead of inferring types the algorithm relies on programmer annotations.
+
+Clojure is a dynamically typed language running on the JVM. It emphasises
+functional programming with immutable data structures
+and provides direct interoperability with JVM languages.
+
+Recently languages have been created to or been modified to support aspects
+of static and dynamic typing.
+Dart is fundamentally dynamically typed but offers a form of optional
+static typing. Haskell recently supports the ability to delay static type errors 
+to runtime, removing the need for the program to type check before running.
+Scala can emulate some features of dynamic languages with Dynamic Scala.
+
+\subsection{What kind of type system does Typed Clojure provide?}
+
+% What are types? No real agreement .. but here is my definition..
+% Every programmer thinks they know what they are. More than one concept
+% in the literature.
+% Static vs. dynamic
+% intrinsic vs. extrinsic
+
+% is it even a type system?
+% type refinements vs ordinary types
+% extrinsic vs intrinsic type systems
+
+% refinements check invariants of programs
+% while ordinary static types check whether programs
+% are basically meaningful - Rowan + Frank P
+% Page 1 Rowan's PhD
+An ordinary static type system is used to check whether programs are basically
+meaningful. Such type systems are \emph{intrinsic}. A language with an intrinsic
+static type system relies on programs being well-typed for runtime semantics to be defined.
+This means such programs must pass the type checker before being run.
+
+A static type system is \emph{extrinsic} when runtime semantics
+are not defined by the type system. In other words, passing a static type checker
+is not essential to running programs. A dynamically typed language can be viewed
+as having a static type system that supports exact one type.
+
+A type checker ensures invariants hold complains if code
+does not conform to its documented type. Powerful type systems like
+Haskell's and Scala's are useful when reasoning about higher-order code.
+
+When a static type checker is not available, which describes the situation
+for most dynamic languages, other techniques are used for checking
+type invariants. Runtime contracts are a popular alternative.
+Clojure takes this approach, providing easy syntax for defining pre-
+and post-conditions \footnote{Inspired by Eiffel}.
+
+% clojure is dynamically typed
+% currently few tools to statically reason about complex higher-order code
+% conduit based on streams, uses continuations
+% core.logic based on monads
+% Static types are useful for reasoning for such things
+% Strong statically typed languages like Haskell use monads regularly
+
+% strength of Clojure is concision
+% typically, concise static languages are built concurrently with their type system
+% Eg. Haskell, ML flavoured languages
+% Counterpoint: Typed Racket preserves racket idioms, adds static typing
+
+\section{Typed Clojure through Examples}
 
@@
 
-\section{Datatypes and Protocols}
+\subsection{Datatypes and Protocols}
 
@@
 
-\section{Polymorphism}
+\subsection{Polymorphism}
 
@@
 
-\lstinline|(U nil (Seqable x))| is a common type in Clojure, read as the union
-of the singleton type \lstinline|nil| and the type \lstinline|(Seqable x)|.
+\lstinline|(U nil (Seqable x))| is a common type in Typed Clojure, read as the union
+of the type \lstinline|nil| and the type \lstinline|(Seqable x)|.
 The vast majority of types for collection processing functions in the Clojure core library feature
-this as an input type, where passing \lstinline|nil| either has some special behaviour 
+it as an input type, where passing \lstinline|nil| either has some special behaviour 
 or is synonymous with passing an empty \lstinline|Seqable|.
 
-\section{Heterogeneous Maps}
+\subsection{Heterogeneous Maps}
 
@@
 
-\section{Multi-Arity Functions}
+\subsection{Multi-Arity Functions}
 
-All Clojure functions support multiple arities, allowing different
-behaviour for different number of arguments.
+Clojure function objects support multiple arities, allowing dispatch
+on the number of supplied arguments.
 
@@
 
-\section{Variable-Arity Functions}
+\subsection{Variable-Arity Functions}
 
@@
 
-\subsection{Uniform Variable-Arity}
+\subsubsection{Uniform Variable-Arity}
 
-\subsection{Non-uniform Variable-Arity}
+\subsubsection{Non-uniform Variable-Arity}
 
-\section{Occurrence Typing}
+\subsection{Occurrence Typing}
 
 Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
-Occurrence typing is used to refine the known types for immutable local bindings 
-down conditional branches.
+Occurrence typing is used to refine types for immutable local bindings 
+as programs follow conditional branches.
 
@@
 
-\section{Java Interoperability}
+\subsection{Java Interoperability}
 
@@
 In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
-argument, so \lstinline|(File. "a")| is type safe. By design, constructors never
-return \emph{null}, so Clojure assigns the return type to be \lstinline|File|.
+argument, so \lstinline|(File. "a")| is type safe. Java constructors never
+return \emph{null}, so Typed Clojure assigns the return type to be \lstinline|File|.
 This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.
@@
 to override all arities of that method.}.
-
-\chapter{Design Goals}
-
-\section{Safer Java Interop}
-\section{Optional Type Checking}
-\section{Explicit Typing}
-\section{No new idioms}
-
-\section{Local Type Inference}
-
-Similar to Typed Racket, type inference is based on Local Type Inference
-by Pierce and Turner.
-
-\section{F-bounded Polymorphism}
-
-Typed Clojure includes support for F-bounds on type variables, as an extension
-to Local Type Inference.
-
-An extra environment from type variables to bounds is kept until after inference,
-after which each unground bound is substituted with the inferred types and the
-subtyping relationships are checked.
-
-\section{Heterogeneous Maps}
-% keyword keys for now, generally encouraged, fast
-% Some heterogeneous maps have string args, eg. from web frameworks
-
-\section{Java Interoperability}
-% Args non-nil, return nilable
-% Constructors and other special methods handled appropriately (eg. Constructor are non-nil return)
-
-\subsection{Primitive Arrays}
-% Primitive arrays are covariant, which is well-known to be statically unsound.
-% We cannot trust the type signature of any Java method or field.
-% Separate read and write types into two parameters.
-
-\subsection{Interaction with null}
-% null is explicitly expressible in my type system
-% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
-%
-
-\subsection{Generic Java}
-% Existential types are needed to fully express Generic Java types in Typed Clojure
-% F-bounded polymorphism supported
-
-\section{Occurrence Typing}
-% interesting paths
-% - count path element
-% - class path element
-% - keyword path element
-
-\section{Variable-arity Polymorphism}
-% Problematic areas
-% - flattened paired arguments. assoc, hash-maps, array-map
-% - complex argument dependencies. comp, partial
-
-\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}
-
-Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
-representing a sequence of types, allowing non-uniform variable parameters.
-This enables \lstinline|map| and other function with complex arguments to be typed.
-Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
-Typed Racket, and is a key feature of Typed Racket's implementation.
-
-Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
-to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
-variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
-associates key-value pairs to a target map. It takes one or more key-value pairs and
-applies them left-to-right to the target map. The key problem from a typing perspective
-is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
-accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
-higher-order function.
-\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
-dotted parameters represent an unrestricted number of types, where we require an even number.
-
-The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
-new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
-Firstly, subtyping between different kinds of dotted and rest
-\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
-parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
-Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
-difficult to express.
-Thirdly, higher-order functions taking functions with dotted arguments commonly 
-The first issue requires careful design and implementation.
-The second and third points require new constructs, in the form of \emph{projections}
-and abstractions over \emph{dotted} parameters, respectively.
-
-\section{Implementation}
-
-\subsection{Composite Propositions}
-
-\subsection{Non-hygienic Macro Expansion and Filters}
-% need to remove bindings as they fall out of scope
-
-\section{Future Work}
-
-\subsection{Multimethods}
-% Problem - typing isa?
-% - probably impossible to type higher order usages of isa?.
-% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
-% Statically proving at least one relevant dispatch method is present.
-% Warn potential method conflicts.
-% Suppress warnings, similar to prefer-method syntax
-
-\subsection{Records}
-% A composition of existing features: Datatypes and Heterogeneous Maps
-
-\subsection{Other Clojure Dialects}
-% unify ASTs for easier porting
-
-\subsection{Linter}
-
-%
-%\begin{mathpar}
-%\inferrule*[right=B-Var]
-%  {\rho(x) = v}
-%  {\rho \vdash x \downarrow v}
-%
-%\inferrule*[right=B-Assoc]
-%  {\rho \vdash e_m \downarrow mi \\
-%   \rho \vdash e_k \downarrow k \\
-%   \rho \vdash e_v \downarrow v \\
-%   \rho \vdash \overrightarrow{e_j}^n \downarrow \overrightarrow{k_j}^n \\
-%   \rho \vdash \overrightarrow{e_m}^n \downarrow \overrightarrow{k_m}^n \\
-%   \delta(add\_entry, mi, k, v, \overrightarrow{k_j, k_m}^n) = mo}
-%  {\rho \vdash (assoc\ e_m\ e_k\ e_v\ \overrightarrow{e_j\ e_m}^n) \downarrow mo}
-
-%\infrule[B-Var]
-%  {\rho(x) = v}
-%  {\rho \vdash x \downarrow v}
-%
-%\infrule[B-Assoc]
-
-%\end{mathpar}
-
-\end{document}

### 2012-09-11 dissertation/lit-review.tex
Rearrange into a file per chapter

--- 2012-05-24 lit-review/lit-review.tex
@@
-\documentclass[12pt, a4paper]{article}
-\usepackage{biblatex}
-\usepackage{fancyvrb}
+% TODO
+% What does this Lit review cover?
+% Why things are included/not included?
+% Background
+% Soft typing
+% See Sam's dissertation
+% Lisp type system background (197x)
+% Why was it so hard?
+% - bidirectional checking vs. global inference
 
-\usepackage{color}
-\usepackage{listings}
-\lstset{ %
-  language=Lisp,                % choose the language of the code
-  columns=fixed,basewidth=.5em,
-  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-  numbers=left,                   % where to put the line-numbers
-  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
-  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-  %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
-  %showspaces=false,               % show spaces adding particular underscores
-  %showstringspaces=false,         % underline spaces within strings
-  %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
-  %tabsize=2,          % sets default tabsize to 2 spaces
-  captionpos=b,           % sets the caption-position to bottom
-  breaklines=true,        % sets automatic line breaking
-  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
-  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
-}
+\chapter{Literature Review}
 
-\bibliography{bibliography}
-\title{A Practical Optional Type System for Clojure, Literary Review}
-\author{Ambrose Bonnaire-Sergeant (20350292)}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-Clojure is a practical, dynamically typed language running on the Java Virtual Machine.
-I intend to design and implement a prototype optional static type 
-system for Clojure, with practical use being a priority. Devising a static type system 
-requires a combination of components that each adds some utility
-without significantly complicating the overall implementation.
-I look at several novel inventions created as a result 
-of the Typed Racket project including
-non-uniform variable-arity polymorphism and occurrence typing.
-Typed Racket and Scala both use flavours of local type inference
-which I compare and judge for applicability to this project.
-Other relevant features of Scala are considered, such as its approach to bounded polymorphism,
-the Scala class hierarchy, and the similarities of its methods of composition to Clojure's.
-\end{abstract}
-
-\tableofcontents
+%\begin{abstract}
+%Clojure is a practical, dynamically typed language running on the Java Virtual Machine.
+%I intend to design and implement a prototype optional static type 
+%system for Clojure, with practical use being a priority. Devising a static type system 
+%requires a combination of components that each adds some utility
+%without significantly complicating the overall implementation.
+%I look at several novel inventions created as a result 
+%of the Typed Racket project including
+%non-uniform variable-arity polymorphism and occurrence typing.
+%Typed Racket and Scala both use flavours of local type inference
+%which I compare and judge for applicability to this project.
+%Other relevant features of Scala are considered, such as its approach to bounded polymorphism,
+%the Scala class hierarchy, and the similarities of its methods of composition to Clojure's.
+%\end{abstract}
 
@@
 project, many of which will reveal themselves during development.
-
-\printbibliography
-
-\end{document}

### 2012-09-11 dissertation/paper.tex
Rearrange into a file per chapter

@@
 \documentclass{cshonours}
-\bibliographystyle{acm}
 
-%\usepackage{biblatex}
+\usepackage{mmm}
+\usepackage{mathpartir}
+\usepackage{clj-grammar}
+\usepackage[style=numeric]{biblatex}
+\addbibresource{bibliography.bib}
 
-\title{Typed Clojure}
-\author{Ambrose Bonnaire-Sergeant}
-
-\keywords{Type systems}
-
-\bibliography{bibliography}
-
-\usepackage{color}
 \usepackage{listings}
@@
   %showtabs=false,                 % show tabs within strings adding particular underscores
-  frame=single,           % adds a frame around the code
+  %frame=single,           % adds a frame around the code
   %tabsize=2,          % sets default tabsize to 2 spaces
@@
 
+\title{Typed Clojure: A Practical Optional Type System for Clojure}
+\author{Ambrose Bonnaire-Sergeant}
+
+\keywords{Type systems, Clojure}
+% TODO What are categories? cshonours requires them
+\categories{}
+
 \begin{document}
+
 \maketitle
 
-\chapter{Introduction}
+\begin{abstract}
+  % FIXME long tradition for attempts
+  There is a long tradition to create static type systems for dynamically-typed
+  languages. Static type checking promises to eliminate many common user errors
+  that are otherwise unnoticed or are caught when the program is run.
 
-This dissertation describes Typed Clojure, an optional static type system for Clojure.
+  Recent attempts have seen several promising and innovative
+  type systems built for dynamically typed languages. Typed Racket is able to
+  statically type check idioms ..
+  We take the lessons learnt from these
+  projects and apply them to the dynamically-typed language Clojure in the form of Typed Clojure.
+\end{abstract}
 
-% Based on Typed Racket
+%\tableofcontents
 
-\chapter{Typed Clojure through Examples}
+\include{intro}
+\include{lit-review}
+\include{design-choices}
 
-% Hello world
-This section introduces Typed Clojure with example code. An attempt is
-made to introduce some Clojure syntax and sematics to those unfamiliar or needing a refresher.
-A basic knowledge of Lisp syntax is assumed.
+% A syntactic approach to type soundness (Wright and Felleisen)
+% operational semantics
+% - See Sam TH's operational semantics
+% TAPL 392
 
-We begin with the obligatory \emph{Hello world} example.
+% potential proofs/rules
+% - hmaps
+% - intersection types (protocols + types)
+% - subtyping
+% - dot methods
+% - f-bounded polymorphism
 
-\begin{lstlisting}[caption=Hello world]
-(ns typed.test.hello-world
-  (:require [typed.core :refer [check-ns]]]))
+\include{implementation}
 
-(println "Hello world")
-\end{lstlisting}
+% Evaluation
+% - successes
+% - future work identified
 
-The \lstinline|clojure.core/ns|
-\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
-invocation declares the current namespace to be \lstinline|typed.test.hello-world|
-and requires the namespace \lstinline|typed.core| to be loaded as a dependency, referring the var
-\lstinline|typed.core/check-ns| to be referenced locally. Other than this
-dependency, this is identical to the untyped \emph{Hello world}.
-% footnote: check-ns used to initiate type checking usually at REPL
-
-More complex code require extra annotations to type check:
-
-\begin{lstlisting}
-(ns typed.test.collatz
-  (:require [typed.core :refer [check-ns ann]]))
-
-(ann collatz [Number -> Number])
-(defn collatz [n]
-  (cond
-    (= 1 n) 
-      1
-    (and (integer? n) 
-         (even? n)) 
-      (collatz (/ n 2))
-    :else 
-      (collatz (inc (* 3 n)))))
-\end{lstlisting}
-\footnote{Example adapted from Sam's dissertation TODO}
-
-In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
-used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
-to be a function from \lstinline{java.lang.Number} to 
-\lstinline{java.lang.Number}
-\footnote{All Classes in the \lstinline|java.lang| package
-are automatically imported in every Clojure namespace (for those familiar with Java, the equivalent of \lstinline|import java.lang.*;|).}.
-
-\section{Datatypes and Protocols}
-
-We can annotate datatype and protocol definitions similarly.
-
-\begin{lstlisting}
-(ns typed.test.deftype
-  (:require [typed.core 
-             :refer [check-ns ann-datatype
-                     tc-ignore ann-protocol AnyInteger]]))
-
-(ann-protocol Age 
-  :methods
-  {age [Age -> AnyInteger]})
-(tc-ignore
-  (defprotocol Age
-    (age [this])))
-
-(ann-datatype Person 
-  [[name :- String]
-   [age :- AnyInteger]])
-(deftype Person [name age]
-  Age
-  (age [this] age))
-
-(age (Person. "Lucy" 34))
-\end{lstlisting}
-
-\lstinline|clojure.core/defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols}
-with a set of methods. \lstinline|typed.core/ann-protocol| annotates a protocol with the types of its methods.
-In this example, we define a protocol \lstinline|typed.test.person/Age| with an \lstinline|age| method.
-The call to \lstinline|clojure.core/defprotocol| also defines a new var \lstinline|typed.test.person/age|, a first-class function
-wrapping the \lstinline|age| method, but taking the target Object as the first parameter. The
-type signature provided with \lstinline|typed.core/ann-protocol|, here \lstinline|[typed.test.person/Age -> typed.core/AnyInteger]|, 
-is for this function.
-
-Invocations of \lstinline|clojure.core/defprotocol| are currently not able to be type checked
-and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.
-
-\lstinline|clojure.core/deftype|
-defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes}
-in the current namespace with a number of fields and methods. 
-\lstinline|typed.core/ann-datatype| annotates a datatype with its field types.
-In this example, we create a datatype \lstinline|typed.test.person.Person|
-with fields \lstinline|name| and \lstinline|age| and implement the \lstinline|age|
-method from protocol \lstinline|typed.test.person/Age|.
-
-Java constructors are invoked in Clojure by suffixing the Class we want to instantiate with a dot.
-Datatypes are implemented as Java Classes with immutable fields (by default) and a single constructor, taking as arguments its fields 
-in the order they are passed to \lstinline|deftype|
-\footnote{When unambiguous, I omit the qualifying namespace/package for the remainder of the chapter.}.
-\lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
-instance, setting the fields to their corresponding positional arguments.
-Typed Clojure checks the datatype constructor to be the equivalent of 
-\lstinline|[String AnyInteger -> Person]|.
-
-Finally, Typed Clojure checks invocations of Protocol methods. It infers \lstinline|Person|
-is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.
-
-\section{Polymorphism}
-
-Typed Clojure supports F-bounded polymorphism. All type variables may have
-upper and lower bounds.
-
-Typed Clojure parameterises some of Clojure's data structures. For example,
-the interface behind Clojure's \emph{seq} abstraction \lstinline|clojure.lang.Seqable| has one 
-covariant parameter\footnote{\lstinline|(Seqable Integer)| being a subtype of \lstinline|(Seqable Number)|
-because Integer is a subtype of Number.}.
-
-\begin{lstlisting}
-...
-(ann to-set 
-     (All [x]
-       [(U nil (Seqable x)) -> (clojure.lang.PersistentHashSet x)]))
-(defn to-set [a]
-  (set a))
-...
-\end{lstlisting}
-
-In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}
-, we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
-\lstinline|All| introduces a set of type variables to the body of a type,
-here \lstinline|x| is used to define a relationship between the input type and return type.
-
-\lstinline|(U nil (Seqable x))| is a common type in Clojure, read as the union
-of the singleton type \lstinline|nil| and the type \lstinline|(Seqable x)|.
-The vast majority of types for collection processing functions in the Clojure core library feature
-this as an input type, where passing \lstinline|nil| either has some special behaviour 
-or is synonymous with passing an empty \lstinline|Seqable|.
-
-\section{Heterogeneous Maps}
-
-Where particularly object-oriented languages reach for objects, Clojure
-utilises maps. Clojure provides a map literal using curly braces. For example,
-\lstinline|{:a 1, :b 2}| is a map with two key-value entries: from Keyword key \lstinline|:a|
-to value \lstinline|1|, and Keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
-whitespace in Clojure and are included for readability.
-
-Typed Clojure provides a heterogeneous map type, restricted to 
-maps with singleton Keyword keys. This restriction is reflected
-in the syntax for defining heterogeneous map types.
-
-\begin{lstlisting}
-...
-(ann config
-     (HMap {:file String,
-            :ns Symbol}))
-(def config
-  {:file "clojure/core.clj",
-   :ns 'clojure.core})
-...
-\end{lstlisting}
-
-This example checks \lstinline|config| to be a heterogeneous map
-with \lstinline|:file| and \lstinline|:ns| keys, with values of
-type \lstinline|String| and \lstinline|Symbol| respectively.
-
-Heterogeneous vector and seq types are also provided.
-
-\section{Multi-Arity Functions}
-
-All Clojure functions support multiple arities, allowing different
-behaviour for different number of arguments.
-
-\begin{lstlisting}
-(ns typed.test.poly
-  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
-  (:import [clojure.lang Seqable]))
-
-(ann repeatedly'
-     (All [x]
-       (Fn [[-> x] -> (Seqable x)]
-           [AnyInteger [-> x] -> (Seqable x)])))
-(defn repeatedly'
-  "Takes a function of no args, presumably with side effects, and
-  returns an infinite (or length n if supplied) lazy sequence of calls
-  to it"
-  ([f] (lazy-seq (cons (f) (repeatedly' f))))
-  ([n f] (take n (repeatedly' f))))
-\end{lstlisting}
-
-This example defines \lstinline|repeatedly'|\footnote{Identical to \lstinline|clojure.core/repeatedly|},
-a multi-arity function taking either a zero-argument function, or an integer and a zero-argument function as arguments.
-The type variable \lstinline|x| links the return type of the generator function to
-the resulting \lstinline|Seqable| instance.
-
-When checking the definition of multi-arity functions, Typed Clojure
-finds the matching function types for a particular arity by number of arguments
-and requires each relevant type to check successfully.
-
-\section{Variable-Arity Functions}
-
-Clojure functions can take an optional \emph{rest} argument.
-
-\subsection{Uniform Variable-Arity}
-
-\subsection{Non-uniform Variable-Arity}
-
-\section{Occurrence Typing}
-
-Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
-Occurrence typing is used to refine the known types for immutable local bindings 
-down conditional branches.
-
-\begin{lstlisting}
-(ann num-vec2 
-     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
-(defn num-vec2 [a b]
-  [(if a a 0) 
-   (if b b 0)])
-\end{lstlisting}
-
-To check this example, occurrence typing infers type information based on the result of each test.
-In Clojure, \lstinline|nil| and \lstinline|false| are false values and all other values are true.
-The test \lstinline|(if a a 0)| follows the \emph{then} branch is \lstinline|a| is not \lstinline|nil|
-or \lstinline|false|. When checking this branch, we can safely refine the type of \lstinline|a| to \lstinline|Number| from
-\lstinline|(U nil Number)|. Similarly, following the \emph{else} branch refines the type of \lstinline|a|
-to \lstinline|nil| from \lstinline|(U nil Number)|.
-
-\section{Java Interoperability}
-
-Typed Clojure currently provides safe interop\footnote{Interop is short for interoperability} with (non-Generic) Java
-\footnote{Compatibility with Generic Java is planned future work}.
-
-The key to safe Java interop is the treatment of Java's \emph{null}.
-\emph{null} is a subtype to all reference types
-represented in Clojure by \lstinline|nil|. Unlike Java's type system
-Clojure explicitly separates \emph{null} and reference types, allowing
-Typed Clojure to express
-\emph{nullable}
-\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
-which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
-positions.
-
-\begin{lstlisting}
-(ns typed.test.interop
-  (:import (java.io File))
-  (:require [typed.core :refer [ann non-nil-return check-ns]]))
-
-(ann f File)
-(def f (File. "a"))
-
-(ann prt (U nil String))
-(def prt (.getParent ^File f))
-
-(non-nil-return java.io.File/getName :all)
-(ann nme String)
-(def nme (.getName ^File f))
-
-\end{lstlisting}
-
-This example shows how Typed Clojure handles \emph{null} while creating and
-using an instance of \emph{java.io.File}\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
-
-Typed Clojure checks calls to Java constructors by requiring the provided
-arguments be acceptable input to at least one constructor for that Class.
-In this case, \emph{java.io.File} has a constructor accepting a \emph{java.lang.String}
-argument, so \lstinline|(File. "a")| is type safe. By design, constructors never
-return \emph{null}, so Clojure assigns the return type to be \lstinline|File|.
-This constructor is equivalent to \lstinline|[String -> File]| in Typed Clojure.
-
-Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
-The \emph{java.io.File} instance method \emph{getParent}
-is equivalent to \lstinline|[-> (U nil String]| in Typed Clojure. This happens to be
-a valid approximation of the method as \emph{getParent} returns \emph{null} 
-``if the pathname does not name a parent directory''\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
-On the other hand, the instance method \emph{getName} always returns an
-instance of \emph{java.lang.String}, so we set the return position of
-\emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
-\footnote{The second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
-returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
-to override all arities of that method.}.
-
-\chapter{Design Goals}
-
-\section{Safer Java Interop}
-\section{Optional Type Checking}
-\section{Explicit Typing}
-\section{No new idioms}
-
-\section{Local Type Inference}
-
-Similar to Typed Racket, type inference is based on Local Type Inference
-by Pierce and Turner.
-
-\section{F-bounded Polymorphism}
-
-Typed Clojure includes support for F-bounds on type variables, as an extension
-to Local Type Inference.
-
-An extra environment from type variables to bounds is kept until after inference,
-after which each unground bound is substituted with the inferred types and the
-subtyping relationships are checked.
-
-\section{Heterogeneous Maps}
-% keyword keys for now, generally encouraged, fast
-% Some heterogeneous maps have string args, eg. from web frameworks
-
-\section{Java Interoperability}
-% Args non-nil, return nilable
-% Constructors and other special methods handled appropriately (eg. Constructor are non-nil return)
-
-\subsection{Primitive Arrays}
-% Primitive arrays are covariant, which is well-known to be statically unsound.
-% We cannot trust the type signature of any Java method or field.
-% Separate read and write types into two parameters.
-
-\subsection{Interaction with null}
-% null is explicitly expressible in my type system
-% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
-%
-
-\subsection{Generic Java}
-% Existential types are needed to fully express Generic Java types in Typed Clojure
-% F-bounded polymorphism supported
-
-\section{Occurrence Typing}
-% interesting paths
-% - count path element
-% - class path element
-% - keyword path element
-
-\section{Variable-arity Polymorphism}
-% Problematic areas
-% - flattened paired arguments. assoc, hash-maps, array-map
-% - complex argument dependencies. comp, partial
-
-\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}
-
-Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
-representing a sequence of types, allowing non-uniform variable parameters.
-This enables \lstinline|map| and other function with complex arguments to be typed.
-Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
-Typed Racket, and is a key feature of Typed Racket's implementation.
-
-Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
-to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
-variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
-associates key-value pairs to a target map. It takes one or more key-value pairs and
-applies them left-to-right to the target map. The key problem from a typing perspective
-is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
-accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
-higher-order function.
-\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
-dotted parameters represent an unrestricted number of types, where we require an even number.
-
-The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
-new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
-Firstly, subtyping between different kinds of dotted and rest
-\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
-parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
-Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
-difficult to express.
-Thirdly, higher-order functions taking functions with dotted arguments commonly 
-The first issue requires careful design and implementation.
-The second and third points require new constructs, in the form of \emph{projections}
-and abstractions over \emph{dotted} parameters, respectively.
-
-\section{Implementation}
-
-\subsection{Composite Propositions}
-
-\subsection{Non-hygienic Macro Expansion and Filters}
-% need to remove bindings as they fall out of scope
-
-\section{Future Work}
-
-\subsection{Multimethods}
-% Problem - typing isa?
-% - probably impossible to type higher order usages of isa?.
-% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
-% Statically proving at least one relevant dispatch method is present.
-% Warn potential method conflicts.
-% Suppress warnings, similar to prefer-method syntax
-
-\subsection{Records}
-% A composition of existing features: Datatypes and Heterogeneous Maps
-
-\subsection{Other Clojure Dialects}
-% unify ASTs for easier porting
-
-\subsection{Linter}
+\include{future-work}
 
@@
 
+\printbibliography[title=References]
+
 \end{document}

### 2012-09-11 dissertation/design-choices.tex
more grammar stuff

@@
 
+%$$
+%\begin{tdisplay}{Test1}
+%\begin{altgrammar}
+%  \s{},\t{} &::= & \int \alt \bool \alt {\proctype {\s{}} {\t{}}} 
+%  &\mbox{Types}\\
+%  \M{}, \N{}, \P{}, \dots &::=& \x{} \alt \abs{\x{}}{\M{}} \alt
+%  \comb{\M1}{\M2}  &\mbox{Raw Terms} \\
+%\end{altgrammar}
+%\end{tdisplay} 
+%$$
+
 $$
-\begin{tdisplay}{Test1}
+\begin{tdisplay}{Syntax of Terms}
 \begin{altgrammar}
-  \s{},\t{} &::= & \int \alt \bool \alt {\proctype {\s{}} {\t{}}} 
-  &\mbox{Types}\\
-  \M{}, \N{}, \P{}, \dots &::=& \x{} \alt \abs{\x{}}{\M{}} \alt
-  \comb{\M1}{\M2}  &\mbox{Raw Terms} \\
+  \exp{} &::=& (\exp{} \overrightarrow{\exp{}})         % function application, multiple arguments
+             \alt \{ \overrightarrow{\exp{}\ \exp{}} \} % map literal
+             &\mbox{Expressions} \\ 
+   \v{} &::=& \k{} \alt \nil{}
+               \alt \{ \overrightarrow{\v{}\ \v{}} \}   % map literal
+              &\mbox{Values}
 \end{altgrammar}
@@
 $$
-\begin{tdisplay}{Syntax of Terms}
+\begin{tdisplay}{Evaluation Contexts}
   \begin{altgrammar}
-  \exp{} &::=&  \x{} \alt \{ \overrightarrow{\exp{}\ \exp{}} \} 
-             \alt [ \overrightarrow{\exp{}} ] 
-             &\mbox{Expressions} \\ 
-  \v{} &::=& \alt \{ \overrightarrow{\v{}\ \v{}} \} 
-             \alt [ \overrightarrow{\v{}} ] 
-  &\mbox{Values} \\ 
+    \E{} &::=& % application rules
+              (\E{}\ \overrightarrow{\exp{}})  % target first
+              \alt (\v{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % arguments next
+              % map rules
+              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
+              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
+              % vector rule
+              \alt [ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}} ] &\mbox{Evaluation Contexts}
   \end{altgrammar}
@@
 {\inferrule[E-Assoc]
-  { {\Gamma \vdash e_1 \hookrightarrow v_1} \\
-    {\Gamma \vdash e_2 \hookrightarrow v_2} \\
-    {\Gamma \vdash e_3 \hookrightarrow v_3} \\
+  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
     {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
-  { {\Gamma \vdash (assoc\ e_1\ e_2\ e_3) \hookrightarrow v_4} }} \\
+  { {\Gamma \vdash (assoc\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
 \\
 {\inferrule[E-Dissoc]
-  { {\Gamma \vdash e_1 \hookrightarrow v_1} \\
-    {\Gamma \vdash e_2 \hookrightarrow v_2} \\
-    {v_3 = v_1\ without\ entry\ with\ key\ v_2}}
- {\Gamma \vdash (dissoc\ e_1\ e_2) \hookrightarrow v_3} } \\
+  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
+    {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
+  {\Gamma \vdash (dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
 \\
-{\inferrule[E-Var]
-  {\Gamma(x) = v}
-  {\Gamma \vdash x \hookrightarrow v}}
+{\inferrule[E-GetMapExist]
+  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
+    {v_1\ has\ entry\ with\ key\ v_2} \\
+    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1} }
+  { {\Gamma \vdash (get\ v_1\ v_2) \hookrightarrow v_3} }} \\
+\\
+{\inferrule[E-GetMapNotExist]
+  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
+    {v_1\ has\ no\ entry\ with\ key\ v_2} \\
+    {v_3 = \nil{}} }
+  { {\Gamma \vdash (get\ v_1\ v_2) \hookrightarrow v_3} }}
 \end{array} 
 \end{tdisplay} 
+$$
+
+$$
+\begin{tdisplay}{Type Syntax}
+\begin{altgrammar}
+  \T{} &::=& \Top \alt \Nil \alt \k{}
+\end{altgrammar}
+\end{tdisplay}{Type Syntax}
 $$

### 2012-09-12 dissertation/design-choices.tex
Added some type rules and type metafunctions

@@
 \begin{altgrammar}
-  \exp{} &::=& (\exp{} \overrightarrow{\exp{}})         % function application, multiple arguments
+  \exp{} &::=& (\c{} \overrightarrow{\exp{}})         % function application, multiple arguments
              \alt \{ \overrightarrow{\exp{}\ \exp{}} \} % map literal
@@
               \alt \{ \overrightarrow{\v{}\ \v{}} \}   % map literal
-              &\mbox{Values}
+              &\mbox{Values} \\
+  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
+              &\mbox{Constants}
 \end{altgrammar}
@@
     \E{} &::=& % application rules
-              (\E{}\ \overrightarrow{\exp{}})  % target first
-              \alt (\v{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % arguments next
+              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
               % map rules
@@
               \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
-              % vector rule
-              \alt [ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}} ] &\mbox{Evaluation Contexts}
+              &\mbox{Evaluation Contexts}
   \end{altgrammar}
@@
     {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
-  { {\Gamma \vdash (assoc\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
+    {{(\assoc{}\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
 \\
@@
     {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
-  {\Gamma \vdash (dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
+  {(dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
 \\
@@
     {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1} }
-  { {\Gamma \vdash (get\ v_1\ v_2) \hookrightarrow v_3} }} \\
+  { {(get\ v_1\ v_2) \hookrightarrow v_3} }} \\
 \\
@@
     {v_3 = \nil{}} }
-  { {\Gamma \vdash (get\ v_1\ v_2) \hookrightarrow v_3} }}
+  { {(get\ v_1\ v_2) \hookrightarrow v_3} }}
 \end{array} 
@@
 \begin{altgrammar}
-  \T{} &::=& \Top \alt \Nil \alt \k{}
+  \T{} &::=& \Top \alt \Bottom \alt \Nil \alt \k{} \alt \{ \overrightarrow{\k{}\ \T{}} \}
+             \alt {\IPersistentMap {\T{}} {\T{}}} \alt ( \cup\ \overrightarrow{\T{}} )
+             \alt ( \cap\ \overrightarrow{\T{}} )
 \end{altgrammar}
-\end{tdisplay}{Type Syntax}
+\end{tdisplay}
+$$
+
+$$
+\begin{tdisplay}{Core Type Rules}
+\begin{array}{c}
+{\inferrule[T-AssocHMap]
+  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\k{}}}} } \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
+    {\vdash {\issubtype {{\T{}}_{m}} {{ \{ \} }} }} \\
+    {\T{} = (update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})} }
+  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
+\\
+{\inferrule[T-AssocHMapPromote]
+  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}}} } \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
+    {\vdash {\issubtype {{\T{}}_{m}} { \{ \}}}} \\
+    {\T{} = (promote\_hmap\ {\T{}}_{m}\ {\T{}}_{k}\ {\T{}}_{v})} }
+  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
+\\
+{\inferrule[T-AssocIPersistentMap]
+  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}} }} \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}} }} \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}} }} }
+  { {\Gamma \vdash {\hastype {(\assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {T}}}}}
+\end{array}
+\end{tdisplay}
+$$
+
+$$
+\begin{tdisplay}{Type Metafunctions}
+\begingroup
+\fontsize{10pt}{12pt}
+\begin{altgrammar}
+  (update\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\k{}}_1\ {\T{}}_1) = 
+                 \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}}\ {\k{}}_1\ {\T{}}_1 \}\ \\
+  (update\_hmap\ (\cap\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v} = 
+                               (\cap\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})} \\
+  (update\_hmap\ (\cup\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v} = 
+                               (\cup\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
+  \\ \\
+  (promote\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\T{}}_{kn}\ {\T{}}_{vn})
+                                    = {\IPersistentMap {(\cup\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
+                                                       {(\cup\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
+  (promote\_hmap\ {\IPersistentMap {{\T{}}_{k}} {{\T{}}_{v}}} \ {\T{}}_{kn}\ {\T{}}_{vn})
+                                    = {\IPersistentMap {(\cup\ {\T{}}_{k}\ {\T{}}_{kn})}
+                                                       {(\cup\ {\T{}}_{v}\ {\T{}}_{vn})}}
+\end{altgrammar}
+\endgroup
+\end{tdisplay}
 $$

### 2012-10-02 dissertation/design-choices.tex
Experiments

@@
 Typed Clojure is designed to be of practical use to Clojure programmers.
-Many of the design goals are similar to Tobin-Hochstadt's \cite{SAMTH:dissertation}
+Many of the design goals are similar to Tobin-Hochstadt's \cite{Tob10}
 for Typed Racket.
@@
 
+\section{Multimethods}
+
+% INTRODUCTION TO CLOJURE MULTIMETHODS
+
+%A multimethod in Clojure differs from most other
+%abstractions of the same name. Clojure multimethods
+%dispatch on the result of an arbitrary function
+%rather than the types or structure of its arguments.
+%
+%\begin{lstlisting}[caption=Clojure Multimethods, label=lst:mm1]
+%(defmulti ToString class)
+%
+%(defmethod ToString java.lang.Integer
+%  [a]
+%  (str "Integer: " a))
+%
+%(defmethod ToString java.lang.Number
+%  [a]
+%  (str "Number: " a))
+%\end{lstlisting}
+%
+%Even with this extra flexibility, multimethods are often used to dispatch on the type of its arguments.
+%Listing \ref{lst:mm1} illustrates a common usage of multimethods: dispatching on the class of the first argument.
+%We use \lstinline|defmulti| to define a new multimethod.
+%The function \lstinline|class| is used as the dispatch function, a one-argument function returning the class of its argument
+%as an instance of \lstinline|java.lang.Class|.
+%
+%A multimethod invocation redirects control flow to exactly one of its \emph{methods} with unresolved ambiguities resulting in
+%a runtime error. The method is selected by first comparing the result of the dispatch function to
+%the dispatch values defined by each method. If one or more methods match, multimethods provide
+%the ability to prefer one method over another.
+
+\section{Clojure Class Hierarchy}
+\section{Protocols and Datatypes}
 \section{Optional Type Checking}
@@
   \begin{altgrammar}
-    \E{} &::=& % application rules
+    \E{} &::=& [ ] % application rules
               \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
@@
 
+% \mapsto for updating maps
 $$
@@
   { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
-    {v_1\ has\ entry\ with\ key\ v_2} \\
-    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1} }
+    {v_2\ v_3\ in\ v_1} }
+%    {v_1\ has\ entry\ with\ key\ v_2} \\
+%    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1}
   { {(get\ v_1\ v_2) \hookrightarrow v_3} }} \\
@@
   \T{} &::=& \Top \alt \Bottom \alt \Nil \alt \k{} \alt \{ \overrightarrow{\k{}\ \T{}} \}
-             \alt {\IPersistentMap {\T{}} {\T{}}} \alt ( \cup\ \overrightarrow{\T{}} )
-             \alt ( \cap\ \overrightarrow{\T{}} )
+             \alt {\IPersistentMap {\T{}} {\T{}}} \alt ( \vee\ \overrightarrow{\T{}} )
+             \alt ( \wedge\ \overrightarrow{\T{}} )
 \end{altgrammar}
@@
     {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
-    {\vdash {\issubtype {{\T{}}_{m}} {{ \{ \} }} }} \\
-    {\T{} = (update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})} }
-  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
+    {\vdash {\issubtype {{\T{}}_{m}} {\TopHMap} }} }
+  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} 
+        {(update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})}}}}} \\
 \\
-{\inferrule[T-AssocHMapPromote]
+{\inferrule[T-AssocPromote]
   { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
@@
     {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
-    {\vdash {\issubtype {{\T{}}_{m}} { \{ \}}}} \\
+    {\vdash {\issubtype {{\T{}}_{m}} {\TopIPersistentMap}}} \\
     {\T{} = (promote\_hmap\ {\T{}}_{m}\ {\T{}}_{k}\ {\T{}}_{v})} }
   { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
-\\
-{\inferrule[T-AssocIPersistentMap]
-  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}} }} \\
-    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}} }} \\
-    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}} }} }
-  { {\Gamma \vdash {\hastype {(\assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {T}}}}}
 \end{array}
@@
                  \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}}\ {\k{}}_1\ {\T{}}_1 \}\ \\
-  (update\_hmap\ (\cap\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v} = 
-                               (\cap\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})} \\
-  (update\_hmap\ (\cup\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v} = 
-                               (\cup\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
+  (update\_hmap\ (\wedge\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
+                               (\wedge\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})}) \\
+  (update\_hmap\ (\vee\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
+                               (\vee\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
   \\ \\
   (promote\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\T{}}_{kn}\ {\T{}}_{vn})
-                                    = {\IPersistentMap {(\cup\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
-                                                       {(\cup\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
+                                    = {\IPersistentMap {(\vee\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
+                                                       {(\vee\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
   (promote\_hmap\ {\IPersistentMap {{\T{}}_{k}} {{\T{}}_{v}}} \ {\T{}}_{kn}\ {\T{}}_{vn})
-                                    = {\IPersistentMap {(\cup\ {\T{}}_{k}\ {\T{}}_{kn})}
-                                                       {(\cup\ {\T{}}_{v}\ {\T{}}_{vn})}}
+                                    = {\IPersistentMap {(\vee\ {\T{}}_{k}\ {\T{}}_{kn})}
+                                                       {(\vee\ {\T{}}_{v}\ {\T{}}_{vn})}}
 \end{altgrammar}
@@
 
-\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}
+$$
+\begin{tdisplay}{Subtyping}
+\begin{array}{c}
+\inferrule[S-Refl] { } {\vdash {\issubtype {\T{}} {\T{}}}} \\
+\inferrule[S-Any] { } {\vdash {\issubtype {\T{}} {\Top}}}
+\inferrule[S-UnionSuper] 
+  {\ } 
+  {\vdash {\issubtype {\T{}} {\Top}}}
+\end{array}
+\end{tdisplay}
+$$
 
-Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
-representing a sequence of types, allowing non-uniform variable parameters.
-This enables \lstinline|map| and other function with complex arguments to be typed.
-Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
-Typed Racket, and is a key feature of Typed Racket's implementation.
-
-Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
-to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
-variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
-associates key-value pairs to a target map. It takes one or more key-value pairs and
-applies them left-to-right to the target map. The key problem from a typing perspective
-is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
-accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
-higher-order function.
-\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
-dotted parameters represent an unrestricted number of types, where we require an even number.
-
-The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
-new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
-Firstly, subtyping between different kinds of dotted and rest
-\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
-parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
-Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
-difficult to express.
-Thirdly, higher-order functions taking functions with dotted arguments commonly 
-The first issue requires careful design and implementation.
-The second and third points require new constructs, in the form of \emph{projections}
-and abstractions over \emph{dotted} parameters, respectively.
+%\subsection{Higher-order Variable-arity Polymorphism (SKETCH)}
+%
+%Practical Variable-arity Polymorphism introduces the concept of a dotted type parameter,
+%representing a sequence of types, allowing non-uniform variable parameters.
+%This enables \lstinline|map| and other function with complex arguments to be typed.
+%Strickland, Tobin-Hochstadt, and Felleisen's approach is solves almost all variable-arity applications in
+%Typed Racket, and is a key feature of Typed Racket's implementation.
+%
+%Idiomatic variable-arity functions in Clojure's are even more sophisticated than Typed Racket, requiring extensions
+%to Typed Racket's approach to type check satisfactorily. The Clojure core library favours functions with
+%variable parameters, especially in a higher-order context. For example, \lstinline|assoc|
+%associates key-value pairs to a target map. It takes one or more key-value pairs and
+%applies them left-to-right to the target map. The key problem from a typing perspective
+%is \lstinline|assoc|'s arguments are flattened, resulting in \lstinline|assoc| only
+%accepting an even number of rest arguments. \lstinline|assoc| is also commonly used as a
+%higher-order function.
+%\lstinline|assoc| cannot be expressed in Typed Racket. For instance,
+%dotted parameters represent an unrestricted number of types, where we require an even number.
+%
+%The nature of Clojure's idiomatic variable-arity functions suggest we must introduce
+%new kinds of dotted parameters. Three main issues must be considered in the context of Typed Clojure.
+%Firstly, subtyping between different kinds of dotted and rest
+%\footnote{A rest parameter consumes zero or more types, rather than being a sequence of types.}
+%parameters can be tricky, only made trickier by adding new kinds of dotted parameters.
+%Secondly, elimination rules for dotted parameters in many frequently used Clojure core functions
+%difficult to express.
+%Thirdly, higher-order functions taking functions with dotted arguments commonly 
+%The first issue requires careful design and implementation.
+%The second and third points require new constructs, in the form of \emph{projections}
+%and abstractions over \emph{dotted} parameters, respectively.

### 2012-10-02 dissertation/experiments.tex
Experiments

\chapter{Experiments}

% 
In this chapter we look at several example of using Typed Clojure and
how well the current prototype handles them.

\section{Java Interoperability}

As Typed Clojure is intended to be useful for practical purposes, it was important
to port existing code that utilized Java interoperability.
My porting effort attempted to follow how a real programmer might port code to Typed Clojure
and I describe my approach for each situation.

I ported a function from clojure.contrib.reflect\footnote{https://github.com/cemerick/pomegranate}, 
a Clojure library relying heavily on Java interoperability.
I chose \lstinline|call-method| for several reasons: it chains several Java calls together,
it uses primitive arrays, it exposes a wart of Clojure with \emph{Named} things,
and \lstinline|null| is a valid value in one place.

Before showing the implementation, there is a brief explanation of the relevant syntax.
Java methods are called using the \emph{dot} operator. If \lstinline|o| is an object \lstinline|(. o m a*)|
calls its method named \lstinline|m| passing \lstinline|a*| arguments. The method can be named first
using the equivalent sugar \lstinline|(.m o a*)|. Also, \lstinline|doto| is convenient 
notation for muliple method calls on the same object, presumably for side effects, and returns the original
object. For example, \lstinline|(doto o (.m1 a*) (m2 a*))| calls methods \lstinline|m1| and \lstinline|m2|
on \lstinline|o| and returns \lstinline|o|.

\begin{lstlisting}[caption=call-method, label=lst:callmethod]
(defn call-method
  "Calls a private or protected method.

  params is a vector of classes which correspond to the arguments to
  the method e

  obj is nil for static methods, the instance object otherwise.

  The method-name is given a symbol or a keyword (something Named)."
  [^Class klass method-name params obj & args]
  (let [method (doto (.getDeclaredMethod klass 
                                         (name method-name)
                                         (into-array Class params))
                 (.setAccessible true))]
    (.invoke method obj (into-array Object args))))
\end{lstlisting}

The original function is modified slightly for readability and is presented in listing \ref{lst:callmethod}.

\begin{lstlisting}[caption=call-method Type Annotation, label=lst:callmethodtype]
(ann call-method 
     [Class Named (IPersistentVector Class) (U nil Object) (U nil Object) * -> (U nil Object)])
\end{lstlisting}

Thankfully this function has up-to-date documention, and from it we can derive an expected type
(listing \ref{lst:callmethodtype}).

Before running the type checker we must convert our array constructors into ones that Typed Clojure
can understand. Array types in Typed Clojure are represented by \lstinline|(Array c t)|, where
the Java class \lstinline|c| is the Java component type,
and the Typed Clojure type \lstinline|t| is the Typed Clojure component type. 
We can pass this array to Java methods accepting
type \lstinline|c[]|, and we can read and write type \lstinline|t| to the array from Typed Clojure.

\begin{itemize}
\item \lstinline|(into-array Class params)| becomes \lstinline|(into-array> Class Class params)|
      which is of type \lstinline|(Array Class Class)|.
\item \lstinline|(into-array Object args)| becomes \lstinline|(into-array> Object (U nil Object) args)|
      which is of type \lstinline|(Array Object (U nil Object))|.
\end{itemize}

The second place in the \lstinline|Array| type constructor allows fine grained control over what is allowed
in the array. The first point above must be type \lstinline|(Array Class Class)| because the 
\lstinline|getDeclaredMethod| method on \lstinline|java.lang.Class| instances requires an array of non-null
\lstinline|Class| objects. On the other hand, the \lstinline|invoke| method takes an array that allows
\emph{null} members, so its type is \lstinline|(Array Object (U nil Object))|.

\section{Red-Black Tree}

### 2012-10-02 dissertation/lit-review.tex
Experiments

@@
 % Background
+% Heterogeneous Maps/record types
 % Soft typing
@@
 
-%\begin{abstract}
-%Clojure is a practical, dynamically typed language running on the Java Virtual Machine.
-%I intend to design and implement a prototype optional static type 
-%system for Clojure, with practical use being a priority. Devising a static type system 
-%requires a combination of components that each adds some utility
-%without significantly complicating the overall implementation.
-%I look at several novel inventions created as a result 
-%of the Typed Racket project including
-%non-uniform variable-arity polymorphism and occurrence typing.
-%Typed Racket and Scala both use flavours of local type inference
-%which I compare and judge for applicability to this project.
-%Other relevant features of Scala are considered, such as its approach to bounded polymorphism,
-%the Scala class hierarchy, and the similarities of its methods of composition to Clojure's.
-%\end{abstract}
+Typed Clojure is related to several existing systems for both typed and untyped languages.
+The first section of this chapter gives an overview of type systems designed for untyped languages.
+The second section explores how existing type systems statically type check multimethods
+and their relevance to Clojure's multimethods.
 
-\section{Introduction}
+\section{Static Type Systems}
+
+% Haskell
+% OCaml
+% Forsythe
+% Java C#
+% Scala
+
+\section{Static Types for Untyped Languages}
+
+There is a history of designing static type systems for untyped languages.
+
+\subsection{Soft Typing}
+
+Soft typing\cite{CF91}
+is an approach for ensuring type safety in untyped languages.
+A soft type system infers types for programs, distinguishing between degrees
+of potential type safety.
+A soft type checker uses this information to 
+preserve type safety by inserting appropriate checks 
+and informs the programmer of potential inconsistencies.
+For example, if the soft type system detects a portion of code is sometimes not type safe,
+the soft type checker inserts a check that throws a runtime error upon unsafe usages.
+Thus, soft type systems differ from traditional type systems in that type inference 
+never fails and an inconsistency always results in a runtime check.
+
+Wright and Cartwright\cite{WC97} developed Soft Scheme, a practical
+soft type system for Scheme. 
+It extended earlier work by Cartwright and Fagan\cite{CF91}
+and Fagan \cite{Fag92}, adding support for practical features such as
+first-class continuations and variable-arity functions.
+Soft Scheme does not require any extra type annotations.
+
+% what challenges did soft typing face?
+
+\subsection{Gradual Typing}
+
+Gradual typing combines static and runtime type checking so programmers
+can the most appropriate one for the situtation.
+
+Typed Racket was developed as a path for module-by-module
+porting of existing untyped Racket modules to a typed sister language\cite{Tob10}.
+Once a module is ported and type checked, it is protected from untyped modules
+by inserting runtime checks.
+
+\section{Interlanguage Interoperability}
+
+This section compares several existing languages that feature interlanguage interoperability.
+
+Clojure is a dynamic functional language hosted on the Java Virtual Machine. It provides 
+interoperability with Java libraries. As Clojure is a dynamically typed language, it does
+not give strong type type guarantees at compile time whether interactions with Java
+are type safe.
+
+Scala is a statically typed language on the Java Virtual Machine offering integrated interoperability with Java, a typed language.
+Scala objects and classes can ``inherit from Java classes and implement Java interfaces''\cite{OCD+}
+with the usual static type guarantees normal Scala code enjoys.
+Scala offers an Option type to safely eliminate null pointers.
+Java Generics are also fully supported by Scala, accounting for Scala support existential types
+
+Typed Racket includes safe interoperability between any combination of typed and untyped 
+Racket modules\cite{Tob10}\cite{TF08}. 
+Interactions with untyped modules are protected by adding runtime checks based on expected types.
+Typed Racket implements a sophisticated blame calculus. It ensures 
+error messages correctly \emph{blame} the source of type errors,
+which can be difficult to determine in the presence of higher-order functions. %TODO reference
+
+\section{Record Types}
+
+This section summarises research in typing OCaml-style records
+which relate to Clojure records and heterogeneous hash maps.
+
+\section{Intersection Types}
+
+Intersection types are an interesting type construct relevant to capturing the complicated
+types manipulated by dynamic languages.
+
+Forsyth\cite{Rey96} was the first language to use intersection types.
+Uses of intersections in Forsyth include representing extensible record types
+and function overloading.
+
+% CIDRE
+% TODO
+CIDRE\cite{Dav05} adds \emph{refinement type checking} to SML.
+
+% PADL '12 paper, Typing the Numeric Tower
+Typed Racket uses \emph{ordered intersection types} to offer a kind of function overloading.
+
+Java allows intersection types only in type parameter bounds. Instead of supporting intersection
+types in the rest of the language, Java support type inheritance by extending Classes and Interfaces.
+
+\section{Statically Typed Multimethods}
 
@@
 for such languages should preserve existing idioms and syntactic abstractions
-until time tested idioms are devised for the new typed language \cite{SAMTH:dissertation}.
+until time tested idioms are devised for the new typed language \cite{Tob10}.
 
-Typed Racket \cite{SAMTH:dissertation} integrates a selection of ideas to
+Typed Racket \cite{Tob10} integrates a selection of ideas to
 create a typed sister language to Racket.
-Both Clojure and Racket are dynamically typed Lisps and unsurprisingly have similar idioms. 
-Consequently, a number of novel Typed Racket features are relevant to designing and implementing 
-a type system for Clojure.
-The inventions of occurrence typing \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
+Clojure and Racket share a key similarity: they are dynamically typed Lisps.
+Consequently, a number of novel Typed Racket features are relevant to the design and implementation
+of a type system for Clojure.
+The inventions of occurrence typing \cite{TF10}
 and a systematic approach to variable-arity polymorphism
-\cite{Strickland:2009:PVP:1532974.1532978}
+\cite{STF09}
 are both applicable to this project.
@@
 results of an arbitrary function; and protocols, a mechanism for single dispatch.
-Millstien and Chambers explore statically typing multiple dispatched multimethods in Dubious \cite{Millstein02modularstatically}.
+Millstien and Chambers explore statically typing multiple dispatched multimethods in Dubious \cite{MS02}.
 Dubious' multimethods are less powerful than Clojure's multimethods but more powerful than protocols,
@@
 
-Clojure and the statically typed language Scala \cite{Odersky06anoverview} 
-have some interesting similarities. The recently redesigned 
-Scala collection hierarchy
-\cite{Odersky06anoverview} \cite{Scala:Collections}
+Clojure and the statically typed language Scala \cite{OCD+} 
+have some interesting similarities. The redesigned 
+of the Scala collection hierarchy
+\cite{OS10}
 bears similarity to Clojure's encompassing ``seq''
@@
 to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
-relevant, particularly considering the potential of using occurrence typing \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
+relevant, particularly considering the potential of using occurrence typing \cite{TF10}
 to prevent erroneous usage of \lstinline|null|.
@@
 Support for complex variable-arity functions is sparse in typed languages,
-Strickland claims \cite{Strickland:2009:PVP:1532974.1532978}
+Strickland, Tobin-Hochstadt, and Felleisen claim \cite{STF09}
 ``no typed language supports them in a systematic and principled manner''.
 
-Strickland et al. invented \cite{Strickland:2009:PVP:1532974.1532978} 
+Strickland et al. invented \cite{STF09} 
 a calculus capable of typing non-uniform variable-arity functions, a version of which
@@
 
-One approach to typing \lstinline|dissoc| could be to include in every hash-map type how to dissociate exactly one
-key from itself, and then treat subsequent entries passed to \lstinline|dissoc| individually in a similar manner.
-This is a complex polymorphic relationship between the parameter and result types useful for typing
-several core Clojure functions.
-
 \section{Type Inference}
@@
 
-Typed Racket uses Local Type Inference \cite{Pierce:2000:LTI:345099.345100}
+Typed Racket uses Local Type Inference \cite{PT00}
 as an inference and checking tool. Pierce and Turner
-\cite{Pierce:2000:LTI:345099.345100} divide Local Type Inference into
+\cite{PT00} divide Local Type Inference into
 two complementary algorithms. \emph{Local type argument synthesis}
@@
 Pierce and Turner split \emph{local type argument synthesis} into two further
-algorithms: bounded, and unbounded quantification \cite{Pierce:2000:LTI:345099.345100}. 
+algorithms: bounded, and unbounded quantification \cite{PT00}. 
 Typed Racket 
-supports unbounded polymorphism \cite{SAMTH:dissertation}, implementing the latter algorithm by Piece et al.
-Scala supports bounded quantification with F-bounded polymorphism \cite{Canning:1989:FPO:99370.99392},
+supports unbounded polymorphism \cite{Tob10}, implementing the latter algorithm by Piece et al.
+Scala supports bounded quantification with F-bounded polymorphism \cite{CCHOM89},
 basing its type argument synthesis on the bounded quantification algorithm.
 
-Pierce and Turner explicitly forbid \cite{Pierce:2000:LTI:345099.345100}
+Pierce and Turner explicitly forbid \cite{PT00}
 attempting to synthesise type variables with interdependent bounds, including
@@
 I am not aware of papers specifically describing Scala's modifications, but they are at least inspired by
-Scala's spiritual ancestors Generic Java \cite{Bracha98gj:extending} and Pizza \cite{Odersky97pizzainto}.
+Scala's spiritual ancestors Generic Java \cite{BOSW98} and Pizza \cite{OW97}.
 
-Hosoya and Pierce \cite{Hosoya99howgood} reiterate two common problems with Local Type Inference:
+Hosoya and Pierce \cite{HP99} reiterate two common problems with Local Type Inference:
 ``hard-to-synthesise arguments'' and ``no best type argument''. The first problem occurs because
@@
 
-Scala's type checking uses Colored Local Type Inference \cite{Odersky:2001:CLT:373243.360207},
-a variant of Local Type Inference \cite{Pierce:2000:LTI:345099.345100} specifically designed to
+Scala's type checking uses Colored Local Type Inference \cite{OZZ01},
+a variant of Local Type Inference \cite{PT00} specifically designed to
 improve inference with certain kinds of Scala pattern matching expressions. It allows
@@
 for describing ``information flow in polymorphic type systems with propagation-based type inference''
-\cite{Odersky:2001:CLT:373243.360207}. If the scope of my research permits it, I plan to investigate
+\cite{OZZ01}. If the scope of my research permits it, I plan to investigate
 using colored types to solve the polymorphic higher-order-function limitation of Local Type Inference
@@
 
-The Local Type Inference paper \cite{Pierce:2000:LTI:345099.345100}
+The Local Type Inference paper \cite{PT00}
 describes two implementations of type variables, for bounded
@@
 Bounded type variables can have subtype constraints, and 
-subsume unbounded type variables \cite{Pierce:2000:LTI:345099.345100}, 
+subsume unbounded type variables \cite{PT00}, 
 as a unbounded variable can be represented as a variable bounded
@@
 is more subtle, due to ``some surprising interactions between bounded quantifiers
-and the \emph{Bot} type'' \cite{Pierce:2000:LTI:345099.345100}, described fully
-by Pierce \cite{Pierce97boundedquantification}.
+and the \emph{Bot} type'' \cite{PT00}, described fully
+by Pierce \cite{Pie97}.
 
-Typed Racket \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
-supports unbounded polymorphism, while Scala \cite{Odersky06anoverview}
+Typed Racket \cite{TF08}
+supports unbounded polymorphism, while Scala \cite{OCD+}
 supports an extended form of bounded polymorphism called
-F-bounded polymorphism \cite{Canning:1989:FPO:99370.99392}, which allows the
+F-bounded polymorphism \cite{CCHOM89}, which allows the
 bound variable to occur in its own bound.
 F-bounded polymorphism is useful in the context of object-oriented abstractions,
-as demonstrated by Odersky \cite{Odersky06anoverview}.
+as demonstrated by Odersky \cite{OCD+}.
 This is one possible explanation why Typed Racket, which is not built on abstractions like Scala,
@@
 attempts to preserve existing Racket idioms and aims type check
-existing Racket code by simply adding top level type annotations \cite{SAMTH:dissertation}.
+existing Racket code by simply adding top level type annotations \cite{Tob10}.
 
-Typed Racket fully expands all macro calls before type checking \cite{SAMTH:dissertation} to
+Typed Racket fully expands all macro calls before type checking \cite{Tob10} to
 avoid the complex semantics of type checking macro definitions, an ongoing research area summarised
- by Herman \cite{Herman10:Theory}.
+ by Herman \cite{Her10}.
 The design of my Clojure type system will follow a similar strategy; only the fully macro-expanded form
@@
 Along with a full static type system, Typed Racket 
-also uses runtime contracts to enforce type invariants at runtime \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}.
+also uses runtime contracts to enforce type invariants at runtime \cite{TF08}.
 Utilising runtime contracts to aid type checking is outside the scope of this project, but would be 
@@
 Two other Typed Racket features that will be explored are recursive types and refinement types  
-\cite{SAMTH:dissertation}. Recursive types allow a type definition to refer to itself, enabling structurally
+\cite{Tob10}. Recursive types allow a type definition to refer to itself, enabling structurally
 recursive types like binary trees. Refinement types let the programmer define
@@
 
-Occurrence typing \cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
-\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561} extends the type 
+Occurrence typing \cite{TF08}
+\cite{TF10} extends the type 
 system with a \emph{proposition environment} that represents 
 the type refinements inferred down a particular path.
-The 2010 paper \cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
+The 2010 paper \cite{TF10}
 reformulated occurrence typing, improving the original 2008 paper
-\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
+\cite{TF08}
 after revealing that ``three years of practical experience has revealed
-serious shortcoming of our type system.''\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}
+serious shortcoming of our type system.''\cite{TF10}
 The new implementation utilizes a simple proof system to solve
@@
 Listing \ref{lst:nil} infers from the Java signature \lstinline|Object getObject()| that
-that \lstinline|a| is of type \lstinline|(Union nil Object)|. This is equivalent to Java's
-\lstinline|Object| class, as \lstinline|null| is a subtype of all reference types. By surrounding
+that \lstinline|a| is of type \lstinline|(U nil Object)|. This is equivalent to Java's
+\lstinline|Object| static type, as \lstinline|null| is a subtype of all reference types. By surrounding
 the call \lstinline|(expects-non-nil a)| with \lstinline|(when a ...)|, we guarantee that
@@
 
-A Clojure multimethod is slightly different to most other
-abstractions of the same name. Here, a multimethod 
-dispatches on the result of an \emph{arbitrary function},
-rather that more traditional multimethods that dispatch on
-the types or structure of its arguments.
+Clojure provides mulimethods as a core language feature. This section discusses systems that statically
+verify type safety for multimethods.
 
-Millstein and Chambers \cite{Millstein02modularstatically}
+Millstein and Chambers\cite{MS02}
 describe Dubious, a simple statically typed core language including multimethods that
@@
 multimethods: ``it is possible for two modules containing arbitrary multimethods to typecheck
-successfully in isolation but generate type errors when linked together.'' \cite{Millstein02modularstatically}
-
-This challenge holds when statically typing Clojure multimethods.
-One additional challenge would be multimethods being imported from 
-untyped modules. This would have to be handled carefully.
-
-If we consider a multimethod as a spread out conditional, a
-novel use of occurrence typing becomes apparent. We can use
-occurrence typing to refine the argument types in the method bodies
-by utilizing the results of the dispatch function. This is simple
-when dispatching by type, but a common idiom in Clojure is to dispatch
-on the value associated with a key in a hash-map.
-
-To satisfy this idiom, we require occurrence typing with more
-finely grained hash-map types (see section~\ref{sec:OccurrenceTyping} for discussion).
+successfully in isolation but generate type errors when linked together.''\cite{MS02}
 
@@
 that is compatible with Java, and colored local type inference.
-
-I intend to design a static type system for Clojure
-that features a satisfying combination of features, individually taken from other
-languages, when combined produce novel and interesting problems.
-I have looked at the limitations of Local Type Inference and identified Colored Local
-Type Inference as potentially improving them.
-I have proposed a novel use of occurrence typing to prevent erroneous use of Java's
-null pointer, and identified a novel problem typing certain Clojure functions with variable
-arguments.
-There are some interesting problems and novel solutions to be discovered while undertaking this
-project, many of which will reveal themselves during development.

### 2012-10-10 dissertation/experiments.tex
Monad experiment + other changes

@@
 
-% 
 In this chapter we look at several example of using Typed Clojure and
 how well the current prototype handles them. 
+We intentially chose examples that could be challenging to type check
+or were particularly useful to the everyday Clojure programmer.
 
@@
 \begin{lstlisting}[caption=call-method, label=lst:callmethod]
+;; call-method pulled from clojure.contrib.reflect, (c) 2010 Stuart Halloway & Contributors
 (defn call-method
@@
 
-Thankfully this function has up-to-date documention, and from it we can derive an expected type
+Thankfully this function has up-to-date documentation, and from it we can derive an expected type
 (listing \ref{lst:callmethodtype}).
@@
 
+Now we run the type checker, which produces a type error.
+
+\begin{lstlisting}
+#<Exception java.lang.Exception: 29: Cannot call instance method java.lang.reflect.AccessibleObject/setAccessible on type (U nil java.lang.reflect.Method)>
+\end{lstlisting}
+
+Because Typed Clojure assumes all methods return nilable Objects, the call to \lstinline|getDeclaredMethod|
+has return type \lstinline|(U nil java.lang.reflect.Method)|. It is not type safe to call \lstinline|setAccessible|
+on this type, so we get a type error.
+
+In this case, Typed Clojure is too conservative: according to its documentation \lstinline|getDeclaredMethod|
+never returns \emph{null}. We add this rule with \lstinline|non-nil-return|.
+
+\begin{lstlisting}
+(non-nil-return java.lang.Class/getDeclaredMethod :all)
+\end{lstlisting}
+
+Running the type checker produces a different type error.
+
+\begin{lstlisting}
+#<Exception java.lang.Exception: Type Error, REPL:32 - (U java.lang.Object nil) is not a subtype of: java.lang.Object>
+\end{lstlisting}
+
+This concerns passing \lstinline|obj| as the first argument to the \lstinline|invoke| method.
+Typed Clojure conservatively defaults method parameter types as non-nullable. 
+Therefore the first parameter of \lstinline|invoke| is \lstinline|Object| by default; \lstinline|obj|
+is \lstinline|(U java.lang.Object nil)|. Again, this is too conservative as the first argument can
+be \emph{null} for static methods, and we use \lstinline|nilable-param| to specify the first argument
+of \lstinline|invoke| may be nil, for the arity of two parameters.
+
+\begin{lstlisting}
+(nilable-param java.lang.reflect.Method/invoke {2 #{0}})
+\end{lstlisting}
+
+Here is the final successfully type checked code.
+\begin{verbatim}Is this a reasonable result?\end{verbatim}
+
+\begin{lstlisting}
+(non-nil-return java.lang.Class/getDeclaredMethod :all)
+(nilable-param java.lang.reflect.Method/invoke {2 #{0}})
+
+(ann call-method [Class Named (IPersistentVector Class) (U nil Object) (U nil Object) * -> (U nil Object)])
+
+;; call-method pulled from clojure.contrib.reflect, (c) 2010 Stuart Halloway & Contributors
+(defn call-method
+  "Calls a private or protected method.
+
+  params is a vector of classes which correspond to the arguments to
+  the method e
+
+  obj is nil for static methods, the instance object otherwise.
+
+  The method-name is given a symbol or a keyword (something Named)."
+  [^Class klass method-name params obj & args]
+  (let [method (doto (.getDeclaredMethod klass 
+                                         (name method-name)
+                                         (into-array> Class Class params))
+                 (.setAccessible true))]
+    (.invoke method obj (into-array> Object (U nil Object) args))))
+\end{lstlisting}
+
 \section{Red-Black Tree}
+
+\section{Monads}
+
+Monads[] are an interesting control structure probably most recognisable from its inclusion
+in the statically-typed language Haskell[].
+
+I chose to port a subset of the Clojure library \emph{algo.monads} to Typed Clojure. The library
+provides support for several kinds of monads.
+
+This library represents a monad as a hash-map with four keys: \lstinline|:m-bind|, \lstinline|:m-result|,
+\lstinline|:m-zero|, and \lstinline|:m-plus|. A valid monad must provide the first two, and the latter
+two may optionally be mapped to the keyword \lstinline|::undefined|\footnote{Keywords prefixed with \lstinline|::|
+are qualified in the current namespace.}.
+
+\begin{lstlisting}[caption=Untyped definition for the identity monad, label=lst:identitymdef]
+(defmonad identity-m
+   "Monad describing plain computations. This monad does in fact nothing
+    at all. It is useful for testing, for combination with monad
+    transformers, and for code that is parameterized with a monad."
+  [m-result identity
+   m-bind   (fn m-result-id [mv f]
+              (f mv))
+  ])
+\end{lstlisting}
+
+\begin{lstlisting}[caption=Macro expansion of identity monad definition, label=lst:identitymexpand]
+(def identity-m
+  (let [m-bind :typed.test.monads/undefined
+        m-result :typed.test.monads/undefined
+        m-zero :typed.test.monads/undefined
+        m-plus :typed.test.monads/undefined
+        m-result identity
+        m-bind (fn m-result-id [mv f] (f mv))]
+    {:m-plus m-plus, :m-zero m-zero, :m-bind m-bind, :m-result m-result}))
+\end{lstlisting}
+
+A monad is defined using the macro \lstinline|defmonad|. The macro expands to code that binds a var to a hash-map
+with the aforementioned keys. For example, listing \ref{lst:identitymdef} expands to 
+the code in listing \ref{lst:identitymexpand}.
+
+\begin{lstlisting}[caption=Type for identity monad, label=lst:identitymtype]
+(ann identity-m
+     '{:m-bind (All [x y]
+                 [x [x -> y] -> y])
+       :m-result (All [x]
+                   [x -> x])
+       :m-zero '::undefined
+       :m-plus '::undefined})
+\end{lstlisting}
+
+Typed Clojure operates directly on fully macro-expanded forms, so it is unsurprising that the type
+annotation for \lstinline|identity-m| reflects its macro-expansion (listing \ref{lst:identitymtype}).
+The \lstinline|:m-bind| entry is comparable to Haskell's monadic \emph{bind},
+and the \lstinline|:m-result| entry is comparable to Haskell's monadic \emph{return}.
+
+\begin{lstlisting}[caption=Maybe monad definition, label=lst:maybemdef]
+(defmonad maybe-m
+  "Monad describing computations with possible failures. Failure is
+  represented by nil, any other value is considered valid. As soon as
+  a step returns nil, the whole computation will yield nil as well."
+  [m-zero   nil
+   m-result (fn m-result-maybe [v] 
+              v)
+   m-bind   (fn m-bind-maybe [mv f]
+              (if (nil? mv) nil (f mv)))
+   m-plus   (fn m-plus-maybe [& mvs]
+              (first 
+                (filter #(not (nil? a))) mvs))
+   ])
+\end{lstlisting}
+
+Instead of explaining the identity monad further, we show a more interesting monad, the maybe monad.
+Its untyped definition is given in listing \ref{lst:maybemdef}.
+
+\begin{lstlisting}[caption=Typed maybe monad definition, label=lst:maybemtyped]
+(ann maybe-m
+    '{:m-bind (All [x y]
+                [(U nil x) [x -> (U nil y)] -> (U nil y)])
+      :m-result (All [x]
+                  [x -> (U nil x)])
+      :m-zero nil
+      :m-plus (All [x]
+                [(U nil x) * -> (U nil x)])})
+(defmonad maybe-m
+  "Monad describing computations with possible failures. Failure is
+  represented by nil, any other value is considered valid. As soon as
+  a step returns nil, the whole computation will yield nil as well."
+  [m-zero   nil
+   m-result (ann-form 
+              (fn m-result-maybe [v] v)
+              (All [x] 
+                [x -> (U nil x)]))
+   m-bind   (ann-form 
+              (fn m-bind-maybe [mv f]
+                (if (nil? mv) nil (f mv)))
+              (All [x y]
+                [(U nil x) [x -> x] -> (U nil x)]))
+   m-plus   (ann-form 
+              (fn m-plus-maybe [& mvs]
+                (first 
+                  (filter
+                    (ann-form 
+                      #(not (nil? %))
+                      [(U nil x) -> boolean :filters {:then (is x 0)}])
+                    mvs)))
+              (All [x]
+                [(U nil x) * -> (U nil x)]))
+   ])
+\end{lstlisting}
+
+A significant difference between typing Haskell style monads in Haskell and typing monads in \emph{algo.monads} with Typed Clojure
+is that monads in Haskell are represented by type classes, while Typed Clojure does not support higher-kinded
+constructs.
+
+\section{Conduit}

### 2012-10-10 dissertation/intro.tex
Monad experiment + other changes

@@
 but recent attempts are noteworthy for their successful use of bidirectional checking.
-Instead of inferring types the algorithm relies on programmer annotations.
+Instead of always attempting to infer types the algorithm relies on programmer annotations
+appearing in some natural places, such as giving the type of each top-level function.
 
-Clojure is a dynamically typed language running on the JVM. It emphasises
-functional programming with immutable data structures
-and provides direct interoperability with JVM languages.
+The Clojure programming language is a dynamically typed language running on the Java Virtual Machine. 
+It emphasises functional programming with immutable data structures
+and provides direct interoperability with languages on the Java Virtual Machine.
+Clojure also is also a family of languages with similar semantics, referred to as Clojure dialects.
+There are several Clojure dialects, notably the aforementioned Clojure for the Java Virtual Machine,
+ClojureCLR which targets the Common Language Runtime, and ClojureScript which compiles to Javascript.
+We refer to Clojure the language as \emph{Clojure}, and the family of Clojure-like languages as
+\emph{Clojure dialects}.
+
+Clojure dialects are widely used for several reasons.
+Performance is a key feature, with most dialects 
+offering access to host-level performance
 
 Recently languages have been created to or been modified to support aspects
-of static and dynamic typing.
+of both static and dynamic typing.
 Dart is fundamentally dynamically typed but offers a form of optional
@@
 % Page 1 Rowan's PhD
+\begin{verbatim}Following the terminology of [Pfenning, Reynolds ..]\end{verbatim}
 An ordinary static type system is used to check whether programs are basically
 meaningful. Such type systems are \emph{intrinsic}. A language with an intrinsic
-static type system relies on programs being well-typed for runtime semantics to be defined.
+static type system has run-time semantics that depends on the types of associated
+variables and expressions during type checking.
+For example, C, Java and ML have intrinsic types.
 This means such programs must pass the type checker before being run.
@@
 A static type system is \emph{extrinsic} when runtime semantics
-are not defined by the type system. In other words, passing a static type checker
+does not depend on the type system. In other words, passing a static type checker
 is not essential to running programs. A dynamically typed language can be viewed
-as having a static type system that supports exact one type.
+as having a trivial static type system that supports exact one type,
+\begin{verbatim} .. a view strongly advocated by Harper[..]
+and common in the literature on static types (see, eg. Pierce[..])
+\end{verbatim}
 
@@
 for most dynamic languages, other techniques are used for checking
-type invariants. Runtime contracts are a popular alternative.
-Clojure takes this approach, providing easy syntax for defining pre-
-and post-conditions \footnote{Inspired by Eiffel}.
+type invariants. Runtime contracts and unit testing are a popular alternatives.
+Clojure adopts these approachs, providing easy syntax for defining pre-
+and post-conditions \footnote{Inspired by Eiffel} and a library for writing unit tests.
+\begin{verbatim} but.. (why a static type system is still desirable\end{verbatim}
 
@@
 \end{lstlisting}
-\footnote{Example adapted from Sam's dissertation TODO}
+\footnote{\emph{Example adapted from Sam's dissertation TODO}}
 
@@
 \footnote{All Classes in the \lstinline|java.lang| package
-are automatically imported in every Clojure namespace (for those familiar with Java, the equivalent of \lstinline|import java.lang.*;|).}.
+are automatically imported in every Clojure namespace (the equivalent of Java's \lstinline|import java.lang.*;|).}.
 
@@
 
-In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}
-, we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
+In this example\footnote{When convenient, namespace declarations are omitted for the remainder of the chapter.}, 
+we define \lstinline|to-set|, aliasing \lstinline|clojure.core/set|.
 \lstinline|All| introduces a set of type variables to the body of a type,

### 2012-10-10 dissertation/lit-review.tex
Monad experiment + other changes

@@
 \chapter{Literature Review}
+
+In this chapter we fit the design of Typed Clojure with literature from related fields.
+Forward-references are given to link the discussed literature with this project
+where there is some relevance or influence.
 
@@
 
+\begin{verbatim}
+  Different levels of Gradual Typing discussed by Siek..
+\end{verbatim}
+
 \section{Interlanguage Interoperability}
@@
 error messages correctly \emph{blame} the source of type errors,
-which can be difficult to determine in the presence of higher-order functions. %TODO reference
+which can be difficult to determine in the presence of higher-order functions. 
+\begin{verbatim}references, Walder etc.\end{verbatim}
 
@@
 
-Forsyth\cite{Rey96} was the first language to use intersection types.
-Uses of intersections in Forsyth include representing extensible record types
+Forsythe, a modern ALGOL dialect by Reynolds\cite{Rey96} was the first wide-spread 
+programming language to use intersection types.
+Uses of intersections in Forsythe include representing extensible record types
 and function overloading.
@@
 % CIDRE
-% TODO
-CIDRE\cite{Dav05} adds \emph{refinement type checking} to SML.
+SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}
+that features a form of intersection types. It is used for function overloading
+and checking invariants involving Refinement types.
+\begin{verbatim}
+Freeman & Pfenning, Refinement types
+\end{verbatim}
 
-% PADL '12 paper, Typing the Numeric Tower
-Typed Racket uses \emph{ordered intersection types} to offer a kind of function overloading.
-
-Java allows intersection types only in type parameter bounds. Instead of supporting intersection
-types in the rest of the language, Java support type inheritance by extending Classes and Interfaces.
+Typed Racket, developed by Tobin-Hochstadt and others \cite{Tob10}, uses \emph{ordered intersection types} 
+to offer a kind of function overloading.
+\begin{verbatim}PADL '12 paper, Typing the Numeric Tower\end{verbatim}
 
 \section{Statically Typed Multimethods}
-
-Designing a static type system for any dynamically typed language requires
-careful consideration. Idiomatic style in dynamically typed languages is usually
-designed to be powerful yet non-intrusive. An effective type system
-for such languages should preserve existing idioms and syntactic abstractions
-until time tested idioms are devised for the new typed language \cite{Tob10}.
-
-Typed Racket \cite{Tob10} integrates a selection of ideas to
-create a typed sister language to Racket.
-Clojure and Racket share a key similarity: they are dynamically typed Lisps.
-Consequently, a number of novel Typed Racket features are relevant to the design and implementation
-of a type system for Clojure.
-The inventions of occurrence typing \cite{TF10}
-and a systematic approach to variable-arity polymorphism
-\cite{STF09}
-are both applicable to this project.
 
@@
 Clojure and the statically typed language Scala \cite{OCD+} 
-have some interesting similarities. The redesigned 
-of the Scala collection hierarchy
-\cite{OS10}
+have some interesting similarities. The redesign
+of the Scala collection hierarchy\cite{OS10}
 bears similarity to Clojure's encompassing ``seq''
@@
 
-\section{Variable-Arity Polymorphism}
+\section{Function Types}
 
-Functions with variable-arity are abundant in dynamically typed languages;
-it is common to have a large standard library of functions that utilize
-variable-arity, and Clojure is no different. 
-Support for complex variable-arity functions is sparse in typed languages,
-Strickland, Tobin-Hochstadt, and Felleisen claim \cite{STF09}
-``no typed language supports them in a systematic and principled manner''.
+There are several different approaches to representing functions in programming languages.
 
-Strickland et al. invented \cite{STF09} 
-a calculus capable of typing non-uniform variable-arity functions, a version of which
-is implemented in Typed Racket. A non-uniform variable-arity function is a function that
-takes a heterogeneous variable parameter list, which usually has a complex relationship 
-with other components of the function.
-For example, Clojure's \lstinline|map| function takes a function and one or more sequences,
+In typed languages like Haskell, functions are as simple as possible, taking a single argument.
+A function with multiple arguments is represented by chaining several single-argument function
+together. This style is characterised by direct syntactic function currying, where applying
+a function to less than its maximum number of arguments results in another function
+that takes the remaining arguments.
+
+In untyped languages like Scheme, functions can take any number of arguments. It is an
+error to supply less than the minimum number of arguments to a function.
+This style features sophisticated support for functions with variable-arity. For example,
+functions can dispatch on the number of arguments provided, and supports a \emph{rest} parameter
+as its last parameter which can accept any number of arguments.
+
+In this regard, Clojure takes an approach identical to Scheme, and supports all the features
+mentioned in the previous paragraph, and none characterised by Haskell-style functions.
+For this reason, we ignore the tradeoffs associated between the two approaches 
+and move directly to literature applicable to typing Scheme-style functions.
+
+\subsection{Variable-Arity Polymorphism}
+
+Strickland et al. invented a type system supporting variable-arity polymorphism\cite{STF09}
+a version of which is included in the current implementation of Typed Racket.
+Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
+of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
+which are used to check definitions and usages of functions with non-trivial rest parameters
+
+For example in Clojure, the function \lstinline|map| takes a function and one or more sequences,
 and returns the result of applying the function argument to each element of the sequences pair-wise.
@@
 ;=> (3.1 5.2)
-\end{lstlisting}
+\end{lstlisting}\footnote{Line comments in Clojure begin with \lstinline|;| and comments to the end of the line. We use \lstinline|;=>| to mean \emph{evaluates to}.}
 
@@
 of \lstinline|Number|, so the expression is well typed.
-
-Clojure presents some opportunities to extend Strickland's calculus. It is idiomatic
-to utilize hash-maps as ad-hoc objects in Clojure, and it would be useful to track the keys associated
-with a hash-map.  The fundamental hash-map manipulation functions \lstinline|assoc| and \lstinline|dissoc|,
-which associate and dissociate entries to (immutable) hash-maps, both take a variable number of arguments.
-
-\begin{lstlisting}[caption=Using \lstinline|dissoc| with variable arguments, label=lst:assoc1]
-(dissoc {:a 1, :b 2} ; => {:a 1, :b 2}
-        :a           ; => {:b 2}
-        :b}          ; => {}
-;=> {}
-\end{lstlisting}
 
@@
 \begin{lstlisting}[caption=Bidirectional checking algorithm with Typed Clojure pseudocode, label=lst:bidir]
-(map (fn [a :- Long, b :- Float]
+(map (fn [[a :- Long] [b :- Float]]
        (+ a b))
@@
 
-The Local Type Inference paper \cite{PT00}
-describes two implementations of type variables, for bounded
+Local type inference by Pierce and Turner\cite{PT00}
+describe two implementations of type variables, for bounded
 and unbounded type variables. The bounded implementation is presented

### 2012-10-10 dissertation/paper.tex
Experiments
Monad experiment + other changes

@@
 
-\title{Typed Clojure: A Practical Optional Type System for Clojure}
+\title{A Practical Optional Type System for Clojure}
 \author{Ambrose Bonnaire-Sergeant}
@@
   % FIXME long tradition for attempts
-  There is a long tradition to create static type systems for dynamically-typed
-  languages. Static type checking promises to eliminate many common user errors
+  Since the first wide-spread programming languages like Fortran, LISP and ALGOL
+  there has been a division between languages that perform type checking at compile-time
+  and those that don't.
+  A recent trend is to combine the advantages of both kinds of languages by adding optional
+  static type systems to languages without static type checking,
+  which promises to eliminate many common user errors
   that are otherwise unnoticed or are caught when the program is run.
 
+\begin{verbatim}
   Recent attempts have seen several promising and innovative
@@
   statically type check idioms ..
+\end{verbatim}
   We take the lessons learnt from these
@@
 % - future work identified
+
+\include{experiments}
 

### 2012-10-14 dissertation/experiments.tex
Add implementation

@@
 
-\begin{lstlisting}[caption=Macro expansion of identity monad definition, label=lst:identitymexpand]
-(def identity-m
-  (let [m-bind :typed.test.monads/undefined
-        m-result :typed.test.monads/undefined
-        m-zero :typed.test.monads/undefined
-        m-plus :typed.test.monads/undefined
-        m-result identity
-        m-bind (fn m-result-id [mv f] (f mv))]
-    {:m-plus m-plus, :m-zero m-zero, :m-bind m-bind, :m-result m-result}))
-\end{lstlisting}
-
 A monad is defined using the macro \lstinline|defmonad|. The macro expands to code that binds a var to a hash-map

### 2012-10-14 dissertation/implementation.tex
Add implementation

@@
 \chapter{Implementation}
+
+This chapter discusses the implementation of the Typed Clojure prototype type 
+system\footnote{We refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter.}.
+concentrating on significant challenges that were identified and overcome.
+Many aspects of Typed Clojure's design follows the implementation of Typed Racket
+
+\section{Type Checking Procedure}
+
+Typed Clojure and Typed Racket differ significantly in how they interact with
+the Clojure and Racket respectively to perform type checking.
+Typed Racket is implemented as a language on the
+Racket platform, which provides highly sophisticated and extensible macro facilities.
+This allows Typed Racket to be entirely implemented with macros.
+Instead, Typed Clojure is implemented as a library that utilises abstract syntax trees (AST)
+generated by analyze, a library developed for this project.
+
+There are several stages to type checking in Typed Clojure.
+Type checking is typically initiated at the read-eval-print-loop prompt (REPL),
+for example \lstinline|(check-ns 'my.ns)| checks the namespace \lstinline|my.ns|.
+Before type checking begins, all global type definitions in the namespace
+are added to the global type environment by compiling the namespace.
+These type definitions include
+
+\begin{itemize}
+  \item type alias definitions
+  \item var, protocol, datatype, and Java Class annotations
+  \item Java method annotations, such as \lstinline|nilable-param| and \lstinline|non-nil-return|
+\end{itemize}
+
+An AST is then generated from the code contained in the target namespace.
+This AST is then recursively descended and is type checked using local type inference.
+Currently only one error is reported at a time, and type checking stops if a type error
+is found.
+
+\section{Occurrence typing}
+
+Typed Clojure's implementation of occurrence typing is ported and extended from Typed Racket.
+Occurrence typing plays several roles in Typed Clojure.
+First, occurrence typing is used to update the type environment at every conditional branch.
+Second, it is used to calculate whether branches are reachable: a branch is not reachable
+if updating the type environment results in a variable of the bottom type.
+Third, paths are used extensively in Typed Clojure. Typed Racket uses \emph{car} and \emph{cdr} paths
+to track which component of a cons cell to update. Typed Clojure adds paths for traversing heterogeneous maps,
+inferring length information, and \lstinline|first| and \lstinline|rest| paths for sequences.
+
+\section{Variable-arity Polymorphism}
+
+Variable-arity polymorphism in Typed Clojure is directly ported from Typed Racket.
+This was the most complicated part of the prototype. At the center of the implementation
+is manipulating dotted type variables, which can represent a sequence of types.
+
+\section{Portability to other Clojure Dialects}
+
+Typed Clojure was built for the Clojure programming language, whose compiler
+and data structures are implementated in Java. Clojurescript is the first major Clojure dialect
+to be written in Clojure, and it is likely future dialects of Clojure will follow this example. 
+Where Clojure uses Java Classes and Interfaces, 
+Clojurescript's compiler and data structures are written in terms of Clojure's
+two core abstractions: protocols and datatypes.
+It would desirable to port Typed Clojure to similar  Clojure dialects while keeping
+the core of the implementation constant.
+
+A significant portion of Typed Clojure is theoretically platform independent but there are
+challenges to targeting new dialects, including incompatible host interoperability and
+non-standardised abstract syntax trees.
+The first issue is predictable due to a core philosophy of Clojure dialects: 
+host interoperability is non-portable\footnote{See the complete Clojure rationale: http://clojure.org/rationale}.
+Each dialect of Clojure has a unique host interopability story and Typed Clojure should cater for them
+separately.
+The second issue is potentially resolvable.
 

### 2012-10-14 dissertation/paper.tex
Add implementation

--- 2012-09-11 dissertation/paper.tex
@@
 \begin{abstract}
-  % FIXME long tradition for attempts
-  There is a long tradition to create static type systems for dynamically-typed
-  languages. Static type checking promises to eliminate many common user errors
-  that are otherwise unnoticed or are caught when the program is run.
 
-  Recent attempts have seen several promising and innovative
-  type systems built for dynamically typed languages. Typed Racket is able to
-  statically type check idioms ..
-  We take the lessons learnt from these
-  projects and apply them to the dynamically-typed language Clojure in the form of Typed Clojure.
+Since the first wide-spread programming languages 
+like Fortran, LISP, and ALGOL
+
+there has been a division between languages 
+that perform type checking at compile-time and those that don't.
+
+A recent trend is to combine the advantages of both kinds of languages by adding optional static type systems to languages without static type checking. 
+This promises to eliminate many common user errors that are otherwise unnoticed or are caught when the program is run, while conserving the flexibility of the language. 
+A major challenge in these combinations is continuing to support the idiomatic style the original language, and some recent attempts appear to have finally produced type systems that have overcome this obstacle.
+Most notably, Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type check most idiomatic Racket code.
+
+Clojure is a recent increasingly popular Lisp dialect with an emphasis on immutability, and interoperability and execution on popular platforms like the Java Virtual Machine (JVM).
+We take the lessons learnt from several projects including Typed Racket and the advanced statically-typed JVM language Scala, and apply them to Clojure in the form of an optional static type system.
+In this seminar I will describe my work on building a prototype static type system for Clojure.
+This prototype is sufficient to allow experiments with existing Clojure code that is sufficiently complicated
+that type checking increases confidence that the code is correct.
+My findings show that it appears to be both practical and useful to design and implement an optional static type system for the
+Clojure programming language.
+
 \end{abstract}
 
-%\tableofcontents
+\tableofcontents
 

### 2012-10-16 dissertation/experiments.tex
Lots of stuff

@@
 
-Monads[] are an interesting control structure probably most recognisable from its inclusion
-in the statically-typed language Haskell[].
+Monads are an interesting control structure most used in functional programming languages,
+as described by Wadler\cite{Wad95}.
+They are most recognisable from its inclusion in the statically-typed language Haskell\cite{Mar10},
+where monads are relied on for many features including global state, file output, and exceptions.
 
-I chose to port a subset of the Clojure library \emph{algo.monads} to Typed Clojure. The library
-provides support for several kinds of monads.
+I chose to port the Clojure Contrib library \emph{algo.monads}\cite{Hic12} to Typed Clojure. The library
+provides several kinds of monads, monad transformers, and monadic functions.
+Macros are used to provide pleasant syntax for consumers of the library.
+
+\subsection{Monad Definitions}
+\label{sec:monaddef}
 
@@
 A monad is defined using the macro \lstinline|defmonad|. The macro expands to code that binds a var to a hash-map
-with the aforementioned keys. For example, listing \ref{lst:identitymdef} expands to 
-the code in listing \ref{lst:identitymexpand}.
+with the aforementioned keys.
 
 \begin{lstlisting}[caption=Type for identity monad, label=lst:identitymtype]
+(def-alias Undefined '::undefined)
 (ann identity-m
@@
                    [x -> x])
-       :m-zero '::undefined
-       :m-plus '::undefined})
+       :m-zero Undefined
+       :m-plus Undefined})
 \end{lstlisting}
 
-Typed Clojure operates directly on fully macro-expanded forms, so it is unsurprising that the type
-annotation for \lstinline|identity-m| reflects its macro-expansion (listing \ref{lst:identitymtype}).
-The \lstinline|:m-bind| entry is comparable to Haskell's monadic \emph{bind},
-and the \lstinline|:m-result| entry is comparable to Haskell's monadic \emph{return}.
+A direct typing of the identity monad would look like listing \ref{lst:identitymtype}.
+This signature assigns the expected type for the var \lstinline|identity-m| 
+to be a heterogeneous map type with the minimum entries for monads: \lstinline|:m-bind|
+and \lstinline|:m-result|. The syntax for heterogeneous map types requires a \lstinline|'| prefix,
+and any number of key-value pairs are given between curly braces. Heterogeneous map keys must be keywords,
+so the syntax is made more convenient by omitting the usually required \lstinline|'| prefix
+for keyword types.
 
-\begin{lstlisting}[caption=Maybe monad definition, label=lst:maybemdef]
-(defmonad maybe-m
-  "Monad describing computations with possible failures. Failure is
-  represented by nil, any other value is considered valid. As soon as
-  a step returns nil, the whole computation will yield nil as well."
-  [m-zero   nil
-   m-result (fn m-result-maybe [v] 
-              v)
-   m-bind   (fn m-bind-maybe [mv f]
-              (if (nil? mv) nil (f mv)))
-   m-plus   (fn m-plus-maybe [& mvs]
-              (first 
-                (filter #(not (nil? a))) mvs))
+Monad types in languages with advanced type systems are often abstracted using type constructors.
+This allows reasoning about monadic code while keeping the particular monad abstract, which is a desirable result,
+so Typed Clojure was extended to support user definable type constructors.
+
+\begin{lstlisting}[caption=An abstract definition of a monad., label=lst:monadalias]
+(def-alias Monad 
+  (TFn [[m :kind (TFn [[x :variance :covariant]] Any)]]
+    '{:m-bind (All [x y]
+                [(m x) [x -> (m y)] -> (m y)])
+      :m-result (All [x]
+                  [x -> (m x)])
+      :m-zero Undefined
+      :m-plus Undefined}))
+\end{lstlisting}
+
+Listing \ref{lst:monadalias} captures the abstract definition of a monad.
+A \lstinline|Monad| is a type constructor parameterised by \lstinline|m|, which is
+a type constructor taking a single argument \lstinline|x| (a type) and returning a type 
+(written \lstinline|Any|)\footnote{Haskell-like syntax is helpful here, the kind \lstinline|(TFn [[x :variance :covariant]] Any)|
+is approximately \lstinline|* -> *|. A more streamlined syntax is planned for future work.}.
+The body of the type constructor uses \lstinline|m| abstractly. Typed Clojure
+ensures the correct number of arguments are passed and recognises the declared variances
+for each parameter, In this case the first argument of \lstinline|m| is declared
+a covariant position. When instantiated, \lstinline|m| must also be a type operator
+of one covariant argument returning a type.
+
+\begin{lstlisting}[caption=Identity monad using user defined type constructors, label=lst:identityctors]
+(ann identity-m (Monad (TFn [[x :variance :covariant]] x)))
+\end{lstlisting}
+
+We can now express the type of the identity monad more abstractly (listing \ref{lst:identityctors}).
+The purpose of the monad seems more apparent just from reading its type.
+In this case, the fact that identity monad has no effect is reflected
+by its type constructor returning exactly its argument.
+
+\begin{lstlisting}[caption=Type checked identity monad definition, label=lst:identitydeftyped]
+(defmonad identity-m
+   "Monad describing plain computations. This monad does in fact nothing
+    at all. It is useful for testing, for combination with monad
+    transformers, and for code that is parameterized with a monad."
+  [m-result identity
+   m-bind   (ann-form
+              (fn m-result-id [mv f]
+                (f mv))
+              (All [x y]
+                [x [x -> y] -> y]))
   ])
@@
 
-Instead of explaining the identity monad further, we show a more interesting monad, the maybe monad.
-Its untyped definition is given in listing \ref{lst:maybemdef}.
+The final type checked definition for the identity monad is given in listing \ref{lst:identitydeftyped}.
+In this case, just the monadic bind required annotation.
 
-\begin{lstlisting}[caption=Typed maybe monad definition, label=lst:maybemtyped]
-(ann maybe-m
-    '{:m-bind (All [x y]
-                [(U nil x) [x -> (U nil y)] -> (U nil y)])
-      :m-result (All [x]
-                  [x -> (U nil x)])
-      :m-zero nil
-      :m-plus (All [x]
-                [(U nil x) * -> (U nil x)])})
-(defmonad maybe-m
-  "Monad describing computations with possible failures. Failure is
-  represented by nil, any other value is considered valid. As soon as
-  a step returns nil, the whole computation will yield nil as well."
-  [m-zero   nil
-   m-result (ann-form 
-              (fn m-result-maybe [v] v)
-              (All [x] 
-                [x -> (U nil x)]))
-   m-bind   (ann-form 
-              (fn m-bind-maybe [mv f]
-                (if (nil? mv) nil (f mv)))
-              (All [x y]
-                [(U nil x) [x -> x] -> (U nil x)]))
-   m-plus   (ann-form 
-              (fn m-plus-maybe [& mvs]
-                (first 
-                  (filter
-                    (ann-form 
-                      #(not (nil? %))
-                      [(U nil x) -> boolean :filters {:then (is x 0)}])
-                    mvs)))
-              (All [x]
-                [(U nil x) * -> (U nil x)]))
-   ])
+\begin{lstlisting}[caption=Several monad types, label=lst:severalmonads]
+; Maybe monad
+(ann maybe-m (MonadPlusZero
+               (TFn [[x :variance :covariant]] 
+                 (U nil x))))
+
+; Sequence monad (called "list monad" in Haskell)
+(ann sequence-m (MonadPlusZero 
+                  (TFn [[x :variance :covariant]]
+                    (Seqable x))))
+
+; State monad
+(def-alias State
+  (TFn [[r :variance :covariant]
+        [s :variance :invariant]]
+    [s -> '[r s]]))
+(ann state-m (All [s]
+               (Monad (TFn [[x :variance :covariant]]
+                        (State x s)))))
 \end{lstlisting}
 
-A significant difference between typing Haskell style monads in Haskell and typing monads in \emph{algo.monads} with Typed Clojure
-is that monads in Haskell are represented by type classes, while Typed Clojure does not support higher-kinded
-constructs.
+Listing \ref{lst:severalmonads} shows several monad types. The alias \lstinline|MonadPlusZero|
+is identical to \lstinline|Monad|, except both \lstinline|:m-zero| and \lstinline|:m-plus| are defined.
+The definitions were type checked by adding type annotations in appropriate places, in a similar fashion
+to the identity monad. A significant obstacle was discovered in while type checking monad definitions,
+relating to filtering sequences with negative type information, discussed in section \ref{sec:filterneg}.
+
+\subsection{Monad Transformer Definitions}
+
+The initial motivation for adding user defined type constructors to Typed Clojure
+(as discussed in section \ref{sec:monaddef}) was to type check
+monad transformer definitions.
+Monad transformers are always parameterised by a monad type constructor, and should work for all
+monad type constructors. Keeping the monad type constructor abstract allows us to determine
+whether a monad transformer works \emph{for any} monads.
 

### 2012-10-16 dissertation/intro.tex
Add implementation
Lots of stuff

@@
 Clojure also is also a family of languages with similar semantics, referred to as Clojure dialects.
-There are several Clojure dialects, notably the aforementioned Clojure for the Java Virtual Machine,
+There are several Clojure dialects, notably the aforementioned Clojure for the Java Virtual Machine (JVM),
 ClojureCLR which targets the Common Language Runtime, and ClojureScript which compiles to Javascript.
-We refer to Clojure the language as \emph{Clojure}, and the family of Clojure-like languages as
-\emph{Clojure dialects}.
+Where it is clear to do so, we refer to Clojure the programming language as \emph{Clojure}, 
+and the family of Clojure-like languages as \emph{Clojure dialects}.
 
 Clojure dialects are widely used for several reasons.
-Performance is a key feature, with most dialects 
-offering access to host-level performance
+Performance is a key feature, for example the JVM implementation of Clojure
+offers ways to access Java-like speed for certain operations.
 
@@
 type invariants. Runtime contracts and unit testing are a popular alternatives.
-Clojure adopts these approachs, providing easy syntax for defining pre-
+Clojure adopts these approaches, providing easy syntax for defining pre-
 and post-conditions \footnote{Inspired by Eiffel} and a library for writing unit tests.
@@
 
+% Clojure community reactions
+% GSOC
+% Conj talk
+% 
+% Motivation: Conduit
+
 \section{Typed Clojure through Examples}
@@
 This section introduces Typed Clojure with example code. An attempt is
-made to introduce some Clojure syntax and sematics to those unfamiliar or needing a refresher.
-A basic knowledge of Lisp syntax is assumed.
+made to introduce some Clojure syntax and semantics to those unfamiliar or needing a refresher.
+A basic knowledge of Lisp syntax is handy, but a brief tutorial is given
+for newcomers.
+
+\subsection{Preliminary: Lisp Syntax}
+
+The core of understanding Lisp syntax can be summarised by these points
+
+\begin{itemize}
+  \item Invocations are always wrapped in a pair of balanced parenthesis 
+  \item Parenthesis start to the left of the operator
+\end{itemize}
+
+For example, the Java expressions \emph{1 + 2 / 3} is written in Lisp pseudocode \lstinline|(/ (+ 1 2) 3)|
+and \emph{numberCrunch(1, 2)} written \lstinline|(numberCrunch 1 2)|.
+
+\subsection{Simple Examples}
 
@@
 
-The \lstinline|clojure.core/ns|
+Here \lstinline|clojure.core/ns|
 \footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
-invocation declares the current namespace to be \lstinline|typed.test.hello-world|
-and requires the namespace \lstinline|typed.core| to be loaded as a dependency, referring the var
-\lstinline|typed.core/check-ns| to be referenced locally. Other than this
+declares the current namespace to be 
+\lstinline|typed.test.hello-world|
+and requires the namespace \lstinline|typed.core| to be loaded as a dependency,
+referring \lstinline|typed.core/check-ns| to be referenced as \lstinline|check-ns|. Other than this
 dependency, this is identical to the untyped \emph{Hello world}.
-% footnote: check-ns used to initiate type checking usually at REPL
+By running \lstinline|check-ns| we can verify that this namespace type checks.
 
@@
 \end{lstlisting}
-\footnote{\emph{Example adapted from Sam's dissertation TODO}}
+\footnote{Example adapted from Tobin-Hochstadt \cite{Tob10}}
 
@@
 to be a function from \lstinline{java.lang.Number} to 
-\lstinline{java.lang.Number}
-\footnote{All Classes in the \lstinline|java.lang| package
+\lstinline{java.lang.Number}\footnote{All Classes in the \lstinline|java.lang| package
 are automatically imported in every Clojure namespace (the equivalent of Java's \lstinline|import java.lang.*;|).}.

### 2012-10-16 dissertation/lit-review.tex
Lots of stuff

@@
 It extended earlier work by Cartwright and Fagan\cite{CF91}
-and Fagan \cite{Fag92}, adding support for practical features such as
+and Fagan \cite{Fag91}, adding support for practical features such as
 first-class continuations and variable-arity functions.
@@
 
-% what challenges did soft typing face?
+Type systems for Scheme have since moved away from soft typing
+to to other approaches like gradual typing.
+Alas, the best reference for commentary on this transition 
+appears to be slides from a talk by Felleisen\cite{Fell09},
+a leader in this area of research for over 20 years.
+Felleisen says that while Soft Scheme discovered type problems, 
+it suffers from incomprehensible
+errors that required PhD-level expertise to decipher.
+
+\begin{verbatim}
+  Bidirectionality
+\end{verbatim}
 
@@
 Gradual typing combines static and runtime type checking so programmers
-can the most appropriate one for the situtation.
+can the most appropriate one for the situation.
 
@@
 Clojure provides two dispatching mechanisms: multimethods, a mechanism to dispatch on the
-results of an arbitrary function; and protocols, a mechanism for single dispatch.
+results of an arbitrary function; and protocols, a mechanism for single dispatch, a common
+strategy that dispatches on the type of the first provided argument.
 Millstien and Chambers explore statically typing multiple dispatched multimethods in Dubious \cite{MS02}.
@@
 
-Clojure provides mulimethods as a core language feature. This section discusses systems that statically
+Clojure provides multimethods as a core language feature. This section discusses systems that statically
 verify type safety for multimethods.
@@
 
+\section{Higher Kinded Programming}
+
+\begin{verbatim}
+Many advanced type systems provide support for ..
+Forward reference to algo.monad experiment.
+\end{verbatim}
+
 \section{Conclusion}

### 2012-10-16 dissertation/paper.tex
Lots of stuff

--- 2012-09-11 dissertation/paper.tex
@@
 \usepackage{clj-grammar}
-\usepackage[style=numeric]{biblatex}
+\usepackage[style=alphabetic]{biblatex}
 \addbibresource{bibliography.bib}
@@
 \begin{abstract}
-  % FIXME long tradition for attempts
-  There is a long tradition to create static type systems for dynamically-typed
-  languages. Static type checking promises to eliminate many common user errors
-  that are otherwise unnoticed or are caught when the program is run.
 
-  Recent attempts have seen several promising and innovative
-  type systems built for dynamically typed languages. Typed Racket is able to
-  statically type check idioms ..
-  We take the lessons learnt from these
-  projects and apply them to the dynamically-typed language Clojure in the form of Typed Clojure.
+Traditionally there has been a clear distinction between 
+programming languages that perform type checking at compile-time, and those that do not.
+In particular, the early wide-spread programming languages ALGOL and LISP
+could be divided this way: ALGOL performs compile-time (or static) type checking,
+and LISP does not.
+Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
+when the program is run.
+On the other hand, a language without static type checking can often be designed 
+to be more flexible and forgiving.
+
+A recent trend is to combine the advantages of both kinds of languages by adding optional static 
+type systems to languages without static type checking. 
+A major challenge in these combinations is continuing to support the idiomatic style of the original 
+language, and some recent attempts appear to have finally produced type systems that have overcome this obstacle.  
+Most notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
+check most idiomatic Racket code.
+
+Clojure is a recent and increasingly popular programming language that is designed to support 
+functional programming and immutability by default. Implementations of Clojure are hosted 
+on popular platforms and provide direct interoperability with the host platform.
+There are several implementations of Clojure, each targeting different runtimes including
+the Java Virtual Machine, the Common Language Runtime, and Javascript virtual machines.
+
+This paper describes my work on designing an optional static type system for Clojure based on
+the lessons learnt from several projects, primarily Typed Racket.
+A prototype type system is developed for Clojure running on the Java Virtual Machine.
+This prototype is sufficient to allow experiments with existing Clojure code that is sufficiently complicated
+that type checking increases confidence that the code is correct.
+For example, a significant portion of algo.monads, a Clojure Contrib library for monadic programming, 
+is able to be type checked. 
+On the positive side, most monad, monad transformer, and monadic function definitions can be type checked,
+usually by adding type annotations in natural places like function definitions.
+Negatively, simple monad comprehensions can be type checked, but require awkward type annotations.
+
+There is significant future work to fully type check all Clojure features and idioms.
+For example, multi-method definitions and functions accepting an even number of variable arguments
+are troublesome. 
+Also, there are desirable features from the Typed Racket project that are missing, such
+as automatic runtime contract generation and a sophisticated blame system, 
+both which significantly improve error messages when mixing typed and untyped code.
+I conclude that it appears to be both practical and useful to design and implement an optional static type system for the
+Clojure programming language.
+
 \end{abstract}
 
-%\tableofcontents
+\tableofcontents
 
@@
 
+\chapter{Conclusion}
+
+\begin{verbatim}
+  Programming is an error-prone task, and verification tools are important.
+  Dynamic languages rely on several techniques for verification, but static
+  type checking is not often supported.
+\end{verbatim}
+
 \printbibliography[title=References]

### 2012-10-17 dissertation/design-choices.tex
Lots of stuff
Implementation work

@@
 Typed Clojure is designed to be of practical use to Clojure programmers.
-Many of the design goals are similar to Tobin-Hochstadt's \cite{Tob10}
-for Typed Racket.
+Many of the design goals are similar to Tobin-Hochstadt's 
+for Typed Racket \cite{Tob10}. Some of the main differences are described in this chapter.
 
@@
 
-In Java, \lstinline|null| is a subtype to all reference
+In Java, \lstinline|null| is included in all reference
 types, but \lstinline|null| is not expressible as a static type. In other words,
@@
 Scala provides the Option type for safe \lstinline|null| elimination, forcing pattern matching
-where \lstinline|null| might occur.
+where \lstinline|null| might occur, and
 
@@
 % Some heterogeneous maps have string args, eg. from web frameworks
+
+As a key idiomatic feature of Clojure, we treat heterogeneous maps more formally.
+This also serves as a prototype for formalising other aspects.
+Here we give a formal syntax, operational semantics and type system using inference
+rules, as is standard in the programming language community (see, eg., Pierce\cite{Pie02}).
 

### 2012-10-17 dissertation/implementation.tex
Lots of stuff
Implementation work

@@
 This chapter discusses the implementation of the Typed Clojure prototype type 
-system\footnote{We refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter.}.
-concentrating on significant challenges that were identified and overcome.
-Many aspects of Typed Clojure's design follows the implementation of Typed Racket
+system\footnote{We refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter.}
+in some detail concentrating on significant challenges that were identified and overcome.
+Many aspects of Typed Clojure's design follows the implementation of Typed Racket,
+which is reflected in this chapter.
 
@@
 
-Typed Clojure and Typed Racket differ significantly in how they interact with
-the Clojure and Racket respectively to perform type checking.
+Typed Clojure and Typed Racket differ significantly in how they perform type checking.
 Typed Racket is implemented as a language on the
 Racket platform, which provides highly sophisticated and extensible macro facilities.
-This allows Typed Racket to be entirely implemented with macros.
+Interestingly, this allows Typed Racket to be entirely implemented with macros.
 Instead, Typed Clojure is implemented as a library that utilises abstract syntax trees (AST)
-generated by analyze, a library developed for this project.
+generated by analyze\footnote{https://github.com/frenchy64/analyze}, a library I developed for this project.
+This strategy follows common practice for Clojure projects, which favours providing modular libraries 
+over modifying the language.
+
+This section goes into details on the implementation of Typed Clojure.
+First, a high-level overview is given on the type checking procedure.
+Then the interfaces to particular high-level functions are discussed.
+
+\subsection{General Overview}
 
@@
 
-\section{Occurrence typing}
+\subsection{Bidirectional Checking}
+
+The interface to the bidirectional checking algorithm is \lstinline|typed.core/check|,
+which takes an expression, represented as an AST generated from \emph{analyze}, and an optional expected type for
+the given expression. If the expected type is present, the bidirectional algorithm \emph{checks}
+that the expected type matches the actual type of the expression.
+If the expected type is omitted, a type is instead \emph{synthesised} for the expression.
+The algorithm is based on Pierce and Turner's  Local Type Inference\cite{PT00}
+and the implementation is similar in shape to Typed Racket's\cite{TF08}, in that
+one function with an extra ``expected type'' argument is preferred over two complementary
+functions, one for checking and one for synthesis.
+
+The main difference between Typed Clojure's and Typed Racket's bidirectional checking
+algorithm is the representation for expressions. Typed Racket relies on pre-existing
+Racket features like syntax objects for expression representation. Clojure instead
+leans towards abstract syntax tree representation, despite its Lisp heritages.
+In terms of the bidirectional check, the difference is mostly cosmetic.
+
+% go into more detail
+% - update env
+% - what is an environment? Bindings + propositions
+% - LOC
+% - link to github
+% - how occurrence typing works
+%   - update env
+%   - how to calculate reachability
+
+\subsection{Occurrence typing}
 
@@
 First, occurrence typing is used to update the type environment at every conditional branch.
-Second, it is used to calculate whether branches are reachable: a branch is not reachable
-if updating the type environment results in a variable of the bottom type.
-Third, paths are used extensively in Typed Clojure. Typed Racket uses \emph{car} and \emph{cdr} paths
-to track which component of a cons cell to update. Typed Clojure adds paths for traversing heterogeneous maps,
-inferring length information, and \lstinline|first| and \lstinline|rest| paths for sequences.
+Second, it is used to calculate whether branches are reachable.
+Third, paths are used extensively in Typed Clojure. 
+The implementation of these features are discussed in this section.
+
+The basic idea of occurrence typing involves keeping a separate environment of \emph{propositions}
+that relate bindings to types. These propositions are then used to update the type environment.
+In Typed Clojure there are several types of propositions, referred to as \emph{filters}.
+They are based on the theory by Tobin-Hochstadt and Felleisen\cite{TF10}, and are ported directly
+from Typed Racket.
+
+\begin{itemize}
+  \item TopFilter and BotFilter represent the trivially true and trivially false propositions respectively.
+  \item TypeFilter and NotTypeFilter represent a positive or negative association of a binding name
+        to a type. The syntax \lstinline|(is t name)|
+  \item AndFilter and OrFilter represent logical conjunction and disjunction of propositions.
+\end{itemize}
+
+Propositions can optionally carry \emph{path} information
+represented by a sequence of \emph{path elements}, which signify which part
+of the binding's type to update. For example, Typed Racket uses \emph{car} and \emph{cdr} path elements
+to track which component of a cons type to update.
+Paths are particularly useful in Typed Clojure. There are path elements for traversing heterogeneous map types (KeyPE),
+inferring length information (CountPE), and \lstinline|first| and \lstinline|rest| paths for sequences.
+These additions do not appear to introduce any significant new complexities related to paths.
+
+\section{Polymorphic Type Inference}
+
+The polymorphic type inference algorithm is directly ported from Typed Racket with slight extensions
+for bounded variables,
+and is reminiscent to Pierce and Turner's Local Type Inference\cite{PT00}.
+A common entry point for inferring type variables to polymorphic function invocations
+\lstinline|typed.core/infer|.
+
+\lstinline|infer| is invoked like \lstinline|(infer X Y S T R expected)|, where
+
+\begin{itemize}
+  \item \lstinline|X| is a map from type variable names
+        to their bounds (representing the type variables in scope),
+  \item \lstinline|Y| is a map from type variable names
+        to their bounds (representing the \emph{dotted} type variables in scope),
+  \item \lstinline|S| and \lstinline|T| are sequences of types of equal length
+  \item \lstinline|R| is a type,
+  \item \lstinline|expected| is the expected type for R, or the value \lstinline|nil|,
+\end{itemize}
+
+and returns a \emph{substitution} that satisfies the following conditions:
+
+\begin{itemize}
+  \item Pairwise, each \lstinline|S| is a below \lstinline|T|,
+  \item \lstinline|R| is below \lstinline|expected|, if \lstinline|expected| is provided.
+\end{itemize}
+
+A substitution maps type variables to an upper and lower bound. 
+It is valid to replace all occurrences of the type variables named in the substitution 
+with a type between their associated bounds.
+For example, usually the substitution is applied to \lstinline|R| by the caller of \lstinline|infer|.
+Whether to substitute the upper or lower bound is determined by the variance
+of the type variable in relation to the type being substituted. 
+The specific algorithm is described in Local Type Inference.
+
+\section{F-Bounded Polymorphism}
+
+A feature not present in Typed Racket is bounded polymorphism.
+Several changes were needed to support bounded polymorphism. In every position
+where a set of type variables was required, it was replaced by a map
+of type variables to bounds.
+
+Bounds consist of an upper and lower type bound, or a \emph{kind bound}.
+Kind bounds are experimental following the inclusion of user definable type constructors
+(motivated in section \ref{sec:monads}).
+They are a stub for a more comprehensive treatment of higher-kinded operators
+such as that described by Moors, Piessens, and Odersky for Scala\cite{MPO08}.
+At present, a type variable can only be instantiated to a type between
+its upper and lower bounds, or, if a kind bound is defined instead, 
+to a kind below the kind bound.
+
+F-bounded polymorphism allows type variables to refer to themselves in their bounds.
 
@@
 is manipulating dotted type variables, which can represent a sequence of types.
+
+It also required changes to the polymorphic type inference algorithm.
+The constraint-generation portion
 
@@
 host interoperability is non-portable\footnote{See the complete Clojure rationale: http://clojure.org/rationale}.
-Each dialect of Clojure has a unique host interopability story and Typed Clojure should cater for them
+Each dialect of Clojure has a unique host interoperability story and Typed Clojure should cater for them
 separately.
-The second issue is potentially resolvable.
+The second issue is potentially resolvable either by enforcing a standard representation for 
+abstract syntax trees across Clojure implementations, or developing a library that provided
+a common interface to abstract syntax trees for each Clojure implementation.
 
@@
 %      false))
+
+\section{Filtering on Negative Propositions}
+\label{sec:filterneg}

### 2012-10-17 dissertation/intro.tex
Implementation work

@@
 In the last decade it has become increasingly common to enhance
-dynamically typed languages with static type systems. This idea not new,
-but recent attempts are noteworthy for their successful use of bidirectional checking.
+dynamically typed languages with static type systems. 
+This idea not new, but recent attempts are noteworthy for their successful use of bidirectional checking\footnote{These 
+slides by Felleisen, a practitioner in this area for over 20 years, present a history of optional type checking
+for Lisp/Scheme that spans around 40 years\cite{Fell09}}.
 Instead of always attempting to infer types the algorithm relies on programmer annotations
@@
 
-The Clojure programming language is a dynamically typed language running on the Java Virtual Machine. 
+The Clojure programming language is a dynamically typed dialect of Lisp invented
+by Hickey, designed to run on popular platforms\footnote{http://clojure.org/}.
 It emphasises functional programming with immutable data structures
-and provides direct interoperability with languages on the Java Virtual Machine.
-Clojure also is also a family of languages with similar semantics, referred to as Clojure dialects.
-There are several Clojure dialects, notably the aforementioned Clojure for the Java Virtual Machine (JVM),
-ClojureCLR which targets the Common Language Runtime, and ClojureScript which compiles to Javascript.
-Where it is clear to do so, we refer to Clojure the programming language as \emph{Clojure}, 
-and the family of Clojure-like languages as \emph{Clojure dialects}.
+and provides direct interoperability with its host platform.
+Notable implementations of Clojure exist for the Java Virtual Machine (JVM),
+the Common Language Runtime, and for Javascript virtual machines.
+At the current time, Clojure on the JVM is the most mature implementation,
+and therefore this project focuses on the JVM implementation.
 
-Clojure dialects are widely used for several reasons.
+Clojure has attracted wide-spread users in part by concentrating on pragmatism.
 Performance is a key feature, for example the JVM implementation of Clojure
 offers ways to access Java-like speed for certain operations.
+Also, Clojure's excellent host interoperability offers Clojure programmers
+access to existing libraries for their platform, such as the extensive Java library ecosystem.
+By coupling pragmatic necessities with elegant features like Lisp, functional programming,
+and immutability by default, Clojure is a compelling general purpose programming language.
 
@@
 of both static and dynamic typing.
-Dart is fundamentally dynamically typed but offers a form of optional
-static typing. Haskell recently supports the ability to delay static type errors 
-to runtime, removing the need for the program to type check before running.
-Scala can emulate some features of dynamic languages with Dynamic Scala.
+Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers optional type annotations
+that do not affect runtime semantics. 
+Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
+a well-known dynamic language.
+
+When a static type checker is not available, which describes the situation
+for most dynamic languages, other techniques are used for checking
+type invariants. For example, ``design by contract'' is often used,
+introduced by Meyer for the Eiffel language\cite{Mey92},
+in which the programmer defines contracts that are enforced at runtime.
+Unit testing is also a popular verification technique in dynamic languages.
+Clojure adopts these approaches, providing easy syntax for defining 
+Eiffel-inspired pre- and post-conditions and a library for writing unit tests.
+
+Static type systems, however, are still desirable for certain situations.
+Powerful type systems like ML's\cite{Mil97} and Haskell's\cite{Mar10} have proved particularly useful
+for writing in complicated programming styles. For example,
+Haskell's advanced static type system helps the programmer write correct monadic code,
+especially in more complicated situations like combining monads via monad transformers.
+
+\subsection{Why implement an optional type system for Clojure?}
+
+The initial motivation for implementing an optional type system
+for Clojure occurred during an anecdotal account of the struggles
+of a Clojure programmer. In a heroic effort, he managed to 
+implement a library for conduits (or \emph{arrows}) in Clojure, 
+which probably surpass monads in complexity, and is usually reserved
+for languages with advanced type systems like Haskell.
+
+He highlighted a strong desire for a type system for several reasons.
+Firstly, to verify the correctness
+of the library (before noting that unit tests and runtime contracts
+were insufficient for this task).
+Secondly, to aid him while writing the library.
+
+Since then, the Clojure community has shown interest in this work.
+I developed this project as a Google Summer of Code 2012 project,
+after it was selected by the Clojure community.
+I have also been invited to speak on this project at the Clojure Conj
+2012 conference, a major annual Clojure conference.
 
 \subsection{What kind of type system does Typed Clojure provide?}
-
-% What are types? No real agreement .. but here is my definition..
-% Every programmer thinks they know what they are. More than one concept
-% in the literature.
-% Static vs. dynamic
-% intrinsic vs. extrinsic
-
-% is it even a type system?
-% type refinements vs ordinary types
-% extrinsic vs intrinsic type systems
 
@@
 % Page 1 Rowan's PhD
-\begin{verbatim}Following the terminology of [Pfenning, Reynolds ..]\end{verbatim}
+
+There are many concepts associated with \emph{types} and \emph{type systems} in both the
+literature and informal discourse.
+A programmer who uses dynamically-typed languages may have a drastically different notion
+of what a type is than, say, a programmer preferring languages with advanced static type systems.
+There is some debate as to whether optional static type systems like Typed Clojure
+can even be called a type system. We choose to
+follow the terminology of Pfenning\cite{Pfe08} and Reynolds\cite{Rey01},
+where such optional type systems are \emph{extrinsic} type systems, and more
+traditional type systems are \emph{intrinsic} type systems.
+
 An ordinary static type system is used to check whether programs are basically
-meaningful. Such type systems are \emph{intrinsic}. A language with an intrinsic
+meaningful. Pfenning and Reynolds call these type systems \emph{intrinsic}. A language with an intrinsic
 static type system has run-time semantics that depends on the types of associated
 variables and expressions during type checking.
-For example, C, Java and ML have intrinsic types.
-This means such programs must pass the type checker before being run.
+For example, C, Java, and ML have intrinsic types.
+This means programs written in these languages must pass the type checker before being run.
 
@@
 as having a trivial static type system that supports exact one type,
-\begin{verbatim} .. a view strongly advocated by Harper[..]
-and common in the literature on static types (see, eg. Pierce[..])
-\end{verbatim}
-
-A type checker ensures invariants hold complains if code
-does not conform to its documented type. Powerful type systems like
-Haskell's and Scala's are useful when reasoning about higher-order code.
-
-When a static type checker is not available, which describes the situation
-for most dynamic languages, other techniques are used for checking
-type invariants. Runtime contracts and unit testing are a popular alternatives.
-Clojure adopts these approaches, providing easy syntax for defining pre-
-and post-conditions \footnote{Inspired by Eiffel} and a library for writing unit tests.
-\begin{verbatim} but.. (why a static type system is still desirable\end{verbatim}
-
-% clojure is dynamically typed
-% currently few tools to statically reason about complex higher-order code
-% conduit based on streams, uses continuations
-% core.logic based on monads
-% Static types are useful for reasoning for such things
-% Strong statically typed languages like Haskell use monads regularly
-
-% strength of Clojure is concision
-% typically, concise static languages are built concurrently with their type system
-% Eg. Haskell, ML flavoured languages
-% Counterpoint: Typed Racket preserves racket idioms, adds static typing
-
-% Clojure community reactions
-% GSOC
-% Conj talk
-% 
-% Motivation: Conduit
+a view strongly advocated by Harper\cite{Har12}
+and common in the literature on static types (for example, Pierce\cite{Pie02}).
 
@@
 % Hello world
-This section introduces Typed Clojure with example code. An attempt is
+This section introduces Typed Clojure with example code. 
+Typed Clojure is developed for the JVM implementation of Clojure, therefore
+the rest of this chapter uses that implementation.
+An attempt is
 made to introduce some Clojure syntax and semantics to those unfamiliar or needing a refresher.
@@
 
-The core of understanding Lisp syntax can be summarised by these points
+The core of understanding Lisp syntax when coming from a popular language
+like Java or Javascript can be summarised by these points.
 
 \begin{itemize}
-  \item Invocations are always wrapped in a pair of balanced parenthesis 
-  \item Parenthesis start to the left of the operator
+  \item Operators are always in prefix position.
+  \item Invocations are always wrapped in a pair of balanced parenthesis.
+  \item Parenthesis start to the left of the operator.
 \end{itemize}
@@
 
+Clojure also adds other syntax:
+
+\begin{itemize}
+  \item Prefixing \lstinline|:| to a symbol defines a \emph{keyword}, often used for map keys. eg. \lstinline|:my-keyword|.
+  \item Square brackets delimit vector literals. eg. \lstinline|[1 2]| is a 2 place vector.
+  \item Curly brackets define map literals. eg. \lstinline|{:a 1 :b 2}| is a map from 
+        \lstinline|:a| to \lstinline|1| and \lstinline|:b| to \lstinline|2|.
+  \item Commas are always optional, and treated as whitespace.
+\end{itemize}
+
 \subsection{Simple Examples}
@@
 
-\begin{lstlisting}[caption=Hello world]
+\begin{lstlisting}[caption=Typed Hello world, label=lst:helloworld]
 (ns typed.test.hello-world
@@
 
-Here \lstinline|clojure.core/ns|
-\footnote{All vars from the \lstinline|clojure.core| namespace are referred implicitly in new namespaces (equivalent to \lstinline|(:require [clojure.core :refer :all])|).}
-declares the current namespace to be 
-\lstinline|typed.test.hello-world|
-and requires the namespace \lstinline|typed.core| to be loaded as a dependency,
-referring \lstinline|typed.core/check-ns| to be referenced as \lstinline|check-ns|. Other than this
-dependency, this is identical to the untyped \emph{Hello world}.
-By running \lstinline|check-ns| we can verify that this namespace type checks.
+At this point, it is worth understanding Clojure's namespacing feature.
+Clojure code is always executed in a \emph{namespace}, and each file of Clojure code should 
+have a \lstinline|ns| declaration with the namespace name and its dependencies,
+which switches the current namespace and executes the given dependency commands.
+There is one special namespace, \lstinline|clojure.core| which is
+loaded with every namespace, implicitly ``referring'' all its vars in the namespace.
+For example, \lstinline|ns| refers to the var \lstinline|clojure.core/ns|,
+similarly \lstinline|println| refers to \lstinline|clojure.core/println|.
+
+The example in listing \ref{lst:helloworld} declares a dependency to 
+\lstinline|typed.core|, Typed Clojure's main namespace. It also refers the var \lstinline|typed.core/check-ns|
+into scope\footnote{\lstinline|check-ns| is the top level function for type checking a namespace}.
+Other than this dependency, this is identical to the untyped \emph{Hello world}.
 
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Annotating vars in Typed Clojure]
 (ns typed.test.collatz
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Annotating protocols and datatypes in Typed Clojure]
 (ns typed.test.deftype
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Polymorphism in Typed Clojure]
 ...
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Heterogeneous map types in Typed Clojure]
 ...
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Annotating multi-arity functions in Typed Clojure]
 (ns typed.test.poly
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure]
 (ann num-vec2 
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Java interoperability with Typed Clojure]
 (ns typed.test.interop

### 2012-10-17 dissertation/lit-review.tex
Implementation work

@@
 
-Typed Clojure is related to several existing systems for both typed and untyped languages.
-The first section of this chapter gives an overview of type systems designed for untyped languages.
-The second section explores how existing type systems statically type check multimethods
-and their relevance to Clojure's multimethods.
+%Typed Clojure is related to several existing systems for both typed and untyped languages.
 
@@
 
-\section{Intersection Types}
+\section{Intersection and Union Types}
 
-Intersection types are an interesting type construct relevant to capturing the complicated
-types manipulated by dynamic languages.
+Intersection and union types are interesting type constructs relevant to capturing the complicated
+types common in dynamic languages.
+An intersection of types \lstinline|a| and \lstinline|b|, written \lstinline|(I a b)| in Typed Clojure,
+can be used safely in both positions expecting type \lstinline|a|,
+and positions expecting type \lstinline|b|.
+Intuitively, \lstinline|(I a b)| is both of type \lstinline|a| \emph{and} \lstinline|b|.
+A union of types \lstinline|a| and \lstinline|b| is \emph{either} of type \lstinline|a| or
+\lstinline|b|, written \lstinline|(U a b)| in Typed Clojure,
+and can be used safely in positions expecting a type that is \emph{both} type \lstinline|a|
+and \lstinline|b|.
+For example, \lstinline|(U Number Boolean)| cannot be used in positions expecting \lstinline|Number|.
+
+Several interesting projects have used intersection or union types.
 
@@
 
-% CIDRE
 SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}
 that features a form of intersection types. It is used for function overloading
-and checking invariants involving Refinement types.
+and checking invariants involving refinement types.
 \begin{verbatim}
@@
 
-Typed Racket, developed by Tobin-Hochstadt and others \cite{Tob10}, uses \emph{ordered intersection types} 
-to offer a kind of function overloading.
-\begin{verbatim}PADL '12 paper, Typing the Numeric Tower\end{verbatim}
+St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
+describe the \emph{ordered intersection types} used in Typed Racket\cite{St12}
+that provide a kind of function overloading.
+Typed Clojure takes a similar approach
+\begin{verbatim}
+described in chapter
+\end{verbatim}
+Typed Racket also uses union types
 
-\section{Statically Typed Multimethods}
-
-Clojure provides two dispatching mechanisms: multimethods, a mechanism to dispatch on the
-results of an arbitrary function; and protocols, a mechanism for single dispatch, a common
-strategy that dispatches on the type of the first provided argument.
-Millstien and Chambers explore statically typing multiple dispatched multimethods in Dubious \cite{MS02}.
-Dubious' multimethods are less powerful than Clojure's multimethods but more powerful than protocols,
-and Clojure's dispatch functions can be extended at runtime, unlike Dubious' multimethods.
-
-Clojure and the statically typed language Scala \cite{OCD+} 
-have some interesting similarities. The redesign
-of the Scala collection hierarchy\cite{OS10}
-bears similarity to Clojure's encompassing ``seq''
-model, revealing an opportunity to test potential typed extensions to the Clojure 
-class hierarchy.
+\section{Java Interoperability}
 
@@
 to prevent erroneous usage of \lstinline|null|.
+\begin{verbatim}
+eg.
+\end{verbatim}
 
@@
 A function with multiple arguments is represented by chaining several single-argument function
-together. This style is characterised by direct syntactic function currying, where applying
+together, or by using lists, or using tuples. This style is characterised by direct syntactic function currying, where applying
 a function to less than its maximum number of arguments results in another function

### 2012-10-17 dissertation/paper.tex
Implementation work

@@
   %showtabs=false,                 % show tabs within strings adding particular underscores
-  %frame=single,           % adds a frame around the code
+  frame=single,           % adds a frame around the code
   %tabsize=2,          % sets default tabsize to 2 spaces
@@
 A prototype type system is developed for Clojure running on the Java Virtual Machine.
-This prototype is sufficient to allow experiments with existing Clojure code that is sufficiently complicated
-that type checking increases confidence that the code is correct.
+Several experiments are conducted using this prototype, particularly involving
+sufficiently complicated existing Clojure code where type checking increases confidence that the code correct.
 For example, a significant portion of algo.monads, a Clojure Contrib library for monadic programming, 
@@
 There is significant future work to fully type check all Clojure features and idioms.
-For example, multi-method definitions and functions accepting an even number of variable arguments
+For example, multimethod definitions and functions accepting an even number of variable arguments
 are troublesome. 
@@
 as automatic runtime contract generation and a sophisticated blame system, 
-both which significantly improve error messages when mixing typed and untyped code.
+both which are designed to improve error messages when mixing typed and untyped code in similar systems.
 I conclude that it appears to be both practical and useful to design and implement an optional static type system for the

### 2012-10-19 dissertation/implementation.tex
Intro to variable-arity

@@
 
-A substitution maps type variables to an upper and lower bound. 
+A substitution maps type variables to a type.
 It is valid to replace all occurrences of the type variables named in the substitution 
-with a type between their associated bounds.
-For example, usually the substitution is applied to \lstinline|R| by the caller of \lstinline|infer|.
-Whether to substitute the upper or lower bound is determined by the variance
-of the type variable in relation to the type being substituted. 
-The specific algorithm is described in Local Type Inference.
+with their associated type.
+For example, substitution are often applied to \lstinline|R| by the caller of \lstinline|infer|
+to eliminate the type variables in \lstinline|X| and \lstinline|Y|.
 
@@
 
-F-bounded polymorphism allows type variables to refer to themselves in their bounds.
+F-bounded polymorphism allows type variables to refer to themselves in their type bounds.
+Bounds are checked after a substitution is generated, guaranteeing no substitution
+can violate type variable bounds. To support F-bounds, the substitution being checked is applied to
+the lower and upper bounds for each type variable, and the type associated with the type variable 
+in the substitution is checked to be between these bounds.
 
@@
 
-It also required changes to the polymorphic type inference algorithm.
-The constraint-generation portion
+It also required changes to the polymorphic type inference, where each
+reference to a type variable required a special case for a dotted type variable.
+For example, the constraint-generation algorithm for Local Type Inference
+features extra kinds of constraints for dotted variables.
+Strickland, Tobin-Hochstadt, and Felleisen elaborate on the particular changes
+required for the Typed Racket implementation\cite{STF09}.
 
@@
 
+\section{Proposition Inference for Functions}
+\label{sec:filterneg}
+
+A feature not yet implemented in Typed Racket is the ability
+to infer new propositions based on existing propositions of a function.
+This feature was added to Typed Clojure to support filtering
+a sequence based on negative information, such as filtering values
+that are \emph{not} \lstinline|nil|.
+
+\begin{lstlisting}[caption=Type annotation for filter, label=lst:filterann]
+(ann clojure.core/filter 
+  (All [x y]
+    [[x -> Any :filters {:then (is y 0)}] (U nil (Seqable x)) -> (Seqable y)]))
+\end{lstlisting}
+
+To better understand the problem, listing \ref{lst:filterann} presents the type of \lstinline|filter|,
+which takes a function \lstinline|f| and a sequence \lstinline|s| as arguments, and returns a sequence that 
+contains each element in \lstinline|s| such that applying \lstinline|f| to the element returns a true value.
+The \lstinline|:filters| syntax requires some explanation. 
+Function types support an optional \emph{filter set} attached to its return type, written as a map
+with \lstinline|:then| and/or \lstinline|:else| keys (if omitted, they default to the trivially true proposition which has no effect). 
+The ``then-proposition'' and ``else-proposition'' are added
+to the type environment when the return value is a true and false value respectively.
+For example, the filter set \lstinline|{:then (is y 0)}| is read ``if the return value is a true value, then the first argument
+must be of type y, otherwise if it is a false value, nothing interesting is enforced''
+The type given for \lstinline|filter| works because the type variable
+\lstinline|y| occurs in both the ``then-proposition'' of the first argument and the return type 
+\lstinline|(Seqable y)|.
+
+\begin{lstlisting}[caption=Troublesome filter, label=lst:filtertrouble]
+(filter (fn [a] (not (nil? a))) coll)|
+\end{lstlisting}
+
+The difficulty starts with something like listing \ref{lst:filtertrouble},
+where the inferred filter set for the first argument to \lstinline|filter|
+is \lstinline|{:then (! nil 0) :else (is nil 0)}|\footnote{(! nil 0) is the proposition that the first argument is \emph{not}
+of type \lstinline|nil|.}. The ``then-proposition'' \lstinline|(! nil 0)| does not fit with
+\lstinline|(is y 0)| that \lstinline|filter| expects.
+
+We can sometimes get around this if we already have a predicate with a positive
+``then-proposition''. For example, if we are filtering out \lstinline|nil| values
+from a sequence of type \lstinline|(Seqable (U Number nil))|, we can replace
+listing \ref{lst:filtertrouble} with \lstinline|(filter number? coll)|,
+where \lstinline|number?| has the filter set \lstinline|{:then (is Number 0) :else (! Number 0)}|.
+This does not work, however, when filtering a sequence of type like \lstinline|(Seqable (U x nil))|
+for some unspecified \lstinline|x| because there is no built-in predicate with ``then-proposition''
+\lstinline|(is x 0)|.
+
+\begin{lstlisting}[caption=Filtering with negative propositions, label=lst:filtergood]
+  (filter (ann-form 
+            #(not (nil? %))
+            [(U nil x) -> boolean :filters {:then (is x 0)}])
+          mvs)
+\end{lstlisting}
+
+Instead, we generate new propositions using a technique 
+suggested by Tobin-Hochstadt\cite{Tob12},
+that follows from the occurrence typing calculus defined 
+by Tobin-Hochstadt and Felleisen\cite{TF10}.
+First the filter set for functions are inferred as usual.
+Then the ``then-proposition'' is applied to the environment and new propositions
+are added to the filter set for each variable in the type environment that changes.
+The same procedure is followed for the ``else-proposition''.
+
+Using this technique, the anonymous function in listing 
+\ref{lst:filtergood}\footnote{The Clojure syntax \lstinline|#(not (nil? \%))| 
+is equivalent to \lstinline|(fn [a] (not (nil? a)))|}
+has the filter set \lstinline|{:then (& (! nil 0) (is x 0)) :else (is nil 0)}|
+which is good enough to infer the \lstinline|filter|ed result as \lstinline|(Seqable x)|.
+
 \section{Non-hygienic Macro Expansion and Filters}
@@
 %      false))
-
-\section{Filtering on Negative Propositions}
-\label{sec:filterneg}

### 2012-10-19 dissertation/intro.tex
Intro to variable-arity

@@
 \section{Motivation}
-
-% Type Systems for lisp?
-% Why type systems?
-% Why Clojure?
 
@@
 of both static and dynamic typing.
-Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers optional type annotations
-that do not affect runtime semantics. 
+Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers a simple form of optional
+static typing that specifically do not affect runtime semantics.
 Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
@@
 
-Typed Clojure supports F-bounded polymorphism. All type variables may have
-upper and lower bounds.
+Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff\cite{CCHOM89}. 
+F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
+can be restricted by \emph{bounds}.
+In particular, F-bounded polymorphism allows type variable bounds to recursively refer to the
+variable being bounded. Typed Clojure supports upper and lower type variable bounds.
 
@@
 
+\subsection{Singleton Types}
+
+Following Typed Racket, singleton types for certain values are provided
+in Typed Clojure.
+A singleton type is a type with a single member, like \lstinline|1|,
+\lstinline|:a|, or \lstinline|"a"|\footnote{Strings are delimited by " in Clojure}.
+Typed Clojure provides syntax for singleton types, either by passing
+the value literal to the \lstinline|Value| primitive, or by prefixing
+a quote (').
+
+\begin{lstlisting}[caption=Singleton Types, label=lst:singletoneeg]
+(ann k ':my-keyword)
+(def k :my-keyword)
+\end{lstlisting}
+
+Listing \ref{lst:singletoneeg} shows a simple example of using
+singleton types in Typed Clojure.
+Singleton types are are discussed further in section \ref{sec:intersectionunion}.
+
 \subsection{Heterogeneous Maps}
 
+A novel feature of Clojure compared to other dialects of Lisp (including Racket)
+is an emphasis on key-value paired, unordered \emph{maps}.
 Where particularly object-oriented languages reach for objects, Clojure
 utilises maps. Clojure provides a map literal using curly braces. For example,
-\lstinline|{:a 1, :b 2}| is a map with two key-value entries: from Keyword key \lstinline|:a|
-to value \lstinline|1|, and Keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
-whitespace in Clojure and are included for readability.
+\lstinline|{:a 1, :b 2}| is a map value with two key-value entries: from keyword key \lstinline|:a|
+to value \lstinline|1|, and keyword key \lstinline|:b| to value \lstinline|2|. Note that commas are always
+whitespace in Clojure and are included occasionally for readability.
 
-Typed Clojure provides a heterogeneous map type, restricted to 
-maps with singleton Keyword keys. This restriction is reflected
+Typed Clojure provides a heterogeneous map type which captures this common
+``maps as objects'' pattern. A heterogeneous map type has only \emph{positive}
+information on the types of key-value entries. In other words, it conveys
+whether a particular key is present, but not whether it is absent.
+The implications of this is discussed in section \ref{ref:designhmap}.
+Heterogeneous maps only support keys that are singleton Keyword types. This restriction is reflected
 in the syntax for defining heterogeneous map types.
@@
 ...
-(ann config
-     (HMap {:file String,
-            :ns Symbol}))
+(ann config '{:file String, :ns Symbol}))
 (def config
@@
 
-\subsection{Multi-Arity Functions}
-
-Clojure function objects support multiple arities, allowing dispatch
-on the number of supplied arguments.
-
-\begin{lstlisting}[caption=Annotating multi-arity functions in Typed Clojure]
-(ns typed.test.poly
-  (:require [typed.core :refer [ann AnyInteger check-ns cf]])
-  (:import [clojure.lang Seqable]))
-
-(ann repeatedly'
-     (All [x]
-       (Fn [[-> x] -> (Seqable x)]
-           [AnyInteger [-> x] -> (Seqable x)])))
-(defn repeatedly'
-  "Takes a function of no args, presumably with side effects, and
-  returns an infinite (or length n if supplied) lazy sequence of calls
-  to it"
-  ([f] (lazy-seq (cons (f) (repeatedly' f))))
-  ([n f] (take n (repeatedly' f))))
-\end{lstlisting}
-
-This example defines \lstinline|repeatedly'|\footnote{Identical to \lstinline|clojure.core/repeatedly|},
-a multi-arity function taking either a zero-argument function, or an integer and a zero-argument function as arguments.
-The type variable \lstinline|x| links the return type of the generator function to
-the resulting \lstinline|Seqable| instance.
-
-When checking the definition of multi-arity functions, Typed Clojure
-finds the matching function types for a particular arity by number of arguments
-and requires each relevant type to check successfully.
-
 \subsection{Variable-Arity Functions}
 
-Clojure functions can take an optional \emph{rest} argument.
+Functions in Clojure are multi-arity, which means a function
+can be defined with several function bodies (or \emph{arities})
+and which arity is executed depends on the number of arguments passed
+to the function. A function can have any number of arities with fixed parameters, and at most one arity with
+variable-parameters. Each arity must have a unique number of
+parameters and have a lower number of parameters than the ``variable arity'', if present.
+
+Strickland, Tobin-Hochstadt, and Felleisen invented a calculus and corresponding
+implementation in Typed Racket for variable-arity polymorphism \cite{STF09}
+that is sufficient to handle uniform and non-uniform variable-arity functions.
+Typed Clojure includes an implementation of the most immediately useful parts of variable-arity
+polymorphism using algorithms, nomenclature, and implementation based on this work.
 
@@
 
-\subsubsection{Non-uniform Variable-Arity}
+A function with uniform variable parameters can treat its variable parameter
+as a homogeneous list. 
+Strickland et al. attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
+in a function type, where \emph{T} is some type, and we take an identical approach.
+For example, \lstinline|+| in Clojure accepts any number of arguments
+of type \lstinline|Number|, represented by the type \lstinline|[Number * -> Number]|.
+
+\begin{lstlisting}[caption=Typing multi-arity functions, label=lst:noteq]
+(ann not= (Fn [Any -> boolean]
+              [Any Any -> boolean]
+              [Any Any Any * -> boolean]))
+(defn not=
+  "Same as (not (= obj1 obj2))"
+  ([x] false)
+  ([x y] (not (= x y)))
+  ([x y & more]
+   (not (apply = x y more))))
+\end{lstlisting}
+
+It is common to find Clojure library functions that define seemingly redundant
+function arities for performance reasons.
+Listing \ref{lst:noteq} defines the multi-arity function \lstinline|not=|,
+taken from the Clojure standard library \lstinline|clojure.core| that uses this pattern.
+\lstinline|not=| has three arities in its definition, including
+one that takes a variable number of arguments.
+The \lstinline|Fn| type constructor builds an \emph{ordered function intersection type} from
+function types. Each arity must have at least one matching function type associated with it.
+The two ``fixed arities'' are given familiar function types, with one and two fixed parameters respectively.
+The type given for the arity with variable arguments \lstinline|[Any Any Any * -> boolean]|
+uses a starred pre-type to signify any number of arguments of type \lstinline|Any|
+can be provided to the right of its fixed arguments.
+
+\subsubsection{Non-uniform Variable-Arity Functions}
+
+Where \emph{uniform} variable-arity function types use \emph{starred pre-types}, \emph{non-uniform}
+variable-arity function types use \emph{dotted pre-types}.
+Typed Clojure supports usages of \emph{non-uniform} variable-arity functions,
+where the variable parameter is a heterogeneous list, represented by a \emph{dotted type variable}.
+
+For example, the variable argument function \lstinline|clojure.core/map| takes a function and one or more sequences,
+and returns the result of applying the function argument to each element of the sequences in a pair-wise fashion.
+Its type is given in listing \ref{lst:maptype}.
+
+\begin{lstlisting}[caption=Type signature for \lstinline|clojure.core/map|, label=lst:maptype]
+(ann clojure.core/map
+     (All [c a b ...]
+       [[a b ... b -> c] (U nil (Seqable a)) (U nil (Seqable b)) ... b -> (LazySeq c)]))
+\end{lstlisting}
+
+By adding \lstinline|...| after the last type variable in an \lstinline|All| binder
+we can introduce a \emph{dotted type variable} into scope, which is a placeholder for a sequence of types.
+A \emph{dotted pre-type} \lstinline|T ... b| over \emph{base} \lstinline|T| (a type) and \emph{bound}
+\lstinline|b| (a dotted type variable)
+serves as a placeholder for this sequence of types.
+Dotted pre-types must appear to the right of all fixed parameters in a function type,
+and cannot be mixed with other kinds of variable parameters like starred pre-types.
+When a sequence of types of length \emph{n} is associated with a dotted pre-type, 
+the dotted pre-type is expanded to \emph{n} copies of \lstinline|T|.
+One other special property of a dotted pre-type is that the bound \lstinline|b|
+is \emph{in scope} as a \emph{normal} type variable in its base \lstinline|T|.
+
+To demonstrate dotted pre-types, we use \lstinline|typed.core/inst| to instantiate
+\lstinline|map|. \lstinline|inst| takes a polymorphic expression and a number of types
+that are satisfactory for instantiating the polymorphic type and returns an expression
+of the instantiated polymorphic type.
+
+The instantiation
+
+\begin{lstlisting}
+(inst map Number boolean String)
+\end{lstlisting}
+
+returns an expression of type
+
+\begin{lstlisting}
+[[boolean String Integer -> Number] (U (Seqable boolean) nil) (U (Seqable String) nil) (U (Seqable Integer) nil) -> (LazySeq Number)]
+\end{lstlisting}
+
+However, if sufficient types are given, the instantiation for \lstinline|map| can be inferred.
+The invocation
+
+\begin{lstlisting}
+(map (ann-form (fn [a b c] c) 
+               [boolean String Number -> Number]) 
+     [true false] ["mystr" "astr"] [1 4])
+\end{lstlisting}
+
+uses \lstinline|typed.core/ann-form| to assign an expected to
+the first argument,
+which is sufficient to infer the result type
+
+\begin{lstlisting}
+(LazySeq Number)
+\end{lstlisting}
 
@@
 
-Typed Clojure uses occurrence typing to improve inference\footnote{TODO occurrence typing reference}.
-Occurrence typing is used to refine types for immutable local bindings 
-as programs follow conditional branches.
+It is common in Clojure, like other dynamically-typed languages, to
+encode implicit type invariants in conditional tests.
+In listing \ref{lst:numvec2} conditional tests are used
+to refine the type of the bindings \lstinline|a| and \lstinline|b|.
+Occurrence typing is a technique useful for capturing these kinds
+of type invariants (occurrence typing is discussed in further detail in section \ref{sec:OccurrenceTyping}).
 
-\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure]
+\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure, label=lst:numvec2]
 (ann num-vec2 
@@
 
-Typed Clojure currently provides safe interop\footnote{Interop is short for interoperability} with (non-Generic) Java
+Typed Clojure currently provides safe interoperability
 \footnote{Compatibility with Generic Java is planned future work}.
 
-The key to safe Java interop is the treatment of Java's \emph{null}.
+The key to safe Java interoperability is the treatment of Java's \emph{null}.
 \emph{null} is a subtype to all reference types

### 2012-10-19 dissertation/lit-review.tex
Intro to variable-arity

@@
 
-\section{Intersection and Union Types}
+\section{Intersection, Union, and Singleton Types}
+\label{sec:intersection_types}
+
+% Hiyashi Singleton types
+% Intersection types - CDV81
+% Unions in soft typing CF91
+% - Static Type Inference in a Dynamically Typed Lanugage
+%  - cited by CF91, which has unions
 
@@
 types common in dynamic languages.
-An intersection of types \lstinline|a| and \lstinline|b|, written \lstinline|(I a b)| in Typed Clojure,
-can be used safely in both positions expecting type \lstinline|a|,
+An expression of type \lstinline|(I a b)|, an intersection type including types \lstinline|a|
+and \lstinline|b| in Typed Clojure, can be used safely in both positions expecting type \lstinline|a|,
 and positions expecting type \lstinline|b|.
-Intuitively, \lstinline|(I a b)| is both of type \lstinline|a| \emph{and} \lstinline|b|.
-A union of types \lstinline|a| and \lstinline|b| is \emph{either} of type \lstinline|a| or
-\lstinline|b|, written \lstinline|(U a b)| in Typed Clojure,
-and can be used safely in positions expecting a type that is \emph{both} type \lstinline|a|
-and \lstinline|b|.
+An expression of type \lstinline|(U a b)|, a union type including types \lstinline|a| in Typed Clojure,
+and expressions of this type can be used safely in positions 
+that expect a type that is \emph{either} type \lstinline|a|
+or \lstinline|b|.
 For example, \lstinline|(U Number Boolean)| cannot be used in positions expecting \lstinline|Number|.
+% TODO Intersection example
 
@@
 
-SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}
-that features a form of intersection types. It is used for function overloading
-and checking invariants involving refinement types.
+Refinement types add a level of types refining an existing type system
+in order to type check more detailed properties and invariants than standard static type systems
+(described by Freeman and Pfenning\cite{FP91}).
+Intersection types are critical here to allow more than one property or invariant
+to be expressed for a function.
+SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}.
 \begin{verbatim}
-Freeman & Pfenning, Refinement types
+Refinement types are similar to this project ... how?
 \end{verbatim}
@@
 \end{verbatim}
-Typed Racket also uses union types
+Typed Racket also uses disjoint union types,
 

### 2012-10-21 dissertation/design-choices.tex
Intro to variable-arity
Fixed citation breaking spaces. Latest corrections of intro

@@
 Many of the design goals are similar to Tobin-Hochstadt's 
-for Typed Racket \cite{Tob10}. Some of the main differences are described in this chapter.
+for Typed Racket~\cite{Tob10}. Some of the main differences are described in this chapter.
 
@@
 \section{Heterogeneous Maps}
+\label{ref:designhmap}
 % keyword keys for now, generally encouraged, fast
 % Some heterogeneous maps have string args, eg. from web frameworks
+
+\begin{verbatim}
+HMaps only convey positive information, not negative.
+Map concatenation is not approached, see lit review, many
+complications. Instead, only support assoc, dissoc, get.
+\end{verbatim}
 
@@
 Here we give a formal syntax, operational semantics and type system using inference
-rules, as is standard in the programming language community (see, eg., Pierce\cite{Pie02}).
+rules, as is standard in the programming language community (see, eg., Pierce~\cite{Pie02}).
 

### 2012-10-21 dissertation/experiments.tex
Implementation work
Fixed citation breaking spaces. Latest corrections of intro

@@
 \section{Monads}
+\label{sec:monads}
 
-Monads are an interesting control structure most used in functional programming languages,
-as described by Wadler\cite{Wad95}.
-They are most recognisable from its inclusion in the statically-typed language Haskell\cite{Mar10},
+Monads are an interesting control structure used in functional programming languages,
+as described by Wadler~\cite{Wad95}.
+They are most recognisable from its inclusion in the statically-typed language Haskell~\cite{Mar10},
 where monads are relied on for many features including global state, file output, and exceptions.
 
-I chose to port the Clojure Contrib library \emph{algo.monads}\cite{Hic12} to Typed Clojure. The library
+I chose to port the Clojure Contrib library \emph{algo.monads}~\cite{Hic12} to Typed Clojure. The library
 provides several kinds of monads, monad transformers, and monadic functions.

### 2012-10-21 dissertation/implementation.tex
Fixed citation breaking spaces. Latest corrections of intro

@@
 If the expected type is omitted, a type is instead \emph{synthesised} for the expression.
-The algorithm is based on Pierce and Turner's  Local Type Inference\cite{PT00}
-and the implementation is similar in shape to Typed Racket's\cite{TF08}, in that
+The algorithm is based on Pierce and Turner's  Local Type Inference~\cite{PT00}
+and the implementation is similar in shape to Typed Racket's~\cite{TF08}, in that
 one function with an extra ``expected type'' argument is preferred over two complementary
@@
 In Typed Clojure there are several types of propositions, referred to as \emph{filters}.
-They are based on the theory by Tobin-Hochstadt and Felleisen\cite{TF10}, and are ported directly
+They are based on the theory by Tobin-Hochstadt and Felleisen~\cite{TF10}, and are ported directly
 from Typed Racket.
@@
 for bounded variables,
-and is reminiscent to Pierce and Turner's Local Type Inference\cite{PT00}.
+and is reminiscent to Pierce and Turner's Local Type Inference~\cite{PT00}.
 A common entry point for inferring type variables to polymorphic function invocations
@@
 They are a stub for a more comprehensive treatment of higher-kinded operators
-such as that described by Moors, Piessens, and Odersky for Scala\cite{MPO08}.
+such as that described by Moors, Piessens, and Odersky for Scala~\cite{MPO08}.
 At present, a type variable can only be instantiated to a type between
@@
 Strickland, Tobin-Hochstadt, and Felleisen elaborate on the particular changes
-required for the Typed Racket implementation\cite{STF09}.
+required for the Typed Racket implementation~\cite{STF09}.
 
@@
 Instead, we generate new propositions using a technique 
-suggested by Tobin-Hochstadt\cite{Tob12},
+suggested by Tobin-Hochstadt~\cite{Tob12},
 that follows from the occurrence typing calculus defined 
-by Tobin-Hochstadt and Felleisen\cite{TF10}.
+by Tobin-Hochstadt and Felleisen~\cite{TF10}.
 First the filter set for functions are inferred as usual.

### 2012-10-21 dissertation/intro.tex
Fixed citation breaking spaces. Latest corrections of intro

@@
 \emph{It is practical and useful to design and implement an optional typing system 
-for the Clojure programming language that allows Clojure programmers to continue 
+for the Clojure programming language using bidirectional checking that allows Clojure programmers to continue 
 using idioms and style found in current Clojure code.}
@@
 dynamically typed languages with static type systems. 
-This idea not new, but recent attempts are noteworthy for their successful use of bidirectional checking\footnote{These 
-slides by Felleisen, a practitioner in this area for over 20 years, present a history of optional type checking
-for Lisp/Scheme that spans around 40 years\cite{Fell09}}.
-Instead of always attempting to infer types the algorithm relies on programmer annotations
-appearing in some natural places, such as giving the type of each top-level function.
+This is idea not new, but recent attempts are noteworthy for 
+their broad success in matching many of the advantages of statically typed
+languages, notably due to the use of \emph{bidirectional checking}.
+Instead of always attempting to infer types, this algorithm relies on programmer annotations
+appearing in some natural places such as giving the type of each top-level function.
 
@@
 offers ways to access Java-like speed for certain operations.
-Also, Clojure's excellent host interoperability offers Clojure programmers
+Also, Clojure's extensive host interoperability offers Clojure programmers
 access to existing libraries for their platform, such as the extensive Java library ecosystem.
-By coupling pragmatic necessities with elegant features like Lisp, functional programming,
+By coupling pragmatic necessities with elegant features like Lisp-style macros, functional programming,
 and immutability by default, Clojure is a compelling general purpose programming language.
 
-Recently languages have been created to or been modified to support aspects
+Recently a number of languages have been created to or been modified to support aspects
 of both static and dynamic typing.
@@
 a well-known dynamic language.
+Typed Racket~\cite{TF08,Tob10} goes further by offering safe interoperability
+between typed and untyped modules by generating appropriate runtime assertions based
+on expected static types.
 
@@
 type invariants. For example, ``design by contract'' is often used,
-introduced by Meyer for the Eiffel language\cite{Mey92},
+introduced by Meyer for the Eiffel language~\cite{Mey92},
 in which the programmer defines contracts that are enforced at runtime.
@@
 
-Static type systems, however, are still desirable for certain situations.
-Powerful type systems like ML's\cite{Mil97} and Haskell's\cite{Mar10} have proved particularly useful
-for writing in complicated programming styles. For example,
-Haskell's advanced static type system helps the programmer write correct monadic code,
+Static type systems, however, are still desirable for many situations.
+Powerful type systems like ML's~\cite{Mil97} and Haskell's~\cite{Mar10} have proved particularly useful
+when complicated programming styles are required. For example,
+Haskell's advanced static type system helps the programmer write correct monadic code
+(as detailed by Wadler~\cite{Wad95})
 especially in more complicated situations like combining monads via monad transformers.
@@
 The initial motivation for implementing an optional type system
-for Clojure occurred during an anecdotal account of the struggles
-of a Clojure programmer. In a heroic effort, he managed to 
-implement a library for conduits (or \emph{arrows}) in Clojure, 
-which probably surpass monads in complexity, and is usually reserved
+for Clojure was an anecdotal account of the struggles
+of a Clojure programmer. In an apparently heroic effort, he managed to 
+implement a Clojure library for conduits, an advanced form of ``pipes'',
+using arrows, a generalisation of monads.
+\footnote{CONDUIT REFERENCE TODO}
+Conduits also significantly surpass monads in complexity and are usually reserved
 for languages with advanced type systems like Haskell.
@@
 He highlighted a strong desire for a type system for several reasons.
-Firstly, to verify the correctness
-of the library (before noting that unit tests and runtime contracts
-were insufficient for this task).
+Firstly, to verify the correctness of the library.
+Without a static type system, it is a significant task
+to verify such a implementation correct due to heavy use
+of higher-order functions.
 Secondly, to aid him while writing the library.
 
-Since then, the Clojure community has shown interest in this work.
-I developed this project as a Google Summer of Code 2012 project,
-after it was selected by the Clojure community.
+Since then, the Clojure community has shown significant interest in this work.
+I also developed this project as a Google Summer of Code 2012 project,
+after it was selected by the Clojure community as a Clojure Google Summer of Code
+project for 2012.
+\footnote{CITE GSOC, or FOOTNOTE}
 I have also been invited to speak on this project at the Clojure Conj
-2012 conference, a major annual Clojure conference.
+2012 conference in November, the main international conference related to Clojure.
 
@@
 can even be called a type system. We choose to
-follow the terminology of Pfenning\cite{Pfe08} and Reynolds\cite{Rey01},
+follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey01},
 where such optional type systems are \emph{extrinsic} type systems, and more
 traditional type systems are \emph{intrinsic} type systems.
+This distinction has a long history, originating in work in the
 
@@
 as having a trivial static type system that supports exact one type,
-a view strongly advocated by Harper\cite{Har12}
-and common in the literature on static types (for example, Pierce\cite{Pie02}).
+a view strongly advocated by Harper~\cite{Har12}
+and common in the literature on static types (for example, Pierce~\cite{Pie02}).
 
@@
 \end{lstlisting}
-\footnote{Example adapted from Tobin-Hochstadt \cite{Tob10}}
+\footnote{Example adapted from Tobin-Hochstadt~\cite{Tob10}}
 
@@
 
-Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff\cite{CCHOM89}. 
+Typed Clojure supports F-bounded polymorphism, first introduced by Canning, Cook, Hill and Olthoff~\cite{CCHOM89}. 
 F-bounded polymorphism is an extension of bounded polymorphism, where polymorphic type variables
@@
 Strickland, Tobin-Hochstadt, and Felleisen invented a calculus and corresponding
-implementation in Typed Racket for variable-arity polymorphism \cite{STF09}
+implementation in Typed Racket for variable-arity polymorphism~\cite{STF09}
 that is sufficient to handle uniform and non-uniform variable-arity functions.

### 2012-10-21 dissertation/lit-review.tex
Fixed citation breaking spaces. Latest corrections of intro

@@
 
+% cite: Dynamic Typing in Polymorphic languages
+% - coming from the other direction
+% - extending static languages to be more dynamic
+
+% A more recent reference in the same area is "Typed Open Programming" a PhD dissertation from 2007 - at least have a look at 1.3.2 for a good summary of work on dynamics in statically typed languages since 1991 - probably it's good to cite this as a good recent survey of the area.
 \section{Static Types for Untyped Languages}
@@
 
-Soft typing\cite{CF91}
+Soft typing~\cite{CF91}
 is an approach for ensuring type safety in untyped languages.
@@
 
-Wright and Cartwright\cite{WC97} developed Soft Scheme, a practical
+Wright and Cartwright~\cite{WC97} developed Soft Scheme, a practical
 soft type system for Scheme. 
-It extended earlier work by Cartwright and Fagan\cite{CF91}
-and Fagan \cite{Fag91}, adding support for practical features such as
+It extended earlier work by Cartwright and Fagan~\cite{CF91}
+and Fagan~\cite{Fag91}, adding support for practical features such as
 first-class continuations and variable-arity functions.
@@
 Alas, the best reference for commentary on this transition 
-appears to be slides from a talk by Felleisen\cite{Fell09},
+appears to be slides from a talk by Felleisen~\cite{Fell09},
 a leader in this area of research for over 20 years.
@@
 Typed Racket was developed as a path for module-by-module
-porting of existing untyped Racket modules to a typed sister language\cite{Tob10}.
+porting of existing untyped Racket modules to a typed sister language~\cite{Tob10}.
 Once a module is ported and type checked, it is protected from untyped modules
@@
 Scala is a statically typed language on the Java Virtual Machine offering integrated interoperability with Java, a typed language.
-Scala objects and classes can ``inherit from Java classes and implement Java interfaces''\cite{OCD+}
+Scala objects and classes can ``inherit from Java classes and implement Java interfaces''~\cite{OCD+}
 with the usual static type guarantees normal Scala code enjoys.
@@
 Typed Racket includes safe interoperability between any combination of typed and untyped 
-Racket modules\cite{Tob10}\cite{TF08}. 
+Racket modules~\cite{Tob10,TF08}. 
 Interactions with untyped modules are protected by adding runtime checks based on expected types.
@@
 \section{Intersection, Union, and Singleton Types}
-\label{sec:intersection_types}
+\label{sec:intersectionunion}
 
@@
 %  - cited by CF91, which has unions
+
+% Unions, intersections, and singletons all fit best with extrinsic types
+% - this is one reason why they are similar
 
@@
 
-Forsythe, a modern ALGOL dialect by Reynolds\cite{Rey96} was the first wide-spread 
+Forsythe, a modern ALGOL dialect by Reynolds~\cite{Rey96} was the first wide-spread 
 programming language to use intersection types.
@@
 in order to type check more detailed properties and invariants than standard static type systems
-(described by Freeman and Pfenning\cite{FP91}).
+(described by Freeman and Pfenning~\cite{FP91}).
 Intersection types are critical here to allow more than one property or invariant
 to be expressed for a function.
-SML CIDRE is a refinement type checker for Standard ML by Davies \cite{Dav05}.
+SML CIDRE is a refinement type checker for Standard ML by Davies~\cite{Dav05}.
 \begin{verbatim}
@@
 St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
-describe the \emph{ordered intersection types} used in Typed Racket\cite{St12}
+describe the \emph{ordered intersection types} used in Typed Racket~\cite{St12}
 that provide a kind of function overloading.
@@
 to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
-relevant, particularly considering the potential of using occurrence typing \cite{TF10}
+relevant, particularly considering the potential of using occurrence typing~\cite{TF10}
 to prevent erroneous usage of \lstinline|null|.
@@
 \subsection{Variable-Arity Polymorphism}
+\label{sec:variablearity}
 
-Strickland et al. invented a type system supporting variable-arity polymorphism\cite{STF09}
+Strickland et al. invented a type system supporting variable-arity polymorphism~\cite{STF09}
 a version of which is included in the current implementation of Typed Racket.
@@
 
-Typed Racket uses Local Type Inference \cite{PT00}
-as an inference and checking tool. Pierce and Turner
-\cite{PT00} divide Local Type Inference into
+Typed Racket uses Local Type Inference~\cite{PT00}
+as an inference and checking tool. Pierce and Turner~\cite{PT00}
+divide Local Type Inference into
 two complementary algorithms. \emph{Local type argument synthesis}
@@
 Pierce and Turner split \emph{local type argument synthesis} into two further
-algorithms: bounded, and unbounded quantification \cite{PT00}. 
+algorithms: bounded, and unbounded quantification~\cite{PT00}. 
 Typed Racket 
-supports unbounded polymorphism \cite{Tob10}, implementing the latter algorithm by Piece et al.
-Scala supports bounded quantification with F-bounded polymorphism \cite{CCHOM89},
+supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Piece et al.
+Scala supports bounded quantification with F-bounded polymorphism~\cite{CCHOM89},
 basing its type argument synthesis on the bounded quantification algorithm.
 
-Pierce and Turner explicitly forbid \cite{PT00}
+Pierce and Turner explicitly forbid~\cite{PT00}
 attempting to synthesise type variables with interdependent bounds, including
@@
 I am not aware of papers specifically describing Scala's modifications, but they are at least inspired by
-Scala's spiritual ancestors Generic Java \cite{BOSW98} and Pizza \cite{OW97}.
+Scala's spiritual ancestors Generic Java~\cite{BOSW98} and Pizza~\cite{OW97}.
 
-Hosoya and Pierce \cite{HP99} reiterate two common problems with Local Type Inference:
+Hosoya and Pierce~\cite{HP99} reiterate two common problems with Local Type Inference:
 ``hard-to-synthesise arguments'' and ``no best type argument''. The first problem occurs because
@@
 
-Scala's type checking uses Colored Local Type Inference \cite{OZZ01},
-a variant of Local Type Inference \cite{PT00} specifically designed to
+Scala's type checking uses Colored Local Type Inference~\cite{OZZ01},
+a variant of Local Type Inference~\cite{PT00} specifically designed to
 improve inference with certain kinds of Scala pattern matching expressions. It allows
@@
 and missing parts of the type. They are generally useful
-for describing ``information flow in polymorphic type systems with propagation-based type inference''
-\cite{OZZ01}. If the scope of my research permits it, I plan to investigate
+for describing ``information flow in polymorphic type systems with propagation-based type inference''~\cite{OZZ01}. 
+If the scope of my research permits it, I plan to investigate
 using colored types to solve the polymorphic higher-order-function limitation of Local Type Inference
@@
 
-Local type inference by Pierce and Turner\cite{PT00}
+Local type inference by Pierce and Turner~\cite{PT00}
 describe two implementations of type variables, for bounded
@@
 Bounded type variables can have subtype constraints, and 
-subsume unbounded type variables \cite{PT00}, 
+subsume unbounded type variables~\cite{PT00}, 
 as a unbounded variable can be represented as a variable bounded
@@
 is more subtle, due to ``some surprising interactions between bounded quantifiers
-and the \emph{Bot} type'' \cite{PT00}, described fully
-by Pierce \cite{Pie97}.
+and the \emph{Bot} type''~\cite{PT00}, described fully
+by Pierce~\cite{Pie97}.
 
-Typed Racket \cite{TF08}
-supports unbounded polymorphism, while Scala \cite{OCD+}
+Typed Racket~\cite{TF08}
+supports unbounded polymorphism, while Scala~\cite{OCD+}
 supports an extended form of bounded polymorphism called
-F-bounded polymorphism \cite{CCHOM89}, which allows the
+F-bounded polymorphism~\cite{CCHOM89}, which allows the
 bound variable to occur in its own bound.
 F-bounded polymorphism is useful in the context of object-oriented abstractions,
-as demonstrated by Odersky \cite{OCD+}.
+as demonstrated by Odersky~\cite{OCD+}.
 This is one possible explanation why Typed Racket, which is not built on abstractions like Scala,
@@
 attempts to preserve existing Racket idioms and aims type check
-existing Racket code by simply adding top level type annotations \cite{Tob10}.
+existing Racket code by simply adding top level type annotations~\cite{Tob10}.
 
-Typed Racket fully expands all macro calls before type checking \cite{Tob10} to
+Typed Racket fully expands all macro calls before type checking~\cite{Tob10} to
 avoid the complex semantics of type checking macro definitions, an ongoing research area summarised
- by Herman \cite{Her10}.
+ by Herman~\cite{Her10}.
 The design of my Clojure type system will follow a similar strategy; only the fully macro-expanded form
@@
 Along with a full static type system, Typed Racket 
-also uses runtime contracts to enforce type invariants at runtime \cite{TF08}.
+also uses runtime contracts to enforce type invariants at runtime~\cite{TF08}.
 Utilising runtime contracts to aid type checking is outside the scope of this project, but would be 
@@
 Two other Typed Racket features that will be explored are recursive types and refinement types  
-\cite{Tob10}. Recursive types allow a type definition to refer to itself, enabling structurally
+~\cite{Tob10}. Recursive types allow a type definition to refer to itself, enabling structurally
 recursive types like binary trees. Refinement types let the programmer define
@@
 
-Occurrence typing \cite{TF08}
-\cite{TF10} extends the type 
+Occurrence typing~\cite{TF08,TF10} extends the type 
 system with a \emph{proposition environment} that represents 
 the type refinements inferred down a particular path.
-The 2010 paper \cite{TF10}
+The 2010 paper~\cite{TF10}
 reformulated occurrence typing, improving the original 2008 paper
-\cite{TF08}
+~\cite{TF08}
 after revealing that ``three years of practical experience has revealed
-serious shortcoming of our type system.''\cite{TF10}
+serious shortcoming of our type system.''~\cite{TF10}
 The new implementation utilizes a simple proof system to solve
@@
 
-Millstein and Chambers\cite{MS02}
+Millstein and Chambers~\cite{MS02}
 describe Dubious, a simple statically typed core language including multimethods that
@@
 multimethods: ``it is possible for two modules containing arbitrary multimethods to typecheck
-successfully in isolation but generate type errors when linked together.''\cite{MS02}
+successfully in isolation but generate type errors when linked together.''~\cite{MS02}
 

### 2012-10-21 dissertation/intro.tex
2nd page of revisions from rowan

@@
 Also, Clojure's extensive host interoperability offers Clojure programmers
-access to existing libraries for their platform, such as the extensive Java library ecosystem.
+access to existing libraries for their platform, such as the vast Java library ecosystem.
 By coupling pragmatic necessities with elegant features like Lisp-style macros, functional programming,
@@
 Without a static type system, it is a significant task
-to verify such a implementation correct due to heavy use
+to verify such an implementation as correct due to heavy use
 of higher-order functions.
@@
 traditional type systems are \emph{intrinsic} type systems.
-This distinction has a long history, originating in work in the
+This distinction has a long history, originating in work in the $\lambda$-calculus 
+by Church \footnote{CHURCH FROM ROWANS DISSERTATION} and Curry \footnote{CURRY FROM ROWANS DISSERTATION},
+leading to intrinsic types and extrinsic types also being called Church types and
+Curry types often in the literature.
 
@@
 meaningful. Pfenning and Reynolds call these type systems \emph{intrinsic}. A language with an intrinsic
-static type system has run-time semantics that depends on the types of associated
-variables and expressions during type checking.
-For example, C, Java, and ML have intrinsic types.
+static type system has a run-time semantics that depends on the types associated
+with variables and expressions during type checking.
+For example, C, Java, ML, and Haskell have intrinsic types.
 This means programs written in these languages must pass the type checker before being run.
@@
 A static type system is \emph{extrinsic} when runtime semantics
-does not depend on the type system. In other words, passing a static type checker
+does not depend on a static type system. In other words, passing a static type checker
 is not essential to running programs. A dynamically typed language can be viewed
-as having a trivial static type system that supports exact one type,
-a view strongly advocated by Harper~\cite{Har12}
+as having a trivial static type system that supports exactly one type,
+a view advocated by Harper~\cite{Har12}
 and common in the literature on static types (for example, Pierce~\cite{Pie02}).
@@
 
-For example, the Java expressions \emph{1 + 2 / 3} is written in Lisp pseudocode \lstinline|(/ (+ 1 2) 3)|
+For example, the Java expressions \emph{(1 + 2) / 3} is written in Lisp pseudocode \lstinline|(/ (+ 1 2) 3)|
 and \emph{numberCrunch(1, 2)} written \lstinline|(numberCrunch 1 2)|.
@@
         \lstinline|:a| to \lstinline|1| and \lstinline|:b| to \lstinline|2|.
-  \item Commas are always optional, and treated as whitespace.
+  \item Commas are always optional and treated as whitespace, but often used to show the intended structure.
 \end{itemize}
@@
 For example, \lstinline|ns| refers to the var \lstinline|clojure.core/ns|,
-similarly \lstinline|println| refers to \lstinline|clojure.core/println|.
+similarly \lstinline|println| refers to \lstinline|clojure.core/println|
+(vars are global bindings in Clojure).
 
@@
 \lstinline|typed.core|, Typed Clojure's main namespace. It also refers the var \lstinline|typed.core/check-ns|
-into scope\footnote{\lstinline|check-ns| is the top level function for type checking a namespace}.
+into scope (\lstinline|check-ns| is the top level function for type checking a namespace).
 Other than this dependency, this is identical to the untyped \emph{Hello world}.
 
-More complex code require extra annotations to type check:
+More complex code may require extra annotations to type check:
 
-\begin{lstlisting}[caption=Annotating vars in Typed Clojure]
+\begin{lstlisting}[caption=Annotating vars in Typed Clojure (adapted from a Typed Scheme/Racket example by Tobin-Hochstadt~\cite{Tob10})]
 (ns typed.test.collatz
@@
 \end{lstlisting}
-\footnote{Example adapted from Tobin-Hochstadt~\cite{Tob10}}
 
-In this example, we define a new var \lstinline|typed.test.collatz/collatz|. Typed Clojure requires all 
-used vars to be annotated. Here \lstinline|typed.core/ann| annotates \lstinline|typed.test.collatz/collatz|
-to be a function from \lstinline{java.lang.Number} to 
-\lstinline{java.lang.Number}\footnote{All Classes in the \lstinline|java.lang| package
-are automatically imported in every Clojure namespace (the equivalent of Java's \lstinline|import java.lang.*;|).}.
+(\lstinline|(defn [...] body)| expands to \lstinline|(def (fn [...] body))|
+where \lstinline|def| defines a new var and \lstinline|fn| creates a function value).
+
+In this example, we define a new var \lstinline|collatz|
+(when unambiguous, I omit the qualifying namespace/package for the remainder of the chapter).
+Typed Clojure requires all 
+used vars to be annotated. Here \lstinline|typed.core/ann|, a Typed Clojure macro for annotating vars, 
+annotates \lstinline|collatz| to be a function from \lstinline|Number| to 
+\lstinline|Number| (\lstinline|Number| refers to \lstinline|java.lang.Number|,
+due to every Clojure namespace implicitly importing all Classes in the \lstinline|java.lang| package,
+the equivalent of Java's \lstinline|import java.lang.*;|).
 
@@
 
+As well as \lstinline|def| and \lstinline|defn| definitions,
+Clojure programmers typically include datatype and protocol
+definitions. 
+Protocols are similar to Java interfaces\footnote{JAVA INTERFACE REFERENCE} and datatypes to classes implementing interfaces.
 We can annotate datatype and protocol definitions similarly.
@@
 
-\lstinline|clojure.core/defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols}
-with a set of methods. \lstinline|typed.core/ann-protocol| annotates a protocol with the types of its methods.
-In this example, we define a protocol \lstinline|typed.test.person/Age| with an \lstinline|age| method.
-The call to \lstinline|clojure.core/defprotocol| also defines a new var \lstinline|typed.test.person/age|, a first-class function
-wrapping the \lstinline|age| method, but taking the target Object as the first parameter. The
-type signature provided with \lstinline|typed.core/ann-protocol|, here \lstinline|[typed.test.person/Age -> typed.core/AnyInteger]|, 
+\lstinline|defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols.}
+with a set of methods. \lstinline|ann-protocol| is a Typed Clojure macro for annotating 
+a protocol with the types of its methods.
+In this example, we define a protocol \lstinline|Age| with an \lstinline|age| method,
+which is really a first-class function taking the target object as the first parameter. The
+type signature provided with \lstinline|ann-protocol|, here \lstinline|[Age -> AnyInteger]|
+which is the type of the function taking an \lstinline|Age| and returning an \lstinline|AnyInteger|,
 is for this function.
 
-Invocations of \lstinline|clojure.core/defprotocol| are currently not able to be type checked
+Invocations of \lstinline|defprotocol| are currently not able to be type checked
 and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.
 
-\lstinline|clojure.core/deftype|
-defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes}
+\lstinline|deftype|
+defines a new Clojure datatype\footnote{See http://clojure.org/datatypes for a full description of datatypes.}
 in the current namespace with a number of fields and methods. 
-\lstinline|typed.core/ann-datatype| annotates a datatype with its field types.
+\lstinline|typed.core/ann-datatype| is a Typed Clojure form for annotating datatypes, including its field types.
 In this example, we create a datatype \lstinline|typed.test.person.Person|
-with fields \lstinline|name| and \lstinline|age| and implement the \lstinline|age|
-method from protocol \lstinline|typed.test.person/Age|.
+(datatype defintions generate a Java Class, where the current namespace is used as a starting point for its
+qualifying package)
+with fields \lstinline|name| and \lstinline|age| and 
+extend the \lstinline|Age| protocol by implementing the \lstinline|age| method.
 
@@
 in the order they are passed to \lstinline|deftype|
-\footnote{When unambiguous, I omit the qualifying namespace/package for the remainder of the chapter.}.
 \lstinline|(Person. "Lucy" 34)| constructs a new \lstinline|Person|
@@
 is an instance of \lstinline|Age| from the datatype definition, therefore \lstinline|(age (Person. "Lucy" 34))| is type-safe.
+Since generally we only need to annotate definitions and not uses, the number of annotations is relatively small.
 
@@
 variable being bounded. Typed Clojure supports upper and lower type variable bounds.
+\footnote{F-BOUND EXAMPLE NEEDED}
 
-Typed Clojure parameterises some of Clojure's data structures. For example,
-the interface behind Clojure's \emph{seq} abstraction \lstinline|clojure.lang.Seqable| has one 
-covariant parameter\footnote{\lstinline|(Seqable Integer)| being a subtype of \lstinline|(Seqable Number)|
+For accurate type checking, Typed Clojure parameterises some of Clojure's data structures. For example,
+the interface behind Clojure's \emph{seq} abstraction for sequences \lstinline|clojure.lang.Seqable| has one 
+covariant parameter\footnote{Covariant means that \lstinline|(Seqable Integer)| is a subtype of \lstinline|(Seqable Number)|
 because Integer is a subtype of Number.}.
@@
 or is synonymous with passing an empty \lstinline|Seqable|.
+The ability to specify types that explicitly include or exclude \lstinline|nil| is one of the strengths
+of Typed Clojure, and an aspect where it is more expressive than standard type systems like that for Java.
 
@@
 
-A novel feature of Clojure compared to other dialects of Lisp (including Racket)
-is an emphasis on key-value paired, unordered \emph{maps}.
-Where particularly object-oriented languages reach for objects, Clojure
-utilises maps. Clojure provides a map literal using curly braces. For example,
+Clojure emphasises key-value paired, unordered \emph{maps} for many uses where other languages might use objects
+Clojure provides a map literal using curly braces. For example,
 \lstinline|{:a 1, :b 2}| is a map value with two key-value entries: from keyword key \lstinline|:a|
@@
 
-Typed Clojure provides a heterogeneous map type which captures this common
+Typed Clojure provides heterogeneous map types which captures this common
 ``maps as objects'' pattern. A heterogeneous map type has only \emph{positive}
@@
 
-Heterogeneous vector and seq types are also provided.
+Heterogeneous vector and seq types are also provided and work similarly,
+except for their lack of keys like maps, hence are just a sequence of types.
 
@@
 
-Functions in Clojure are multi-arity, which means a function
-can be defined with several function bodies (or \emph{arities})
+Functions in Clojure are multi-arity (a very commonly used feature) which means a function
+can be defined with several sets of function arguments and bodies (or \emph{arities})
 and which arity is executed depends on the number of arguments passed

### 2012-10-22 dissertation/future-work.tex
Example for infer

@@
 \section{Multimethods}
+\label{sec:mmfuture}
+
+\begin{verbatim}
+Multimethods in Clojure are an idiomatic and often used feature.
+It would be essential for an optional static type system for
+Clojure to support
+There are at least two hurdles to supporting multimethods.
+\end{verbatim}
+
 % Problem - typing isa?
@@
 
+A Clojure record is a composite of datatypes and maps; it is a datatype
+with fields that can be treated like a map (records support map operations like \lstinline|get|,
+\lstinline|assoc|, and \lstinline|dissoc|).
+Records have some quirks that may prove difficult to
+
 \subsection{Other Clojure Dialects}

### 2012-10-22 dissertation/implementation.tex
Example for infer

@@
 This chapter discusses the implementation of the Typed Clojure prototype type 
-system\footnote{We refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter.}
+system (we refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter)
 in some detail concentrating on significant challenges that were identified and overcome.
-Many aspects of Typed Clojure's design follows the implementation of Typed Racket,
+Many aspects of Typed Clojure's design follow the implementation of Typed Racket,
 which is reflected in this chapter.
@@
 
-Typed Clojure and Typed Racket differ significantly in how they perform type checking.
+Typed Clojure and Typed Racket differ significantly in how type checking
+is integrated into their programming environments.
 Typed Racket is implemented as a language on the
@@
 Instead, Typed Clojure is implemented as a library that utilises abstract syntax trees (AST)
-generated by analyze\footnote{https://github.com/frenchy64/analyze}, a library I developed for this project.
+generated by analyze~\cite{Analyze2012}, a library I developed for this project.
 This strategy follows common practice for Clojure projects, which favours providing modular libraries 
@@
   \item type alias definitions
-  \item var, protocol, datatype, and Java Class annotations
+  \item global variable, protocol, datatype, and Java Class annotations
   \item Java method annotations, such as \lstinline|nilable-param| and \lstinline|non-nil-return|
@@
 The algorithm is based on Pierce and Turner's  Local Type Inference~\cite{PT00}
-and the implementation is similar in shape to Typed Racket's~\cite{TF08}, in that
+and the implementation is similar in form to Typed Racket's~\cite{TF08}, in that
 one function with an extra ``expected type'' argument is preferred over two complementary
@@
 Racket features like syntax objects for expression representation. Clojure instead
-leans towards abstract syntax tree representation, despite its Lisp heritages.
-In terms of the bidirectional check, the difference is mostly cosmetic.
+leans towards abstract syntax tree representation, despite its Lisp heritage.
+In terms of the bidirectional checking, the difference is mostly cosmetic.
 
@@
   \item TypeFilter and NotTypeFilter represent a positive or negative association of a binding name
-        to a type. The syntax \lstinline|(is t name)|
-  \item AndFilter and OrFilter represent logical conjunction and disjunction of propositions.
+        to a type. The syntax \lstinline|(is t name)| means a proposition that the binding called
+        \lstinline|name| is of type \lstinline|t| (corresponding to TypeFilter).
+        The syntax \lstinline|(! t name)| means a proposition that the binding called \lstinline|name|
+        is \emph{not} of type \lstinline|t| (corresponding to NotTypeFilter).
+  \item AndFilter and OrFilter represent logical conjunction of propositions 
+    (written \lstinline[mathescape]|(& $\overrightarrow{p}$)|), and
+    logical disjunction of propositions
+    (written \lstinline[mathescape]{(| $\overrightarrow{p}$)}) for propositions \lstinline|p|.
+
 \end{itemize}
@@
 inferring length information (CountPE), and \lstinline|first| and \lstinline|rest| paths for sequences.
-These additions do not appear to introduce any significant new complexities related to paths.
+These additions do not appear to introduce any major new complexities related to paths.
 
@@
 for bounded variables,
-and is reminiscent to Pierce and Turner's Local Type Inference~\cite{PT00}.
-A common entry point for inferring type variables to polymorphic function invocations
-\lstinline|typed.core/infer|.
+and is directly based on Pierce and Turner's Local Type Inference~\cite{PT00}.
+A common entry point for inferring type variables for polymorphic function invocations
+is \lstinline|typed.core/infer|.
 
@@
         to their bounds (representing the \emph{dotted} type variables in scope),
-  \item \lstinline|S| and \lstinline|T| are sequences of types of equal length
-  \item \lstinline|R| is a type,
+  \item \lstinline|S| and \lstinline|T| are sequences of types of equal length,
+        (usually the types of the actual arguments provided and the types of the parameters
+        of the polymorphic function),
+  \item \lstinline|R| is a result type (usually the return type of the polymorphic function),
   \item \lstinline|expected| is the expected type for R, or the value \lstinline|nil|,
@@
 to eliminate the type variables in \lstinline|X| and \lstinline|Y|.
+
+\lstinline|infer| is almost always used when invoking polymorphic functions.
+\lstinline|clojure.core/constantly| is a polymorphic function with type \lstinline|(All [x y] [x -> [y * -> x]])|
+(read as a function taking \lstinline|x| and returning a function that takes any number
+of \lstinline|y|'s and returns \lstinline|x|).
+For instance, \lstinline|((constantly true) 'any 'number)| results in the value \lstinline|true|.
+
+Type checking the invocation \lstinline|(constantly true)|
+calls \lstinline|infer| roughly like
+
+\begin{lstlisting}
+(infer {'x no-bounds 'y no-bounds}
+       {}
+       [(parse-type 'true)]
+       [(make-F 'x)]
+       (with-frees [(make-F 'x) (make-F 'y)]
+         (parse-type '[y * -> x]))
+       nil)
+\end{lstlisting}
+
+where the internal Typed Clojure bindings
+
+\begin{itemize}
+  \item \lstinline|no-bounds| is the type variable bounds with upper bound as \lstinline|Any|
+    and lower bound as \lstinline|Nothing|,
+  \item \lstinline|parse-type| is a function that takes type \emph{syntax} and converts it to a type (its argument must be quoted), 
+  \item \lstinline|make-F| is a function that takes a name symbol and returns a type variable type of that name, and
+  \item \lstinline|with-frees| is a macro that brings the type variables named in its first argument into scope in its second argument.
+\end{itemize}
+
+This returns a substitution that replaces occurrences of \lstinline|x| with \lstinline|true|
+and \lstinline|y| with \lstinline|Any|.
+This helps infer a result type of \lstinline|(constantly true)|
+as \lstinline|[Any * -> true]|.
 
@@
 required for the Typed Racket implementation~\cite{STF09}.
+
+\begin{verbatim}
+  Say more about the changes and why so much work was involved. Or at least
+  the remaining changes were more tedious than anything.
+\end{verbatim}
 
@@
 First the filter set for functions are inferred as usual.
-Then the ``then-proposition'' is applied to the environment and new propositions
-are added to the filter set for each variable in the type environment that changes.
+To collect new propositions, the ``then-proposition'' is applied to the type environment (which maps
+local bindings to types). Any types associated with bindings that are changed after this
+are represented as new propositions, which are added to the ``then-proposition'' for this filter set.
 The same procedure is followed for the ``else-proposition''.
@@
 \section{Non-hygienic Macro Expansion and Filters}
+
 % need to remove bindings as they fall out of scope

### 2012-10-22 dissertation/intro.tex
Example for infer

@@
 of both static and dynamic typing.
-Dart\footnote{http://www.dartlang.org/} is dynamically typed but offers a simple form of optional
+Dart~\cite{Dart2012} (created by Google) is dynamically typed but offers a simple form of optional
 static typing that specifically do not affect runtime semantics.
-Typescript\footnote{http://www.typescriptlang.org/} adds an optional type system to Javascript,
+Typescript~\cite{TypeS2012} (created by Microsoft) adds an optional type system to Javascript,
 a well-known dynamic language.
-Typed Racket~\cite{TF08,Tob10} goes further by offering safe interoperability
+Typed Racket~\cite{TF08,Tob10} (by Tobin-Hochstadt et al.) goes further by offering safe interoperability
 between typed and untyped modules by generating appropriate runtime assertions based
@@
 as a homogeneous list. 
-Strickland et al. attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
+Strickland et al.\ attaches a \emph{starred pre-type} \lstinline|T *| to the right of the fixed arguments
 in a function type, where \emph{T} is some type, and we take an identical approach.

### 2012-10-22 dissertation/lit-review.tex
Example for infer

@@
 
-Strickland et al. invented a type system supporting variable-arity polymorphism~\cite{STF09}
+Strickland et al.\ invented a type system supporting variable-arity polymorphism~\cite{STF09}
 a version of which is included in the current implementation of Typed Racket.
@@
 Typed Racket 
-supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Piece et al.
+supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Piece et al.\ .
 Scala supports bounded quantification with F-bounded polymorphism~\cite{CCHOM89},
@@
 for describing ``information flow in polymorphic type systems with propagation-based type inference''~\cite{OZZ01}. 
-If the scope of my research permits it, I plan to investigate
-using colored types to solve the polymorphic higher-order-function limitation of Local Type Inference
-(see listing \ref{lst:hts}).
+Colored Local Type Inference is a candidate for
 
@@
 
+After some investigation, typing multimethods with Typed Clojure is assigned as future work
+(see section \label{sec:mmfuture}).
+
 \section{Higher Kinded Programming}
 
-\begin{verbatim}
-Many advanced type systems provide support for ..
-Forward reference to algo.monad experiment.
-\end{verbatim}
+Many advanced type systems provide support for \emph{higher kinds},
+which are ``another level up'' from types.
+For example, a type \lstinline|Number| is distinguished from a
+\emph{type constructor} \lstinline|Monad| (see section \ref{sec:monads}), which is a type level \emph{function}
+(similar to the difference between values and functions on the value level, but on the type level).
+
+F-$\omega$ is a typed $\lambda$-calculus with support for higher kinds,
+specifically type constructors (see Pierce~\cite{Pie02}).
+Haskell~\cite{Mar10} distinguishes between types and type constructors,
+the latter of which is an essential part of its monad library 
+(see section \ref{sec:monads} for a monad library ported to Typed Clojure
+that uses type constructors similarly).
 
@@
 occurrence typing, local type inference and variable-arity polymorphism.
-Similarly, Scala features F-bounded polymorphism, a class hierarchy
+Scala features F-bounded polymorphism, a class hierarchy
 that is compatible with Java, and colored local type inference.
+Typed Clojure takes inspiration from these, and similar, projects.

### 2012-10-22 dissertation/design-choices.tex
Example for infer
All of Rowan's corrections

@@
 The majority of the design and implementation of Typed Clojure is based on Typed Racket.
+This is a deliberate choice: the purpose of this work is to produce
+a practical tool, and there is already good evidence that Typed Racket
+is practical and makes good tradeoffs and design choices.
+Our differences are mostly due to the differences between Clojure and Racket.
 
@@
 \subsection{Interaction with null}
-% null is explicitly expressible in my type system
-% In Java, null is subtype of all reference types, any reference type must be tested to prevent NPE
-%
+
+It is a goal of Typed Clojure to statically prevent many Java \emph{NullPointerException}
+exceptions.
+\emph{null} is explicitly expressible in Typed Clojure as the type \listline|nil|.
+This enables more expressive types that specify where \emph{null} is and is not allowed.
+Section \ref{sec:javainterop} gives an example of how this could be useful 
+when using Java interoperability and gives details on the different ways Typed Clojure.
+
+%===> Occurrence typing~\cite{TF10} also plays a crucial role in Typed Clojure's
 
@@
 % F-bounded polymorphism supported
-
-\section{Multimethods}
-
-% INTRODUCTION TO CLOJURE MULTIMETHODS
-
-%A multimethod in Clojure differs from most other
-%abstractions of the same name. Clojure multimethods
-%dispatch on the result of an arbitrary function
-%rather than the types or structure of its arguments.
-%
-%\begin{lstlisting}[caption=Clojure Multimethods, label=lst:mm1]
-%(defmulti ToString class)
-%
-%(defmethod ToString java.lang.Integer
-%  [a]
-%  (str "Integer: " a))
-%
-%(defmethod ToString java.lang.Number
-%  [a]
-%  (str "Number: " a))
-%\end{lstlisting}
-%
-%Even with this extra flexibility, multimethods are often used to dispatch on the type of its arguments.
-%Listing \ref{lst:mm1} illustrates a common usage of multimethods: dispatching on the class of the first argument.
-%We use \lstinline|defmulti| to define a new multimethod.
-%The function \lstinline|class| is used as the dispatch function, a one-argument function returning the class of its argument
-%as an instance of \lstinline|java.lang.Class|.
-%
-%A multimethod invocation redirects control flow to exactly one of its \emph{methods} with unresolved ambiguities resulting in
-%a runtime error. The method is selected by first comparing the result of the dispatch function to
-%the dispatch values defined by each method. If one or more methods match, multimethods provide
-%the ability to prefer one method over another.
 
@@
 Similar to Typed Racket, type inference is based on Local Type Inference
-by Pierce and Turner.
+by Pierce and Turner~\cite{PT00}.
 
@@
 
-Typed Clojure includes support for F-bounds on type variables, as an extension
+Typed Clojure includes support for F-bounds on type variables\cite{CCHOM89}, as an extension
 to Local Type Inference.
@@
 
+% what about get? Core Type Rules
+
 $$

### 2012-10-22 dissertation/future-work.tex
All of Rowan's corrections

--- 2012-09-11 dissertation/future-work.tex
@@
 
+\section{Contracts and Blame}
+\label{sec:contractsblame}
+
 \section{Multimethods}
+\label{sec:mmfuture}
+
+\begin{verbatim}
+Multimethods in Clojure are an idiomatic and often used feature.
+It would be essential for an optional static type system for
+Clojure to support
+There are at least two hurdles to supporting multimethods.
+\end{verbatim}
+
 % Problem - typing isa?
@@
 \section{Records}
-% A composition of existing features: Datatypes and Heterogeneous Maps
+
+A Clojure record is a composite of datatypes and maps; it is a datatype
+with fields that can be treated like a map (records support map operations like \lstinline|get| (lookup),
+\lstinline|assoc| (add entries), and \lstinline|dissoc| (remove entries)).
+One interesting property from the perspective of static typing is that if
+a entry corresponding to a field of the underlying datatype is removed
+with \lstinline|dissoc|, then the resulting type is a plain map: it loses
+its record type.
+Correspondingly, a record keeps its type if any other entry is removed.
+
+Records are used frequently in Clojure code, so it is desirable
+to support them to some degree. Future work is planned to investigate solutions
+to satisfactorily capture their subtle semantics in a practical way.
 
@@
 
+\begin{verbatim}
+Other Clojure implementations
+\end{verbatim}
+
+\subsection{Java Generics}
+
+\begin{verbatim}
+Currently, Typed Clojure does not interact with Java generics,
+which provide
+instead 
+It may be desirable to 
+\end{verbatim}
+
 \subsection{Linter}

### 2012-10-22 dissertation/intro.tex
All of Rowan's corrections

@@
 to refine the type of the bindings \lstinline|a| and \lstinline|b|.
-Occurrence typing is a technique useful for capturing these kinds
+Occurrence typing\cite{TF10} is a technique useful for capturing these kinds
 of type invariants (occurrence typing is discussed in further detail in section \ref{sec:OccurrenceTyping}).
@@
 \subsection{Java Interoperability}
+\label{sec:javainterop}
 
-Typed Clojure currently provides safe interoperability
-\footnote{Compatibility with Generic Java is planned future work}.
+Typed Clojure supports integration with Java's type system during
+interoperability from Clojure. This offers many of the same advantages
+as using Java's type system from Java, for example it is a type
+error to pass arguments of the wrong type to methods.
+Typed Clojure also makes different decisions to Java's type system,
+particularly in the treatment of Java's \emph{null}.
 
-The key to safe Java interoperability is the treatment of Java's \emph{null}.
-\emph{null} is a subtype to all reference types
-represented in Clojure by \lstinline|nil|. Unlike Java's type system
-Clojure explicitly separates \emph{null} and reference types, allowing
-Typed Clojure to express
-\emph{nullable}
-\footnote{A type is \emph{nullable} if it may also be an instance of \emph{null},
-which is expressed in Typed Clojure by creating a union of the reference type and \lstinline|nil|.}
-positions.
+From the perspective of static types in Java, \emph{null} is included in all reference types.
+\emph{null} is represented in Clojure by the value \lstinline|nil|. Unlike Java's type system
+Typed Clojure explicitly separates \emph{null} and reference types giving Typed Clojure 
+more accurate types: we can directly express \emph{nullable} and \emph{not nullable} types.
+(A type is \emph{nullable} if it may also be \emph{null},
+which is expressed in Typed Clojure by creating a union of the type and \lstinline|nil|.
+A type is \emph{not nullable} if it doesn't include \emph{null}).
 
@@
 This example shows how Typed Clojure handles \emph{null} while creating and
-using an instance of \emph{java.io.File}\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
+using an instance of \emph{java.io.File}.
 
@@
 a valid approximation of the method as \emph{getParent} returns \emph{null} 
-``if the pathname does not name a parent directory''\footnote{See Javadoc: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
+``if the pathname does not name a parent directory''\footnote{See Javadoc for \emph{java.io.File}: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
 On the other hand, the instance method \emph{getName} always returns an
@@
 \emph{getName} to non-nil with \lstinline|typed.core/non-nil-return|
-\footnote{The second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
+(the second parameter to \lstinline|non-nil-return| specifies which arities to assume non-nil 
 returns, accepting either a set of parameter counts of the relevant arities, or \lstinline|:all|
-to override all arities of that method.}.
+to override all arities of that method).
+
+Such annotations are in a sense assumptions; should they turn out to be wrong, Typed Clojure
+will infer incorrect types. Typed Racket generates runtime contracts based on static types to ensure
+that type errors are caught at runtime, and the original source of the error is blamed~\cite{WF09}.
+A similar system for verifying that method annotations like \lstinline|non-nil-return|
+actually describe correct types is planned as future work
+(see section \ref{sec:contractsblame}).

### 2012-10-22 dissertation/lit-review.tex
All of Rowan's corrections

@@
 In this chapter we fit the design of Typed Clojure with literature from related fields.
-Forward-references are given to link the discussed literature with this project
+Forward-references are often given to link the discussed literature with this later parts
+of this dissertation
 where there is some relevance or influence.
@@
 
-% cite: Dynamic Typing in Polymorphic languages
+% cite: Dynamic Typing in Polymorphic languages (1991?)
 % - coming from the other direction
@@
 
-There is a history of designing static type systems for untyped languages.
+This section briefly reviews work on designing static type systems for untyped languages.
 
@@
 
+% Rowan
+% You've missed some of the steps. Most importantly the move to program analysis
+% techniques like Set Based Analysis (SBA) - which suffered from modularity issues
+% due to the lack of declared intended types.
+% Maybe quote Felleisen - "problem: Mr Spidey and friends infer brittle and large
+% types; errors remain difficult to explain and fix"
+
 Type systems for Scheme have since moved away from soft typing
 to to other approaches like gradual typing.
-Alas, the best reference for commentary on this transition 
+Alas, the only reasonably complete reference for commentary on this transition 
 appears to be slides from a talk by Felleisen~\cite{Fell09},
@@
 it suffers from incomprehensible
-errors that required PhD-level expertise to decipher.
+error reporting that required PhD-level expertise to decipher.
 
@@
 A function with multiple arguments is represented by chaining several single-argument function
-together, or by using lists, or using tuples. This style is characterised by direct syntactic function currying, where applying
+together, or by using lists, or using tuples. This first style is characterised by direct syntactic function currying, where applying
 a function to less than its maximum number of arguments results in another function
 that takes the remaining arguments.
+Using tuples instead requires that the same number of arguments be supplied,
+while using lists allows any number, but requires that all arguments have the same type.
 
@@
 ;=> (3.1 5.2)
-\end{lstlisting}\footnote{Line comments in Clojure begin with \lstinline|;| and comments to the end of the line. We use \lstinline|;=>| to mean \emph{evaluates to}.}
+\end{lstlisting}
+
+(Line comments in Clojure begin with \lstinline|;| and comments to the end of the line. We use \lstinline|;=>| to mean \emph{evaluates to}).
 
@@
 system with a \emph{proposition environment} that represents 
-the type refinements inferred down a particular path.
-The 2010 paper~\cite{TF10}
-reformulated occurrence typing, improving the original 2008 paper
-~\cite{TF08}
-after revealing that ``three years of practical experience has revealed
-serious shortcoming of our type system.''~\cite{TF10}
-The new implementation utilizes a simple proof system to solve
-propositional logic statements, in terms of type predicates and selectors.
+the information on the types of bindings down conditional branches.
+These propositions are then used to update the types associated
+with bindings in the \emph{type environment} down branches
+so binding occurrences are given different types 
+depending on the branches they appear in, and the conditionals
+that lead to that branch.
+(See section \ref{sec:occurenceimpl} for details on how Typed Clojure uses occurrence typing).
 
@@
 
-An exciting application of occurrence typing as yet unexplored is facilitating null-safe interop with Java.
+An exciting application of occurrence typing as yet unexplored is facilitating null-safe interoperability with Java.
 By declaring \lstinline|nil| (Clojure's value of Java's \lstinline|null|) to \emph{not} be a subtype of reference types,
-we can begin to statically disallow potential inconsistent usages of \lstinline|nil| as part of the type system.
+we can begin to statically disallow potentially inconsistent usages of \lstinline|nil| as part of the type system.
 
@@
 that \lstinline|a| is of type \lstinline|(U nil Object)|. This is equivalent to Java's
-\lstinline|Object| static type, as \lstinline|null| is a subtype of all reference types. By surrounding
+\lstinline|Object| static type, as \lstinline|null| is a member of all reference types. By surrounding
 the call \lstinline|(expects-non-nil a)| with \lstinline|(when a ...)|, we guarantee that
@@
 be \lstinline|nil| the body of the \lstinline|when|, refining \lstinline|a|'s type to \lstinline|Object|
-from \lstinline|(Union nil Object)|.
+from \lstinline|(U nil Object)|.
 
-Occurrence typing is a relatively simple, time worn technique used successfully 
+Occurrence typing is a relatively simple technique used successfully 
 in Typed Racket. Clojure is similar enough to Racket for occurrence typing to work
-without issues, and has good potential to enable null-safe Java interop.
+without issues, and has good potential to help programmers avoid using
+\lstinline|nil| incorrectly
 
@@
 For example, a type \lstinline|Number| is distinguished from a
-\emph{type constructor} \lstinline|Monad| (see section \ref{sec:monads}), which is a type level \emph{function}
+\emph{type constructor} \lstinline|Monad| (see section \ref{sec:monads}), which is a type level function
 (similar to the difference between values and functions on the value level, but on the type level).

### 2012-10-24 dissertation/design-choices.tex
Filling in the last bits. Mostly design and future work

@@
 
+This chapter describes the design choices made for Typed Clojure.
 Typed Clojure is designed to be of practical use to Clojure programmers.
-Many of the design goals are similar to Tobin-Hochstadt's 
-for Typed Racket~\cite{Tob10}. Some of the main differences are described in this chapter.
+To this end, many of the design choices are borrowed from other existing projects.
+Some of the main design choices that specifically concern Typed Clojure are given more
+more detailed descriptions.
 
@@
 
-The majority of the design and implementation of Typed Clojure is based on Typed Racket.
+The majority of the design and implementation of Typed Clojure is based on Typed Racket~\cite{Tob10}.
 This is a deliberate choice: the purpose of this work is to produce
@@
 Our differences are mostly due to the differences between Clojure and Racket.
+This section details some of these differences.
 
 \subsection{Occurrence Typing}
-% interesting paths
-% - count path element
-% - class path element
-% - keyword path element
+
+Occurrence typing, described by Tobin-Hochstadt and Felleisen~\cite{TF10},
+plays a major role in capturing many common Clojure idioms.
+One important idiom it helps Typed Clojure to understand 
+is the \emph{sequence abstraction}, which
+is made up of the \lstinline|clojure.core| functions \lstinline|seq|, \lstinline|first|, and \lstinline|rest|.
+
+\lstinline|seq| takes an argument that is either a sequence or \lstinline|nil|,
+and returns a non-empty sequence if passed a non-empty sequence, or returns
+\lstinline|nil| if passed an empty sequence or \lstinline|nil|.
+It is a common idiom to test \lstinline|(seq coll)| in a conditional
+to determine whether the sequence \lstinline|coll| is empty:
+if \lstinline|(seq coll)| is a true value, then \lstinline|coll|
+is non-empty, otherwise it should be considered empty.
+Occurrence typing allows us to capture this idiom as shown in
+listing \ref{lst:seqtype}.
+
+\begin{lstlisting}[caption=Type assiciated with clojure.core/seq, label=lst:seqtype]
+(All [x]
+  [(U nil (Seqable x)) -> (U nil (Seqable x))
+   :filters {:then (is (CountRange 1) 0)
+             :else (| (is nil 0)
+                      (is (ExactCount 0) 0))}])
+\end{lstlisting}
+
+This function type takes an argument of type \lstinline|(U nil (Seqable x))| (the syntax \lstinline[mathescape]{(U $\overrightarrow{t}$)}
+constructs a union of types \lstinline|t|), and returns a type \lstinline|(U nil (Seqable x))|.
+Occurrence typing allows us to attach propositions to the return
+type of a function. Roughly, these say if \lstinline|(seq coll)|
+returns a true value, then \lstinline|coll| is of type
+\lstinline|(CountRange 1)| (a type representing a sequence of length 1 or greater),
+otherwise \lstinline|coll| is \emph{either} of type \lstinline|nil| or of type \lstinline|(ExactCount 0)| (a type
+representing an empty sequence).
+(See section \ref{sec:filterneg} for a more detailed explanation of the filter syntax).
+
+Correspondingly, \lstinline|first| which gets the first member of a sequence.
+can be given a more accurate (and surprisingly subtle) type.
+\lstinline|first| is overloaded in a similar fashion to \lstinline|seq|.
+\lstinline|first| takes an argument that is either a sequence or \lstinline|nil|.
+If it is passed \lstinline|nil| or an empty sequence, it returns \lstinline|nil|. Otherwise, if it is passed
+non-empty sequence, it returns the first member of the sequence.
+Listing \ref{lst:firsttype} shows the type of \lstinline|first|.
+
+\begin{lstlisting}[caption=Type assiciated with clojure.core/first, label=lst:firsttype]
+(All [x]
+  (Fn [(U nil (I (Seqable x) (ExactCount 0))) -> nil]
+      [(I (Seqable x) (CountRange 1)) -> x]
+      [(U nil (Seqable x)) -> (U nil x)])))
+\end{lstlisting}
+
+The function type has three arities, the last of which can be considered a ``default'' arity.
+The first two arities correspond exactly to the two cases described for when 
+\lstinline|first| is passed an empty sequence or \lstinline|nil|, and
+when \lstinline|first| is passed a non-empty sequence
+(the syntax \lstinline[mathescape]{(I $\overrightarrow{t}$)}
+constructs an intersection of types \lstinline|t|).
+
+These two functions interact via occurrence typing.
+For example, listing \ref{lst:seqocc} tests
+\lstinline|(seq coll)| which means occurrences of \lstinline|coll|
+in the ``then'' branch are non-empty and those in the
+``else'' branch are empty.
+We can safely call \lstinline|(.method (first coll))| in the ``then''
+branch (assuming elements of \lstinline|coll| have a no-argument method called \lstinline|method|)
+always avoiding a \emph{NullPointerException}.
+
+\begin{lstlisting}[caption=Example of using the sequence abstraction with occurrence typing, label=lst:seqocc]
+...
+(let [coll (possibly-empty-coll)
+  (if (seq coll)
+    (.method (first coll))
+    nil))
+...
+\end{lstlisting}
 
 \subsection{Variable-arity Polymorphism}
+
+%Tobin-Hochstadt
+
 % Problematic areas
@@
 Dialects of Clojure provide fast and unrestricted access to their host platform. 
-This is a intentional source of incompatibility between dialects.
-For example, Clojure is hosted on the Java Virtual Machine (JVM), ClojureCLR on the Common Language Runtime (CLR),
+This is a intentional source of incompatibility between Clojure implementations.
+For example, Clojure is hosted on the Java Virtual Machine (JVM), ClojureCLR on the Common Language Runtime,
 and ClojureScript on Javascript VMs. These are very different platforms, and it is unlikely
@@
 
-Typed Clojure attempts to improve some shortcomings of Java's type system.
+Typed Clojure attempts to improve some shortcomings of Java's static type system.
 
@@
 Scala provides the Option type for safe \lstinline|null| elimination, forcing pattern matching
-where \lstinline|null| might occur, and
-
-\begin{lstlisting}[caption=null elimination in Scala]
-TODO
-\end{lstlisting}
-
-Clojure
+where \lstinline|null| might occur. This approach is not appropriate for Typed Clojure
+as pattern matching is not idiomatic in Clojure.
 
 \subsection{Primitive Arrays}
+
 % Primitive arrays are covariant, which is well-known to be statically unsound.
@@
 \subsection{Interaction with null}
+\label{sec:interactionnull}
 
@@
 exceptions.
-\emph{null} is explicitly expressible in Typed Clojure as the type \listline|nil|.
-This enables more expressive types that specify where \emph{null} is and is not allowed.
+\emph{null} is explicitly expressible in Typed Clojure as the singleton type \lstinline|nil|,
+representing the \emph{value} \lstinline|nil|.
+This enables more expressive types that specify where \emph{nil} is and is not allowed.
 Section \ref{sec:javainterop} gives an example of how this could be useful 
@@
 
-%===> Occurrence typing~\cite{TF10} also plays a crucial role in Typed Clojure's
+Occurrence typing~\cite{TF10} also plays a crucial role in Typed Clojure's
+treatment of \lstinline|nil|.
+It allows us to track whether particular bindings are possibly \lstinline|nil|
+down conditional branches, and makes it practical to throw type errors 
+where \lstinline|nil| is not allowed (see section \label{sec:OccurrenceTyping} for examples).
+This is not practical in Java's type system because \emph{null} is not directly
+expressible, which makes it impractical to error when a \emph{NullPointerException}
+exception is possible: according to the type system, almost every operation
+could involve \emph{null}!
+Eliminating \emph{NullPointerException} exceptions is entirely the programmers
+responsibility in Java.
+Typed Clojure helps the programmer by statically proving certain misuses of \lstinline|nil| are impossible.
 
-\subsection{Generic Java}
-% Existential types are needed to fully express Generic Java types in Typed Clojure
-% F-bounded polymorphism supported
+The design choice of potentially misusing \lstinline|nil| results in a type error
+needs further work to be completely satisfying.
+Certain cases of type inference should be improved to avoid redundant code
+that ``pleases the type checker''.
+For example, we currently do not know how to infer \lstinline|(filter identity coll)|
+as returning a sequence of non-nil elements, for any sequence \lstinline|coll|.
+(see section \ref{sec:filterneg}).
+It is also possible for code interacting with Java to introduce unexpected \emph{null}
+values. Currently, we let the programmer define a set of unenforced \emph{assumptions}
+where \emph{null} is allowed in Java fields and methods (see section \label{sec:javainterop}).
+Adding runtime assertions that enforce these assumptions would allow more
+accurate errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
+interfacing with untyped Racket code.
 
-\section{Clojure Class Hierarchy}
+\section{Clojure Hierarchy}
+
+The Clojure hierarchy is a set of Java classes and interfaces, mostly in the
+\emph{clojure.lang} package, that define Clojure's core 
+interfaces and data structures.
+Typed Clojure parameterises some of these classes in a way that resembles
+Java or Scala generics.
+
+Clojure's hierarchy was accommodating to this process.
+For example, \lstinline{clojure.lang.Seqable} is the main interface behind
+Clojure's sequence abstraction. Sequences are immutable collections,
+so it was natural to parameterise \lstinline{Seqable} with one covariant
+argument (covariant means \lstinline|(Seqable Integer)| is
+a subtype of \lstinline|(Seqable Number)| because \lstinline|Integer|
+is a subtype of \lstinline|Number|).
+All ``seqable'' classes in the Clojure hierarchy implement \lstinline|Seqable|,
+so updating them to use the parameterised \lstinline|Seqable| was a matter of
+searching for classes that had \lstinline|Seqable| as an ancestor and replacing
+the unparameterised \lstinline|Seqable| with the corresponding parameterised 
+version.
+
+Defining new ``parameterised'' versions of existing classes and interfaces do 
+not affect anything outside Typed Clojure.
+They are only for use by Typed Clojure.
+
 \section{Protocols and Datatypes}
-\section{Optional Type Checking}
+
+Protocols and datatypes are two of Clojure's main means of abstraction.
+
+%section \ref{sec:datatypes}
+
 \section{Local Type Inference}
+
+Typed Clojure takes a
 
@@
 rules, as is standard in the programming language community (see, eg., Pierce~\cite{Pie02}).
-
-% Operational semantics
-
-\subsection{Operational Semantics}
-
-%$$
-%\begin{tdisplay}{Test1}
-%\begin{altgrammar}
-%  \s{},\t{} &::= & \int \alt \bool \alt {\proctype {\s{}} {\t{}}} 
-%  &\mbox{Types}\\
-%  \M{}, \N{}, \P{}, \dots &::=& \x{} \alt \abs{\x{}}{\M{}} \alt
-%  \comb{\M1}{\M2}  &\mbox{Raw Terms} \\
-%\end{altgrammar}
-%\end{tdisplay} 
-%$$
-
-$$
-\begin{tdisplay}{Syntax of Terms}
-\begin{altgrammar}
-  \exp{} &::=& (\c{} \overrightarrow{\exp{}})         % function application, multiple arguments
-             \alt \{ \overrightarrow{\exp{}\ \exp{}} \} % map literal
-             &\mbox{Expressions} \\ 
-  \v{} &::=& \k{} \alt \nil{}
-              \alt \{ \overrightarrow{\v{}\ \v{}} \}   % map literal
-              &\mbox{Values} \\
-  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
-              &\mbox{Constants}
-\end{altgrammar}
-\end{tdisplay}
-$$
-
-$$
-\begin{tdisplay}{Evaluation Contexts}
-  \begin{altgrammar}
-    \E{} &::=& [ ] % application rules
-              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
-              % map rules
-              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
-              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
-              &\mbox{Evaluation Contexts}
-  \end{altgrammar}
-\end{tdisplay}
-$$
-
-% \mapsto for updating maps
-$$
-\begin{tdisplay}{Operational Semantics}
-\begin{array}{c}
-{\inferrule[E-Assoc]
-  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
-    {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
-    {{(\assoc{}\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
-\\
-{\inferrule[E-Dissoc]
-  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
-    {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
-  {(dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
-\\
-{\inferrule[E-GetMapExist]
-  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
-    {v_2\ v_3\ in\ v_1} }
-%    {v_1\ has\ entry\ with\ key\ v_2} \\
-%    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1}
-  { {(get\ v_1\ v_2) \hookrightarrow v_3} }} \\
-\\
-{\inferrule[E-GetMapNotExist]
-  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
-    {v_1\ has\ no\ entry\ with\ key\ v_2} \\
-    {v_3 = \nil{}} }
-  { {(get\ v_1\ v_2) \hookrightarrow v_3} }}
-\end{array} 
-\end{tdisplay} 
-$$
-
-$$
-\begin{tdisplay}{Type Syntax}
-\begin{altgrammar}
-  \T{} &::=& \Top \alt \Bottom \alt \Nil \alt \k{} \alt \{ \overrightarrow{\k{}\ \T{}} \}
-             \alt {\IPersistentMap {\T{}} {\T{}}} \alt ( \vee\ \overrightarrow{\T{}} )
-             \alt ( \wedge\ \overrightarrow{\T{}} )
-\end{altgrammar}
-\end{tdisplay}
-$$
-
-% what about get? Core Type Rules
-
-$$
-\begin{tdisplay}{Core Type Rules}
-\begin{array}{c}
-{\inferrule[T-AssocHMap]
-  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
-    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\k{}}}} } \\
-    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
-    {\vdash {\issubtype {{\T{}}_{m}} {\TopHMap} }} }
-  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} 
-        {(update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})}}}}} \\
-\\
-{\inferrule[T-AssocPromote]
-  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
-    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}}} } \\
-    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
-    {\vdash {\issubtype {{\T{}}_{m}} {\TopIPersistentMap}}} \\
-    {\T{} = (promote\_hmap\ {\T{}}_{m}\ {\T{}}_{k}\ {\T{}}_{v})} }
-  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
-\end{array}
-\end{tdisplay}
-$$
-
-$$
-\begin{tdisplay}{Type Metafunctions}
-\begingroup
-\fontsize{10pt}{12pt}
-\begin{altgrammar}
-  (update\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\k{}}_1\ {\T{}}_1) = 
-                 \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}}\ {\k{}}_1\ {\T{}}_1 \}\ \\
-  (update\_hmap\ (\wedge\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
-                               (\wedge\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})}) \\
-  (update\_hmap\ (\vee\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
-                               (\vee\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
-  \\ \\
-  (promote\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\T{}}_{kn}\ {\T{}}_{vn})
-                                    = {\IPersistentMap {(\vee\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
-                                                       {(\vee\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
-  (promote\_hmap\ {\IPersistentMap {{\T{}}_{k}} {{\T{}}_{v}}} \ {\T{}}_{kn}\ {\T{}}_{vn})
-                                    = {\IPersistentMap {(\vee\ {\T{}}_{k}\ {\T{}}_{kn})}
-                                                       {(\vee\ {\T{}}_{v}\ {\T{}}_{vn})}}
-\end{altgrammar}
-\endgroup
-\end{tdisplay}
-$$
-
-$$
-\begin{tdisplay}{Subtyping}
-\begin{array}{c}
-\inferrule[S-Refl] { } {\vdash {\issubtype {\T{}} {\T{}}}} \\
-\inferrule[S-Any] { } {\vdash {\issubtype {\T{}} {\Top}}}
-\inferrule[S-UnionSuper] 
-  {\ } 
-  {\vdash {\issubtype {\T{}} {\Top}}}
-\end{array}
-\end{tdisplay}
-$$
 

### 2012-10-24 dissertation/experiments.tex
Filling in the last bits. Mostly design and future work

@@
 \section{Java Interoperability}
+\label{sec:experimentinterop}
 
@@
 
-Here is the final successfully type checked code.
-\begin{verbatim}Is this a reasonable result?\end{verbatim}
+The final successfully type checked code is presented in listing \ref{lst:callmethodfinal}.
 
-\begin{lstlisting}
+\begin{lstlisting}[caption=Type Annotated code for call-method, label=lst:callmethodfinal]
 (non-nil-return java.lang.Class/getDeclaredMethod :all)
@@
 
-\section{Red-Black Tree}
+\section{Red-black trees}
+
+This experiment involved porting an implementation of red-black trees
+used as an experiment for SML CIDRE by Davies~\cite{Dav05}.
+SML CIDRE is a sort-checker for Standard ML that supports refinement
+types.
+The red-black tree implementation was ported to work with SML CIDRE
+to statically check red-black tree invariants.
+It is a particularly good experiment for SML CIDRE because it generates an unusually large
+amount of intersection types.
+
+Intersection types in SML CIDRE are heavily optimised via memoisation.
+Because of this, SML CIDRE is able to check this experiment with little trouble.
+The current Typed Clojure prototype is not as successful, and appears to hang
+during type checking.
+The reasons are not entirely clear, but it is likely the number of intersection types
+are too large for the current prototype to handle, which has no optimisations
+around intersection types.
+
+This experiment was intended to stress test Typed Clojure to an extreme, and
+it is unsurprising that Typed Clojure was not up to the task.
+One interesting point was noticed when porting it.
+The SML CIDRE version of the red-black tree experiment
+uses SML CIDRE datasorts to represent the red-black tree invariants.
+The Typed Clojure version entirely uses plain hash-maps, a common Clojure idiom.
+It was noticed that the heterogeneous map types that Typed Clojure provides would be sufficient to
+represent the red-black tree invariants.
 
@@
 
+Listing \ref{lst:maybet} shows the full code for type checking the definition
+of the maybe monad transformer.
+Typing this example was a matter of assigning types to anonymous functions
+(similar to standard monad definitions) and manually instantiating 
+polymorphic types with \lstinline|inst|
+that parameterise over higher kinds, which cannot be inferred by Typed Clojure.
+
+\begin{lstlisting}[caption={Maybe monad transformer in Typed Clojure}, label=lst:maybet]
+(ann maybe-t
+     (All [[m :kind (TFn [[x :variance :covariant]] Any)]]
+       (Fn 
+         [(AnyMonad m) -> (MonadPlusZero (TFn [[y :variance :covariant]]
+                                           (m (U nil y))))]
+         [(AnyMonad m) nil -> (MonadPlusZero (TFn [[y :variance :covariant]]
+                                               (m (U nil y))))]
+         [(AnyMonad m) nil (U ':m-plus-default ':m-plus-from-base)
+          -> (MonadPlusZero (TFn [[y :variance :covariant]]
+                              (m (U nil y))))])))
+(defn maybe-t
+  "Monad transformer that transforms a monad m into a monad in which
+   the base values can be invalid (represented by nothing, which defaults
+   to nil). The third argument chooses if m-zero and m-plus are inherited
+   from the base monad (use :m-plus-from-base) or adopt maybe-like
+   behaviour (use :m-plus-from-transformer). The default is :m-plus-from-base
+   if the base monad m has a definition for m-plus, and
+   :m-plus-from-transformer otherwise."
+  ([m] ((inst maybe-t m) m nil :m-plus-default))
+  ([m nothing] ((inst maybe-t m) m nothing :m-plus-default))
+  ([m nothing which-m-plus]
+   (monad-transformer m which-m-plus
+     [m-result (with-monad m m-result)
+      m-bind   (with-monad m
+                 (ann-form
+                   (fn m-bind-maybe-t [mv f]
+                     (m-bind
+                       mv
+                       (ann-form
+                         (fn [x]
+                           (if (nil? x)
+                             (m-result nothing) 
+                             (f x)))
+                         [(U nil a) -> (m (U nil b))])))
+                   (All [a b]
+                     [(m (U nil a)) [a -> (m (U nil b))] -> (m (U nil b))])))
+      m-zero   (with-monad m (m-result nothing))
+      m-plus   (with-monad m
+                 (ann-form
+                   (fn m-plus-maybe-t [& mvs]
+                     (if (empty? mvs)
+                       (m-result nothing)
+                       (m-bind (first mvs)
+                               (ann-form
+                                 (fn [v]
+                                   (if (= v nothing)
+                                     (apply m-plus-maybe-t (rest mvs))
+                                     (m-result v)))
+                                 [(U x nil) -> (m (U x nil))]))))
+                   (All [x] [(m (U x nil)) * -> (m (U x nil))])))
+      ])))
+
+\end{lstlisting}
+
 \section{Conduit}

### 2012-10-24 dissertation/future-work.tex
Filling in the last bits. Mostly design and future work

--- 2012-09-11 dissertation/future-work.tex
@@
 
+\section{Contracts and Blame}
+\label{sec:contractsblame}
+
+A key part of Typed Racket~\cite{Tob10} is its contract and blame system.
+They enable safe interoperability with untyped Racket by generating runtime
+contracts in key places based on static types. Typed Racket also 
+includes a sophisticated blame system based on Wadler\cite{WF09}
+to provide more accurate error reporting that ensures typed modules
+``can't be blamed''.
+
+There are two kinds of interoperability in Typed Clojure
+that could utilize these features. First, the set of ``assumptions''
+given by the programmer about Java interoperability like those introduced by \lstinline|non-nil-return| 
+and \lstinline|nilable-param| (see section \ref{sec:experimentinterop})
+could be checked by adding runtime contracts.
+Secondly, when typed namespaces import untyped functions, the static type
+assigned to the untyped function could be enforced by wrapping the
+function in a runtime contract. Typed Racket uses this second approach
+when importing untyped Racket functions, and it results in better error messages.
+
 \section{Multimethods}
-% Problem - typing isa?
-% - probably impossible to type higher order usages of isa?.
-% Dispatch function must infer latent filters for multi-arity functions. (Typed Racket only infers single arg I think)
-% Statically proving at least one relevant dispatch method is present.
-% Warn potential method conflicts.
-% Suppress warnings, similar to prefer-method syntax
+\label{sec:mmfuture}
+
+Multimethods in Clojure are an idiomatic and often used feature.
+It would be essential for an optional static type system for Clojure to support
+The major hurdle for type checking multimethod definitions is the
+interaction between the multimethod dispatch mechanism and occurrence typing.
+
+\lstinline|clojure.core/isa?| is the core of multimethod dispatch.
+It is a function that takes two arguments and is \lstinline|true| if
+the first argument is a member of the second argument, otherwise \lstinline|false|.
+It gets quite complicated quickly even with common usages of \lstinline|isa?|.
+Simple calls involving only \emph{Class} objects return true if the left class
+includes the right class in a Java type relationship, eg. \lstinline|(isa? Integer Number)| is \lstinline|true|.
+For example, \lstinline|(isa? [Integer Double] [Number Number])| is \lstinline|true|
+because the left and right arguments are vectors of classes of the same length
+that are subtypes by \lstinline|isa?| pairwise.
+
+It is also unclear whether it is feasible to perform comprehensiveness and
+ambiguity prediction tests on multimethod usages. I predict that they will be
+very difficult to pull off satisfactorily because of the ``openess'' of multimethods,
+and we will fall back on the runtime errors that multimethods already throw.
 
 \section{Records}
-% A composition of existing features: Datatypes and Heterogeneous Maps
 
-\subsection{Other Clojure Dialects}
-% unify ASTs for easier porting
+A Clojure record is a composite of datatypes and maps; it is a datatype
+with fields that can be treated like a map: records support map operations like \lstinline|get| (lookup),
+\lstinline|assoc| (add entries), and \lstinline|dissoc| (remove entries).
+One interesting property from the perspective of static typing is that if
+a entry corresponding to a field of the underlying datatype is removed
+with \lstinline|dissoc|, then the resulting type is a plain map: it loses
+its record type.
+Correspondingly, a record keeps its type if any other entry is removed.
+
+Records are used frequently in Clojure code, so it is desirable
+to support them to some degree. Future work is planned to investigate solutions
+to satisfactorily capture their subtle semantics in a practical way.
+
+\subsection{Porting to other Clojure implementations}
+
+It would be desirable to port Typed Clojure to other Clojure implementations.
+Each implementation would bring its own set of challenges in
+
+\subsection{Extensions to heterogeneous map types}
+
+\subsection{Length information and intersection types}
+
+%\subsection{Java Generics}
 

### 2012-10-24 dissertation/implementation.tex
All of Rowan's corrections
Filling in the last bits. Mostly design and future work

@@
 \subsection{Occurrence typing}
+\label{sec:occurenceimpl}
 
@@
     (written \lstinline[mathescape]{(| $\overrightarrow{p}$)}) for propositions \lstinline|p|.
-
 \end{itemize}
@@
 
-A substitution maps type variables to a type.
+A substitution maps type variables to types.
 It is valid to replace all occurrences of the type variables named in the substitution 
@@
 
-\begin{verbatim}
-  Say more about the changes and why so much work was involved. Or at least
-  the remaining changes were more tedious than anything.
-\end{verbatim}
+Porting Typed Racket's variable-arity polymorphism implementation was
+tedious because some of the relevant internal functions interact
+in strange ways with the rest of Typed Racket. My reaction was that Typed Racket
+was initially designed without variable-arity polymorphism and was added
+without major changes to other components. Typed Clojure was
+developed with the same design so full variable-arity polymorphism
+implementation could be ported without change.
 
@@
 a common interface to abstract syntax trees for each Clojure implementation.
-
-\section{Composite Propositions}
 
@@
 
-\section{Non-hygienic Macro Expansion and Filters}
+Further work in this area is needed when filtering on a non-anonymous function.
+For example, it is not clear how to infer the common idiom \lstinline|(filter identity coll)|
+as returning a sequence of non-nil elements, for any sequence \lstinline|coll|.
+Inferring new propositions for already existing functions like \lstinline|identity|
+does not fit with Tobin-Hochstadt and Felleisen's calculus~\cite{TF10},
+confirmed by Tobin-Hochstadt~\cite{Tob12} as future work in this area.
+
+%\section{Non-hygienic Macro Expansion and Filters}
 

### 2012-10-24 dissertation/intro.tex
Filling in the last bits. Mostly design and future work

@@
 dynamically typed languages with static type systems. 
-This is idea not new, but recent attempts are noteworthy for 
-their broad success in matching many of the advantages of statically typed
-languages, notably due to the use of \emph{bidirectional checking}.
+This is idea not new~\cite{Fag91,CF91,AM91}, but recent attempts~\cite{Tob10}
+are noteworthy for their broad success in matching many of the advantages of statically typed
+languages, notably due to the use of bidirectional checking~\cite{PT00}.
 Instead of always attempting to infer types, this algorithm relies on programmer annotations
@@
 The Clojure programming language is a dynamically typed dialect of Lisp invented
-by Hickey, designed to run on popular platforms\footnote{http://clojure.org/}.
+by Hickey \footnote{\url{http://clojure.org/}}, designed to run on popular platforms.
 It emphasises functional programming with immutable data structures
@@
 of a Clojure programmer. In an apparently heroic effort, he managed to 
-implement a Clojure library for conduits, an advanced form of ``pipes'',
-using arrows, a generalisation of monads.
-\footnote{CONDUIT REFERENCE TODO}
+implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
+an advanced form of ``pipes'', using arrows, a generalisation of monads.
 Conduits also significantly surpass monads in complexity and are usually reserved
@@
 after it was selected by the Clojure community as a Clojure Google Summer of Code
-project for 2012.
-\footnote{CITE GSOC, or FOOTNOTE}
+project for 2012~\footnote{\url{http://www.google-melange.com/gsoc/project/google/gsoc2012/ambrosebs/10001}}.
 I have also been invited to speak on this project at the Clojure Conj
@@
 
-% refinements check invariants of programs
-% while ordinary static types check whether programs
-% are basically meaningful - Rowan + Frank P
 % Page 1 Rowan's PhD
@@
 This distinction has a long history, originating in work in the $\lambda$-calculus 
-by Church \footnote{CHURCH FROM ROWANS DISSERTATION} and Curry \footnote{CURRY FROM ROWANS DISSERTATION},
+by Church~\cite{Chu40} and Curry~\cite{Cur34},
 leading to intrinsic types and extrinsic types also being called Church types and
@@
 and common in the literature on static types (for example, Pierce~\cite{Pie02}).
+
+\section{Contributions}
+
+In this dissertation we develop an optional type system for Clojure.
+We focus on the features needed to make a practical type system that
+existing Clojure programmers can use.
+We also developed a prototype based on Typed Racket~\cite{Tob10}.
+
+The contributions of the work described in this dissertation are as follows:
+
+\begin{itemize}
+  \item We describe a novel use of occurrence typing~\cite{TF10} for eliminating
+        occurrences of Java's \emph{null}. We make \emph{null} directly expressible
+        in Typed Clojure's type system as \lstinline|nil|. This makes it reasonable for it to be a
+        static type error if \emph{null} is potentially misused, an improvement over Java's static type system
+        where \emph{null} cannot be expressed.
+
+  \item We show how a combination of intersection types and occurrence typing enables us to accurately type check
+        the most common usages of Clojure's sequence abstraction. This was particularly subtle, as 
+        many base operations in the sequence abstraction take and/or return \lstinline|nil|, and
+        each case could imply different information about a sequence.
+
+  \item We develop a prototype type checker for Clojure running on the Java Virtual Machine
+        which is able to type check many common Clojure idioms.
+        We show how this type checker can be used to aid Clojure programmers use sophisticated
+        programming styles by porting most of an existing library for monadic programming.
+
+  \item This prototype is extended to support several features that are not present in Typed Racket.
+        We support filtering sequences on anonymous functions that convey only negative type information
+        (eg. non-nil), capturing a common Clojure idiom.
+
+  \item We capture the Clojure idiom of using maps with known keyword fields by using heterogeneous map types.
+        These types are sufficient for accurately associating and dissociating known keyword keys,
+        and accurately looking up values associated with keyword keys in heterogeneous map types that have
+        that keyword key present.
+        This is sufficient for type checking many common idioms involving maps.
+
+  \item We identify major future issues in typing Clojure code. Multimethod definitions are problematic
+        to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
+        Records have subtle semantics for dissociating keys, like dissociating a base key of a record returns
+        a plain map.
+%        Heterogeneous map types 
+\end{itemize}
+
+Together,
+
+These contributions support my thesis by using the prototype to check practical code
+and
 
@@
 \subsection{Datatypes and Protocols}
+\label{sec:datatypes}
 
@@
 definitions. 
-Protocols are similar to Java interfaces\footnote{JAVA INTERFACE REFERENCE} and datatypes to classes implementing interfaces.
+Protocols are similar to Java interfaces~\cite{InterfaceTutorial} and datatypes are similar to classes implementing interfaces.
 We can annotate datatype and protocol definitions similarly.

### 2012-10-24 dissertation/lit-review.tex
Filling in the last bits. Mostly design and future work

@@
 
-\section{Static Type Systems}
+\section{Dynamic Typing in Typed Languages}
 
-% Haskell
-% OCaml
-% Forsythe
-% Java C#
-% Scala
+This section 
+\cite{ACPR94}
 
-% cite: Dynamic Typing in Polymorphic languages (1991?)
+% cite: Dynamic Typing in Polymorphic languages (1994)
 % - coming from the other direction
@@
 % A more recent reference in the same area is "Typed Open Programming" a PhD dissertation from 2007 - at least have a look at 1.3.2 for a good summary of work on dynamics in statically typed languages since 1991 - probably it's good to cite this as a good recent survey of the area.
-\section{Static Types for Untyped Languages}
+\section{Static Typing for Untyped Languages}
 
@@
 
-This section summarises research in typing OCaml-style records
-which relate to Clojure records and heterogeneous hash maps.
+OCaml-style extensible record types has been the subject of extensive research~\cite{Wan89,CM91,HP91}.
+Typed Clojure's heterogeneous map types show some resemblance to extensible
+record types, but a survey of the literature did not reveal research flexible
+enough to capture common usages of maps in Clojure.
+This most likely implies that designing such a system for Typed Clojure would be hard.
+For this reason, heterogeneous map types are kept as simple as possible while still being useful.
 
@@
 
-% Hiyashi Singleton types
+% Hayashi Singleton types
 % Intersection types - CDV81
@@
 or \lstinline|b|.
-For example, \lstinline|(U Number Boolean)| cannot be used in positions expecting \lstinline|Number|.
-% TODO Intersection example
+For example, \lstinline|(U Number Integer)| cannot be used in positions expecting \lstinline|Integer|,
+but \lstinline|(I Number Integer)| can be used in positions expecting \lstinline|Integer|.
+
+Intersection, union, and singleton types fit best as extrinsic types.
+Hayashi~\cite{Hay91} describes two type theories, ATT which includes
+intersection, union, and singleton types, and ATTT which further extends ATT
+to include refinement types, which can be classified as extrinsic types.
+Typed Clojure can also be considered an extrinsic type system which supports
+union, intersection, and singleton types, but in less sophisticated forms
+than ATT and ATTT.
+In particular, singleton types in ATTT a much more advanced than Typed Clojure,
+which offers only singleton types for values like \lstinline|:a| and \lstinline|nil|
+and, unlike ATTT, no way to parameterise over singletons.
 
 Several interesting projects have used intersection or union types.
+Intersection types were originally introduced by Coppo, Dezani-Ciancaglini, and Venneri~\cite{CDV81}.
+The use union types to type check dynamically typed languages with soft typing systems
+dates back to Cartwright and Fagan~\cite{CF91}.
 
-Forsythe, a modern ALGOL dialect by Reynolds~\cite{Rey96} was the first wide-spread 
+Forsythe, a modern ALGOL dialect by Reynolds~\cite{Rey81,Rey96} was the first wide-spread 
 programming language to use intersection types.
@@
 (described by Freeman and Pfenning~\cite{FP91}).
+Refinement types are reminiscent of Typed Clojure, in that Typed Clojure
+adds a level of types on top of Clojure.
 Intersection types are critical here to allow more than one property or invariant
@@
 SML CIDRE is a refinement type checker for Standard ML by Davies~\cite{Dav05}.
-\begin{verbatim}
-Refinement types are similar to this project ... how?
-\end{verbatim}
 
@@
 that provide a kind of function overloading.
-Typed Clojure takes a similar approach
-\begin{verbatim}
-described in chapter
-\end{verbatim}
-Typed Racket also uses disjoint union types,
+Typed Clojure takes a similar approach for its representation of functions.
+Typed Racket uses (disjoint) union types.
 
@@
 to prevent erroneous usage of \lstinline|null|.
-\begin{verbatim}
-eg.
-\end{verbatim}
 
@@
 Typed Clojure takes inspiration from these, and similar, projects.
+
+SML CIDRE \cite{Dav05} was also a secondary influence - particularly it's use of bidirectional 
+checking of extrinsic types including subtyping, intersection types, and it's aim to capture program invariants.  
+However, SML CIDRE starts from the intrinsically typed language SML which already has a rich static 
+type system, which leads to many different considerations compared to starting from a dynamically typed language.  
+Much of this influence was via a student-supervisor relationship, making it particularly hard to pin down the specifics.

### 2012-10-24 dissertation/paper.tex
All of Rowan's corrections
Filling in the last bits. Mostly design and future work

@@
 \usepackage{clj-grammar}
+\usepackage{url}
 \usepackage[style=alphabetic]{biblatex}
@@
 programming languages that perform type checking at compile-time, and those that do not.
-In particular, the early wide-spread programming languages ALGOL and LISP
+In particular, the early wide-spread programming languages ALGOL and Lisp
 could be divided this way: ALGOL performs compile-time (or static) type checking,
-and LISP does not.
+and Lisp does not, and most recent languages can be similarly classified.
 Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
@@
 On the other hand, a language without static type checking can often be designed 
-to be more flexible and forgiving.
+to be more flexible and forgiving, making it easy, for example, to have a \emph{map} function that can map functions
+with any number of arguments (as in most Lisp dialects).
 
-A recent trend is to combine the advantages of both kinds of languages by adding optional static 
+A recent trend is to aim to combine the advantages of both kinds of languages by adding optional static 
 type systems to languages without static type checking. 
 A major challenge in these combinations is continuing to support the idiomatic style of the original 
-language, and some recent attempts appear to have finally produced type systems that have overcome this obstacle.  
-Most notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
+language while performing accurate enough checking to catch similar errors to standard statically
+typed languages. Some recent attempts appear to have finally produced type systems that have overcome this obstacle.
+Notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
 check most idiomatic Racket code.
+These optional type systems have some important differences from standard static
+type systems, for example they often include \emph{union} types.
 
@@
 
-This paper describes my work on designing an optional static type system for Clojure based on
+This dissertation describes my work on designing an optional static type system for Clojure based on
 the lessons learnt from several projects, primarily Typed Racket.
-A prototype type system is developed for Clojure running on the Java Virtual Machine.
+This work includes designing and building a type checker for Clojure running on the Java Virtual Machine.
 Several experiments are conducted using this prototype, particularly involving
-sufficiently complicated existing Clojure code where type checking increases confidence that the code correct.
-For example, a significant portion of algo.monads, a Clojure Contrib library for monadic programming, 
+existing Clojure code that is sufficiently complicated that
+type checking increases confidence that the code is correct.
+For example, nearly all of \emph{algo.monads}, a Clojure Contrib library for monadic programming, 
 is able to be type checked.
-On the positive side, most monad, monad transformer, and monadic function definitions can be type checked,
+Most monad, monad transformer, and monadic function definitions can be type checked,
 usually by adding type annotations in natural places like function definitions.
-Negatively, simple monad comprehensions can be type checked, but require awkward type annotations.
 
@@
 both which are designed to improve error messages when mixing typed and untyped code in similar systems.
-I conclude that it appears to be both practical and useful to design and implement an optional static type system for the
+Overall, the work described in this dissertation leads to the conclusion that it appears to 
+be both practical and useful to design and implement an optional static type system for the
 Clojure programming language.
@@
 
-% Evaluation
-% - successes
-% - future work identified
-
 \include{experiments}
@@
 
-%
-%\begin{mathpar}
-%\inferrule*[right=B-Var]
-%  {\rho(x) = v}
-%  {\rho \vdash x \downarrow v}
-%
-%\inferrule*[right=B-Assoc]
-%  {\rho \vdash e_m \downarrow mi \\
-%   \rho \vdash e_k \downarrow k \\
-%   \rho \vdash e_v \downarrow v \\
-%   \rho \vdash \overrightarrow{e_j}^n \downarrow \overrightarrow{k_j}^n \\
-%   \rho \vdash \overrightarrow{e_m}^n \downarrow \overrightarrow{k_m}^n \\
-%   \delta(add\_entry, mi, k, v, \overrightarrow{k_j, k_m}^n) = mo}
-%  {\rho \vdash (assoc\ e_m\ e_k\ e_v\ \overrightarrow{e_j\ e_m}^n) \downarrow mo}
+\chapter{Conclusion}
 
-%\infrule[B-Var]
-%  {\rho(x) = v}
-%  {\rho \vdash x \downarrow v}
-%
-%\infrule[B-Assoc]
+Whether a language is dynamically typed or statically typed is starting to
+blur. Recent languages have managed to support the advantages of both styles:
+the safety of a statically typed language while retaining the idioms found in
+dynamically typed languages.
 
-%\end{mathpar}
-
-\chapter{Conclusion}
+Extensions like Typed Clojure
 
@@
 
+\include{appendix}
+
 \end{document}

### 2012-10-24 dissertation/design-choices.tex
Hmap design

@@
 
-\begin{verbatim}
-HMaps only convey positive information, not negative.
-Map concatenation is not approached, see lit review, many
-complications. Instead, only support assoc, dissoc, get.
-\end{verbatim}
+Heterogeneous maps in Typed Clojure are simple enough to avoid
+the complications associated with concatenable records (section \ref{sec:recordtypes})
+while still supporting many Clojure idioms.
+They support adding new entries with keyword keys, removing entries,
+and lookup of keyword keys.
+When using these operations with non-keyword keys, heterogeneous maps
+are treated like plain hash-maps.
 
-As a key idiomatic feature of Clojure, we treat heterogeneous maps more formally.
-This also serves as a prototype for formalising other aspects.
+They do not support concatenation (merging), as heterogeneous maps
+only hold positive information on the keys with entries, not
+which keys are \emph{absent}.
+This makes concatenation unsafe, as keys to the right of a
+concatentation can ``overwrite'' keys to the left. With no information
+on the absence of keys, concatenation becomes an unsound operation.
+
+We formalised some aspects of the behaviour of heterogeneous map types in Appendix~\ref{sec:hmaptheory}.
+This also serves as a prototype for formalising other aspects in the future.
 Here we give a formal syntax, operational semantics and type system using inference

### 2012-10-24 dissertation/future-work.tex
Hmap design

@@
 \chapter{Future Work}
+
+\section{Paired Arguments}
+
+It is not clear how to handle several key Clojure functions
+that accept an even number of arguments.
+For example, valid usages of \lstinline|assoc| are
+\lstinline[mathescape]{(assoc m k v $\overrightarrow{k v}$)},
+which takes three arguments and any number of paired arguments.
+
+Strickland, Tobin-Hochstadt, and Felleisen's calculus~\cite{STF09}
+is insufficient to express this pattern.
+Devising and integrating types that can express this pattern 
+is set as future work.
 
@@
 It would be desirable to port Typed Clojure to other Clojure implementations.
-Each implementation would bring its own set of challenges in
+Each implementation would bring its own set of challenges with interoperability.
+The core of Typed Clojure would preserved with each port changing at least 
+its interoperability with its host platform. Section \label{sec:portability}
+discusses potential problems with the porting process, and how
+they might be solved.
 
-\subsection{Extensions to heterogeneous map types}
-
-\subsection{Length information and intersection types}
+%\subsection{Extensions to heterogeneous map types}
 
@@
 
-\subsection{Linter}
+%\subsection{Linter}

### 2012-10-24 dissertation/lit-review.tex
Hmap design

@@
 
-This section 
-\cite{ACPR94}
+Typed Clojure adds static type checking to an existing untyped language.
+Coming from the other direction, Albardi, Cardelli, Pierce and Didier~\cite{ACPR94}
+extend a static language to be more dynamic.
+This work centres around the type \emph{Dynamic}, which represents
+a static type in a dynamically typed languages.
+
+Rossberg~\cite{Ros07} describes the advantages of having a type
+\emph{Dynamic} over other approaches to making statically typed languages
+more like dynamic ones. He says that type \emph{Dynamic} models
+allows aspects of dynamic style to be represented in statically
+typed languages without compromising the type system.
+However, he claims type \emph{Dynamic} is rarely used in practical languages
+because they are too inconvenient for many cases and they destroy
+a valuable property of polymorphic, statically typed languages.
 
@@
 
-% Rowan
-% You've missed some of the steps. Most importantly the move to program analysis
-% techniques like Set Based Analysis (SBA) - which suffered from modularity issues
-% due to the lack of declared intended types.
-% Maybe quote Felleisen - "problem: Mr Spidey and friends infer brittle and large
-% types; errors remain difficult to explain and fix"
-
-Type systems for Scheme have since moved away from soft typing
-to to other approaches like gradual typing.
+Type systems for Scheme have since moved away from soft typing.
 Alas, the only reasonably complete reference for commentary on this transition 
@@
 
-\begin{verbatim}
-  Bidirectionality
-\end{verbatim}
+\subsection{Program Analysis}
+
+Scheme then moved to program analysis techniques like Set Based Analysis (SBA).
+They, however, suffered from modularity issues due to the lack of declared intended types.
+MrSpidey\footnote{\url{http://www.plt-scheme.org/software/mrspidey/}}
+is a static type checker for Racket that uses uses SBA.
+Felleisen states the main problem with these kinds of systems: ``Mr Spidey and friends infer brittle and large
+types; errors remain difficult to explain and fix''~\cite{Fell09}.
 
@@
 can the most appropriate one for the situation.
+A key feature of gradual type systems is its use of \emph{bidirectional checking}
+like~\cite{PT00}, which requires type annotations in natural places like functions.
+Type errors in gradual type systems are often more manageable than in Soft Typing and SBA,
+often more comparable to type errors in statically typed languages.
 
+Gradually typed languages have different degrees of runtime checking.
 Typed Racket was developed as a path for module-by-module
@@
 by inserting runtime checks.
-
-\begin{verbatim}
-  Different levels of Gradual Typing discussed by Siek..
-\end{verbatim}
+Typescript~\cite{TypeS2012}, a gradual type system for Javascript, 
+does not add runtime checks to untyped interactions.
+Runtime checks affect performance, but give better errors and ensure
+type errors result in the program failing quicker.
 
@@
 Interactions with untyped modules are protected by adding runtime checks based on expected types.
-Typed Racket implements a sophisticated blame calculus. It ensures 
+Typed Racket implements a sophisticated blame calculus, based on Wadler~\cite{WF09}. It ensures 
 error messages correctly \emph{blame} the source of type errors,
 which can be difficult to determine in the presence of higher-order functions.
-\begin{verbatim}references, Walder etc.\end{verbatim}
 
 \section{Record Types}
+\label{sec:recordtypes}
 
-OCaml-style extensible record types has been the subject of extensive research~\cite{Wan89,CM91,HP91}.
+O'Caml-style extensible record types has been the subject of extensive research 
+(eg. Wand~\cite{Wan89}, Cardelli and Mitchell~\cite{CM91}, Harper and Pierce~\cite{HP91}).
 Typed Clojure's heterogeneous map types show some resemblance to extensible
@@
 \label{sec:intersectionunion}
-
-% Hayashi Singleton types
-% Intersection types - CDV81
-% Unions in soft typing CF91
-% - Static Type Inference in a Dynamically Typed Lanugage
-%  - cited by CF91, which has unions
-
-% Unions, intersections, and singletons all fit best with extrinsic types
-% - this is one reason why they are similar
 

### 2012-10-24 dissertation/design-choices.tex
final dissertation (unreviewed by Rowan)

@@
 
-%Tobin-Hochstadt
+Strickland, Tobin-Hochstadt, and Felleisen's describe variable-arity polymorphism~\cite{STF09},
+which has been implemented in Typed Racket.
+Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
+as described in section \ref{sec:impvariablearity}
 
-% Problematic areas
-% - flattened paired arguments. assoc, hash-maps, array-map
-% - complex argument dependencies. comp, partial
+Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
+and \lstinline|partial| cannot be typed by Strickland et al./ .
+Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
+be used in a higher-order context.
+Expressing the types of these functions is set as future work (section \ref{sec:futurearity}).
 
@@
 \subsection{Primitive Arrays}
+\label{designarrays}
 
-% Primitive arrays are covariant, which is well-known to be statically unsound.
-% We cannot trust the type signature of any Java method or field.
-% Separate read and write types into two parameters.
+Primitive arrays in Java are covariant, which is well-known to be statically unsound~\cite{OW97,BK99}.
+This mean that at compile time we cannot completely trust the type signature of any Java method or field that involves 
+an array.
+
+We handle this by introducing a type \lstinline|(Array c t)|, which is an array has the Java component type \lstinline|c| 
+array that can read and write values of type \lstinline|t|.
+Internally, \lstinline|Array| types expand to \lstinline|(Array3 c w r)|,
+which is an array has the Java component type \lstinline|c|,
+that can write type \lstinline|w| (which is contravariant in \lstinline|Array|) and read type \lstinline|r| 
+(which is covariant in \lstinline|Array|).
+\lstinline|(Array c t)| expands to \lstinline|(Array3 c t t)|.
+
+\lstinline|Array3| is useful when an array is passed to/from Java or some untyped Clojure code.
+As soon as an array leaves the safety of Typed Clojure, it can be considered to have type
+\lstinline|(Array3 c Nothing t)|. It is illegal to write to an array of this type in Typed Clojure
+without first checking its component type, but reads remain legal.
+
+This scheme should be considered experimental, and is not completely fleshed out or implemented.
+Notably, the component type \lstinline|c| seems redundant, and Typed Clojure currently
+does not provide syntax to ``cast'' an unwriteable array to a be writeable.
 
@@
 Protocols and datatypes are two of Clojure's main means of abstraction.
+Protocols, have a set of unimplemented methods, are typed by
+giving the type for each method.
+Datatype definitions are type checked by ensuring the methods it
+implements match the types defined by the Java interface or protocol
+it is extending.
 
-%section \ref{sec:datatypes}
+Common usages of datatypes are checkable, like datatype
+constructors, field accesses, and protocol method invocation.
+These are described in section \ref{sec:datatypes}.
+Mutable fields in datatypes are not yet supported, but is
+also a rarely used feature.
 
@@
 
-Typed Clojure takes a
-
-Similar to Typed Racket, type inference is based on Local Type Inference
+Type inference in Typed Clojure is based on Local Type Inference
 by Pierce and Turner~\cite{PT00}.
+The main extension to Local Type Inference in Typed Clojure
+is support for \emph{dotted variables} for variable-arity polymorphism.
+This implementation is lifted directly from Typed Racket.
 
@@
 to Local Type Inference. 
-
-An extra environment from type variables to bounds is kept until after inference,
-after which each unground bound is substituted with the inferred types and the
-subtyping relationships are checked.
+Type variables can have upper and lower bounds, and F-bounds allow type variable bounds
+to recursively refer to the type variable being bounded.
+This is similar to Java and Scala, which both support F-bounded polymorphism.
 
@@
 
-Heterogeneous maps in Typed Clojure are simple enough to avoid
+Heterogeneous maps types in Typed Clojure are simple enough to avoid
 the complications associated with concatenable records (section \ref{sec:recordtypes})
@@
 
-They do not support concatenation (merging), as heterogeneous maps
+Heterogeneous map types do not support concatenation (merging), as they
 only hold positive information on the keys with entries, not
@@
 This makes concatenation unsafe, as keys to the right of a
-concatentation can ``overwrite'' keys to the left. With no information
+concatenation can ``overwrite'' keys to the left. With no information
 on the absence of keys, concatenation becomes an unsound operation.

### 2012-10-24 dissertation/experiments.tex
final dissertation (unreviewed by Rowan)

@@
 \section{Conduit}
+
+Conduits are an advanced form of ``pipes'' using arrows, and a generalisation of monads~\cite{ConduitRef}.
+\emph{Conduit} is a Clojure library developed by Duey and 
+contributors~\footnote{Github home of Conduit: \url{https://github.com/jduey/conduit}} 
+that supports programming with conduits.
+
+The library had to be modified and simplified considerably to be able to type check.
+Many existing conduits had troublesome variable-parameters, reminiscent of the
+issues of assigning types to \lstinline|assoc| and \lstinline|partial|
+(which is future work, see section \label{sec:futurearity}).
+In these cases, simplified versions of the conduits were created that 
+took fixed arguments.
+
+An interesting property of the types assigned to conduits (which
+are internally functions of a single argument) is that it is
+a recursive type. This is not in itself interesting, but Typed Clojure
+implements conduit types as \emph{type functions} with variance,
+which required the variance of the conduit types to be known in advance.
+
+Listing \ref{lst:conduittypes} shows the type constructor for conduit,
+\lstinline|==>|. Notice the body of \lstinline|==>| includes a reference
+to itself in a position where variance must be known in advance.
+\lstinline|declare-alias-kind| is used to declare the kind of an alias.
+Once an alias with an already declared kind is defined with \lstinline|def-alias|,
+the defined kind must match the declared kind, otherwise a type error is thrown.
+(The specifics of the types in listing \ref{lst:conduittypes} are not relevant to the rest of this discussion,
+so they are not explained).
+
+\begin{lstlisting}[caption=Types for Conduits, label=lst:conduittypes]
+...
+
+(def-alias Result
+  (TFn [[x :variance :covariant]]
+    (U nil ;stream is closed
+       '[] ;abort/skip
+       '[x];consume/continue
+       )))
+
+(def-alias Cont
+  (TFn [[in :variance :covariant]
+        [out :variance :invariant]]
+    [(U nil [(Result in) -> (Result out)]) -> (Result out)]))
+
+(declare-alias-kind ==> (TFn [[in :variance :contravariant]
+                              [out :variance :invariant]] Any))
+
+(def-alias ==>
+  (TFn [[in :variance :contravariant] 
+        [out :variance :invariant]]
+    [in -> '[(U nil (==> in out)) (Cont out out)]]))
+
+...
+\end{lstlisting}
+
+The resulting port of \emph{Conduit} was unsatisfying for real world use.
+The simplifications made in order to type check the library reversed
+any effort \emph{Conduit} made to conform to common Clojure idioms.
+The future work on improving variable-argument types should reveal whether
+this library can be ported satisfactorily.

### 2012-10-24 dissertation/lit-review.tex
final dissertation (unreviewed by Rowan)

@@
 
-\section{Java Interoperability}
+\section{Java Interoperability in Statically Typed Languages}
 
-Scala and Clojure are both languages with a focus on host interoperability, specifically
-to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
-relevant, particularly considering the potential of using occurrence typing~\cite{TF10}
-to prevent erroneous usage of \lstinline|null|.
+Typed Clojure features integration with Java's static type system
+to help verify interoperability with Java as correct.
+Section \ref{sec:interactionnull} presents Typed Clojure's treatment of
+Java's \emph{null} pointer. 
+Section \ref{designarrays} shows how Typed Clojure deals with shortcomings
+of Java's static type system relating to primitive arrays.
+
+Other statically typed languages support Java interoperability with
+similar goals, most completely, Scala.
+Scala~\cite{OCD+} is tightly integrated with Java. It manages interactions
+with Java's \emph{null} pointer by using an \emph{Option} type, which (if
+used correctly) provides strong elimination for \emph{null}.
+\emph{null} is still not expressible as a type in Scala, however.
+Arrays are parameterised by a non-variant (or invariant) parameter,
+which are checked using \emph{conservative approximation}~\cite{OSV08}. For example,
+this checks that covariant parameters are only used in covariant positions.
 
@@
 Typed Racket 
-supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Piece et al.\ .
+supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Pierce and Turner.
 Scala supports bounded quantification with F-bounded polymorphism~\cite{CCHOM89},

### 2012-10-24 dissertation/paper.tex
final dissertation (unreviewed by Rowan)

@@
 
-Whether a language is dynamically typed or statically typed is starting to
-blur. Recent languages have managed to support the advantages of both styles:
+Whether a language is dynamically typed or statically typed is not always a
+dividing classification. Recent languages have managed to support the advantages of both styles:
 the safety of a statically typed language while retaining the idioms found in
@@
 
-Extensions like Typed Clojure
+This dissertation describes an optional static typing system
+that attempts to bring the advantages of static typing to the dynamically type language
+Clojure, running on the Java Virtual Machine. 
+By reusing many of the design choices made in similar projects
+like Typed Racket, we are able to design and implement a prototype type system
+\emph{Typed Clojure} that can statically check many Clojure idioms.
 
-\begin{verbatim}
-  Programming is an error-prone task, and verification tools are important.
-  Dynamic languages rely on several techniques for verification, but static
-  type checking is not often supported.
-\end{verbatim}
+Typed Clojure is intended to be of practical use to Clojure programmers.
+We show that Typed Clojure help verify Clojure code written in sophisticated
+programming styles is correct by porting most of a wide-spread Clojure library for monadic programming.
+Similarly, we show that Typed Clojure helps verify code using Java interoperability is correct, and is
+a useful tool when code involves usages of Java's \emph{null}.
+
+There is still significant future work in order to type check all Clojure idioms,
+but the work already carried out suggests that an optional type system for Clojure
+like Typed Clojure is both practical and useful tool.
 

### 2012-10-24 dissertation/design-choices.tex
Half of Rowan's corrections

@@
 down conditional branches, and makes it practical to throw type errors 
-where \lstinline|nil| is not allowed (see section \label{sec:OccurrenceTyping} for examples).
+where \lstinline|nil| is not allowed (see section \ref{sec:OccurrenceTyping} for examples).
 This is not practical in Java's type system because \emph{null} is not directly
@@
 values. Currently, we let the programmer define a set of unenforced \emph{assumptions}
-where \emph{null} is allowed in Java fields and methods (see section \label{sec:javainterop}).
+where \emph{null} is allowed in Java fields and methods (see section \ref{sec:javainterop}).
 Adding runtime assertions that enforce these assumptions would allow more

### 2012-10-24 dissertation/experiments.tex
Half of Rowan's corrections

@@
 issues of assigning types to \lstinline|assoc| and \lstinline|partial|
-(which is future work, see section \label{sec:futurearity}).
+(which is future work, see section \ref{sec:futurearity}).
 In these cases, simplified versions of the conduits were created that 

### 2012-10-24 dissertation/intro.tex
Half of Rowan's corrections

@@
 dynamically typed languages with static type systems. 
-This is idea not new~\cite{Fag91,CF91,AM91}, but recent attempts~\cite{Tob10}
+This is idea not new (see work by Fagan, Cartwright, Aiken and Murphy~\cite{Fag91,CF91,AM91}), but recent attempts
 are noteworthy for their broad success in matching many of the advantages of statically typed
-languages, notably due to the use of bidirectional checking~\cite{PT00}.
+languages (see work by Tobin-Hochstadt~\cite{Tob10}), 
+notably due to the use of bidirectional checking (like work by Pierce and Turner~\cite{PT00}).
 Instead of always attempting to infer types, this algorithm relies on programmer annotations
@@
 \begin{itemize}
+  \item We develop a prototype type checker for Clojure running on the Java Virtual Machine
+        which is able to type check many common Clojure idioms.
+        We show how this type checker can be used to aid Clojure programmers use sophisticated
+        programming styles by porting most of an existing library for monadic programming.
+
+  \item This prototype is extended to support several features that are not present in Typed Racket.
+        We support filtering sequences on anonymous functions that convey only negative type information
+        (eg. non-nil), capturing a common Clojure idiom.
+
   \item We describe a novel use of occurrence typing~\cite{TF10} for eliminating
@@
 
-  \item We develop a prototype type checker for Clojure running on the Java Virtual Machine
-        which is able to type check many common Clojure idioms.
-        We show how this type checker can be used to aid Clojure programmers use sophisticated
-        programming styles by porting most of an existing library for monadic programming.
-
-  \item This prototype is extended to support several features that are not present in Typed Racket.
-        We support filtering sequences on anonymous functions that convey only negative type information
-        (eg. non-nil), capturing a common Clojure idiom.
-
   \item We capture the Clojure idiom of using maps with known keyword fields by using heterogeneous map types.
@@
 
-  \item We identify major future issues in typing Clojure code. Multimethod definitions are problematic
+  \item We identify the main future issues in typing Clojure code. Multimethod definitions are problematic
         to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
@@
         a plain map.
-%        Heterogeneous map types 
 \end{itemize}
 
-Together,
-
-These contributions support my thesis by using the prototype to check practical code
-and
+Together these contributions support my thesis and motivate further work in this area.
 
@@
 variable being bounded. Typed Clojure supports upper and lower type variable bounds.
-\footnote{F-BOUND EXAMPLE NEEDED}
+This would be required mainly for future work involving Java Generics, like typing
+invocations of \emph{java.util.Collection/max}\footnote{\url{http://docs.oracle.com/javase/1.5.0/docs/api/java/util/Collections.html#max(java.util.Collection)}}.
 

### 2012-10-24 dissertation/lit-review.tex
Half of Rowan's corrections

@@
 Typed Clojure adds static type checking to an existing untyped language.
-Coming from the other direction, Albardi, Cardelli, Pierce and Didier~\cite{ACPR94}
+Coming from the other direction, Abadi, Cardelli, Pierce and R\'{e}my~\cite{ACPR94}
 extend a static language to be more dynamic.
-This work centres around the type \emph{Dynamic}, which represents
-a static type in a dynamically typed languages.
+This work centres around adding a type \emph{Dynamic}, which represents
+a dynamic type in a statically typed language.
 
@@
 \emph{Dynamic} over other approaches to making statically typed languages
-more like dynamic ones. He says that type \emph{Dynamic} models
+more like dynamic ones and surveys work in this area. 
+He says that type \emph{Dynamic}
 allows aspects of dynamic style to be represented in statically
@@
 However, he claims type \emph{Dynamic} is rarely used in practical languages
-because they are too inconvenient for many cases and they destroy
-a valuable property of polymorphic, statically typed languages.
+because it is too inconvenient in real programs.
+He claims to have fixed these issues, but a practical language has yet to integrated 
+his changes.
 
@@
 
-Wright and Cartwright~\cite{WC97} developed Soft Scheme, a practical
+Wright and Cartwright~\cite{WC97} developed Soft Scheme, a
 soft type system for Scheme. 
@@
 Type systems for Scheme have since moved away from soft typing.
-Alas, the only reasonably complete reference for commentary on this transition 
+Alas, the only reasonably complete account of this transition 
 appears to be slides from a talk by Felleisen~\cite{Fell09},
 a leader in this area of research for over 20 years.
-Felleisen says that while Soft Scheme discovered type problems, 
+Felleisen comments that while Soft Scheme discovered type problems, 
 it suffers from incomprehensible
@@
 
-Intersection, union, and singleton types fit best as extrinsic types.
+Intersection, union, and singleton types are best considered as extrinsic types.
 Hayashi~\cite{Hay91} describes two type theories, ATT which includes
@@
 Several interesting projects have used intersection or union types.
-Intersection types were originally introduced by Coppo, Dezani-Ciancaglini, and Venneri~\cite{CDV81}.
-The use union types to type check dynamically typed languages with soft typing systems
+Intersection types were originally introduced by Coppo, Dezani-Ciancaglini, and Venneri~\cite{CDV81}
+for the $\lambda$-calculus.
+The use of union types to type check dynamically typed languages with soft typing systems
 dates back to Cartwright and Fagan~\cite{CF91}.
@@
 
-Refinement types add a level of types refining an existing type system
+Refinement types add a level of extrinsic types refining an existing intrinsic type system
 in order to type check more detailed properties and invariants than standard static type systems
 (described by Freeman and Pfenning~\cite{FP91}).
-Refinement types are reminiscent of Typed Clojure, in that Typed Clojure
-adds a level of types on top of Clojure.
+Refinement types have similarities with Typed Clojure, in that Typed Clojure
+adds a level of extrinsic types on top of Clojure.
 Intersection types are critical here to allow more than one property or invariant
@@
 Typed Clojure takes a similar approach for its representation of functions.
-Typed Racket uses (disjoint) union types.
+Typed Racket uses union types which are used in Typed Clojure in very similar ways,
+like for creating ML-like ``datatypes''.
 
@@
 
-Typed Racket fully expands all macro calls before type checking~\cite{Tob10} to
-avoid the complex semantics of type checking macro definitions, an ongoing research area summarised
+Typed Racket fully expands all macro calls before type checking~\cite{Tob10},
+avoiding the complex semantics of type checking macro definitions, an ongoing research area summarised
  by Herman~\cite{Her10}.
-The design of my Clojure type system will follow a similar strategy; only the fully macro-expanded form
-will be type checked. Type checking macro definitions are outside the scope of this project, and would
-be exceptionally hard.
+Typed Clojure follows a similar strategy; only the fully macro-expanded form
+will be type checked. Type checking macro definitions are outside the scope of this project.
 
 Along with a full static type system, Typed Racket 
-also uses runtime contracts to enforce type invariants at runtime~\cite{TF08}.
+also uses runtime contracts to enforce type invariants at runtime
+at the interface with untyped code ~\cite{TF08}.
 Utilising runtime contracts to aid type checking is outside the scope of this project, but would be 
@@
 After some investigation, typing multimethods with Typed Clojure is assigned as future work
-(see section \label{sec:mmfuture}).
+(see section \ref{sec:mmfuture}).
 
@@
 
-F-$\omega$ is a typed $\lambda$-calculus with support for higher kinds,
+F$_\omega$ is a typed $\lambda$-calculus with support for higher kinds,
 specifically type constructors (see Pierce~\cite{Pie02}).

### 2012-10-24 dissertation/design-choices.tex
Add proposal, acknowledgements

@@
 
-Strickland, Tobin-Hochstadt, and Felleisen's describe variable-arity polymorphism~\cite{STF09},
+Strickland, Tobin-Hochstadt, and Felleisen describe variable-arity poly\-morphism ~\cite{STF09},
 which has been implemented in Typed Racket.
@@
 Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
-and \lstinline|partial| cannot be typed by Strickland et al./ .
+and \lstinline|partial| cannot be typed by Strickland et al.\ .
 Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
@@
 
-\section{Clojure Hierarchy}
+\section{Clojure type hierarchy}
 
-The Clojure hierarchy is a set of Java classes and interfaces, mostly in the
+The Clojure type hierarchy is a set of Java classes and interfaces, mostly in the
 \emph{clojure.lang} package, that define Clojure's core 
@@
 
-Clojure's hierarchy was accommodating to this process.
+Clojure's type hierarchy was accommodating to this process.
 For example, \lstinline{clojure.lang.Seqable} is the main interface behind
@@
 is support for \emph{dotted variables} for variable-arity polymorphism.
-This implementation is lifted directly from Typed Racket.
+This implementation is ported directly from Typed Racket.
 
@@
 They support adding new entries with keyword keys, removing entries,
-and lookup of keyword keys.
+and lookup of keyword keys with the type associated with each key in
+a map tracked separately, similar to type systems for record types 
+(see work by Harper and Pierce~\cite{HP91}).
 When using these operations with non-keyword keys, heterogeneous maps
@@
 This makes concatenation unsafe, as keys to the right of a
-concatenation can ``overwrite'' keys to the left. With no information
+concatenation can ``overwrite'' keys to the left with different types. With no information
 on the absence of keys, concatenation becomes an unsound operation.

### 2012-10-24 dissertation/experiments.tex
Add proposal, acknowledgements

@@
 
-In this chapter we look at several example of using Typed Clojure and
+In this chapter work through several examples of using Typed Clojure and
 how well the current prototype handles them. 
@@
 monad type constructors. Keeping the monad type constructor abstract allows us to determine
-whether a monad transformer works \emph{for any} monads.
+whether a monad transformer works \emph{for all} monads.
 
@@
 
-Conduits are an advanced form of ``pipes'' using arrows, and a generalisation of monads~\cite{ConduitRef}.
+Conduits are an advanced form of ``pipes'' using arrows, a generalisation of monads~\cite{ConduitRef}.
 \emph{Conduit} is a Clojure library developed by Duey and 
@@
 An interesting property of the types assigned to conduits (which
-are internally functions of a single argument) is that it is
-a recursive type. This is not in itself interesting, but Typed Clojure
+are internally functions of a single argument) is that they are 
+recursive types. This is not in itself interesting, but Typed Clojure
 implements conduit types as \emph{type functions} with variance,
@@
 The simplifications made in order to type check the library reversed
-any effort \emph{Conduit} made to conform to common Clojure idioms.
+any effort that was invested into \emph{Conduit} to conform to common Clojure idioms.
 The future work on improving variable-argument types should reveal whether
-this library can be ported satisfactorily.
+this library can be ported satisfactorily. It is desirable to port
+this library because implementations and usages of conduits result in very abstract code
+which static type systems can help verify as correct.

### 2012-10-24 dissertation/future-work.tex
final dissertation (unreviewed by Rowan)
Add proposal, acknowledgements

@@
 
-\section{Paired Arguments}
+\section{Variable-Arity Polymorphism}
+\label{sec:futurearity}
 
@@
 
-A key part of Typed Racket~\cite{Tob10} is its contract and blame system.
+A key part of Typed Racket~\cite{Tob10} is its contract and blame systems.
 They enable safe interoperability with untyped Racket by generating runtime
@@
 The core of Typed Clojure would preserved with each port changing at least 
-its interoperability with its host platform. Section \label{sec:portability}
+its interoperability with its host platform. Section \ref{sec:portability}
 discusses potential problems with the porting process, and how

### 2012-10-24 dissertation/implementation.tex
Hmap design
final dissertation (unreviewed by Rowan)
Add proposal, acknowledgements

@@
 which is reflected in this chapter. 
+(The code is available on Github at \url{https://github.com/frenchy64/typed-clojure}).
 
@@
 
-\lstinline|infer| is almost always used when invoking polymorphic functions.
-\lstinline|clojure.core/constantly| is a polymorphic function with type \lstinline|(All [x y] [x -> [y * -> x]])|
+\lstinline|infer| is almost always used when invoking polymorphic functions like,
+for example, \lstinline|constantly|, which has type \lstinline|(All [x y] [x -> [y * -> x]])|
 (read as a function taking \lstinline|x| and returning a function that takes any number
 of \lstinline|y|'s and returns \lstinline|x|).
-For instance, \lstinline|((constantly true) 'any 'number)| results in the value \lstinline|true|.
+For instance, \lstinline|((constantly true) 'any 'number)|, results in the value \lstinline|true|.
 
@@
 \section{Variable-arity Polymorphism}
+\label{sec:impvariablearity}
 
@@
 \section{Portability to other Clojure Dialects}
+\label{sec:portability}
 

### 2012-10-24 dissertation/paper.tex
Half of Rowan's corrections
Add proposal, acknowledgements

@@
 \keywords{Type systems, Clojure}
-% TODO What are categories? cshonours requires them
-\categories{}
+% TODO What are categories? cshonours requires them - but commented out now
+%\categories{}
 
@@
 
+\begin{acknowledgements}
+
+The past year has included by far my most enjoyable and motivating programming-related experiences.
+
+Last November I attended my first conference, Clojure Conj 2011 (Raleigh, NC),
+where I delivered a talk on logic programming with Clojure and met some of my programming heroes.
+A week later my supervisor Rowan Davies invited me to visit Carnegie-Mellon University
+where he was working on his PhD project. Most memorably, I was invited to a meeting between
+Rowan and his PhD supervisor Frank Pfenning, which was a great thrill.
+
+This trip was funded by donations from the programming community. Thank you to all who helped me get there:
+this work was directly inspired from conversations during this period.
+
+Many thanks go to my supervisor Rowan. His influence as a lecturer inspired my exploration of
+functional programming languages. He has exceeded my expectations as an honours supervisor,
+often dedicating time to assist me on public holidays and weekends.
+Rowan's experience with refinement types brought a unique perspective to this project
+which was very welcome.
+
+I'd also like to thank: Sam Tobin-Hochstadt for his consistent assistance and patience;
+David Nolen for mentoring this project as part of Google Summer of Code 2012;
+and Chas Emerick, without whom I would have never considered attending Clojure Conj 2011
+(and this work probably would not have happened).
+Also, Daniel Spiewak, Dan Friedman and William Byrd, Jim Duey, Baishampayan Ghose, and
+Steve Strickland, thank you for your help and inspiration.
+Thank you Rich Hickey for creating Clojure, it is a thrill to build tools that make
+this great language even better.
+
+Fittingly, I have been invited to talk on Typed Clojure at Clojure Conj
+2012 in November, which I am highly anticipating.
+
+Finally, thanks to my family, friends and to my girlfriend Tamara for your love and support.
+
+\end{acknowledgements}
+
 \tableofcontents
@@
 Typed Clojure is intended to be of practical use to Clojure programmers.
-We show that Typed Clojure help verify Clojure code written in sophisticated
-programming styles is correct by porting most of a wide-spread Clojure library for monadic programming.
+We show that Typed Clojure helps verify the absence of errors in Clojure code written in sophisticated
+programming styles by porting most of a wide-spread Clojure library for monadic programming.
 Similarly, we show that Typed Clojure helps verify code using Java interoperability is correct, and is
-a useful tool when code involves usages of Java's \emph{null}.
+a useful tool to show misusing of Java's \emph{null}.
 

### 2012-10-24 dissertation/intro.tex
Fix incomplete references

@@
 can even be called a type system. We choose to
-follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey01},
+follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey02},
 where such optional type systems are \emph{extrinsic} type systems, and more

### 2012-10-24 dissertation/paper.tex
Fix incomplete references
Thanks

@@
 where I delivered a talk on logic programming with Clojure and met some of my programming heroes.
-A week later my supervisor Rowan Davies invited me to visit Carnegie-Mellon University
+A week later my supervisor Rowan Davies invited me to visit Carnegie Mellon University
 where he was working on his PhD project. Most memorably, I was invited to a meeting between
@@
 
-Finally, thanks to my family, friends and to my girlfriend Tamara for your love and support.
+Finally, thanks to my family, friends and to my girlfriend Tamara, each for your love and support.
 

### 2012-10-25 dissertation/appendix.tex
Final commit before marking

--- 2012-05-24 dissertation-proposal/research-proposal.tex
@@
-\documentclass[12pt, a4paper]{article}
+\appendix
+\chapter{Heterogeous map type theory prototype}
+\label{sec:hmaptheory}
 
-\usepackage{biblatex}
-\bibliography{bibliography}
+\section{Operational Semantics}
 
-\setlength{\oddsidemargin}{0.5cm}
-\setlength{\evensidemargin}{0.5cm}
-\setlength{\topmargin}{-1.6cm}
-\setlength{\leftmargin}{0.5cm}
-\setlength{\rightmargin}{0.5cm}
-\setlength{\textheight}{24.00cm} 
-\setlength{\textwidth}{15.00cm}
-\parindent 0pt
-\parskip 5pt
-\pagestyle{plain}
+%$$
+%\begin{tdisplay}{Test1}
+%\begin{altgrammar}
+%  \s{},\t{} &::= & \int \alt \bool \alt {\proctype {\s{}} {\t{}}} 
+%  &\mbox{Types}\\
+%  \M{}, \N{}, \P{}, \dots &::=& \x{} \alt \abs{\x{}}{\M{}} \alt
+%  \comb{\M1}{\M2}  &\mbox{Raw Terms} \\
+%\end{altgrammar}
+%\end{tdisplay} 
+%$$
 
-\title{Research Proposal}
-\author{}
-\date{}
+$$
+\begin{tdisplay}{Syntax of Terms}
+\begin{altgrammar}
+  \exp{} &::=& (\c{} \overrightarrow{\exp{}})         % function application, multiple arguments
+             \alt \{ \overrightarrow{\exp{}\ \exp{}} \} % map literal
+             &\mbox{Expressions} \\ 
+  \v{} &::=& \k{} \alt \nil{}
+              \alt \{ \overrightarrow{\v{}\ \v{}} \}   % map literal
+              &\mbox{Values} \\
+  \c{} &::=& \assoc{} \alt \dissoc{} \alt \get{}
+              &\mbox{Constants}
+\end{altgrammar}
+\end{tdisplay}
+$$
 
-\newcommand{\namelistlabel}[1]{\mbox{#1}\hfil}
-\newenvironment{namelist}[1]{%1
-\begin{list}{}
-    {
-        \let\makelabel\namelistlabel
-        \settowidth{\labelwidth}{#1}
-        \setlength{\leftmargin}{1.1\labelwidth}
-    }
-  }{%1
-\end{list}}
+$$
+\begin{tdisplay}{Evaluation Contexts}
+  \begin{altgrammar}
+    \E{} &::=& [ ] % application rules
+              \alt (\c{}\ \overrightarrow{\v{}}\ \E{}\ \overrightarrow{\exp{}}) % eval arguments left-to-right
+              % map rules
+              \alt \{\overrightarrow{\v{}\ \v{}}\ \E{}\ \exp{}\ \overrightarrow{\exp{}\ \exp{}} \} % key first
+              \alt \{\overrightarrow{\v{}\ \v{}}\ \v{}\ \E{}\ \overrightarrow{\exp{}\ \exp{}} \}   % value next
+              &\mbox{Evaluation Contexts}
+  \end{altgrammar}
+\end{tdisplay}
+$$
 
-\begin{document}
-\maketitle
+% \mapsto for updating maps
+$$
+\begin{tdisplay}{Operational Semantics}
+\begin{array}{c}
+{\inferrule[E-Assoc]
+  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
+    {v_4 = v_1\ with\ entry\ v_2\ to\ v_3} }
+    {{(\assoc{}\ v_1\ v_2\ v_3) \hookrightarrow v_4} }} \\
+\\
+{\inferrule[E-Dissoc]
+  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
+    {v_3 = v_1\ without\ entry\ indexed\ by\ v_2} }
+  {(dissoc\ v_1\ v_2) \hookrightarrow v_3} } \\
+\\
+{\inferrule[E-GetMapExist]
+  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
+    {v_2\ v_3\ in\ v_1} }
+%    {v_1\ has\ entry\ with\ key\ v_2} \\
+%    {v_3 = corresponding\ value\ in\ entry\ with\ key\ v_2\ in\ v_1}
+  { {(get\ v_1\ v_2) \hookrightarrow v_3} }} \\
+\\
+{\inferrule[E-GetMapNotExist]
+  { {v_1 = \{ \overrightarrow{\v{}\ \v{}} \} } \\
+    {v_1\ has\ no\ entry\ with\ key\ v_2} \\
+    {v_3 = \nil{}} }
+  { {(get\ v_1\ v_2) \hookrightarrow v_3} }}
+\end{array} 
+\end{tdisplay} 
+$$
 
-\begin{namelist}{xxxxxxxxxxxx}
-\item[{\bf Title:}]
-	An Optional Static Type System for Clojure
-\item[{\bf Author:}]
-	Ambrose Bonnaire-Sergeant
-\item[{\bf Supervisor:}]
-	Rowan Davies
-\end{namelist}
+$$
+\begin{tdisplay}{Type Syntax}
+\begin{altgrammar}
+  \T{} &::=& \quad \Top \alt \Bottom \alt \Nil \alt \k{} \alt \{ \overrightarrow{\k{}\ \T{}} \} 
+             \alt {\IPersistentMap {\T{}} {\T{}}} \\
+             && \alt ( \vee\ \overrightarrow{\T{}} )
+                \alt ( \wedge\ \overrightarrow{\T{}} )
+\end{altgrammar}
+\end{tdisplay}
+$$
+
+% what about get? Core Type Rules
+
+$$
+\begin{tdisplay}{Core Type Rules}
+\begin{array}{c}
+{\inferrule[T-AssocHMap]
+  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\k{}}}} } \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
+    {\vdash {\issubtype {{\T{}}_{m}} {\TopHMap} }} }
+  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} 
+        {(update\_hmap\ {{\T{}}_{m}}\ {{\k{}}}\ {{\T{}}_{v}})}}}}} \\
+\\
+{\inferrule[T-AssocPromote]
+  { {\Gamma \vdash {\hastype {{\exp{}}_{m}} {{\T{}}_{m}}} } \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{k}} {{\T{}}_{k}}} } \\
+    {\Gamma \vdash {\hastype {{\exp{}}_{v}} {{\T{}}_{v}}} } \\
+    {\vdash {\issubtype {{\T{}}_{m}} {\TopIPersistentMap}}} \\
+    {\T{} = (promote\_hmap\ {\T{}}_{m}\ {\T{}}_{k}\ {\T{}}_{v})} }
+  { {\Gamma \vdash {\hastype {( \assoc{}\ {\exp{}}_{m}\ {\exp{}}_{k}\ {\exp{}}_{v})} {\T{}}}}}} \\
+\end{array}
+\end{tdisplay}
+$$
+
+$$
+\begin{tdisplay}{Type Metafunctions}
+\begingroup
+\fontsize{10pt}{12pt}
+\begin{altgrammar}
+  (update\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\k{}}_1\ {\T{}}_1) = 
+                 \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}}\ {\k{}}_1\ {\T{}}_1 \}\ \\
+  (update\_hmap\ (\wedge\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
+                               (\wedge\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})}) \\
+  (update\_hmap\ (\vee\ \overrightarrow{{\T{}}})\ {\T{}}_{k}\ {\T{}}_{v}) = 
+                               (\vee\ \overrightarrow{(update\_hmap\ {\T{}}\ {\T{}}_{k}\ {\T{}}_{v})})
+  \\ \\
+  (promote\_hmap\ \{ \overrightarrow{{\T{}}_{k}\ {\T{}}_{v}} \}\ {\T{}}_{kn}\ {\T{}}_{vn})
+                                    = {\IPersistentMap {(\vee\ \overrightarrow{{\T{}}_{k}}\ {\T{}}_{kn})}
+                                                       {(\vee\ \overrightarrow{{\T{}}_{v}}\ {\T{}}_{vn})}} \\
+  (promote\_hmap\ {\IPersistentMap {{\T{}}_{k}} {{\T{}}_{v}}} \ {\T{}}_{kn}\ {\T{}}_{vn})
+                                    = {\IPersistentMap {(\vee\ {\T{}}_{k}\ {\T{}}_{kn})}
+                                                       {(\vee\ {\T{}}_{v}\ {\T{}}_{vn})}}
+\end{altgrammar}
+\endgroup
+\end{tdisplay}
+$$
+
+%$$
+%\begin{tdisplay}{Subtyping}
+%\begin{array}{c}
+%\inferrule[S-Refl] { } {\vdash {\issubtype {\T{}} {\T{}}}} \\
+%\inferrule[S-Any] { } {\vdash {\issubtype {\T{}} {\Top}}}
+%\inferrule[S-UnionSuper] 
+%  {\ } 
+%  {\vdash {\issubtype {\T{}} {\Top}}}
+%\end{array}
+%\end{tdisplay}
+%$$
+
+\chapter{Dissertation Proposal}
 
@@
 Tobin-Hochstadt notes that ``untyped scripts are difficult to 
-maintain over the long run'' \cite{TypedScheme:2010} because types
+maintain over the long run''~\cite{Tob10} because types
 contain valuable design information.
-He developed Typed Scheme
-\cite{TypedScheme:2010} 
+He developed Typed Scheme~\cite{Tob10} 
 to safely and incrementally port existing Scheme code, a dynamic language, to
@@
 is also a Lisp, which makes it a good candidate to test the generality of
-ideas developed for Typed Scheme \cite{TypedScheme:2010}.
+ideas developed for Typed Scheme~\cite{Tob10}.
 
@@
 It will be based on the lessons learnt throught the development
-of Typed Scheme, and as a response to Tobin-Hochstadt's \cite{TypedScheme:2010}
+of Typed Scheme, and as a response to Tobin-Hochstadt's~\cite{Tob10}
 suggestion to add type systems to other existing dynamic languages.
@@
 type system for Clojure should understand them to some degree.
-There is some experience statically typing multimethods
-\cite{Millstein02modularstatically}, 
+There is some experience statically typing multimethods~\cite{MS02}, 
 but this is a challenge and may not be addressed in the
@@
 support for non-uniform variable arity polymorphism. 
-Strickland, Tobin-Hochstadt and Felleisen \cite{Strickland:2009:PVP:1532974.1532978}.
+Strickland, Tobin-Hochstadt and Felleisen~\cite{STF09}.
 describe their approach for non-uniform variable arity polymorphism, as used
@@
 
-Occurence typing is type checking technique developed for Typed Scheme 
-\cite{Tobin-Hochstadt:2008:DIT:1328897.1328486}
-and improved for Typed Racket
-\cite{Tobin-Hochstadt:2010:LTU:1932681.1863561}.
+Occurence typing is type checking technique developed for Typed Scheme~\cite{TF08}
+and improved for Typed Racket~\cite{TF10}.
 It helps the type checker understand common programming idioms 
@@
 
-Ensuring vigorous type safety is an important aspect of Typed Scheme 
-\cite{TypedScheme:2010} ,
+Ensuring vigorous type safety is an important aspect of Typed Scheme~\cite{Tob10},
 especially when interacting between untyped and typed modules.
@@
 % outlining your plan for the year.
-
-\printbibliography
-
-\end{document}

### 2012-10-25 dissertation/lit-review.tex
Add proposal, acknowledgements
Final commit before marking

@@
 Typed Clojure adds static type checking to an existing untyped language.
-Coming from the other direction, Abadi, Cardelli, Pierce and R\'{e}my~\cite{ACPR94}
+Coming from the other direction, Abadi, Cardelli, Pierce and R\'{e}my~\cite{ACPR95}
 extend a static language to be more dynamic.
@@
 
-Strickland et al.\ invented a type system supporting variable-arity polymorphism~\cite{STF09}
+Strickland et al.\ invented a type system supporting variable-arity polymorphism ~\cite{STF09}
 a version of which is included in the current implementation of Typed Racket.

### 2012-10-25 dissertation/paper.tex
Final commit before marking

@@
 (and this work probably would not have happened).
-Also, Daniel Spiewak, Dan Friedman and William Byrd, Jim Duey, Baishampayan Ghose, and
+Also, Daniel Spiewak, Dan Friedman and William Byrd, Jim Duey, Baishampayan Ghose, Sam Aaron, and
 Steve Strickland, thank you for your help and inspiration.

### 2012-10-25 dissertation/intro.tex
Wording, typo caught by Bronsa

@@
 Next, we see how Typed Clojure's default behaviour treats method return positions as nullable.
-The \emph{java.io.File} instance method \emph{getParent}
-is equivalent to \lstinline|[-> (U nil String]| in Typed Clojure. This happens to be
+By default, the \emph{java.io.File} instance method \emph{getParent}
+is assigned the type \lstinline|[-> (U nil String)]| in Typed Clojure. This happens to be
 a valid approximation of the method as \emph{getParent} returns \emph{null} 
 ``if the pathname does not name a parent directory''\footnote{See Javadoc for \emph{java.io.File}: http://docs.oracle.com/javase/1.4.2/docs/api/java/io/File.html}.
-On the other hand, the instance method \emph{getName} always returns an
+On the other hand, the \emph{java.io.File} instance method \emph{getName} always returns an
 instance of \emph{java.lang.String}, so we set the return position of

### 2012-10-27 dissertation/intro.tex
Fix wording on hmaps, suggested by nDuff

@@
 
-Clojure emphasises key-value paired, unordered \emph{maps} for many uses where other languages might use objects
-Clojure provides a map literal using curly braces. For example,
+It is idiomatic in Clojure to use plain maps with known keyword keys as lightweight objects.
+Clojure provides a hash-map literal (using curly braces), making this idiom particularly
+convenient.
+For example,
 \lstinline|{:a 1, :b 2}| is a map value with two key-value entries: from keyword key \lstinline|:a|

### 2012-10-28 dissertation/design-choices.tex
Edit Java design section for clarity

@@
 plays a major role in capturing many common Clojure idioms.
-One important idiom it helps Typed Clojure to understand 
-is the \emph{sequence abstraction}, which
-is made up of the \lstinline|clojure.core| functions \lstinline|seq|, \lstinline|first|, and \lstinline|rest|.
+One such idiom is the \emph{sequence abstraction}, which
+consists of the \lstinline|clojure.core| functions \lstinline|seq|, \lstinline|first|, and \lstinline|rest|.
 
@@
 representing an empty sequence).
-(See section \ref{sec:filterneg} for a more detailed explanation of the filter syntax).
+(See Section \ref{sec:filterneg} for a more detailed explanation of the filter syntax).
 
@@
 Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
-as described in section \ref{sec:impvariablearity}
+as described in Section \ref{sec:impvariablearity}
 
@@
 be used in a higher-order context.
-Expressing the types of these functions is set as future work (section \ref{sec:futurearity}).
+Expressing the types of these functions is set as future work (Section \ref{sec:futurearity}).
 
@@
 
-% what is this paragraph saying?
-Dialects of Clojure provide fast and unrestricted access to their host platform. 
-This is a intentional source of incompatibility between Clojure implementations.
-For example, Clojure is hosted on the Java Virtual Machine (JVM), ClojureCLR on the Common Language Runtime,
-and ClojureScript on Javascript VMs. These are very different platforms, and it is unlikely
-a Clojure dot call will be portable. There is no attempt at reconciling
-host interoperability differences, and it is up to the programmer to decide
-how to best abstract over different hosts.
+Clojure implementations are designed to target popular platforms, 
+providing fast, unrestricted access to their respective hosts.
+Notable implementations target
+the Java Virtual Machine (via Clojure), the Common Language Runtime (via ClojureCLR),
+and Javascript virtual machines (via ClojureScript). 
+Host interoperability is the main source of incompatibility between Clojure implementations.
+Aside from syntactic matters, Clojure makes no attempt at reconciling
+host interoperability differences:
+it is up to the programmer to decide how to best write portable Clojure code.
 
-Typed Clojure targets the JVM hosted Clojure.
-Clojure embraces the JVM as a host by sharing its runtime type system and encouraging direct
+Typed Clojure is designed for Clojure, which runs on the Java Virtual Machine (JVM).
+Clojure embraces the JVM as a host by sharing its runtime type system and providing direct
 interaction with libraries written in other JVM languages. Most commonly, Clojure programmers
-reach to Java libraries. Java is statically typed language. It follows that any interaction with
-Java will already have an annotated Java type. Typed Clojure can use these annotations
-to statically type check these interactions.
+use libraries written in Java. Java is a statically typed language, and it follows that any interaction with
+Java will already have an annotated Java type. Typed Clojure uses Java's static type annotations
+to help statically type check interactions with Java from Clojure.
 
-Typed Clojure attempts to improve some shortcomings of Java's static type system.
-
-Java's static type system does not provide a type-safe construct for eliminating
-the \lstinline|null| pointer. Instead, Java programmers often rely on either testing
-for \lstinline|null| or prior domain knowledge.
-
-\begin{lstlisting}[caption=null elimination in Java]
-...
-Object a = nextObject();
-if (a != null)
-  parseNextObject(a);
-else
-  throw new Exception("null Pointer Found!);
-...
-\end{lstlisting}
-
-In Java, \lstinline|null| is included in all reference
-types, but \lstinline|null| is not expressible as a static type. In other words,
-when a Java type claims a reference type, you should also assume it can also be of 
-type \lstinline|null|.
-
-Scala provides the Option type for safe \lstinline|null| elimination, forcing pattern matching
-where \lstinline|null| might occur. This approach is not appropriate for Typed Clojure
-as pattern matching is not idiomatic in Clojure.
+This section details where Typed Clojure attempts to improve some shortcomings of Java's static type system.
 
@@
 
-\subsection{Interaction with null}
+\subsection{Interaction with \emph{null}}
 \label{sec:interactionnull}
 
-It is a goal of Typed Clojure to statically prevent many Java \emph{NullPointerException}
-exceptions.
-\emph{null} is explicitly expressible in Typed Clojure as the singleton type \lstinline|nil|,
-representing the \emph{value} \lstinline|nil|.
-This enables more expressive types that specify where \emph{nil} is and is not allowed.
+As of Java 7, Java does not provide a type-safe construct for eliminating
+occurrences of \emph{null}. Instead, Java programmers rely on ad-hoc approaches,
+like testing for the presence of \emph{null} (like Listing \ref{lst:nullelim}) or prior domain knowledge.
+One solution to this problem is an Option type (also known a Maybe type).
+Scala provides \emph{scala.Option} for this purpose. The programmer provides two branches of code,
+(respectively to execute if an expression is \emph{null}, and if it is not)
+and the appropriate branch is chosen at runtime.
+There is no such construct in idiomatic Clojure, so this solution is inappropriate:
+a satisfactory solution should reuse current Clojure idioms.
+
+\begin{lstlisting}[caption=null elimination in Java, label=lst:nullelim]
+...
+Object a = nextObject();
+if (a != null)
+  parseNextObject(a);
+else
+  throw new Exception("null Pointer Found!);
+...
+\end{lstlisting}
+
+It is a goal of Typed Clojure to statically prevent misuses of Java's \emph{null}.
+At runtime, Clojure represents \emph{null} as the value \lstinline|nil|;
+Typed Clojure assigns this value the static type \lstinline|nil|
+(a singleton type containing just the value \lstinline|nil|).
+This enables more expressive types that explicitly communicate where \lstinline|nil|,
+and therefore \emph{null}), is allowed.
 Section \ref{sec:javainterop} gives an example of how this could be useful 
-when using Java interoperability and gives details on the different ways Typed Clojure.
+when using Java interoperability.
 
-Occurrence typing~\cite{TF10} also plays a crucial role in Typed Clojure's
-treatment of \lstinline|nil|.
-It allows us to track whether particular bindings are possibly \lstinline|nil|
-down conditional branches, and makes it practical to throw type errors 
-where \lstinline|nil| is not allowed (see section \ref{sec:OccurrenceTyping} for examples).
-This is not practical in Java's type system because \emph{null} is not directly
-expressible, which makes it impractical to error when a \emph{NullPointerException}
-exception is possible: according to the type system, almost every operation
-could involve \emph{null}!
-Eliminating \emph{NullPointerException} exceptions is entirely the programmers
-responsibility in Java.
-Typed Clojure helps the programmer by statically proving certain misuses of \lstinline|nil| are impossible.
+The crucial aspect of this approach is that misusing \emph{null} results in a \emph{static type error}.
+This differs significantly from Java: according to Java's static type system,
+it is implicit that almost every operation could involve \emph{null}!
+Occurrence typing~\cite{TF10} plays a crucial role in making this a practical
+design choice, by understanding existing Clojure idioms related to \emph{null} elimination (see Section \ref{sec:OccurrenceTyping}).
+Eliminating occurrences of \emph{null} in Java is the programmer's responsibility.
+Instead, Typed Clojure helps the programmer by statically proving many misuses of \emph{null} are impossible.
 
-The design choice of potentially misusing \lstinline|nil| results in a type error
-needs further work to be completely satisfying.
-Certain cases of type inference should be improved to avoid redundant code
-that ``pleases the type checker''.
+This decision, however, needs further work to be completely satisfying.
+In particular, type inference should be improved in certain common situations.
 For example, we currently do not know how to infer \lstinline|(filter identity coll)|
-as returning a sequence of non-nil elements, for any sequence \lstinline|coll|.
-(see section \ref{sec:filterneg}).
-It is also possible for code interacting with Java to introduce unexpected \emph{null}
-values. Currently, we let the programmer define a set of unenforced \emph{assumptions}
-where \emph{null} is allowed in Java fields and methods (see section \ref{sec:javainterop}).
+as returning a sequence of non-\lstinline|nil| elements, for any sequence \lstinline|coll|
+(see Section \ref{sec:filterneg}).
+Also, Typed Clojure lets the programmer define a set of unenforced \emph{assumptions}
+where \emph{null} is allowed in Java fields and methods (see Section \ref{sec:javainterop}),
+which is potentially brittle: there is nothing stopping the Java code from introducing unexpected \emph{null} values at runtime.
 Adding runtime assertions that enforce these assumptions would allow more
-accurate errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
+accurate and earlier errors, an approach similar to that taken by Typed Racket~\cite{Tob10} when
 interfacing with untyped Racket code.
@@
 constructors, field accesses, and protocol method invocation.
-These are described in section \ref{sec:datatypes}.
+These are described in Section \ref{sec:datatypes}.
 Mutable fields in datatypes are not yet supported, but is
@@
 Heterogeneous maps types in Typed Clojure are simple enough to avoid
-the complications associated with concatenable records (section \ref{sec:recordtypes})
+the complications associated with concatenable records (see Section \ref{sec:recordtypes})
 while still supporting many Clojure idioms.

### 2012-10-28 dissertation/intro.tex
Improve Occurrence typing example

@@
 encode implicit type invariants in conditional tests.
-In listing \ref{lst:numvec2} conditional tests are used
-to refine the type of the bindings \lstinline|a| and \lstinline|b|.
-Occurrence typing\cite{TF10} is a technique useful for capturing these kinds
-of type invariants (occurrence typing is discussed in further detail in section \ref{sec:OccurrenceTyping}).
+In listing \ref{lst:numvec2}, conditional tests are used
+to refine the types of the bindings \lstinline|a| and \lstinline|b|.
+Occurrence typing~\cite{TF10} is a technique useful for capturing these kinds
+of type invariants (occurrence typing is discussed in further detail in Section \ref{sec:OccurrenceTyping}).
 
@@
      [(U nil Number) (U nil Number) -> (Vector* Number Number)])
-(defn num-vec2 [a b]
-  [(if a a 0) 
-   (if b b 0)])
+(defn num-vec2 [x y]
+  [(if x x 0) 
+   (if y y 0)])
 \end{lstlisting}
 
-To check this example, occurrence typing infers type information based on the result of each test.
-In Clojure, \lstinline|nil| and \lstinline|false| are false values and all other values are true.
-The test \lstinline|(if a a 0)| follows the \emph{then} branch is \lstinline|a| is not \lstinline|nil|
-or \lstinline|false|. When checking this branch, we can safely refine the type of \lstinline|a| to \lstinline|Number| from
-\lstinline|(U nil Number)|. Similarly, following the \emph{else} branch refines the type of \lstinline|a|
-to \lstinline|nil| from \lstinline|(U nil Number)|.
+To check Listing \ref{lst:numvec2}, occurrence typing infers type information based on the results of conditional tests.
+A conditional in Clojure is written \lstinline|(if test then else)|, where \lstinline|then|
+is executed if \lstinline|test| is a true value, or \lstinline|else| is executed if \lstinline|test|
+is a false value.
+In Clojure, \lstinline|nil| and \lstinline|false| are the only false values: all other values are true values.
+\lstinline|(if x x 0)| in Listing \ref{lst:numvec2} chooses a branch based on the truth value of the test expression \lstinline|x|.
+Before this expression, the binding \lstinline|x| is known to be of type \lstinline|(U nil Number)|.
+There are two cases to consider when checking this conditional: when the test \lstinline|x| is a true value, and when it is a false value.
+If \lstinline|x| is a true value, then the binding \lstinline|x| must be of type \lstinline|Number|.
+If \lstinline|x| is a false value, then the binding \lstinline|x| must be of type \lstinline|nil|.
+Occurrence typing understands these invariants and applies them when checking each branch:
+occurrences of \lstinline|x| in the first branch are considered to be of type \lstinline|Number|, and
+occurrences of \lstinline|x| in the second branch are considered to be of type \lstinline|nil|.
+This infers \lstinline|[(if x x 0) (if y y 0)])| to be of type
+\lstinline|(Vector* Number Number)| (the type for a two-place heterogeneous vector of \lstinline|Number|s).
 

### 2012-10-29 dissertation/design-choices.tex
A bunch of corrections

@@
 
-We handle this by introducing a type \lstinline|(Array c t)|, which is an array has the Java component type \lstinline|c| 
-array that can read and write values of type \lstinline|t|.
+We handle this by introducing a type \lstinline|(Array c t)|, 
+the type of the array that has the Java component type \lstinline|c| 
+and that can read and write values of type \lstinline|t|.
 Internally, \lstinline|Array| types expand to \lstinline|(Array3 c w r)|,
-which is an array has the Java component type \lstinline|c|,
-that can write type \lstinline|w| (which is contravariant in \lstinline|Array|) and read type \lstinline|r| 
-(which is covariant in \lstinline|Array|).
-\lstinline|(Array c t)| expands to \lstinline|(Array3 c t t)|.
+the type of the array that has the Java component type \lstinline|c|,
+that can write values of type \lstinline|w| (which is contravariant in \lstinline|Array3|), 
+and that can read values of type \lstinline|r| (which is covariant in \lstinline|Array3|).
+Concretely, \lstinline|(Array c t)| expands to \lstinline|(Array3 c t t)|.
 
@@
 
-Heterogeneous maps types in Typed Clojure are simple enough to avoid
+Heterogeneous map types in Typed Clojure are simple enough to avoid
 the complications associated with concatenable records (see Section \ref{sec:recordtypes})

### 2012-10-29 dissertation/experiments.tex
A bunch of corrections

@@
 
-In this chapter work through several examples of using Typed Clojure and
-how well the current prototype handles them. 
-We intentially chose examples that could be challenging to type check
+In this chapter we work through several examples of using Typed Clojure and
+gauge how well the current prototype handles them. 
+We intentionally chose examples that could be challenging to type check,
 or were particularly useful to the everyday Clojure programmer.
@@
 
-I ported a function from clojure.contrib.reflect\footnote{https://github.com/cemerick/pomegranate}, 
+I ported a function from clojure.contrib.reflect\footnote{https://github.com/richhickey/clojure-contrib},
 a Clojure library relying heavily on Java interoperability.
@@
 
-The original function is modified slightly for readability and is presented in listing \ref{lst:callmethod}.
+The original function is modified slightly for readability and is presented in Listing \ref{lst:callmethod}.
 
@@
 Thankfully this function has up-to-date documentation, and from it we can derive an expected type
-(listing \ref{lst:callmethodtype}).
+(Listing \ref{lst:callmethodtype}).
 
@@
 
-The final successfully type checked code is presented in listing \ref{lst:callmethodfinal}.
+The final successfully type checked code is presented in Listing \ref{lst:callmethodfinal}.
 
@@
 
-A direct typing of the identity monad would look like listing \ref{lst:identitymtype}.
+A direct typing of the identity monad would look like Listing \ref{lst:identitymtype}.
 This signature assigns the expected type for the var \lstinline|identity-m| 
@@
 
-We can now express the type of the identity monad more abstractly (listing \ref{lst:identityctors}).
+We can now express the type of the identity monad more abstractly (Listing \ref{lst:identityctors}).
 The purpose of the monad seems more apparent just from reading its type.
@@
 
-The final type checked definition for the identity monad is given in listing \ref{lst:identitydeftyped}.
+The final type checked definition for the identity monad is given in Listing \ref{lst:identitydeftyped}.
 In this case, just the monadic bind required annotation.
@@
 The definitions were type checked by adding type annotations in appropriate places, in a similar fashion
-to the identity monad. A significant obstacle was discovered in while type checking monad definitions,
-relating to filtering sequences with negative type information, discussed in section \ref{sec:filterneg}.
+to the identity monad. A significant obstacle was discovered while type checking monad definitions,
+related to filtering sequences with negative type information (discussed in Section \ref{sec:filterneg}).
 
@@
 The initial motivation for adding user defined type constructors to Typed Clojure
-(as discussed in section \ref{sec:monaddef}) was to type check
+(as discussed in Section \ref{sec:monaddef}) was to type check
 monad transformer definitions.
@@
 issues of assigning types to \lstinline|assoc| and \lstinline|partial|
-(which is future work, see section \ref{sec:futurearity}).
+(which is future work, see Section \ref{sec:futurearity}).
 In these cases, simplified versions of the conduits were created that 
@@
 the defined kind must match the declared kind, otherwise a type error is thrown.
-(The specifics of the types in listing \ref{lst:conduittypes} are not relevant to the rest of this discussion,
+(The specifics of the types in Listing \ref{lst:conduittypes} are not relevant to the rest of this discussion,
 so they are not explained).

### 2012-10-29 dissertation/future-work.tex
A bunch of corrections

@@
 It would be essential for an optional static type system for Clojure to support
+multimethods if it is to be of practical use.
 The major hurdle for type checking multimethod definitions is the

### 2012-10-29 dissertation/implementation.tex
A bunch of corrections

@@
 Typed Clojure was built for the Clojure programming language, whose compiler
-and data structures are implementated in Java. Clojurescript is the first major Clojure dialect
+and data structures are implemented in Java. ClojureScript is the first major Clojure dialect
 to be written in Clojure, and it is likely future dialects of Clojure will follow this example. 
 Where Clojure uses Java Classes and Interfaces, 
-Clojurescript's compiler and data structures are written in terms of Clojure's
+ClojureScript's compiler and data structures are written in terms of Clojure's
 two core abstractions: protocols and datatypes.
-It would desirable to port Typed Clojure to similar  Clojure dialects while keeping
+It would be desirable to port Typed Clojure to such Clojure dialects while keeping
 the core of the implementation constant.

### 2012-10-29 dissertation/intro.tex
A bunch of corrections

@@
 dynamically typed languages with static type systems. 
-This is idea not new (see work by Fagan, Cartwright, Aiken and Murphy~\cite{Fag91,CF91,AM91}), but recent attempts
+This idea is not new (see work by Fagan, Cartwright, Aiken and Murphy~\cite{Fag91,CF91,AM91}), but recent attempts
 are noteworthy for their broad success in matching many of the advantages of statically typed
@@
 
-Recently a number of languages have been created to or been modified to support aspects
+Recently a number of languages have been created or modified to support aspects
 of both static and dynamic typing.
@@
 
-It is common in Clojure, like other dynamically-typed languages, to
+It is common in Clojure, like other dynamically typed languages, to
 encode implicit type invariants in conditional tests.

### 2012-10-29 dissertation/lit-review.tex
A bunch of corrections

@@
 than ATT and ATTT.
-In particular, singleton types in ATTT a much more advanced than Typed Clojure,
+In particular, singleton types in ATTT are much more advanced than in Typed Clojure,
 which offers only singleton types for values like \lstinline|:a| and \lstinline|nil|
@@
 
-Other statically typed languages support Java interoperability with
-similar goals, most completely, Scala.
+Other statically typed languages have similar goals
+to support Java interoperability, like, most completely, Scala.
 Scala~\cite{OCD+} is tightly integrated with Java. It manages interactions
@@
 \lstinline|(+ a b)| as \lstinline|Float|. We now have sufficient information to 
-synthesise the type of listing \ref{lst:bidir} to be \lstinline|(List Float)|.
+synthesise the type of listing \ref{lst:bidir} to be \lstinline|(LazySeq Float)|.
 
@@
 for describing ``information flow in polymorphic type systems with propagation-based type inference''~\cite{OZZ01}. 
-Colored Local Type Inference is a candidate for
+Colored Local Type Inference is a candidate for future extensions 
+to Typed Clojure's inference.
 
@@
 Still, unbounded type variables have an advantage: their implementations are
-are simpler in the presence of a \emph{Bottom} type. 
+simpler in the presence of a \emph{Bottom} type. 
 The constraint resolution algorithm for bounded variables
@@
 Typed Racket is a statically typed sister language of Racket. It
-attempts to preserve existing Racket idioms and aims type check
+attempts to preserve existing Racket idioms and aims to type check
 existing Racket code by simply adding top level type annotations~\cite{Tob10}.

### 2012-10-29 dissertation/paper.tex
A bunch of corrections

@@
 This dissertation describes an optional static typing system
-that attempts to bring the advantages of static typing to the dynamically type language
+that attempts to bring the advantages of static typing to the dynamically typed language
 Clojure, running on the Java Virtual Machine. 

### 2012-11-01 dissertation/experiments.tex
Wording

@@
 
-As Typed Clojure is intended to be useful for practical purposes, it was important
-to port existing code that utilized Java interoperability.
-My porting effort attempted to follow how a real programmer might port code to Typed Clojure
-and I describe my approach for each situation.
+It is idiomatic and common in Clojure to interface with existing Java code
+via Java interoperability.
+Typed Clojure is intended to be useful for practical purposes, so it is important
+to understand this common feature.
+To test Typed Clojure's approach, 
+existing code utilising Java interoperability was ported.
+My porting effort attempted to follow how a real programmer might port code to Typed Clojure;
+I describe this process step-by-step.
 
-I ported a function from clojure.contrib.reflect\footnote{https://github.com/richhickey/clojure-contrib},
-a Clojure library relying heavily on Java interoperability.
-I chose \lstinline|call-method| for several reasons: it chains several Java calls together,
-it uses primitive arrays, it exposes a wart of Clojure with \emph{Named} things,
-and \lstinline|null| is a valid value in one place.
+This section describes porting a function from \emph{clojure.contrib.reflect}
+\footnote{\emph{clojure.contrib} Github project: https://github.com/richhickey/clojure-contrib},
+a Clojure library that relies heavily on Java interoperability.
+I chose to port \lstinline|call-method| for several reasons: it chains several Java calls together,
+it uses primitive arrays, and \lstinline|null| is a valid value in one place.
 
 Before showing the implementation, there is a brief explanation of the relevant syntax.
-Java methods are called using the \emph{dot} operator. If \lstinline|o| is an object \lstinline|(. o m a*)|
-calls its method named \lstinline|m| passing \lstinline|a*| arguments. The method can be named first
-using the equivalent sugar \lstinline|(.m o a*)|. Also, \lstinline|doto| is convenient 
-notation for muliple method calls on the same object, presumably for side effects, and returns the original
-object. For example, \lstinline|(doto o (.m1 a*) (m2 a*))| calls methods \lstinline|m1| and \lstinline|m2|
-on \lstinline|o| and returns \lstinline|o|.
+Java methods are called using the \emph{dot} operator. If \lstinline|o| is an object,
+then \lstinline[mathescape]|(. o m $\overrightarrow{a}$)|
+calls its method named \lstinline|m|, passing \lstinline[mathescape]|$\overrightarrow{a}$| as arguments. 
+Syntactic sugar allows the method to be named first:
+\lstinline[mathescape]|(.m o $\overrightarrow{a}$)|
+is equivalent to
+\lstinline[mathescape]|(. o m $\overrightarrow{a}$)|.
+Also, \lstinline|doto| is convenient 
+notation for chaining multiple method calls to the same object, presumably for side effects, and returns the original
+object. For example, \lstinline[mathescape]|(doto o (.m1 $\overrightarrow{a}$) (.m2 $\overrightarrow{b}$))| 
+calls methods \lstinline|m1| (passing \lstinline[mathescape]|$\overrightarrow{a}$| as arguments)
+and \lstinline|m2| (passing \lstinline[mathescape]|$\overrightarrow{b}$| as arguments)
+on \lstinline|o| in order, and returns \lstinline|o| with any side effects applied to it.
 
@@
 
+In particular we change two invocations of \lstinline|into-array|.
+
 \begin{itemize}
-\item \lstinline|(into-array Class params)| becomes \lstinline|(into-array> Class Class params)|
+\item \lstinline|(into-array Class params)|, which creates an array of type \lstinline|Class[]| and populates it
+with the elements of \lstinline|params|,
+ becomes \lstinline|(into-array> Class Class params)|,
       which is of type \lstinline|(Array Class Class)|.
-\item \lstinline|(into-array Object args)| becomes \lstinline|(into-array> Object (U nil Object) args)|
+\item \lstinline|(into-array Object args)|, which creates an array of type \lstinline|Object[]|
+and populates it with the elements of \lstinline|args|,
+becomes \lstinline|(into-array> Object (U nil Object) args)|
       which is of type \lstinline|(Array Object (U nil Object))|.

### 2012-11-04 poster/poster.tex
Poster draft

\documentclass[landscape,final,a0paper,fontscale=0.25]{baposter}

\usepackage{url}

\usepackage{graphicx}
\graphicspath{{images/}{../images/}}

\usepackage{listings}
\lstset{ %
  language=Lisp,                % choose the language of the code
  columns=fixed,basewidth=.5em,
  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
  %numbers=left,                   % where to put the line-numbers
  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
  %numbersep=5pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
  %showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  %showtabs=false,                 % show tabs within strings adding particular underscores
  frame=single,           % adds a frame around the code
  %tabsize=2,          % sets default tabsize to 2 spaces
  captionpos=t,           % sets the caption-position to bottom
  breaklines=true,        % sets automatic line breaking
  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
}

\begin{document}

\definecolor{lightorange}{rgb}{0.9,0.4,0}
\definecolor{lightestorange}{rgb}{1,0.8,0.5}
\definecolor{darkorange}{rgb}{0.2,0.1,0}
%%
\begin{poster}%
  % Poster Options
  {
  % Show grid to help with alignment
  grid=false,
  % Column spacing
  colspacing=1em,
  % Color style
  bgColorOne=lightestorange,
  bgColorTwo=white,
  borderColor=darkorange,
  headerColorOne=darkorange,
  headerColorTwo=lightorange,
  headerFontColor=white,
  boxColorOne=lightestorange,
  boxColorTwo=lightorange,
  columns=3,
  % Format of textbox
  textborder=faded,
  % Format of text header
  eyecatcher=true,
  headerborder=closed,
  headerheight=0.1\textheight,
%  textfont=\sc, An example of changing the text font
  headershape=roundedright,
  headershade=shadelr,
  headerfont=\Large\bf\textsc, %Sans Serif
  textfont={\setlength{\parindent}{1.5em}},
  boxshade=plain,
%  background=shade-tb,
  background=none,
  linewidth=2pt
  }
  % Eye Catcher
  {\includegraphics{images/1000px-UWA_crest}}

  % Title
  { A Practical Optional Type System for Clojure }
  % Authors
  { Ambrose Bonnaire-Sergeant <abonnairesergeant@gmail.com>\\
    \emph{Supervised by Rowan Davies <rowan@csse.uwa.edu.au>}}
  % University logo
  { University of Western Australia }
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Now define the boxes that make up the poster
%%%---------------------------------------------------------------------------
%%% Each box has a name and can be placed absolutely or relatively.
%%% The only inconvenience is that you can only specify a relative position 
%%% towards an already declared box. So if you have a box attached to the 
%%% bottom, one to the top and a third one which should be in between, you 
%%% have to specify the top and bottom boxes before you specify the middle 
%%% box.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{posterbox}[name=problem,column=0,row=0]{Background}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Static and Dynamic Types}

Traditionally there has been a clear distinction between 
programming languages that perform type checking at compile-time, and those that do not.
A recent trend is to aim to combine the advantages of both kinds of languages by adding optional static 
type systems to languages without static type checking.

\section*{Clojure}

Clojure is a recent and increasingly popular programming language that a is dynamically typed dialect of Lisp.
Designed to be a pragmatic language, it supports functional programming and immutability by default,
while being hosted on, and providing direct interoperability with popular platforms.
\end{posterbox}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{posterbox}[name=contribution,column=0,below=problem]{Contributions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We take the lessons learnt from several projects to design \emph{Typed Clojure},
an optional type system for Clojure running on the JVM. 
The main contributions are:
\begin{itemize}
  \item We develop a prototype type checker based on Typed Racket, which is able to type check many Clojure idioms
  \item We describe a novel use of \emph{occurrence typing} for eliminating occurrences of Java's \emph{null}
  \item We show how a combination of occurrence typing and intersection types allows accurrate typing checking of
        the most common usages of Clojure's sequence abstraction 
  \item We capture the Clojure idiom of using maps with known keyword fields by using heterogeneous map types
  \item We identify the main future issues in typing Clojure code.

\end{itemize}

\end{posterbox}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{posterbox}[name=sequenceabs,column=1]{Sequence Abstraction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Some common idioms that Typed Clojure types can express
are usages of Clojure's sequence abstraction, a set of functions providing a common
interface to collections.
These functions include:

\begin{itemize}
\item \lstinline|seq|, which takes a collection (or \lstinline|nil|) and returns
   \begin{itemize}
      \item a true value if the argument is non-empty
      \item a false value if the argument is empty (or \lstinline|nil|)
   \end{itemize}
\item \lstinline|first|, which takes a collection (or \lstinline|nil|) and returns
   \begin{itemize}
      \item the first item of the argument if the argument is non-empty
      \item \lstinline|nil| if the argument is empty (or \lstinline|nil|)
    \end{itemize}
\end{itemize}

By encoding these invariants in the respective types, 
we can type check common combinations of these functions.

\begin{lstlisting}[caption={Assuming \lstinline|coll| is a collection of numbers, we can infer that this expression always returns a number}]
(if (seq coll)
  (first coll)
  0)
\end{lstlisting}

\end{posterbox}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{posterbox}[name=hmaps,column=1,below=sequenceabs]{Heterogeneous Maps}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Using plain maps with known keyword keys is common style in Clojure.
Heterogeneous map types that record the \emph{presence} of keyword keys
are sufficient to type check many usages of this idiom.
Accurate types for lookup, associating, and dissociating of known keyword keys are supported.

\end{posterbox}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{posterbox}[name=nulltreatment,column=1,below=hmaps]{Treatment of \emph{null}}

Clojure represents Java's \emph{null} as the value \lstinline|nil|,
which has the singleton type \lstinline|nil| in Typed Clojure.
This means, unlike Java's static type system, \emph{null} and reference types are separated as static types.
The programmer can then annotate Java methods and fields, specifying where \emph{null} is allowed.

\end{posterbox}

\begin{posterbox}[name=experiments,column=2]{Experiments}

\section*{Monads}

Almost all of a Clojure library for monadic programming (\emph{algo.monads}) was successfully ported
to Typed Clojure. This included most:
\begin{itemize}
  \item Monad definitions
  \item Monad function definitions
  \item Monad transformer defintions
\end{itemize}

This experiment warranted an extension for type functions (functions at the type level), with variance.

\section*{Java Interoperability}

A function relying heavily on Java interoperability was ported successfully.
This function chained several Java calls together, and
Typed Clojure was able to express and check invariants, mostly related to occurrences of \emph{null}
and primitive arrays.

\end{posterbox}

\begin{posterbox}[name=futurework,column=2,below=experiments]{Future Work}

There is significant future work to type check all Clojure idioms:

\begin{itemize}
\item Multimethod definitions should infer types from their dispatch function
\item Implement a blame calculus to improve error messages when using untyped code
\end{itemize}

\end{posterbox}

\begin{posterbox}[name=conclusion,column=2,below=futurework]{Conclusion}
This work has demonstrated that it is both practical and useful to
design and implement an optional type system for Clojure that preserves
current Clojure style.
\end{posterbox}

\begin{posterbox}[name=source,column=2,above=bottom]{Source}
Typed Clojure is available at: \\
\url{https://github.com/frenchy64/typed-clojure}
\end{posterbox}

\end{poster}

\end{document}

### 2012-11-04 poster/poster.tex
Last changes to poster

@@
-\documentclass[landscape,final,a0paper,fontscale=0.25]{baposter}
+\documentclass[landscape,final,a0paper,fontscale=0.277]{baposter}
 
 \usepackage{url}
+\usepackage{relsize}
+
+\usepackage{color} % for colorbox
+%\usepackage[svgnames]{xcolor}
 
 \usepackage{graphicx}
-\graphicspath{{images/}{../images/}}
+\graphicspath{{images/}}
+
+\definecolor{codebg}{HTML}{EEEEEE}
 
@@
   %numbersep=5pt,                  % how far the line-numbers are from the code
-  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
+  backgroundcolor=\color{codebg},  % choose the background color. You must add \usepackage{color}
   %showspaces=false,               % show spaces adding particular underscores
@@
 \definecolor{darkorange}{rgb}{0.2,0.1,0}
+\definecolor{ForestGreen}{rgb}{0.13, 0.54, 0.13}
+\definecolor{navyblue}{rgb}{0,0,0.5}
 %%
@@
   borderColor=darkorange,
-  headerColorOne=darkorange,
-  headerColorTwo=lightorange,
+  headerColorOne=ForestGreen,
+  headerColorTwo=navyblue,
   headerFontColor=white,
-  boxColorOne=lightestorange,
-  boxColorTwo=lightorange,
+  boxColorOne=green,
+  boxColorTwo=blue,
   columns=3,
@@
   % Eye Catcher
-  {\includegraphics{images/1000px-UWA_crest}}
-
+  {\includegraphics[height=7em]{images/clojure-logo10.png}} 
   % Title
-  { A Practical Optional Type System for Clojure }
+  {\bf\textsc{A Practical Optional Type System for Clojure}}
   % Authors
-  { Ambrose Bonnaire-Sergeant <abonnairesergeant@gmail.com>\\
-    \emph{Supervised by Rowan Davies <rowan@csse.uwa.edu.au>}}
+  %{\textsc{\{ Brian.Amberg and Thomas.Vetter \}@unibas.ch}}
+  { \textsc{Ambrose Bonnaire-Sergeant <abonnairesergeant@gmail.com>}
+    \\
+    \textsc{{Supervised by Rowan Davies <rowan@csse.uwa.edu.au>}}}
   % University logo
-  { University of Western Australia }
+  {% The makebox allows the title to flow into the logo, this is a hack because of the L shaped logo.
+    \includegraphics[height=6.0em]{images/1000px-UWA_crest.png}
+  }
+%
+%  % Eye Catcher
+%  {\includegraphics{images/1000px-UWA_crest}}
+%   
+%  % Title
+%  { \textsc}
+%  % Authors
+%  % University logo
+%  { University of Western Australia }
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\section*{Static and Dynamic Types}
+\section*{\textsc{Static and Dynamic Types}}
 
@@
 
-\section*{Clojure}
+\section*{\textsc{Clojure}}
 
-Clojure is a recent and increasingly popular programming language that a is dynamically typed dialect of Lisp.
-Designed to be a pragmatic language, it supports functional programming and immutability by default,
+Clojure is a recent and increasingly popular dynamically typed programming language.
+Designed to be pragmatic, it is a dialect of Lisp supporting functional programming and immutability by default,
 while being hosted on, and providing direct interoperability with popular platforms.
@@
 We take the lessons learnt from several projects to design \emph{Typed Clojure},
-an optional type system for Clojure running on the JVM. 
+an optional type system for Clojure running on the Java Virtual Machine. 
 The main contributions are:
@@
   \item We describe a novel use of \emph{occurrence typing} for eliminating occurrences of Java's \emph{null}
-  \item We show how a combination of occurrence typing and intersection types allows accurrate typing checking of
-        the most common usages of Clojure's sequence abstraction 
-  \item We capture the Clojure idiom of using maps with known keyword fields by using heterogeneous map types
-  \item We identify the main future issues in typing Clojure code.
+  \item We show how a combination of occurrence typing and intersection types allows accurate typing checking of
+        the most common usages of Clojure's \emph{sequence abstraction} 
+  \item We capture the Clojure idiom of using maps with known keyword fields by using \emph{heterogeneous map types}
+  \item We identify the main \emph{future issues} in typing Clojure code.
 
@@
 
+\end{posterbox}
+
+\begin{posterbox}[name=nulltreatment,column=0,above=bottom]{Treatment of \emph{null}}
+
+Clojure represents Java's \emph{null} as the value \lstinline{nil},
+which has the singleton type \colorbox{yellow}{\lstinline|nil|} in Typed Clojure.
+This means, unlike Java's static type system, \emph{null} and reference types are separated as static types.
+The programmer can then annotate Java methods and fields, specifying where \emph{null} is allowed.
+
+\begin{lstlisting}[caption={Annotating Java methods that never return \emph{null}}]
+(non-nil-return java.lang.Object/getClass :all)
+(non-nil-return java.io.File/getName :all)
+\end{lstlisting}
 \end{posterbox}
@@
 
-\begin{lstlisting}[caption={Assuming \lstinline|coll| is a collection of numbers, we can infer that this expression always returns a number}]
+\begin{lstlisting}[caption={Assuming \lstinline|coll| is a collection of numbers currently in local scope, we can infer that this expression always returns a number}]
 (if (seq coll)
@@
 \end{lstlisting}
-
 \end{posterbox}
@@
 
-Using plain maps with known keyword keys is common style in Clojure.
 Heterogeneous map types that record the \emph{presence} of keyword keys
-are sufficient to type check many usages of this idiom.
-Accurate types for lookup, associating, and dissociating of known keyword keys are supported.
+are sufficient to type check the most common Clojure idiom of using plain maps with known keyword keys.
+
+\begin{description}
+  \item[Map Literals] \colorbox{codebg}{\lstinline|\{:a 1 :b 2\}|} creates a map value with two entries, and
+        is of the heterogeneous map type \colorbox{yellow}{\lstinline{'\{:a Number :b Number\}}}
+  \item[Associating keyword keys] \colorbox{codebg}{\lstinline|(assoc \{:a 1 :b 2\} :c 3)|} 
+    associates entry with key \lstinline|:c|, and
+        is of type \colorbox{yellow}{\lstinline{'\{:a Number :b Number :c Number\}}}
+  \item[Dissociating keyword keys] \lstinline|(dissoc {:a 1 :b 2} :a)| dissociates entry with key \lstinline|:a|, and
+        is of type \colorbox{yellow}{\lstinline|'\{:b Number\}|}
+  \item[Keyword lookup] \lstinline|(get {:a 1 :b 2} :a)| returns the value corresponding to the key \lstinline|:a|, and
+        is of type \colorbox{yellow}{\lstinline|Number|}
+\end{description}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{posterbox}[name=nulltreatment,column=1,below=hmaps]{Treatment of \emph{null}}
 
-Clojure represents Java's \emph{null} as the value \lstinline|nil|,
-which has the singleton type \lstinline|nil| in Typed Clojure.
-This means, unlike Java's static type system, \emph{null} and reference types are separated as static types.
-The programmer can then annotate Java methods and fields, specifying where \emph{null} is allowed.
-
+\begin{posterbox}[name=source,column=2]{Source}
+Typed Clojure is available at:\\\url{https://github.com/frenchy64/typed-clojure}
 \end{posterbox}
 
-\begin{posterbox}[name=experiments,column=2]{Experiments}
+\begin{posterbox}[name=experiments,column=2,below=source]{Experiments}
 
-\section*{Monads}
+\section*{\textsc{Monads}}
 
 Almost all of a Clojure library for monadic programming (\emph{algo.monads}) was successfully ported
-to Typed Clojure. This included most:
-\begin{itemize}
-  \item Monad definitions
-  \item Monad function definitions
-  \item Monad transformer defintions
-\end{itemize}
+to Typed Clojure, including most monad, monad function, and monad transformer definitions.
+This experiment warranted an extension for type functions (functions at the type level) with variance.
 
-This experiment warranted an extension for type functions (functions at the type level), with variance.
-
-\section*{Java Interoperability}
+\section*{\textsc{Java Interoperability}}
 
@@
 
-There is significant future work to type check all Clojure idioms:
+There is significant future work, including:
 
 \begin{itemize}
-\item Multimethod definitions should infer types from their dispatch function
 \item Implement a blame calculus to improve error messages when using untyped code
+\item Support type checking multimethod definitions
+\item Support Clojure records, which are datatypes with known keyword keys
+\item Port Typed Clojure to other Clojure implementations
 \end{itemize}
@@
 
-\begin{posterbox}[name=source,column=2,above=bottom]{Source}
-Typed Clojure is available at: \\
-\url{https://github.com/frenchy64/typed-clojure}
+\begin{posterbox}[name=references,column=2,below=conclusion]{References}
+  \smaller
+  \bibliographystyle{ieee}
+  \renewcommand{\section}[2]{\vskip 0.05em}
+  \begin{thebibliography}{1}\itemsep=-0.01em
+    \setlength{\baselineskip}{0.4em}
+    \bibitem{tob10:bnb}
+    Tobin-Hochstadt, S.
+    \newblock {T}yped {S}cheme: {F}rom {S}cripts to {P}rograms (2010)
+    \newblock PhD Dissertation,
+    \newblock Northeastern University
+  \end{thebibliography}
+  \vspace{0.3em}
+
 \end{posterbox}

### 2012-11-04 poster/poster.tex
Small edits to poster

@@
 
-Clojure represents Java's \emph{null} as the value \lstinline{nil},
+Clojure represents Java's \emph{null} as the value \colorbox{codebg}{\lstinline{nil}},
 which has the singleton type \colorbox{yellow}{\lstinline|nil|} in Typed Clojure.
@@
 
-Some common idioms that Typed Clojure types can express
-are usages of Clojure's sequence abstraction, a set of functions providing a common
-interface to collections.
+Typed Clojure types can express the most common usages of Clojure's sequence abstraction, 
+which is a set of functions providing a common interface to collections.
 These functions include:
@@
 \begin{itemize}
-\item \lstinline|seq|, which takes a collection (or \lstinline|nil|) and returns
+\item \colorbox{codebg}{\lstinline|seq|}, which takes a collection (or \colorbox{codebg}{\lstinline|nil|}) and returns
    \begin{itemize}
       \item a true value if the argument is non-empty
-      \item a false value if the argument is empty (or \lstinline|nil|)
+      \item a false value if the argument is empty (or \colorbox{codebg}{\lstinline|nil|})
    \end{itemize}
-\item \lstinline|first|, which takes a collection (or \lstinline|nil|) and returns
+\item \colorbox{codebg}{\lstinline|first|}, which takes a collection (or \colorbox{codebg}{\lstinline|nil|}) and returns
    \begin{itemize}
       \item the first item of the argument if the argument is non-empty
-      \item \lstinline|nil| if the argument is empty (or \lstinline|nil|)
+      \item \colorbox{codebg}{\lstinline|nil|} if the argument is empty (or \colorbox{codebg}{\lstinline|nil|})
     \end{itemize}
@@
 Heterogeneous map types that record the \emph{presence} of keyword keys
-are sufficient to type check the most common Clojure idiom of using plain maps with known keyword keys.
+are sufficient to type check the most common usages of the Clojure idiom of using plain maps with known keyword keys.
 
@@
   \item[Associating keyword keys] \colorbox{codebg}{\lstinline|(assoc \{:a 1 :b 2\} :c 3)|} 
-    associates entry with key \lstinline|:c|, and
+      extends the first argument with a new entry with key \lstinline|:c|, and
         is of type \colorbox{yellow}{\lstinline{'\{:a Number :b Number :c Number\}}}
-  \item[Dissociating keyword keys] \lstinline|(dissoc {:a 1 :b 2} :a)| dissociates entry with key \lstinline|:a|, and
+  \item[Dissociating keyword keys] \colorbox{codebg}{\lstinline|(dissoc \{:a 1 :b 2\} :a)|} dissociates 
+      from the first argument the entry with key \lstinline|:a|, and
         is of type \colorbox{yellow}{\lstinline|'\{:b Number\}|}
-  \item[Keyword lookup] \lstinline|(get {:a 1 :b 2} :a)| returns the value corresponding to the key \lstinline|:a|, and
+  \item[Keyword lookup] \colorbox{codebg}{\lstinline|(get \{:a 1 :b 2\} :a)|} 
+      returns the value corresponding to the key \lstinline|:a| in the first argument, and
         is of type \colorbox{yellow}{\lstinline|Number|}
@@
 This function chained several Java calls together, and
-Typed Clojure was able to express and check invariants, mostly related to occurrences of \emph{null}
+Typed Clojure was able to express and check invariants related to occurrences of \emph{null}
 and primitive arrays.
@@
 
-There is significant future work, including:
+There is significant future work.
 
@@
     Tobin-Hochstadt, S.
-    \newblock {T}yped {S}cheme: {F}rom {S}cripts to {P}rograms (2010)
+    \newblock {T}yped {S}cheme: {F}rom {S}cripts to {P}rograms
     \newblock PhD Dissertation,
-    \newblock Northeastern University
+    \newblock Northeastern University, 2010
   \end{thebibliography}

### 2012-11-05 poster/poster.tex
Prescreened poster
Final poster, printed

@@
   %{\textsc{\{ Brian.Amberg and Thomas.Vetter \}@unibas.ch}}
-  { \textsc{Ambrose Bonnaire-Sergeant <abonnairesergeant@gmail.com>}
+  { \textsc{Ambrose Bonnaire-Sergeant <abonnairesergeant@gmail.com> (20350292)}
     \\
@@
   {% The makebox allows the title to flow into the logo, this is a hack because of the L shaped logo.
-    \includegraphics[height=6.0em]{images/1000px-UWA_crest.png}
+    \includegraphics[width=14em]{images/UWA_logo.png}
   }
@@
 \begin{itemize}
-  \item We develop a prototype type checker based on Typed Racket, which is able to type check many Clojure idioms
-  \item We describe a novel use of \emph{occurrence typing} for eliminating occurrences of Java's \emph{null}
+  \item We develop a prototype type checker based on Typed Racket [1], which is able to type check many Clojure idioms
+  \item We describe a novel use of \emph{occurrence typing} for eliminating erroneous access to Java's \emph{null} pointer
   \item We show how a combination of occurrence typing and intersection types allows accurate typing checking of
@@
 
+  \textsc{\\\smaller Bachelor of Computer Science Honours (Semester 1, 2012)}
 \end{posterbox}

### 2012-11-05 poster/poster.tex
Corrected comma - Edmund Jackson

@@
 Designed to be pragmatic, it is a dialect of Lisp supporting functional programming and immutability by default,
-while being hosted on, and providing direct interoperability with popular platforms.
+while being hosted on, and providing direct interoperability with, popular platforms.
 \end{posterbox}

### 2012-11-24 dissertation/intro.tex
Correct typos in dissertation
Fix defn -> def explanation

@@
 
-(\lstinline|(defn [...] body)| expands to \lstinline|(def (fn [...] body))|
-where \lstinline|def| defines a new var and \lstinline|fn| creates a function value).
+(\lstinline|(defn name [...] body)| expands to \lstinline|(def name (fn [...] body))|,
+where \lstinline|(def name init)| defines a new var in the current namespace 
+called \lstinline|name|, and \lstinline|fn| creates a function value).
 
@@
 singleton types in Typed Clojure.
-Singleton types are are discussed further in section \ref{sec:intersectionunion}.
+Singleton types are discussed further in section \ref{sec:intersectionunion}.
 

### 2012-11-26 dissertation/design-choices.tex
Correct typos in dissertation
Chris' corrections

@@
 To this end, many of the design choices are borrowed from other existing projects.
-Some of the main design choices that specifically concern Typed Clojure are given more
+Some of the main design choices that specifically concern Typed Clojure are given
 more detailed descriptions.
@@
 
-\begin{lstlisting}[caption=Type assiciated with clojure.core/seq, label=lst:seqtype]
+\begin{lstlisting}[caption=Type associated with clojure.core/seq, label=lst:seqtype]
 (All [x]
@@
 
-Correspondingly, \lstinline|first| which gets the first member of a sequence.
+Correspondingly, \lstinline|first| which returns the first member of a sequence,
 can be given a more accurate (and surprisingly subtle) type.
@@
 
-\begin{lstlisting}[caption=Type assiciated with clojure.core/first, label=lst:firsttype]
+\begin{lstlisting}[caption=Type associated with clojure.core/first, label=lst:firsttype]
 (All [x]
@@
 Primitive arrays in Java are covariant, which is well-known to be statically unsound~\cite{OW97,BK99}.
-This mean that at compile time we cannot completely trust the type signature of any Java method or field that involves 
+This means that at compile time we cannot completely trust the type signature of any Java method or field that involves 
 an array.
@@
 Protocols and datatypes are two of Clojure's main means of abstraction.
-Protocols, have a set of unimplemented methods, are typed by
-giving the type for each method.
+Protocols define a set of unimplemented methods, and each must be assigned types.
 Datatype definitions are type checked by ensuring the methods it
-implements match the types defined by the Java interface or protocol
+implements match the types defined by the Java interface or assigned to the protocol method
 it is extending.

### 2012-11-26 dissertation/experiments.tex
Chris' corrections

@@
 ensures the correct number of arguments are passed and recognises the declared variances
-for each parameter, In this case the first argument of \lstinline|m| is declared
+for each parameter. In this case the first argument of \lstinline|m| is declared
 a covariant position. When instantiated, \lstinline|m| must also be a type operator

### 2012-11-26 dissertation/intro.tex
Chris' corrections

@@
 (ns typed.test.hello-world
-  (:require [typed.core :refer [check-ns]]]))
+  (:require [typed.core :refer [check-ns]]))
 
@@
 
-uses \lstinline|typed.core/ann-form| to assign an expected to
+uses \lstinline|typed.core/ann-form| to assign an expected type to
 the first argument,
@@
 Such annotations are in a sense assumptions; should they turn out to be wrong, Typed Clojure
-will infer incorrect types. Typed Racket generates runtime contracts based on static types to ensure
-that type errors are caught at runtime, and the original source of the error is blamed~\cite{WF09}.
+will infer incorrect types. 
+This is somewhat troubling, but type safety can be restored by using other techniques.
+For example, Typed Racket generates runtime contracts based on static types to ensure
+that type errors are caught at runtime, while giving good error messages 
+which indicate (or ``blame''~\cite{WF09}) the original source of the error.
 A similar system for verifying that method annotations like \lstinline|non-nil-return|
-actually describe correct types is planned as future work
+are correct is planned as future work
 (see section \ref{sec:contractsblame}).

### 2012-11-26 dissertation/lit-review.tex
Chris' corrections

@@
 because it is too inconvenient in real programs.
-He claims to have fixed these issues, but a practical language has yet to integrated 
+He claims to have fixed these issues, but a practical language has yet to integrate
 his changes.
@@
 
-Gradual typing combines static and runtime type checking so programmers
-can the most appropriate one for the situation.
+Gradual typing combines static and runtime type checking, so programmers
+can utilise whichever is most appropriate for a given situation.
 A key feature of gradual type systems is its use of \emph{bidirectional checking}
@@
 
-SML CIDRE \cite{Dav05} was also a secondary influence - particularly it's use of bidirectional 
-checking of extrinsic types including subtyping, intersection types, and it's aim to capture program invariants.  
+SML CIDRE \cite{Dav05} was also a secondary influence - particularly its use of bidirectional 
+checking of extrinsic types including subtyping, intersection types, and its aim to capture program invariants.  
 However, SML CIDRE starts from the intrinsically typed language SML which already has a rich static 

### 2012-11-26 dissertation/paper.tex
Chris' corrections

@@
 Similarly, we show that Typed Clojure helps verify code using Java interoperability is correct, and is
-a useful tool to show misusing of Java's \emph{null}.
+a useful tool to show misuse of Java's \emph{null}.
 

### 2012-12-06 dissertation/design-choices.tex
Lyndon corrections + others

@@
 plays a major role in capturing many common Clojure idioms.
-One such idiom is the \emph{sequence abstraction}, which
-consists of the \lstinline|clojure.core| functions \lstinline|seq|, \lstinline|first|, and \lstinline|rest|.
+One such idiom is Clojure's \emph{sequence abstraction}, which
+provides a common interface to all collections.
+For example, \lstinline|seq| tests whether a collection is empty,
+\lstinline|first| returns the first element of a collection,
+and \lstinline|rest| returns the tail of its collection argument as a new sequence.
+While these summaries describe their most common usages, they do not tell the whole story.
 
-\lstinline|seq| takes an argument that is either a sequence or \lstinline|nil|,
-and returns a non-empty sequence if passed a non-empty sequence, or returns
-\lstinline|nil| if passed an empty sequence or \lstinline|nil|.
-It is a common idiom to test \lstinline|(seq coll)| in a conditional
-to determine whether the sequence \lstinline|coll| is empty:
-if \lstinline|(seq coll)| is a true value, then \lstinline|coll|
-is non-empty, otherwise it should be considered empty.
-Occurrence typing allows us to capture this idiom as shown in
-listing \ref{lst:seqtype}.
+\lstinline|seq|'s description might suggest it is better named \lstinline|not-empty?|,
+but \lstinline|seq| is perhaps surprising in two main ways: 
+it accepts \lstinline|nil| as an argument (returning \lstinline|nil|),
+and, rather than returning \lstinline|true| or \lstinline|false|, returns
+a non-empty sequence or \lstinline|nil| respectively.
+
+From a typing perspective, the subtlety of \lstinline|seq| is expressing that \lstinline|(seq coll)|
+
+\begin{itemize}
+  \item is a true value when \lstinline|coll| is a non-empty collection, and
+  \item is a false value when \lstinline|coll| is either \lstinline|nil| or an empty collection.
+\end{itemize}
+
+Occurrence typing allows us to capture this idiom (Listing \ref{lst:seqtype}).
 
@@
 
-This function type takes an argument of type \lstinline|(U nil (Seqable x))| (the syntax \lstinline[mathescape]{(U $\overrightarrow{t}$)}
+This function type (Listing \ref{lst:seqtype}) 
+takes an argument of type \lstinline|(U nil (Seqable x))| (the syntax \lstinline[mathescape]{(U $\overrightarrow{t}$)}
 constructs a union of types \lstinline|t|), and returns a type \lstinline|(U nil (Seqable x))|.
@@
 
-Correspondingly, \lstinline|first| which returns the first member of a sequence,
-can be given a more accurate (and surprisingly subtle) type.
-\lstinline|first| is overloaded in a similar fashion to \lstinline|seq|.
-\lstinline|first| takes an argument that is either a sequence or \lstinline|nil|.
-If it is passed \lstinline|nil| or an empty sequence, it returns \lstinline|nil|. Otherwise, if it is passed
-non-empty sequence, it returns the first member of the sequence.
+\lstinline|first| is overloaded in a similar fashion to \lstinline|seq|:
+\lstinline|(first coll)| returns
+
+\begin{itemize}
+  \item \lstinline|nil| if \lstinline|coll| is \lstinline|nil| or an empty collection, and
+  \item returns the first item of the collection if \lstinline|coll| is a non-empty collection.
+\end{itemize}
+
 Listing \ref{lst:firsttype} shows the type of \lstinline|first|.
+The main thing to grasp from this definition is that three function types are given:
+the first type handling the first case above, the second type handling the second case above,
+and the third arity acting as a base case, general enough to catch both cases.
 
@@
 
-The function type has three arities, the last of which can be considered a ``default'' arity.
-The first two arities correspond exactly to the two cases described for when 
-\lstinline|first| is passed an empty sequence or \lstinline|nil|, and
-when \lstinline|first| is passed a non-empty sequence
-(the syntax \lstinline[mathescape]{(I $\overrightarrow{t}$)}
-constructs an intersection of types \lstinline|t|).
-
-These two functions interact via occurrence typing.
-For example, listing \ref{lst:seqocc} tests
+\lstinline|seq| and \lstinline|first| interact via occurrence typing.
+For example, Listing \ref{lst:seqocc} tests
 \lstinline|(seq coll)| which means occurrences of \lstinline|coll|
@@
 which has been implemented in Typed Racket.
-Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
-as described in Section \ref{sec:impvariablearity}
+Typed Clojure directly ports most of Typed Racket's implementation of variable-arity polymorphism,
+as described in Section \ref{sec:impvariablearity}.
 
 Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
-and \lstinline|partial| cannot be typed by Strickland et al.\ .
+and \lstinline|partial| cannot be typed by Strickland \emph{et al.}.
 Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
@@
 Notably, the component type \lstinline|c| seems redundant, and Typed Clojure currently
-does not provide syntax to ``cast'' an unwriteable array to a be writeable.
+does not provide syntax to ``cast'' an unwriteable array to be writeable.
 
@@
 For example, we currently do not know how to infer \lstinline|(filter identity coll)|
-as returning a sequence of non-\lstinline|nil| elements, for any sequence \lstinline|coll|
+as returning a sequence with the elements of \lstinline|coll| that are not \lstinline|nil| or \lstinline|false|,
+for any sequence \lstinline|coll|
 (see Section \ref{sec:filterneg}).
@@
 Protocols define a set of unimplemented methods, and each must be assigned types.
-Datatype definitions are type checked by ensuring the methods it
-implements match the types defined by the Java interface or assigned to the protocol method
-it is extending.
+Datatype definitions are type checked by ensuring the methods they
+implement match the types defined by the Java interface or assigned to the protocol method
+they are extending.
 

### 2012-12-06 dissertation/experiments.tex
Lyndon corrections + others

@@
 This section describes porting a function from \emph{clojure.contrib.reflect}
-\footnote{\emph{clojure.contrib} Github project: https://github.com/richhickey/clojure-contrib},
+\footnote{\emph{clojure.contrib} Github project: https://github.com/richhickey/clojure-contrib}
+\footnote{Note that while the \emph{clojure.contrib} library as a whole should be considered deprecated, useful
+code still exists within it which apparently has not been migrated elsewhere.},
 a Clojure library that relies heavily on Java interoperability.
@@
 The final successfully type checked code is presented in Listing \ref{lst:callmethodfinal}.
-
+\\\\
+% bump to next page
 \begin{lstlisting}[caption=Type Annotated code for call-method, label=lst:callmethodfinal]
@@
 It is a particularly good experiment for SML CIDRE because it generates an unusually large
-amount of intersection types.
+number of intersection types.
 
@@
 The current Typed Clojure prototype is not as successful, and appears to hang
-during type checking.
-The reasons are not entirely clear, but it is likely the number of intersection types
-are too large for the current prototype to handle, which has no optimisations
-around intersection types.
+during type checking. As this experiment was particularly ambitious,
+diagnosing and fixing this issue is delegated to future work.
 
-This experiment was intended to stress test Typed Clojure to an extreme, and
-it is unsurprising that Typed Clojure was not up to the task.
-One interesting point was noticed when porting it.
+One interesting point was noticed when porting the experiment.
 The SML CIDRE version of the red-black tree experiment
@@
                  (U nil x))))
-
 ; Sequence monad (called "list monad" in Haskell)
@@
                     (Seqable x))))
-
 ; State monad
@@
 
-Listing \ref{lst:maybet} shows the full code for type checking the definition
-of the maybe monad transformer.
-Typing this example was a matter of assigning types to anonymous functions
-(similar to standard monad definitions) and manually instantiating 
-polymorphic types with \lstinline|inst|
-that parameterise over higher kinds, which cannot be inferred by Typed Clojure.
+Finally, Listing \ref{lst:maybet} shows the type assigned to the maybe monad transformer definition.
+The most interesting aspect to notice here is that the monad \lstinline|m| is kept abstract throughout
+the type.
 
-\begin{lstlisting}[caption={Maybe monad transformer in Typed Clojure}, label=lst:maybet]
+\begin{lstlisting}[caption={Maybe monad transformer type}, label=lst:maybet]
 (ann maybe-t
@@
                               (m (U nil y))))])))
-(defn maybe-t
-  "Monad transformer that transforms a monad m into a monad in which
-   the base values can be invalid (represented by nothing, which defaults
-   to nil). The third argument chooses if m-zero and m-plus are inherited
-   from the base monad (use :m-plus-from-base) or adopt maybe-like
-   behaviour (use :m-plus-from-transformer). The default is :m-plus-from-base
-   if the base monad m has a definition for m-plus, and
-   :m-plus-from-transformer otherwise."
-  ([m] ((inst maybe-t m) m nil :m-plus-default))
-  ([m nothing] ((inst maybe-t m) m nothing :m-plus-default))
-  ([m nothing which-m-plus]
-   (monad-transformer m which-m-plus
-     [m-result (with-monad m m-result)
-      m-bind   (with-monad m
-                 (ann-form
-                   (fn m-bind-maybe-t [mv f]
-                     (m-bind
-                       mv
-                       (ann-form
-                         (fn [x]
-                           (if (nil? x)
-                             (m-result nothing) 
-                             (f x)))
-                         [(U nil a) -> (m (U nil b))])))
-                   (All [a b]
-                     [(m (U nil a)) [a -> (m (U nil b))] -> (m (U nil b))])))
-      m-zero   (with-monad m (m-result nothing))
-      m-plus   (with-monad m
-                 (ann-form
-                   (fn m-plus-maybe-t [& mvs]
-                     (if (empty? mvs)
-                       (m-result nothing)
-                       (m-bind (first mvs)
-                               (ann-form
-                                 (fn [v]
-                                   (if (= v nothing)
-                                     (apply m-plus-maybe-t (rest mvs))
-                                     (m-result v)))
-                                 [(U x nil) -> (m (U x nil))]))))
-                   (All [x] [(m (U x nil)) * -> (m (U x nil))])))
-      ])))
-
 \end{lstlisting}
@@
 
-Listing \ref{lst:conduittypes} shows the type constructor for conduit,
+Listing \ref{lst:conduittypes} shows the type constructor for a conduit type,
 \lstinline|==>|. Notice the body of \lstinline|==>| includes a reference
@@
 \begin{lstlisting}[caption=Types for Conduits, label=lst:conduittypes]
-...
-
 (def-alias Result
@@
     [in -> '[(U nil (==> in out)) (Cont out out)]]))
-
-...
 \end{lstlisting}

### 2012-12-06 dissertation/implementation.tex
Lyndon corrections + others

@@
 \begin{itemize}
-  \item Pairwise, each \lstinline|S| is a below \lstinline|T|,
+  \item Pairwise, each \lstinline|S| is a subtype of \lstinline|T|,
   \item \lstinline|R| is below \lstinline|expected|, if \lstinline|expected| is provided.
@@
 and \lstinline|y| with \lstinline|Any|.
-This helps infer a result type of \lstinline|(constantly true)|
-as \lstinline|[Any * -> true]|.
+This helps infer the type of the expression \lstinline|(constantly true)|
+as \lstinline|[Any * -> true]| (where \lstinline|true| is the singleton type containing just the value \lstinline|true|).
 
@@
 tedious because some of the relevant internal functions interact
-in strange ways with the rest of Typed Racket. My reaction was that Typed Racket
+in strange ways with the rest of Typed Racket. My impression was that Typed Racket
 was initially designed without variable-arity polymorphism and was added
@@
 
-To better understand the problem, listing \ref{lst:filterann} presents the type of \lstinline|filter|,
+To better understand the problem, Listing \ref{lst:filterann} presents the type of \lstinline|filter|,
 which takes a function \lstinline|f| and a sequence \lstinline|s| as arguments, and returns a sequence that 
@@
 The \lstinline|:filters| syntax requires some explanation. 
-Function types support an optional \emph{filter set} attached to its return type, written as a map
+Function types support an optional \emph{filter set} attached to the return type, written as a map
 with \lstinline|:then| and/or \lstinline|:else| keys (if omitted, they default to the trivially true proposition which has no effect). 
@@
 \begin{lstlisting}[caption=Troublesome filter, label=lst:filtertrouble]
-(filter (fn [a] (not (nil? a))) coll)|
+(filter (fn [a] (not (nil? a))) coll)
 \end{lstlisting}
 
-The difficulty starts with something like listing \ref{lst:filtertrouble},
+The difficulty starts with something like Listing \ref{lst:filtertrouble},
 where the inferred filter set for the first argument to \lstinline|filter|
@@
 from a sequence of type \lstinline|(Seqable (U Number nil))|, we can replace
-listing \ref{lst:filtertrouble} with \lstinline|(filter number? coll)|,
+Listing \ref{lst:filtertrouble} with \lstinline|(filter number? coll)|,
 where \lstinline|number?| has the filter set \lstinline|{:then (is Number 0) :else (! Number 0)}|.
@@
 
-Using this technique, the anonymous function in listing 
+Using this technique, the anonymous function in Listing 
 \ref{lst:filtergood}\footnote{The Clojure syntax \lstinline|#(not (nil? \%))| 

### 2012-12-06 dissertation/intro.tex
Lyndon corrections + others

@@
 The Clojure programming language is a dynamically typed dialect of Lisp invented
-by Hickey \footnote{\url{http://clojure.org/}}, designed to run on popular platforms.
+by Hickey~\cite{Hic08}, designed to run on popular platforms.
 It emphasises functional programming with immutable data structures
@@
 
+\subsubsection{Better error messages}
+
+Dynamically typed languages like Clojure often sacrifice some of the advantages of
+static typing. In particular, statically typed languages generally offer 
+compile-time type errors where dynamically typed languages either give runtime
+type errors, or none at all.
+
+This is distinction is easily demonstrated by comparing Clojure and Haskell's behaviour at their interactive prompts.
+Listing \ref{lst:cljrtime} shows how Clojure happily compiles a nonsensical function that performs addition
+on a string. It is only until we invoke it that we get a type error.
+
+\begin{lstlisting}[caption=Runtime type errors in Clojure, label=lst:cljrtime]
+user=> (fn [x] (+ "string" x))
+#<user$eval128376$fn__128377 user$eval128376$fn__128377@183ab4e>
+
+user=> ((fn [x] (+ "string" x)) 1)
+#<ClassCastException java.lang.ClassCastException: java.lang.String cannot be cast to java.lang.Number>
+\end{lstlisting}
+
+The error provided by Clojure is unsatisfying:
+
+\begin{itemize}
+  \item The error message does little to help the programmer identify the source of the error.
+  \item It requires at least some knowledge of Clojure's internals to debug.
+  \item We would rather receive the error at the earliest time possible: compile-time.
+\end{itemize}
+
+Listing \ref{lst:hskctime} shows how Haskell, a statically typed language, handles this interaction.
+Haskell detects our programming error at compile-time, disallowing our function to compile.
+
+\begin{lstlisting}[caption=Compile-time type errors in Haskell, label=lst:hskctime]
+Prelude> (\x -> "string" + x)
+
+<interactive>:1:17:
+   No instance for (Num [Char])
+   arising from a use of `+'
+   Possible fix: add an instance declaration for (Num [Char])
+   In the expression: "string" + x
+   In the expression: (\ x -> "string" + x)
+   In an equation for `it': it = (\ x -> "string" + x)
+\end{lstlisting}
+
+Haskell's error here is much preferred over Clojure's:
+
+\begin{itemize}
+  \item The exact source of the error is given in \emph{user code}.
+  \item The error is given as early as possible: compile-time.
+\end{itemize}
+
+By adding static type checking to Clojure, we can achieve errors messages that are closer to
+Haskell's.
+Listing \ref{lst:tcctime} gives an idea of what kind of error messages Typed Clojure actually
+provides in these situations.
+
+\begin{lstlisting}[caption=Compile-time errors in Typed Clojure, label=lst:tcctime]
+user=> (cf (fn> [[a :- Number]] (+ "string" a)))
+#<Exception java.lang.Exception: 5: Static method clojure.lang.Numbers/add could not be applied to arguments:
+Domains: 
+  typed.core/AnyInteger typed.core/AnyInteger
+  java.lang.Number java.lang.Number
+
+Arguments:
+  (Value "string") java.lang.Number
+
+in: (. clojure.lang.Numbers add "string" a)>
+\end{lstlisting}
+
+Typed Clojure lists each of the expected domain types, of which the static method \lstinline|add| of the Java class
+\lstinline|clojure.lang.Numbers| has been assigned two,
+and also lists the types of the given arguments.
+A line number and the source of the error is also provided.
+
+Listing \ref{lst:tcctime} deserves some explanation:
+
+\begin{itemize}
+  \item The given source of the error may be surprising, but it is explained
+        by noting that Clojure ``inlines'' \lstinline|(+ "string" a)| to a Java method call
+        \lstinline|(. clojure.lang.Numbers add "string" a)|.
+        \lstinline|cf| type checks a form.
+  \item \lstinline|add|'s type has been overridden by Typed Clojure to be more specific (see Section \ref{sec:javainterop}).
+  \item \lstinline|cf| type checks a form.
+  \item \lstinline|(fn> [[a :- Number]] ...)| defines a function with its first argument of static type \lstinline|Number|.
+\end{itemize}
+
+\subsubsection{Higher order programming}
+
 The initial motivation for implementing an optional type system
 for Clojure was an anecdotal account of the struggles
-of a Clojure programmer. In an apparently heroic effort, he managed to 
+of a Clojure programmer, Jim Duey. In an apparently heroic effort, Duey managed to 
 implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
@@
 
-He highlighted a strong desire for a type system for several reasons.
+Duey highlighted a strong desire for a type system for several reasons.
 Firstly, to verify the correctness of the library.
@@
 after it was selected by the Clojure community as a Clojure Google Summer of Code
-project for 2012~\footnote{\url{http://www.google-melange.com/gsoc/project/google/gsoc2012/ambrosebs/10001}}.
+project for 2012~\cite{GSOC2012}.
 I have also been invited to speak on this project at the Clojure Conj
@@
 existing Clojure programmers can use.
-We also developed a prototype based on Typed Racket~\cite{Tob10}.
+We also develop a prototype based on Typed Racket~\cite{Tob10}.
 
@@
 
-  \item This prototype is extended to support several features that are not present in Typed Racket.
+  \item This prototype is extended to support several features that are not yet present in Typed Racket.
         We support filtering sequences on anonymous functions that convey only negative type information
         (eg. non-nil), capturing a common Clojure idiom.
+        We enable type checking of very abstract code (like code using monads)
+        by adding \emph{type functions} (functions at the type level).
 
-  \item We describe a novel use of occurrence typing~\cite{TF10} for eliminating
-        occurrences of Java's \emph{null}. We make \emph{null} directly expressible
-        in Typed Clojure's type system as \lstinline|nil|. This makes it reasonable for it to be a
-        static type error if \emph{null} is potentially misused, an improvement over Java's static type system
-        where \emph{null} cannot be expressed.
+  \item We support safer interoperability with Java. We show how techniques developed for
+        Typed Racket, such as occurrence typing~\cite{TF10},  of Java's \emph{null}
+        while still conforming to existing Clojure idioms.
 
-  \item We show how a combination of intersection types and occurrence typing enables us to accurately type check
-        the most common usages of Clojure's sequence abstraction. This was particularly subtle, as 
-        many base operations in the sequence abstraction take and/or return \lstinline|nil|, and
-        each case could imply different information about a sequence.
-
-  \item We capture the Clojure idiom of using maps with known keyword fields by using heterogeneous map types.
-        These types are sufficient for accurately associating and dissociating known keyword keys,
-        and accurately looking up values associated with keyword keys in heterogeneous map types that have
-        that keyword key present.
-        This is sufficient for type checking many common idioms involving maps.
+  \item We can type check several particularly useful Clojure idioms, like common usages of the \emph{sequence abstraction}
+        (Clojure's common interface to all collections) and 
+        common usages of ``untyped'' hash-maps, which have a similar role to records in ML or Haskell,
+        and objects in object-oriented languages.
 
@@
         to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
-        Records have subtle semantics for dissociating keys, like dissociating a base key of a record returns
+        Records have subtle semantics for dissociating keys, specifically: dissociating all base keys of a record returns
         a plain map.
@@
 
-(\lstinline|(defn name [...] body)| expands to \lstinline|(def name (fn [...] body))|,
+(\lstinline|(defn collatz [n] ...)| expands to \lstinline|(def collatz (fn [n] ...))|,
 where \lstinline|(def name init)| defines a new var in the current namespace 
@@
 (when unambiguous, I omit the qualifying namespace/package for the remainder of the chapter).
-Typed Clojure requires all 
-used vars to be annotated. Here \lstinline|typed.core/ann|, a Typed Clojure macro for annotating vars, 
+Typed Clojure type checks at the namespace granularity. All vars that occur
+in a ``typed'' namespace must have a type annotation. 
+Here \lstinline|typed.core/ann|, a Typed Clojure macro for annotating vars, 
 annotates \lstinline|collatz| to be a function from \lstinline|Number| to 

### 2012-12-06 dissertation/lit-review.tex
Lyndon corrections + others

@@
 
-Scheme then moved to program analysis techniques like Set Based Analysis (SBA).
+Scheme then moved to program analysis techniques like Set-Based Analysis~\cite{FF97} (SBA).
 They, however, suffered from modularity issues due to the lack of declared intended types.
@@
 Gradual typing combines static and runtime type checking, so programmers
-can utilise whichever is most appropriate for a given situation.
+can choose the most appropriate one for the situation.
 A key feature of gradual type systems is its use of \emph{bidirectional checking}
@@
 Runtime checks affect performance, but give better errors and ensure
-type errors result in the program failing quicker.
+type errors result in the program failing more quickly.
 
@@
 interoperability with Java libraries. As Clojure is a dynamically typed language, it does
-not give strong type type guarantees at compile time whether interactions with Java
+not give strong type guarantees at compile time that interactions with Java
 are type safe.
@@
 Typed Racket implements a sophisticated blame calculus, based on Wadler~\cite{WF09}. It ensures 
-error messages correctly \emph{blame} the source of type errors,
+error messages always correctly \emph{blame} the source of type errors,
 which can be difficult to determine in the presence of higher-order functions.
@@
 
-O'Caml-style extensible record types has been the subject of extensive research 
+O'Caml-style extensible record types have been the subject of extensive research 
 (eg. Wand~\cite{Wan89}, Cardelli and Mitchell~\cite{CM91}, Harper and Pierce~\cite{HP91}).
@@
 which are checked using \emph{conservative approximation}~\cite{OSV08}. For example,
-this checks that covariant parameters are only used in covariant positions.
+this checks that covariant parameters are used only in covariant positions.
 
@@
 
-In untyped languages like Scheme, functions can take any number of arguments. It is an
-error to supply less than the minimum number of arguments to a function.
+In untyped languages like Scheme, functions can have more complicated arity.
+In this context, functions have a fixed number of required parameters, and an optional \emph{rest}
+parameter, which represents the sequence of the arguments provided after the required parameters.
+It is an error to apply a function to fewer arguments than the functions number of required parameters.
 This style features sophisticated support for functions with variable-arity. For example,
-functions can dispatch on the number of arguments provided, and supports a \emph{rest} parameter
-as its last parameter which can accept any number of arguments.
+Scheme functions can dispatch on the number of arguments provided, and support an optional \emph{rest} parameter
+which can accept any number of arguments.
 
@@
 
-Strickland et al.\ invented a type system supporting variable-arity polymorphism ~\cite{STF09}
+Strickland \emph{et al.} invented a type system supporting variable-arity polymorphism ~\cite{STF09} % doesn't fit on a line
 a version of which is included in the current implementation of Typed Racket.
@@
 
-For example in Clojure, the function \lstinline|map| takes a function and one or more sequences,
-and returns the result of applying the function argument to each element of the sequences pair-wise.
+For example in Clojure, the function \lstinline|map| takes a function and at least one sequence,
+and returns a sequence of the results of applying the function to each subsequent element of the sequences simultaneously.
 
@@
 
-The pseudocode in listing \ref{lst:bidir} show both algorithms in action. Local type argument synthesis is able
+The pseudocode in Listing \ref{lst:bidir} show both algorithms in action. Local type argument synthesis is able
 to infer the type arguments to \lstinline|map| by observing the argument types of the first
@@
 \lstinline|(+ a b)| as \lstinline|Float|. We now have sufficient information to 
-synthesise the type of listing \ref{lst:bidir} to be \lstinline|(LazySeq Float)|.
+synthesise the type of Listing \ref{lst:bidir} to be \lstinline|(LazySeq Float)|.
 
@@
 Typed Racket uses occurrence typing to capture these ad-hoc type refinements.
-For example, listing \ref{lst:occ1} shows occurrence typing following the implications 
+For example, Listing \ref{lst:occ1} shows occurrence typing following the implications 
 of the type predicate \emph{number?} and the selector \emph{first}, and utilises those implications to refine
@@
 that lead to that branch.
-(See section \ref{sec:occurenceimpl} for details on how Typed Clojure uses occurrence typing).
+(See Section \ref{sec:occurenceimpl} for details on how Typed Clojure uses occurrence typing).
 
@@
 we can begin to statically disallow potentially inconsistent usages of \lstinline|nil| as part of the type system.
+\\\\
+% fit next listing onto one page
 
@@
 
-Listing \ref{lst:nil} infers from the Java signature \lstinline|Object getObject()| that
+Listing \ref{lst:nil} infers from the Java signature \lstinline|Object getObject()|
 that \lstinline|a| is of type \lstinline|(U nil Object)|. This is equivalent to Java's
@@
 this by observing \lstinline|nil| is a false value in Clojure, therefore \lstinline|a| cannot
-be \lstinline|nil| the body of the \lstinline|when|, refining \lstinline|a|'s type to \lstinline|Object|
+be \lstinline|nil| in the body of the \lstinline|when|, refining \lstinline|a|'s type to \lstinline|Object|
 from \lstinline|(U nil Object)|.
@@
 After some investigation, typing multimethods with Typed Clojure is assigned as future work
-(see section \ref{sec:mmfuture}).
+(see Section \ref{sec:mmfuture}).
 
@@
 For example, a type \lstinline|Number| is distinguished from a
-\emph{type constructor} \lstinline|Monad| (see section \ref{sec:monads}), which is a type level function
+\emph{type constructor} \lstinline|Monad| (see Section \ref{sec:monads}), which is a type level function
 (similar to the difference between values and functions on the value level, but on the type level).
@@
 the latter of which is an essential part of its monad library 
-(see section \ref{sec:monads} for a monad library ported to Typed Clojure
+(see Section \ref{sec:monads} for a monad library ported to Typed Clojure
 that uses type constructors similarly).

### 2012-12-06 dissertation/paper.tex
Lyndon corrections + others

@@
 There is significant future work to fully type check all Clojure features and idioms.
-For example, multimethod definitions and functions accepting an even number of variable arguments
-are troublesome. 
+For example, multimethod definitions and functions with particular constraints on the number of 
+variable arguments they accept (particularly, functions taking only an even number of variable arguments) are troublesome. 
 Also, there are desirable features from the Typed Racket project that are missing, such
@@
 A week later my supervisor Rowan Davies invited me to visit Carnegie Mellon University
-where he was working on his PhD project. Most memorably, I was invited to a meeting between
-Rowan and his PhD supervisor Frank Pfenning, which was a great thrill.
+where he worked on his PhD project. Most memorably, I was invited to a meeting between
+Rowan and his PhD supervisor Frank Pfenning.
 

### 2012-12-06 dissertation/intro.tex
Add Supervisor. Revise abstract, motivation and acknowledgements

@@
 The initial motivation for implementing an optional type system
-for Clojure was an anecdotal account of the struggles
-of a Clojure programmer, Jim Duey. In an apparently heroic effort, Duey managed to 
+for Clojure was outlined in a discussion with Jim Duey, a Clojure programmer, at Clojure Conj 2011.
+In an apparently heroic effort, Duey managed to
 implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
@@
 
-Since then, the Clojure community has shown significant interest in this work.
-I also developed this project as a Google Summer of Code 2012 project,
-after it was selected by the Clojure community as a Clojure Google Summer of Code
-project for 2012~\cite{GSOC2012}.
-I have also been invited to speak on this project at the Clojure Conj
-2012 conference in November, the main international conference related to Clojure.
+The proliferation of other abstract programming styles, like monads, in statically typed languages
+also suggested that such a static type system could
+help the programmer write and debug such code, while improving confidence that their code is correct.
+
+\subsection{Why does not Clojure feature static type checking?}
+
+Given the advantages of static type checking outlined in the previous section, it is natural to ask:
+why does Clojure not already support static type checking?
+The simplest answer is that Clojure was not designed to support a static type system.
+Instead, features and idioms that are not necessarily compatible with traditional type systems
+were favoured, like variable-arity functions, untyped hash-maps used like objects, and extension of program behaviour
+at runtime.
+
+Without specifically speaking for the designers of Clojure, at the time Clojure was designed
+it was not clear how to reconcile Clojure's feature-set with static type checking in a satisfying way.
+Since then, Typed Racket, which add static type checking to the Racket programming language,
+have matured and gained acceptance.
+Once this occurred, it seemed like only a matter of time until the ideas developed for the Typed Racket project 
+were tested for their applicability to Clojure. 
+As far as we know, this project is the first serious attempt to build an optional static type system for Clojure.
 
@@
 and common in the literature on static types (for example, Pierce~\cite{Pie02}).
+
+\subsection{What is the community reaction to Typed Clojure?}
+
+Since I began developing Typed Clojure, the Clojure community has shown significant interest in this work:
+
+\begin{itemize}
+  \item I developed this project as a Google Summer of Code 2012 project,
+        after it was selected by the Clojure community as a Clojure Google Summer of Code
+        project for 2012~\cite{GSOC2012}.
+  \item I gave a talk on this project at the Clojure Conj
+        2012 conference, the main international programming conference related to Clojure, in November.
+\end{itemize}
 

### 2012-12-06 dissertation/paper.tex
Add Supervisor. Revise abstract, motivation and acknowledgements

@@
 
-\keywords{Type systems, Clojure}
+\keywords{Programming languages, optional type systems, Clojure}
 % TODO What are categories? cshonours requires them - but commented out now
@@
 
-Traditionally there has been a clear distinction between 
-programming languages that perform type checking at compile-time, and those that do not.
-In particular, the early wide-spread programming languages ALGOL and Lisp
-could be divided this way: ALGOL performs compile-time (or static) type checking,
-and Lisp does not, and most recent languages can be similarly classified.
-Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
-when the program is run.
-On the other hand, a language without static type checking can often be designed 
-to be more flexible and forgiving, making it easy, for example, to have a \emph{map} function that can map functions
-with any number of arguments (as in most Lisp dialects).
+%Traditionally there has been a clear distinction between 
+%programming languages that perform type checking at compile-time, and those that do not.
+%In particular, the early wide-spread programming languages ALGOL and Lisp
+%could be divided this way: ALGOL performs compile-time (or static) type checking,
+%and Lisp does not, and most recent languages can be similarly classified.
+%Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
+%when the program is run.
+%On the other hand, a language without static type checking can often be designed 
+%to be more flexible and forgiving, making it easy, for example, to have a \emph{map} function that can map functions
+%with any number of arguments (as in most Lisp dialects).
+%
+%A recent trend is to aim to combine the advantages of both kinds of languages by adding optional static 
+%type systems to languages without static type checking. 
+%A major challenge in these combinations is continuing to support the idiomatic style of the original 
+%language while performing accurate enough checking to catch similar errors to standard statically
+%typed languages. Some recent attempts appear to have finally produced type systems that have overcome this obstacle.
+%Notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
+%check most idiomatic Racket code.
+%These optional type systems have some important differences from standard static
+%type systems, for example they often include \emph{union} types.
+%
 
-A recent trend is to aim to combine the advantages of both kinds of languages by adding optional static 
-type systems to languages without static type checking. 
-A major challenge in these combinations is continuing to support the idiomatic style of the original 
-language while performing accurate enough checking to catch similar errors to standard statically
-typed languages. Some recent attempts appear to have finally produced type systems that have overcome this obstacle.
-Notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
-check most idiomatic Racket code.
-These optional type systems have some important differences from standard static
-type systems, for example they often include \emph{union} types.
+Dynamic programming languages often abandon the advantages of static type checking 
+in favour of their characteristic convenience and flexibility.
+Static type checking eliminates many common user errors at compile-time that
+are otherwise unnoticed, or are caught later in languages without static type checking.
+A recent trend is to aim to combine the advantages of both kinds of languages by adding \emph{optional} static 
+type systems to languages without static type checking, while preserving the languages idioms and style.
 
-Clojure is a recent and increasingly popular programming language that is designed to support 
-functional programming and immutability by default. Implementations of Clojure are hosted 
-on popular platforms and provide direct interoperability with the host platform.
-There are several implementations of Clojure, each targeting different runtimes including
-the Java Virtual Machine, the Common Language Runtime, and Javascript virtual machines.
-
-This dissertation describes my work on designing an optional static type system for Clojure based on
+This dissertation describes my work on designing an optional static type system for the Clojure programming language,
+a dynamically typed dialect of Lisp, based on
 the lessons learnt from several projects, primarily Typed Racket.
@@
 For example, multimethod definitions and functions with particular constraints on the number of 
-variable arguments they accept (particularly, functions taking only an even number of variable arguments) are troublesome. 
+variable arguments they accept (particularly functions taking only an even number of variable arguments) are troublesome. 
 Also, there are desirable features from the Typed Racket project that are missing, such
@@
 both which are designed to improve error messages when mixing typed and untyped code in similar systems.
+
 Overall, the work described in this dissertation leads to the conclusion that it appears to 
@@
 
-Last November I attended my first conference, Clojure Conj 2011 (Raleigh, NC),
+Last November I attended my first programming conference, Clojure Conj 2011 (Raleigh, NC),
 where I delivered a talk on logic programming with Clojure and met some of my programming heroes.
@@
 functional programming languages. He has exceeded my expectations as an honours supervisor,
-often dedicating time to assist me on public holidays and weekends.
+and we plan to continue to collaborate in this area.
 Rowan's experience with refinement types brought a unique perspective to this project
@@
 
-I'd also like to thank: Sam Tobin-Hochstadt for his consistent assistance and patience;
-David Nolen for mentoring this project as part of Google Summer of Code 2012;
-and Chas Emerick, without whom I would have never considered attending Clojure Conj 2011
+I am grateful to Sam Tobin-Hochstadt for his pioneering work in this area of research.
+He has always been happy to offer assistance and advise throughout this project, which always proved invaluable.
+
+I would also like to thank David Nolen for mentoring this project as part of Google Summer of Code 2012,
+and Chas Emerick, without whom I would have never even considered attending Clojure Conj 2011
 (and this work probably would not have happened).
@@
 
-Fittingly, I have been invited to talk on Typed Clojure at Clojure Conj
-2012 in November, which I am highly anticipating.
+Fittingly, I gave a talk on Typed Clojure at Clojure Conj 2012, which was extremely satisfying.
 

### 2012-12-06 dissertation/intro.tex
Fix ref sorting. Adjust listings to not break pages

@@
 provides in these situations. 
+However, it deserves some explanation:
+
+\begin{itemize}
+  \item The given source of the error may be surprising, but it is explained
+        by noting that Clojure ``inlines'' \lstinline|(+ "string" a)| to a Java method call
+        \lstinline|(. clojure.lang.Numbers add "string" a)|.
+  \item \lstinline|add|'s type has been overridden by Typed Clojure to be more specific (see Section \ref{sec:javainterop}).
+  \item \lstinline|cf| type checks a form.
+  \item \lstinline|(fn> [[a :- Number]] ...)| defines a function with its first argument of static type \lstinline|Number|.
+\end{itemize}
 
@@
 user=> (cf (fn> [[a :- Number]] (+ "string" a)))
-#<Exception java.lang.Exception: 5: Static method clojure.lang.Numbers/add could not be applied to arguments:
+#<Exception java.lang.Exception: 5: 
+Static method clojure.lang.Numbers/add could not be applied to arguments:
 Domains: 
@@
 
-Listing \ref{lst:tcctime} deserves some explanation:
-
-\begin{itemize}
-  \item The given source of the error may be surprising, but it is explained
-        by noting that Clojure ``inlines'' \lstinline|(+ "string" a)| to a Java method call
-        \lstinline|(. clojure.lang.Numbers add "string" a)|.
-        \lstinline|cf| type checks a form.
-  \item \lstinline|add|'s type has been overridden by Typed Clojure to be more specific (see Section \ref{sec:javainterop}).
-  \item \lstinline|cf| type checks a form.
-  \item \lstinline|(fn> [[a :- Number]] ...)| defines a function with its first argument of static type \lstinline|Number|.
-\end{itemize}
+Most importantly: the error was caught at \emph{compile-time}.
 
@@
 
-The proliferation of other abstract programming styles, like monads, in statically typed languages
-also suggested that such a static type system could
-help the programmer write and debug such code, while improving confidence that their code is correct.
+Programmers in statically typed languages take advantage of the static type system
+to help verify code written in abstract programming styles, like monads.
+Adding a static type system to Clojure should bring similar benefits to Clojure programmers.
 
-\subsection{Why does not Clojure feature static type checking?}
+\subsection{Why does Clojure not already feature static type checking?}
 
@@
 Instead, features and idioms that are not necessarily compatible with traditional type systems
-were favoured, like variable-arity functions, untyped hash-maps used like objects, and extension of program behaviour
-at runtime.
+were favoured, like variable-arity functions, untyped hash-maps used like objects, and dynamic extension of program behaviour.
 
@@
 Since then, Typed Racket, which add static type checking to the Racket programming language,
-have matured and gained acceptance.
-Once this occurred, it seemed like only a matter of time until the ideas developed for the Typed Racket project 
+has matured and gained acceptance.
+Once this occurred, it seemed only a matter of time until the ideas developed for the Typed Racket project 
 were tested for their applicability to Clojure. 
@@
 
-  \item We support safer interoperability with Java. We show how techniques developed for
-        Typed Racket, such as occurrence typing~\cite{TF10},  of Java's \emph{null}
+  \item We support safer interoperability with Java from Clojure. We show how techniques developed for
+        Typed Racket, such as occurrence typing~\cite{TF10}, help detect misuses of Java's \emph{null}
         while still conforming to existing Clojure idioms.
@@
 We can annotate datatype and protocol definitions similarly.
-
+%bump over the page
+\\\\\\\\\\
 \begin{lstlisting}[caption=Annotating protocols and datatypes in Typed Clojure]
 (ns typed.test.deftype
-  (:require [typed.core 
-             :refer [check-ns ann-datatype
-                     tc-ignore ann-protocol AnyInteger]]))
+  (:require 
+    [typed.core :refer [check-ns ann-datatype defprotocol>
+                        ann-protocol AnyInteger]]))
 
 (ann-protocol Age 
-  :methods
-  {age [Age -> AnyInteger]})
-(tc-ignore
-  (defprotocol Age
-    (age [this])))
+  age [Age -> AnyInteger])
+(defprotocol> Age
+  (age [this]))
 
 (ann-datatype Person 
-  [[name :- String]
-   [age :- AnyInteger]])
+  [name :- String]
+   age :- AnyInteger])
 (deftype Person [name age]
@@
 
-\lstinline|defprotocol| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols.}
-with a set of methods. \lstinline|ann-protocol| is a Typed Clojure macro for annotating 
+\lstinline|defprotocol>| defines a new Clojure protocol\footnote{See http://clojure.org/protocols for a full description of protocols.}
+with a set of methods. It is identical to the usual \lstinline|defprotocol|, but is required when using Typed Clojure
+due to its complicated macroexpansion.
+\lstinline|ann-protocol| is a Typed Clojure macro for annotating 
 a protocol with the types of its methods.
@@
 is for this function.
-
-Invocations of \lstinline|defprotocol| are currently not able to be type checked
-and are ignored by Typed Clojure by passing them to \lstinline|typed.core/tc-ignore|.
 
@@
 in the syntax for defining heterogeneous map types.
-
+%bump
+\\\\
 \begin{lstlisting}[caption=Heterogeneous map types in Typed Clojure]
-...
 (ann config '{:file String, :ns Symbol}))
@@
    :ns 'clojure.core})
-...
 \end{lstlisting}
@@
 
-\subsubsection{Uniform Variable-Arity}
+\subsubsection{Uniform Variable-Arity Functions}
 
@@
 of type \lstinline|Number|, represented by the type \lstinline|[Number * -> Number]|.
+\\\\\\\\
 
+%bump over page
 \begin{lstlisting}[caption=Typing multi-arity functions, label=lst:noteq]
@@
 For example, the variable argument function \lstinline|clojure.core/map| takes a function and one or more sequences,
-and returns the result of applying the function argument to each element of the sequences in a pair-wise fashion.
+and returns a sequence of the results of applying the function to each element of the sequences simultaneously.
 Its type is given in listing \ref{lst:maptype}.
@@
 \begin{lstlisting}
-[[boolean String Integer -> Number] (U (Seqable boolean) nil) (U (Seqable String) nil) (U (Seqable Integer) nil) -> (LazySeq Number)]
+[[boolean String -> Number] (U (Seqable boolean) nil) 
+  (U (Seqable String) nil) -> (LazySeq Number)]
 \end{lstlisting}
@@
 particularly in the treatment of Java's \emph{null}.
+Note that these decisions do not change Java's type system, but only
+how Typed Clojure handles Java interoperability compared to how someone
+familiar with Java might expect.
 
@@
 A type is \emph{not nullable} if it doesn't include \emph{null}).
-
-\begin{lstlisting}[caption=Java interoperability with Typed Clojure]
+%bump over the page
+\\\\
+\begin{lstlisting}[caption=Java interoperability with Typed Clojure, label=lst:interopfile]
 (ns typed.test.interop
@@
 
-This example shows how Typed Clojure handles \emph{null} while creating and
+Listing \ref{lst:interopfile} shows how Typed Clojure handles \emph{null} while creating and
 using an instance of \emph{java.io.File}.

### 2012-12-06 dissertation/paper.tex
Fix ref sorting. Adjust listings to not break pages
Typos

@@
 \usepackage{url}
-\usepackage[style=alphabetic]{biblatex}
+\usepackage[style=alphabetic, sorting=nyt]{biblatex}
 \addbibresource{bibliography.bib}
@@
 A recent trend is to aim to combine the advantages of both kinds of languages by adding \emph{optional} static 
-type systems to languages without static type checking, while preserving the languages idioms and style.
+type systems to languages without static type checking, while preserving the idioms and style of the language.
 
@@
 
-I am grateful to Sam Tobin-Hochstadt for his pioneering work in this area of research.
-He has always been happy to offer assistance and advise throughout this project, which always proved invaluable.
+I am grateful to Sam Tobin-Hochstadt for his excellent work in this area of research.
+He has always been happy to offer assistance and advice throughout this project, which has proved invaluable.
 

### 2013-02-24 icfp2013/interop.tex
Add makefile

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{listings}

\begin{document}

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2013} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Interoperability with a Statically Typed Language with Null values in a Gradually Typed Language}
%\subtitle{Subtitle Text, if any}

\authorinfo{Ambrose Bonnaire-Sergeant}
           {University of Western Australia}
           {abonnairesergeant@gmail.com}
\authorinfo{Rowan Davies}
           {University of Western Australia}
           {rowan@csse.uwa.edu.au}

% me:  I'm playing around with some approaches to a paper:
% Title: "Interoperability with a Statically Typed Language with Null types in a Gradually Typed Language"
% Abstract: Interactions with statically-typed languages from dynamically-typed languages
% can be better verified at compile time by utilising a gradual type system.
% This paper describes how a dynamically-typed language like Clojure, Ruby, or Python
% can better verify interactions with a statically-typed language with Null types
% like Java or C#.
%  Sent at 10:13 PM on Thursday
%  Rowan:  It's  a start.
% Java and C# don't have null types, right?
%  me:  no, I guess I mean nullable
%  Rowan:  I'd just say that they have null values.
%  me:  Ok, I like that.
%  Rowan:  Nullable doesn't have wide enough usage.
%  Sent at 11:39 PM on Thursday
%  Rowan:  Or, "only have types that implicitly include null values".
% You could say Java and C# have dynamic types for null, but that's just going to confuse everyone.
%  me:  Except for the special cases of arrays and primitives for Java.
%  Rowan:  Array variables can be null in Java, I thought?
%  me:  I thought not.
% Will check...
% whoops
% You're correct.
% Do CLR languages have similar edge cases?
%  Sent at 11:44 PM on Thursday
%  Rowan:  They are similar, yes, but the CLR is fully typed, so e.g., generic types aren't erased, the CLR has generic types and type constructors.
% I think the CLR even has variance for type constructors.
%  me:  Ah that's interesting.
%  Sent at 11:46 PM on Thursday
%  Rowan:  I generally think of Java variables as either being a primitive type, or a reference.
% A reference is either null, or refers to a value that has a runtime type consistent with the compile time type.
%  Sent at 11:49 PM on Thursday
%  me:  Got it.
%  Rowan:  So, it's not that an array can be null - it's the variable that's null.
%  me:  Ah.
% Yes that makes sense.
%  Rowan:  Oops, I meant, it's the reference that's null.
% Since, e.g., the reference could be an element of an array rather than a variable.
%  me:  Is there a difference between variable and reference?
% In Java?
% I don't understand that.
%  Rowan:  a[3] = null;
% Does that set a variable to null?
% A: No, it sets a reference.
% This assumes we've declared   int[] a;
% Or        int a[10];
% Oops, not int, I mean:
% java.lang.Object a[10];
%  Sent at 11:55 PM on Thursday
%  me:  
% Ok, makes sense.
%  Sent at 11:57 PM on Thursday
%  Rowan:  Then   a  consists of 10 references, each can be null, but there's only one variable.
%  Sent at 11:58 PM on Thursday
%  Rowan:  Anyway, back to the abstract:   " Interactions with statically-typed languages from dynamically-typed languages
% can be better verified at compile time by utilising a gradual type system."
%  Sent at 11:59 PM on Thursday
%  Rowan:  It's not wise to state this without convincing evidence, I'd expect most referees to be skeptical, and I don't think we can be convincing within the space of the abstract.  Plus it's not clear what "better" is relative too.
% And it's not clear why you'd want to verify them.
% I'd start with sentence or two that frame the issue that's being addressed.
%  Sent at 12:03 AM on Friday
%  Rowan:  Actually, the set of questions Chris McDonald recommends probably applies:
% - what is the problem?
% - why is it interesting, challenging, important?
% - what have others done before?
% - what did you do?
% - how did you evaluate it?
% - what contribution have you made?
% - how would you continue this in the future?
% Start with "what is the problem?"
%  Sent at 12:05 AM on Friday
%  me:  The problem is dynamic languages ignore existing type information which can help avoid common programming errors at compile time, if utilised.
%  Sent at 12:07 AM on Friday
%  Rowan:  "Modern libraries such as the Java API are sufficiently large and complicated that correctly interfacing with them from other languages can be a difficult and error aspect of software development."
% Mine is maybe not quite saying the right thing, but I think it starts at about the level of generality.
%  me:  Ok, I've got the idea.
%  Rowan:  error -> error prone
%  Sent at 12:10 AM on Friday
%  Rowan:  I like the phrase "error prone" in the context of static types - it's not that you can't get it right, or that you won't eventually fix it, it's that there's good potential to get it wrong - without static checking you have to be much more careful.
%  Sent at 12:13 AM on Friday
%  me:  Ok, sounds convincing.
%  Sent at 12:16 AM on Friday
%  me:  2. The interesting part is it can be made less error prone in dynamically-typed languages.
% 
%  Sent at 12:18 AM on Friday
%  Rowan:  Yeah, hmmm.   "One specific area for potential errors is that many popular languages like Java allow all object references to be null, while generally libraries like those in the Java API have detailed specifications of their behaviour with null values."
% It'd be good to include "dynamic languages" here, but maybe in sentence 3 is okay.
%  Sent at 12:30 AM on Friday
%  Rowan:  There's some tension here - we're proposing more static checking than Java, but in the context of a dynamic langage.
%  Sent at 12:31 AM on Friday
%  Rowan:  The question is whether to make it clear we consider that Java should have non-nullable reference types, or we leave that alone.
%  Sent at 12:32 AM on Friday
%  me:  Ah. Good question.
%  Sent at 12:35 AM on Friday
%  me:  Our solution would be irrelevant for languages without null values.
%  Sent at 12:38 AM on Friday
%  Rowan:  It sometimes works to be frank about such things - somewhere in the intro we could say "Arguably Java itself should include non-nullable reference types, but we put that to one side since our main concern here is the practical issue of interoperability which arises with current languages/systems such as Clojure for the JVM.
% "
% Right, without null values there's no issue.
% But, that's different to having null values, but having separate types for when null values are allowed and when they aren't.
%  me:  Yes.
%  Rowan:  When I said "... Java should have non-nullable reference types", I was thinking that there would be still be nullable types also.
%  Sent at 12:42 AM on Friday
%  me:  Oh.
% Fair enough.
% Scala does this I think.
%  Rowan:  Right, yes.
%  Sent at 12:44 AM on Friday

\maketitle

\begin{abstract}
Modern libraries such as the Java API are sufficiently large and complicated that 
correctly interfacing with them from other languages can be a difficult and 
error prone aspect of software development.

In particular,

%Dynamically-typed languages 
%
%Interactions with statically-typed languages from dynamically-typed languages
%can be better verified at compile time by utilising a gradual type system.
%This paper describes how a dynamically-typed language, like Clojure, Ruby, or Python,
%extended with a gradual type system
%can better verify interactions with a statically-typed language with Null types
%like Java or C#.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
%term1, term2

\keywords
%keyword1, keyword2

\section{Introduction}

\section{Static null-safety}

\section{Array covariance}

Array covariance refers to the subtyping relationship between different array types.
In Java, arrays are parameterised by one covariant parameter, such that
t1[] <: t2[] for all t1 <: t2. This is well known to be statically unsound,
mainly because arrays are \emph{mutable}.

The main advantages of covariant array types are code reuse
and understandability of types.
The disadvantage is loss of static type soundness.
Our approach avoids the pitfalls of array covariance
while facilitating code reuse, at the cost of more complex array types.

\begin{lstlisting}[language=java]
public static void unsoundModify(Number[] ar) {
  Object[] bad_ar = (Object[])ar;
  bad_ar[0] = new java.util.Observable();
}
\end{lstlisting}

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

### 2013-02-25 icfp2013/interop.tex
array covariance stuff

@@
 
-\usepackage{amsmath}
+%\usepackage{amsmath}
 \usepackage{listings}
+
+\lstset{ %
+  language=Lisp,                % choose the language of the code
+  columns=fixed,basewidth=.5em,
+  basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
+  %numbers=left,                   % where to put the line-numbers
+  %numberstyle=\small\ttfamily,      % the size of the fonts that are used for the line-numbers
+  %stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
+  %numbersep=5pt,                  % how far the line-numbers are from the code
+  %backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
+  %showspaces=false,               % show spaces adding particular underscores
+  showstringspaces=false,         % underline spaces within strings
+  %showtabs=false,                 % show tabs within strings adding particular underscores
+  frame=single,           % adds a frame around the code
+  %tabsize=2,          % sets default tabsize to 2 spaces
+  captionpos=t,           % sets the caption-position to bottom
+  breaklines=true,        % sets automatic line breaking
+  breakatwhitespace=true,    % sets if automatic breaks should only happen at whitespace
+  %escapeinside={\%*}{*)},          % if you want to add a comment within your code
+}
 
@@
 
+% - what is the problem?
+% - why is it interesting, challenging, important?
+% - what have others done before?
+% - what did you do?
+% - how did you evaluate it?
+% - what contribution have you made?
+% - how would you continue this in the future?
+
 \terms
@@
 In Java, arrays are parameterised by one covariant parameter, such that
-t1[] <: t2[] for all t1 <: t2. This is well known to be statically unsound,
-mainly because arrays are \emph{mutable}.
+\lstinline|t1[] <: t2[]| for all \lstinline|t1 <: t2|. This is well known to be statically unsound,
+because arrays are \emph{mutable}.
 
-The main advantages of covariant array types are code reuse
-and understandability of types.
+The main advantages of covariant array types are increased code reuse
+and more straightforward array types.
 The disadvantage is loss of static type soundness.
 Our approach avoids the pitfalls of array covariance
-while facilitating code reuse, at the cost of more complex array types.
+while facilitating code reuse, at the cost of slightly more complex types.
 
-\begin{lstlisting}[language=java]
+Listing \ref{lst:bad_java_array} demonstrates how array covariance allows
+obviously ill-typed code to pass the static type checker.
+By up-casting our \lstinline|Number[]| array to \lstinline|Object[]|,
+the static type checker allows subtypes of \lstinline|Object|
+to be written to it.
+The problem occurs when writing to the up-casted array:
+\lstinline|java.util.Observable| cannot be written to arrays with tag \lstinline|Number[]|.
+The compiler instead inserts a runtime check which always results in a runtime error.
+
+\begin{lstlisting}[language=java, label=lst:bad_java_array, caption=Statically unsound Java array manipulation]
 public static void unsoundModify(Number[] ar) {
@@
 
-\appendix
-\section{Appendix Title}
+We want the type checker to reject obviously wrong code at compile time.
+To achieve this, we use a \emph{bivariant} array type
+\lstinline|(Array2 write read)| which has separate type parameters for
+reading from and writing to the array. We refer to \lstinline|Array2| as bivariant (or invariant)
+because it is contravariant in its first parameter and covariant in its second.
 
-This is the text of the appendix, if you need one.
+\begin{verb}
+(Array2 S T) <: (Array2 S' T') 
+  when S' <: S and T <: T'
 
-\acks
+(Array2 T T) is abbreviated (Array T)
+\end{verb}
 
-Acknowledgments, if needed.
+Listing \ref{lst:modify_array} uses \lstinline|Array2|
+to track the types that can be read or written to an array.
+We upcast an \lstinline|(Array Number)| (an array that can hold and write \lstinline|Number|s)
+to \lstinline|(Array Number Object)| (an array that can hold \lstinline|Object|s and write
+\lstinline|Number|s).
+Most importantly, the upcast preserves which types we can write to the array.
+
+The type assigned to \lstinline|aset|, the Clojure function for array modification, is slightly more interesting
+(Listing \ref{lst:aset_ann}).
+It takes an array, an array index and a value, and writes the value to the corresponding position of the array,
+and returns the value written.
+
+\begin{lstlisting}[language=lisp, label=lst:modify_array, caption=A type error with bivariant array types in core.typed]
+(ann modify-array [(Array Number) -> nil])
+(defn modify-array [ar]
+  (let [ar2 (ann-form ar (Array Number Object))]
+    (aset ar2 0 (new java.util.Observable))
+    nil))
+\end{lstlisting}
+
+\begin{lstlisting}[language=Lisp, label=lst:aset_ann, caption=Type for aset]
+(All [i o]
+  [(Array2 i o) AnyInteger i -> o]))
+\end{lstlisting}
+
+Listing \ref{lst:array_error} shows the static type error that results when type checking Listing \ref{lst:modify_array}.
+We retains enough information to determine we cannot write instances of
+\lstinline|java.util.Observable| to \lstinline|(Array2 Number Object)|.
+
+\begin{lstlisting}[language=lisp, label=lst:array_error, caption=Static type error for ill-typed array set]
+#<Exception java.lang.Exception: 44: Polymorphic static method clojure.lang.RT/aset could not be applied to arguments:
+Domains: 
+  (Array2 i o) clojure.core.typed/AnyInteger i
+
+Arguments:
+  (Array2 java.lang.Number java.lang.Object) int java.util.Observable
+
+in: (clojure.lang.RT/aset ar2 (clojure.lang.RT/intCast 0) (new java.util.Observable))>
+\end{lstlisting}
+
+%\appendix
+%\section{Appendix Title}
+%
+%This is the text of the appendix, if you need one.
+%
+%\acks
+%
+%Acknowledgments, if needed.
 
@@
 
-\bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\begin{thebibliography}{}
-\softraggedright
-
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{thebibliography}
+%\bibliographystyle{abbrvnat}
+%
+%% The bibliography should be embedded for final submission.
+%
+%\begin{thebibliography}{}
+%\softraggedright
+%
+%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
+%P. Q. Smith, and X. Y. Jones. ...reference text...
+%
+%\end{thebibliography}
 

### 2014-07-18 paper-template/paper.tex
template

--- 2013-02-25 icfp2013/interop.tex
@@
 \usepackage{listings}
+\usepackage{biblatex}
 
@@
 
-\title{Interoperability with a Statically Typed Language with Null values in a Gradually Typed Language}
+\title{Title}
 %\subtitle{Subtitle Text, if any}
@@
            {abonnairesergeant@gmail.com}
-\authorinfo{Rowan Davies}
-           {University of Western Australia}
-           {rowan@csse.uwa.edu.au}
-
-% me:  I'm playing around with some approaches to a paper:
-% Title: "Interoperability with a Statically Typed Language with Null types in a Gradually Typed Language"
-% Abstract: Interactions with statically-typed languages from dynamically-typed languages
-% can be better verified at compile time by utilising a gradual type system.
-% This paper describes how a dynamically-typed language like Clojure, Ruby, or Python
-% can better verify interactions with a statically-typed language with Null types
-% like Java or C#.
-%  Sent at 10:13 PM on Thursday
-%  Rowan:  It's  a start.
-% Java and C# don't have null types, right?
-%  me:  no, I guess I mean nullable
-%  Rowan:  I'd just say that they have null values.
-%  me:  Ok, I like that.
-%  Rowan:  Nullable doesn't have wide enough usage.
-%  Sent at 11:39 PM on Thursday
-%  Rowan:  Or, "only have types that implicitly include null values".
-% You could say Java and C# have dynamic types for null, but that's just going to confuse everyone.
-%  me:  Except for the special cases of arrays and primitives for Java.
-%  Rowan:  Array variables can be null in Java, I thought?
-%  me:  I thought not.
-% Will check...
-% whoops
-% You're correct.
-% Do CLR languages have similar edge cases?
-%  Sent at 11:44 PM on Thursday
-%  Rowan:  They are similar, yes, but the CLR is fully typed, so e.g., generic types aren't erased, the CLR has generic types and type constructors.
-% I think the CLR even has variance for type constructors.
-%  me:  Ah that's interesting.
-%  Sent at 11:46 PM on Thursday
-%  Rowan:  I generally think of Java variables as either being a primitive type, or a reference.
-% A reference is either null, or refers to a value that has a runtime type consistent with the compile time type.
-%  Sent at 11:49 PM on Thursday
-%  me:  Got it.
-%  Rowan:  So, it's not that an array can be null - it's the variable that's null.
-%  me:  Ah.
-% Yes that makes sense.
-%  Rowan:  Oops, I meant, it's the reference that's null.
-% Since, e.g., the reference could be an element of an array rather than a variable.
-%  me:  Is there a difference between variable and reference?
-% In Java?
-% I don't understand that.
-%  Rowan:  a[3] = null;
-% Does that set a variable to null?
-% A: No, it sets a reference.
-% This assumes we've declared   int[] a;
-% Or        int a[10];
-% Oops, not int, I mean:
-% java.lang.Object a[10];
-%  Sent at 11:55 PM on Thursday
-%  me:  
-% Ok, makes sense.
-%  Sent at 11:57 PM on Thursday
-%  Rowan:  Then   a  consists of 10 references, each can be null, but there's only one variable.
-%  Sent at 11:58 PM on Thursday
-%  Rowan:  Anyway, back to the abstract:   " Interactions with statically-typed languages from dynamically-typed languages
-% can be better verified at compile time by utilising a gradual type system."
-%  Sent at 11:59 PM on Thursday
-%  Rowan:  It's not wise to state this without convincing evidence, I'd expect most referees to be skeptical, and I don't think we can be convincing within the space of the abstract.
-Plus it's not clear what "better" is relative too.
-% And it's not clear why you'd want to verify them.
-% I'd start with sentence or two that frame the issue that's being addressed.
-%  Sent at 12:03 AM on Friday
-%  Rowan:  Actually, the set of questions Chris McDonald recommends probably applies:
-% - what is the problem?
-% - why is it interesting, challenging, important?
-% - what have others done before?
-% - what did you do?
-% - how did you evaluate it?
-% - what contribution have you made?
-% - how would you continue this in the future?
-% Start with "what is the problem?"
-%  Sent at 12:05 AM on Friday
-%  me:  The problem is dynamic languages ignore existing type information which can help avoid common programming errors at compile time, if utilised.
-%  Sent at 12:07 AM on Friday
-%  Rowan:  "Modern libraries such as the Java API are sufficiently large and complicated that correctly interfacing with them from other languages can be a difficult and error aspect of software development."
-% Mine is maybe not quite saying the right thing, but I think it starts at about the level of generality.
-%  me:  Ok, I've got the idea.
-%  Rowan:  error -> error prone
-%  Sent at 12:10 AM on Friday
-%  Rowan:  I like the phrase "error prone" in the context of static types - it's not that you can't get it right, or that you won't eventually fix it, it's that there's good potential to get it wrong - without static checking you have to be much more careful.
-%  Sent at 12:13 AM on Friday
-%  me:  Ok, sounds convincing.
-%  Sent at 12:16 AM on Friday
-%  me:  2. The interesting part is it can be made less error prone in dynamically-typed languages.
-% 
-%  Sent at 12:18 AM on Friday
-%  Rowan:  Yeah, hmmm.
-"One specific area for potential errors is that many popular languages like Java allow all object references to be null, while generally libraries like those in the Java API have detailed specifications of their behaviour with null values."
-% It'd be good to include "dynamic languages" here, but maybe in sentence 3 is okay.
-%  Sent at 12:30 AM on Friday
-%  Rowan:  There's some tension here - we're proposing more static checking than Java, but in the context of a dynamic langage.
-%  Sent at 12:31 AM on Friday
-%  Rowan:  The question is whether to make it clear we consider that Java should have non-nullable reference types, or we leave that alone.
-%  Sent at 12:32 AM on Friday
-%  me:  Ah. Good question.
-%  Sent at 12:35 AM on Friday
-%  me:  Our solution would be irrelevant for languages without null values.
-%  Sent at 12:38 AM on Friday
-%  Rowan:  It sometimes works to be frank about such things - somewhere in the intro we could say "Arguably Java itself should include non-nullable reference types, but we put that to one side since our main concern here is the practical issue of interoperability which arises with current languages/systems such as Clojure for the JVM.
-% "
-% Right, without null values there's no issue.
-% But, that's different to having null values, but having separate types for when null values are allowed and when they aren't.
-%  me:  Yes.
-%  Rowan:  When I said "... Java should have non-nullable reference types", I was thinking that there would be still be nullable types also.
-%  Sent at 12:42 AM on Friday
-%  me:  Oh.
-% Fair enough.
-% Scala does this I think.
-%  Rowan:  Right, yes.
-%  Sent at 12:44 AM on Friday
 
@@
 \begin{abstract}
-Modern libraries such as the Java API are sufficiently large and complicated that 
-correctly interfacing with them from other languages can be a difficult and 
-error prone aspect of software development.
-
-In particular,
-
-%Dynamically-typed languages 
-%
-%Interactions with statically-typed languages from dynamically-typed languages
-%can be better verified at compile time by utilising a gradual type system.
-%This paper describes how a dynamically-typed language, like Clojure, Ruby, or Python,
-%extended with a gradual type system
-%can better verify interactions with a statically-typed language with Null types
-%like Java or C#.
 \end{abstract}
